---
title: "Vehicles dataset"
author: "Dominika Lukašáková, Lukáš Lovás"
date: "2026-04-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("D:/Škola/Ing/2.semester/OZNAL_projekt")
#setwd("C:/school/oznal/OZNAL_Project")
```

# Introduction

In this project, we work with a dataset of vehicles that contains various technical and categorical features describing their characteristics, such as engine specifications and fuel type. The target variable of interest is combined fuel economy (`comb08`), measured in miles per gallon (MPG). Understanding and accurately predicting fuel efficiency is important for evaluating vehicle performance and environmental impact.

We formulate the following hypothesis:

**Fuel economy can be accurately predicted based on vehicle characteristics, such as engine specifications, transmission type, vehicle size and other attributes.**

We test this hypothesis by applying a range of regression models, evaluating their predictive performance, and comparing their ability to capture patterns in the data. To investigate this problem, we focus on **Scenario 2** and **Scenario 3**. We also include Scenario 5, although it is not a primary focus, as the dataset is not highly multidimensional (23 features after preprocessing).

# EDA

Vehicles dataset\
Source: <https://fueleconomy.gov/feg/ws/index.shtml>

Based on the column descriptions, we could potentially predict fuel consumption (MPG).

-   **MPG (Miles per gallon)** – a measure of fuel economy in automobiles, higher number= higher efficiency\
-   **Target variable:** `comb08`

```{r}
library(MASS)
library(tidyverse)
library(ggcorrplot)
library(patchwork)
library(rsample)
library(recipes)
library(stringr)
library(forcats)
library(tidymodels)
library(broom)
library(glmnet)
library(pls)
library(leaps)
library(rpart)
library(ranger)
library(kknn)
library(vip)
library(parsnip)
library(factoextra)

rm(list = ls())
data <- read_csv("data/vehicles.csv")
data
View(data)
```

## Basic Structure

First, we check dataset dimensions and data types.

```{r}
dim(data)
```

The dataset has 49,846 rows and 84 columns.

```{r}
glimpse(data)
```

```{r}
data %>%
  summarise(across(everything(), class))
```

The dataset contains columns of type: character, double, logical. Columns createdOn and modifiedOn are dates stored as strings. The dataset has some missing values.

## Missing Values

```{r}
sum(is.na(data))
```

The dataset contains 538 466 missing values.

# Identify columns with missing values

```{r}
missing_summary <- data %>%
  summarise(across(everything(), ~mean(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_pct") %>%
  arrange(desc(missing_pct))

missing_summary %>%
  print(n = 84)
```

Columns with more than 60% missing values:

-   `c240bDscr`, `c240Dscr`, `sCharger`, `rangeA`, `fuelType2`, `guzzler`, `evMotor`,
-   `atvType`, `tCharger`, `trans_dscr`, `startStop`, `mfrCode`

We will not use these columns and remove them later.

-   Column `eng_dscr` has \~37% missing values
-   Columns `cylinders`, `displ`, `drive`, `trany`, `fuelType1` contain only a small amount of missing data (up to \~3%), so they could potentially be imputed.

## Unique Values

Next we identify columns with unique values.

```{r}
data %>%
  summarise(across(everything(), n_distinct)) %>%
  pivot_longer(everything(), 
               names_to = "variable", 
               values_to = "n_unique") %>%
  arrange(desc(n_unique)) %>%
  print(n = 84)
```

From these we filter out non-informative columns. We remove variables with nearly all unique values, as they act as identifiers and do not provide generalizable patterns for prediction. Also variables with very low variability (binary or near-constant features) were excluded, as they provide minimal information for explaining variation in the target variable.

```{r}
data %>%
  summarise(across(everything(), n_distinct)) %>%
  pivot_longer(everything(), 
               names_to = "variable", 
               values_to = "n_unique") %>%
  arrange(desc(n_unique)) %>%
  filter(!(n_unique >= 0.9 * nrow(data))) %>%
  filter(!(n_unique <= 2)) %>%
  print(n = 84)
```

-   We should also remove `model` column, because its extremely high number of unique values makes it impractical for modeling, and `engId,` as it also represents an identifier.
-   The dataset contains a large number of variables with high cardinality, indicating a high-dimensional feature space that requires careful feature selection.

## Numerical Variables

First, we check the distribution of numerical variables.

```{r}
num_data <- data %>% 
  select(where(is.numeric)) %>% 
  select(-id, -engId)

summary(num_data)
```

-   Many columns have median = 0 because most vehicles do not have that feature, for example `highwayE` applies only to electric vehicles. So, these variables might be informative for specific subsets, but not for the general fuel consumption prediction.

-   Some columns have min and median = -1, which is unusual and likely represents missing values. The columns are: `co2`, `co2A`, `feScore`, `ghgScore`, `ghgScoreA`.

-   Although column `youSaveSpend` does not have issues in terms of distribution, its definition reveals that it is directly derived from fuel consumption and fuel cost variables. As a result, it contains information closely related to the target variable and could introduce data leakage and artificially inflating model performance.

For further analysis we will remove: - columns with median = 0 or -1 - column `youSaveSpend`

```{r}
num_data <- num_data %>%
  select(where(~ {
    med <- median(., na.rm = TRUE)
    !(med == 0 | med == -1)
  })) %>%
  select(-youSaveSpend)

summary(num_data)
```

-   We are left with 11 numerical columns.

-   The similarity between mean and median values suggests that most variables are not heavily skewed, although there might be some outliers.

```{r}
num_data %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~variable, scales = "free") +
  theme_minimal()
# A warning is displayed because of NA in columns cylinders and displ
```

Most variables show right-skewed distributions, with the majority of observations with lower values and with a long tail toward higher values.

A few variables, such as `barrels08` and `co2TailpipeGpm` appear approximately normally distributed, though still with slight positive skew. In contrast, `cylinders` is a discrete variable, with clear peaks at common engine configurations - 4, 6, and 8 cylinders.

The variable `year` shows a relatively even but slightly increasing distribution toward more recent years, indicating a higher representation of newer vehicles in the dataset.

```{r}
num_data %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "value") %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()
# A warning is displayed because of NA in columns cylinders and displ
```

-   The variable `fuelCost08` shows a wide IQR and numerous outliers.
-   Column `co2TailpipeGpm` has a relatively compact central distribution but still includes several higher outliers.
-   The `year`column has a narrow spread from 1984 to 2026, indicating that most observations are concentrated within a similar time range, so there are no outliers.

We will plot another boxplot without columns `year`, `fuelCost08`, and `co2TailpipeGpm` due to differences in scale:

```{r}
num_data %>%
  select(-year, -fuelCost08, -co2TailpipeGpm) %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "value") %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()
# A warning is displayed because of NA in columns cylinders and displ
```

-   All columns have a significant amount of outsiders on the right side.

The boxplots reveal that most variables are right-skewed, as indicated by longer upper whiskers and a large number of high-value outliers. This pattern is especially clear for fuel economy variables (`city08`, `highway08`, `comb08`, `UCity`, `UHighway`), where a small number of highly efficient vehicles extend the upper range.

Variables such as `barrels08` and `displ` exhibit moderate right skewness with some extreme values. In contrast, `cylinders` is a discrete variable, making its boxplot less informative and leading to apparent outliers due to limited possible values.

Overall, the boxplots highlight skewness and the presence of outliers.

## Correlation Analysis

```{r}
cor_matrix <- num_data %>%
  select(where(is.numeric)) %>%
  cor(use = "pairwise.complete.obs")

cor_matrix
ggcorrplot(cor_matrix)
```

-   Target `comb08` is highly positively correlated with:

    -   `city08`, `highway08`, `UCity`, `UHighway` (up to \~0.996), so the usage of these columns in a model would cause data leakage

-   Target `comb08` is strongly negatively correlated with:

    -   `barrels08`, `co2TailpipeGpm`, `cylinders`, `displ`, `fuelCost08` (up to \~ -0.7)
    -   however we know from the columns description that columns `barrels08`, `co2TailpipeGpm` and `fuelCost08` are derived from fuel consumption, so they would also cause data leakage
    -   `cylinders` and `displ` are suitable predictors but they are highly correlated with each other (\~0.91)

So, for the model, we will keep only column **`displ`**, because it has fewer missing values and a stronger correlation with target (\~ -0.71), and column `year`, because it has a moderate positive correlation (\~0.32) with target and has almost no correlation with `displ`.

## Categorical Variables

```{r}
cat_data <- data %>% 
  select(where(is.character))

ncol(cat_data)

names(cat_data)
```

The dataset has 23 categorical variables.

```{r}
cat_overview <- cat_data %>%
  summarise(across(everything(), ~n_distinct(.))) %>%
  pivot_longer(everything(),
               names_to = "variable",
               values_to = "n_levels")

cat_overview
```

Based on the number of unique values and the description of each column, we will further explore only `fuelType`, `drive`, `VClass` and `trany` columns, because they have rational amount of categories and are related to fuel efficiency.

```{r}
data %>%
  select(fuelType, drive, VClass, trany) %>%
  pivot_longer(everything(),
               names_to = "variable",
               values_to = "category") %>%
  filter(!is.na(category)) %>%
  group_by(variable, category) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(variable) %>%
  slice_max(n, n = 10) %>%
  ggplot(aes(x = reorder(category, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  facet_wrap(~variable, scales = "free") +
  theme_minimal()
```

The `fuelType` variable shows a highly imbalanced distribution, with a few dominant categories (`Regular` and `Premium`) accounting for most observations, while several alternative fuel types appear only rarely. We will group rare categories of the fuelType variable into an "Other" class.

The `drive` variable is moderately imbalanced. `Front-Wheel Drive` and `Rear-Wheel Drive` represent the majority of observations, while several four-wheel-drive variants and two-wheel-drive vehicles occur much less frequently. The presence of multiple related four-wheel-drive labels indicates potential redundancy in category definitions, so merging semantically similar categories could reduce noise.

The `trany` variable contains many transmission categories, with a clear dominance of `Automatic 4-speed` and `Manual 5-speed` types.

The VClass has 3 dominant categories - `Compact Cars`, `Midsize Cars`, and `Subcompact Cars`. The other catogories appear less frequently. Since vehicle class captures structural differences in size, it is expected to be an informative predictor and should be retained in the modeling stage.

```{r}
ggplot(data, aes(x = fuelType, y = comb08)) +
  geom_boxplot() +
  coord_flip()
```

The boxplot shows how different fuel type influences fuel consumption. We can see that the electricity type has higher values then gasoline types.

```{r}
data %>%
  filter(!is.na(drive)) %>%
  ggplot(aes(x = drive, y = comb08)) +
  geom_boxplot() +
  coord_flip()
```

Drive type clearly affects fuel efficiency. `Front-Wheel Drive` vehicles tend to have higher combined MPG, while `4-Wheel or All-Wheel Drive` has smaller values.

```{r}
data %>%
  filter(!is.na(VClass)) %>%
  mutate(VClass = fct_lump(VClass, n = 10)) %>%
  ggplot(aes(x = VClass, y = comb08)) +
  geom_boxplot() +
  coord_flip()
```

Vehicle class has a clear relationship with fuel efficiency. Smaller vehicles (`compact` and `subcompact cars`) tend to achieve higher MPG, while pickups, SUVs, and larger vehicles have smaller MPG values, so they are less fuel efficient.

```{r}
data %>%
  filter(!is.na(trany)) %>%
  mutate(trany = fct_lump(trany, n = 10)) %>%
  ggplot(aes(x = trany, y = comb08)) +
  geom_boxplot() +
  coord_flip()
```

Transmission type strongly differentiates vehicle efficiency. `Automatic (A1)` shows the highest MPG values, while other categories have smaller values.

## Tagret variable

```{r}
summary(data$comb08)
```

The target variable `comb08` (combined fuel efficiency) has strongly right-skewed distribution. Most vehicles have between 15 and 30 MPG, with a median value of 20 MPG. However, a long right tail is observed, extending up to 146 MPG.

```{r}
hist <- ggplot(data, aes(x = comb08)) +
  geom_histogram(bins = 40, fill = "steelblue") +
  theme_minimal()

box <- ggplot(data, aes(y = comb08)) +
  geom_boxplot(fill = "steelblue") +
  coord_flip() +
  theme_minimal()

hist + box + plot_layout(ncol = 1, nrow = 2) 
```

Based on the histogram, we can say that the distribution is not normal. The boxplot confirms the presence of numerous high-value outliers. However, these observations are not errors but represent a distinct technological group of vehicles with significantly higher efficiency.

## Preprocessing and Train/Test Split

### Column Groups to Remove

#### Columns with Too Many Missing Values (`cols_missing`)

These variables have more than 60% missing values and cannot be reliably used in a general model.

```{r}
cols_missing <- c(
  "c240bDscr", "c240Dscr", "sCharger", "rangeA", "fuelType2",
  "guzzler", "evMotor", "atvType", "tCharger", "trans_dscr",
  "startStop", "mfrCode"
)
```

#### Identifier Columns (`cols_ids`)

These variables are identifiers or overly specific strings. Using them would encourage the model to memorize specific vehicles or models.

```{r}
cols_ids <- c("id", "engId", "model")
```

#### Data Leakage Columns (`cols_leakage`)

These variables are either directly equivalent to `comb08` or derived from the same fuel consumption, emissions, or cost calculations. Including them would incorrectly inflate model performance.

```{r}
cols_leakage <- c(
  "city08", "highway08", "UCity", "UHighway",
  "barrels08", "co2TailpipeGpm", "fuelCost08", "youSaveSpend",
  "comb08U", "city08U", "highway08U",
  "combA08", "cityA08", "highwayA08",
  "combA08U", "cityA08U", "highwayA08U",
  "co2", "co2A", "co2TailpipeAGpm",
  "feScore", "ghgScore", "ghgScoreA",
  "phevCity", "phevHwy", "phevComb",
  "combE", "cityE", "highwayE",
  "combinedCD", "combinedUF",
  "cityCD", "cityUF",
  "highwayCD", "highwayUF"
)
```

#### Sparse or Invalid Columns

Many numeric columns have median = 0 or contain encoded missing values (−1). These are typically relevant only for electric or hybrid vehicles. Including them in the general regression dataset would introduce large numbers of zeros, encoded missing values, or derived information.

```{r}
cols_sparse_or_invalid <- c(
  "barrelsA08", "charge120", "charge240", "charge240b",
  "city08U", "cityA08", "cityA08U", "cityCD", "cityE", "cityUF",
  "co2", "co2A", "co2TailpipeAGpm",
  "comb08U", "combA08", "combA08U", "combE",
  "combinedCD", "combinedUF",
  "feScore", "fuelCostA08", "ghgScore", "ghgScoreA",
  "highway08U", "highwayA08", "highwayA08U",
  "highwayCD", "highwayE", "highwayUF",
  "hlv", "hpv", "lv2", "lv4", "pv2", "pv4",
  "range", "rangeCity", "rangeCityA", "rangeHwy", "rangeHwyA",
  "UCityA", "UHighwayA",
  "phevCity", "phevHwy", "phevComb"
)
```

#### Unused Columns

-   `eng_dscr` has \~37% missing values and is a free-text engine description.
-   `createdOn` and `modifiedOn` are metadata with no predictive value.
-   `baseModel` would support memorization of specific model lines rather than generalization.
-   `fuelType` and `fuelType1` are replaced by the new `fuel_group` variable.

```{r}
cols_unused <- c("createdOn", "modifiedOn", "eng_dscr", "baseModel")
```

#### Combine All Removal Lists

```{r}
cols_to_remove <- unique(c(cols_missing, cols_ids, cols_leakage,
                            cols_sparse_or_invalid, cols_unused))

cat("Total columns scheduled for removal:", length(cols_to_remove), "\n")
```

### Electric Vehicles

Electric vehicles do not have a combustion engine and therefore have no `displ` or `cylinders`. Their missing values are structurally meaningful — they signal the absence of an engine, not random data loss. We capture this before any imputation.

-   `is_electric`: TRUE if the vehicle runs on electricity only.
-   `has_combustion_engine`: TRUE if the vehicle has a combustion engine.

For EVs, `displ` and `cylinders` are set to 0 — meaning zero displacement, not a statistical fill-in. Remaining missing values in non-EV rows will be handled later by the recipe.

```{r}
data_fe <- data %>%
  mutate(
    is_electric = (fuelType == "Electricity") | (fuelType1 == "Electricity"),
    is_electric = replace_na(is_electric, FALSE),
    has_combustion_engine = !is_electric,
    displ     = if_else(is_electric, 0, displ),
    cylinders = if_else(is_electric, 0, cylinders)
  )

cat("Electric vehicles:", sum(data_fe$is_electric), "\n")
cat("Combustion engine vehicles:", sum(data_fe$has_combustion_engine), "\n")
```

### Categorical Variable Simplification

#### `fuel_group` — Fuel Type Grouping

We create a new variable `fuel_group` that distributes all fuel categories into six groups that simplify using the `fuel_type` variable. `fuelType` variable contains many rare categories that don't appear often enough. If we used these categories separatelly, the model could become more unstable and less accurate.

```{r}
data_fe <- data_fe %>%
  mutate(
    fuel_group = case_when(
      fuelType %in% c("Regular", "Premium", "Midgrade")       ~ "Gasoline",
      fuelType == "Diesel"                                     ~ "Diesel",
      fuelType == "Electricity"                                ~ "Electricity",
      fuelType %in% c(
        "Premium and Electricity",
        "Regular Gas and Electricity",
        "Premium Gas or Electricity",
        "Regular Gas or Electricity",
        "Electricity and Hydrogen"
      )                                                        ~ "Hybrid/Electric mix",
      fuelType %in% c("Gasoline or E85", "Premium or E85")    ~ "Flex-fuel/E85",
      fuelType %in% c(
        "CNG", "Hydrogen",
        "Gasoline or natural gas",
        "Gasoline or propane"
      )                                                        ~ "Alternative fuel",
      TRUE                                                     ~ "Other"
    )
  )

table(data_fe$fuel_group)
```

#### `drive` — Reduce 4WD/AWD Variants

We identified multiple overlapping 4-wheel and all-wheel drive labels. We group them into three interpretable groups.

```{r}
data_fe <- data_fe %>%
  mutate(drive = case_when(
    str_detect(drive, "4-Wheel|All-Wheel") ~ "AWD/4WD",
    str_detect(drive, "Front-Wheel")       ~ "Front-Wheel Drive",
    str_detect(drive, "Rear-Wheel")        ~ "Rear-Wheel Drive",
    is.na(drive)                           ~ "Unknown",
    TRUE                                   ~ "Other"
  ))

table(data_fe$drive)
```

#### `VClass` — Keep Top 10 Categories

`VClass` has many levels. We leave the 10 most frequent and collapse the rest into "Other".

```{r}
data_fe <- data_fe %>%
  mutate(VClass = fct_lump(as.factor(VClass), n = 10, other_level = "Other"))

table(data_fe$VClass)
```

#### `trany` — Extract Important Information Into More Interpretable Features

Instead of keeping the raw high-cardinality `trany` string, we extract four meaningful features:

-   `transmission_type`: "Automatic", "Manual", or "Other"
-   `is_cvt`: TRUE if the transmission uses "continuously variable gear ratios" - no set number of gears
-   `n_gears`: number of gears; 0 means "not applicable" (CVT or non-standard transmission)
-   `has_discrete_gears`: TRUE if the number of gear steps was successfully extracted from `trany`

```{r}
data_fe <- data_fe %>%
  mutate(
    transmission_type = case_when(
      str_detect(trany, regex("^Automatic", ignore_case = TRUE)) ~ "Automatic",
      str_detect(trany, regex("^Manual",    ignore_case = TRUE)) ~ "Manual",
      TRUE ~ "Other"
    ),
    is_cvt  = str_detect(trany, regex("variable gear ratios", ignore_case = TRUE)),
    is_cvt  = replace_na(is_cvt, FALSE),
    n_gears = as.integer(str_match(trany, "(\\d+)-spd")[, 2]),
    has_discrete_gears = as.integer(!is.na(n_gears)),
    n_gears = replace_na(n_gears, 0)
  )

cat("Transmission type distribution:\n")
print(table(data_fe$transmission_type))
cat("\nCVT vehicles:", sum(data_fe$is_cvt), "\n")
cat("\nn_gears summary:\n")
print(summary(data_fe$n_gears))
cat("\nhas_discrete_gears distribution:\n")
print(table(data_fe$has_discrete_gears))
```

### Build Final Modeling Dataset

We apply all column removals. Pure battery electric vehicles (BEVs) are excluded from the modeling dataset before the train/test split. The target variable `comb08` represents Miles Per Gallon (MPG) for combustion-engine vehicles, but for Electric Vehicles the MPG is converted to MPGe — an artificial conversion. That does not reflect actual fuel consumption, which means including these vehicles that have two structurally different aspects in a single regression target would only harm the prediction.

```{r}
n_ev <- sum(data_fe$is_electric)

vehicles_model <- data_fe %>%
  filter(!is_electric) %>%
  select(-any_of(cols_to_remove)) %>%
  select(-any_of(c("trany", "fuelType", "fuelType1", "is_electric", "has_combustion_engine")))

cat("Electric vehicles excluded:", n_ev, "\n")
cat("Remaining vehicles:        ", nrow(vehicles_model), "\n")
cat("Dimensions of vehicles_model:", nrow(vehicles_model), "×", ncol(vehicles_model), "\n")
cat("\nColumns retained:\n")
print(names(vehicles_model))
```

`cylinders` is kept as a candidate predictor despite its strong correlation with `displ` (\~0.91). Multicollinearity between the two was expected, but LASSO (Scenario 2) selected `penalty = 1e-4` — too small to zero any coefficient — and retained both predictors. With n/p \> 2 000, OLS variance is low enough that the multicollinearity does not destabilize estimates. Both `cylinders` and `displ` also appear as the two strongest PLS Component 1 loadings (Scenario 5), confirming they each carry independent signal: `displ` reflects total engine displacement while `cylinders` captures engine architecture (e.g. a turbocharged 4-cylinder can have large displacement but few cylinders). Dropping either variable based solely on the correlation would discard real predictive information.

### Sanity Checks

```{r}
cat("Duplicate rows in vehicles_model:", sum(duplicated(vehicles_model)), "\n\n")

cat("summary(comb08):\n")
print(summary(vehicles_model$comb08))

cat("\nYear range:", range(vehicles_model$year, na.rm = TRUE), "\n")

cat("\ncomb08 <= 0:", sum(vehicles_model$comb08 <= 0, na.rm = TRUE), "\n")
cat("displ < 0:",    sum(vehicles_model$displ    < 0, na.rm = TRUE), "\n")
cat("cylinders < 0:", sum(vehicles_model$cylinders < 0, na.rm = TRUE), "\n")

missing_before <- colSums(is.na(vehicles_model))
cat("\nColumns with missing values (before recipe):\n")
print(missing_before[missing_before > 0])
cat("Total missing values:", sum(is.na(vehicles_model)), "\n")
```

After excluding BEVs, the remaining `comb08` values reflect actual MPG from combustion-engine and hybrid vehicles. Extreme values at the high end (diesel hybrids, efficient compact cars) are retained as valid observations.

### Train/Test Split

We split the data into 80% training and 20% test. The test set is only used at the very end to evaluate final model performance — we do not touch it during training or tuning. We stratify on `comb08` so both sets have a similar distribution of MPG values.

```{r}
set.seed(123)

data_split <- initial_split(vehicles_model, prop = 0.8, strata = comb08)

train_data <- training(data_split)
test_data  <- testing(data_split)

cat("Train:", nrow(train_data), "×", ncol(train_data), "\n")
cat("Test: ", nrow(test_data),  "×", ncol(test_data),  "\n")
```

### Check Train/Test Distribution

```{r}
cat("Train comb08:\n"); print(summary(train_data$comb08))
cat("Test  comb08:\n"); print(summary(test_data$comb08))
```

```{r}
bind_rows(
  train_data %>% select(comb08) %>% mutate(split = "train"),
  test_data  %>% select(comb08) %>% mutate(split = "test")
) %>%
  ggplot(aes(x = comb08, fill = split)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("train" = "steelblue", "test" = "tomato")) +
  labs(
    title = "Distribution of comb08: Train vs. Test",
    x     = "Combined MPG (comb08)",
    y     = "Density"
  ) +
  theme_minimal()
```

The density curves confirm that stratification produces very similar `comb08` distributions in both sets.

### Cross-Validation Folds

All models require hyperparameter tuning, which involves comparing performance across different parameter values. We use 5-fold cross-validation for this: the training data is split into 5 folds, the model trains on 4 and is evaluated on the remaining one, rotating through all folds. This gives a reliable performance estimate without using the test set.

```{r}
folds <- vfold_cv(train_data, v = 5, strata = comb08)

cat("Number of CV folds:", nrow(folds), "\n")
print(folds)
```

### Preprocessing Recipe

We define preprocessing as a `recipe` to prevent data leakage — all transformations (normalization, imputation, dummy encoding) are learned only from the training data and then applied to the test set using those saved values. This also works correctly inside cross-validation, where each fold fits its own preprocessing independently.

-   `step_unknown()` — handles unseen factor levels that may appear in the test set
-   `step_other()` — collapses any residual rare categories below 1%
-   `step_impute_median(displ, cylinders)` — fills the small number of numeric NAs in engine parameters for non-BEV rows; `n_gears` is intentionally set to 0 in feature engineering and does not need imputation
-   `step_dummy()` — one-hot encodes all nominal predictors; required by linear regression methods
-   `step_zv()` / `step_nzv()` — removes zero- and near-zero-variance columns, consistent with EDA findings on low-variability variables
-   `step_normalize()` — standardizes all numeric predictors; required for ridge, lasso, elastic net, kNN, and PCA/PCR/PLS

```{r}
model_recipe <- recipe(comb08 ~ ., data = train_data) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%
  step_impute_median(displ, cylinders) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_nzv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

model_recipe
```

### Fit and Apply Recipe

The recipe is fitted exclusively on `train_data`. The same fitted statistics (medians, modes, dummy levels) are then applied to both sets.

```{r}
prep_recipe <- prep(model_recipe, training = train_data)

train_processed <- bake(prep_recipe, new_data = NULL)
test_processed  <- bake(prep_recipe, new_data = test_data)
```

### Verify Preprocessing Output

```{r}
cat("train_processed:", nrow(train_processed), "×", ncol(train_processed), "\n")
cat("test_processed: ", nrow(test_processed),  "×", ncol(test_processed),  "\n")

cat("\nMissing values — train_processed:", sum(is.na(train_processed)), "\n")
cat("Missing values — test_processed: ", sum(is.na(test_processed)),  "\n")

cat("\ncomb08 in train_processed:", "comb08" %in% names(train_processed), "\n")
cat("comb08 in test_processed: ", "comb08" %in% names(test_processed),  "\n")

n_predictors <- ncol(train_processed) - 1
cat("\nNumber of predictors after dummy encoding:", n_predictors, "\n")

cat("\nFirst 10 column names:\n")
print(head(names(train_processed), 10))
```

Both processed datasets contain no missing values in predictors, and `comb08` is retained as the regression target. The objects `vehicles_model`, `train_data`, `test_data`, `folds`, `model_recipe`, `prep_recipe`, `train_processed`, and `test_processed` are ready for regression scenarios 2, 3, and 5.

## Scenario 2: Parametric vs. Non-Parametric Models

### Parametric models

Parametric models assume that the relationship between predictors and the response follows a specific, fixed mathematical form — in regression, a linear combination of weighted features.

We compare four parametric regression methods:

1.  **Linear Regression (OLS)** — the classical least-squares baseline, no regularization
2.  **Ridge Regression** — OLS plus an L2 penalty that shrinks all coefficients toward zero
3.  **LASSO** — OLS plus an L1 penalty that can shrink some coefficients to exactly zero, selecting features
4.  **Elastic Net** — a blend of Ridge and LASSO penalties, combining both approaches

All four methods share the same preprocessing recipe defined in the previous section. We use the 5-fold cross-validation folds for hyperparameter tuning and model comparison, and evaluate final performance on the held-out test set.

We define the three evaluation metrics once and reuse them across every model.

```{r}
eval_metrics <- metric_set(rmse, rsq, mae)
```

------------------------------------------------------------------------

#### 1. Linear Regression (OLS)

Ordinary least squares minimizes the sum of squared residuals and imposes no constraint on coefficient size. It is our reference point — any benefit from regularization will appear as an improvement over this baseline.

```{r}
lm_spec <- linear_reg() %>%
  set_engine("lm") %>%
  set_mode("regression")

lm_wf <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(lm_spec)
```

We wrap the recipe and model in a `workflow` so preprocessing is applied automatically during cross-validation and when predicting on new data.

```{r warning=FALSE}
lm_cv <- lm_wf %>%
  fit_resamples(
    resamples = folds,
    metrics   = eval_metrics,
    control   = control_resamples(save_pred = TRUE)
  )

lm_cv_metrics <- collect_metrics(lm_cv)
cat("=== OLS — 5-fold CV ===\n")
cat("RMSE:", round(lm_cv_metrics %>% filter(.metric == "rmse") %>% pull(mean), 4),
    "  ±", round(lm_cv_metrics %>% filter(.metric == "rmse") %>% pull(std_err), 4), "\n")
cat("MAE: ", round(lm_cv_metrics %>% filter(.metric == "mae")  %>% pull(mean), 4), "\n")
cat("R²:  ", round(lm_cv_metrics %>% filter(.metric == "rsq")  %>% pull(mean), 4), "\n")
```

Now we fit on the full training set to inspect which predictors drive the predictions.

```{r}
lm_fit <- lm_wf %>% fit(data = train_data)

lm_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(term != "(Intercept)") %>%
  slice_max(abs(estimate), n = 20) %>%
  ggplot(aes(x = reorder(term, abs(estimate)), y = estimate)) +
  geom_col(aes(fill = estimate > 0)) +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
    labels = c("TRUE" = "Positive", "FALSE" = "Negative"),
    name   = "Direction"
  ) +
  labs(
    title = "OLS: Top 20 Predictors by Absolute Coefficient",
    x     = NULL,
    y     = "Coefficient estimate"
  ) +
  theme_minimal()
```

Blue bars indicate predictors that increase `comb08` (e.g. electric drivetrain), red bars indicate predictors that decrease it (e.g. larger engine displacement).

------------------------------------------------------------------------

#### 2. Ridge Regression

Ridge regression adds an L2 penalty — λ × Σβ² — to the OLS loss. This forces the optimiser to keep coefficients small, reducing variance at the cost of some bias. Ridge never removes a predictor entirely: all coefficients are shrunk toward zero, but remain non-zero. It is especially effective when predictors are correlated — a situation we identified between `displ` and `cylinders` in the EDA.

The regularization strength λ is controlled by `penalty`. We tune it over a log-spaced grid.

```{r}
ridge_spec <- linear_reg(penalty = tune(), mixture = 0) %>%
  set_engine("glmnet") %>%
  set_mode("regression")

ridge_wf <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(ridge_spec)
```

`glmnet` fits the entire regularization path in a single pass, so evaluating 50 penalty values has almost no extra cost compared to evaluating one.

```{r}
ridge_grid <- grid_regular(penalty(range = c(-4, 2)), levels = 50)
```

```{r warning=FALSE}
ridge_cv <- ridge_wf %>%
  tune_grid(
    resamples = folds,
    grid      = ridge_grid,
    metrics   = eval_metrics
  )

autoplot(ridge_cv)
```

At very low penalty (left end), Ridge behaves like OLS. As penalty increases, coefficients shrink and the model becomes more biased but less variable. The optimal point is where RMSE on the validation folds is minimized.

```{r}
best_ridge <- ridge_cv %>% select_best(metric = "rmse")

ridge_best_row <- show_best(ridge_cv, metric = "rmse", n = 1)
cat("=== Ridge — best hyperparameter ===\n")
cat("Best penalty:", round(best_ridge$penalty, 6), "\n")
cat("CV RMSE:     ", round(ridge_best_row$mean, 4),
    "  ±", round(ridge_best_row$std_err, 4), "\n")
```

```{r}
ridge_final_wf <- ridge_wf %>% finalize_workflow(best_ridge)
ridge_fit       <- ridge_final_wf %>% fit(data = train_data)
```

```{r}
ridge_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(term != "(Intercept)") %>%
  slice_max(abs(estimate), n = 20) %>%
  ggplot(aes(x = reorder(term, abs(estimate)), y = estimate)) +
  geom_col(aes(fill = estimate > 0)) +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
    labels = c("TRUE" = "Positive", "FALSE" = "Negative"),
    name   = "Direction"
  ) +
  labs(
    title = paste0("Ridge: Top 20 Predictors (penalty = ",
                   round(best_ridge$penalty, 5), ")"),
    x     = NULL,
    y     = "Coefficient estimate"
  ) +
  theme_minimal()
```

Compared to OLS, Ridge produces smaller coefficient magnitudes across all predictors. The ordering and direction of the most important features generally remains the same, but correlated predictors share their weight more evenly.

------------------------------------------------------------------------

#### 3. LASSO Regression

LASSO (Least Absolute Shrinkage and Selection Operator) uses an L1 penalty — λ × Σ\|β\|. The key geometric property of L1 is that the regularized solution often lands exactly at zero for many coefficients. LASSO therefore performs simultaneous regression and feature selection: it automatically identifies the most informative predictors and removes the rest.

```{r}
lasso_spec <- linear_reg(penalty = tune(), mixture = 1) %>%
  set_engine("glmnet") %>%
  set_mode("regression")

lasso_wf <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(lasso_spec)
```

LASSO typically needs smaller penalties than Ridge to achieve similar shrinkage, so we search a narrower range.

```{r}
lasso_grid <- grid_regular(penalty(range = c(-4, 0)), levels = 50)
```

```{r warning=FALSE}
lasso_cv <- lasso_wf %>%
  tune_grid(
    resamples = folds,
    grid      = lasso_grid,
    metrics   = eval_metrics
  )

autoplot(lasso_cv)
```

```{r}
best_lasso <- lasso_cv %>% select_best(metric = "rmse")

lasso_best_row <- show_best(lasso_cv, metric = "rmse", n = 1)
cat("=== LASSO — best hyperparameter ===\n")
cat("Best penalty:", round(best_lasso$penalty, 6), "\n")
cat("CV RMSE:     ", round(lasso_best_row$mean, 4),
    "  ±", round(lasso_best_row$std_err, 4), "\n")
```

```{r}
lasso_final_wf <- lasso_wf %>% finalize_workflow(best_lasso)
lasso_fit       <- lasso_final_wf %>% fit(data = train_data)
```

One of LASSO's main advantages is sparsity. Let's see how many predictors it retained.

```{r}
lasso_coefs <- lasso_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(term != "(Intercept)")

cat("=== LASSO — sparsity ===\n")
cat("Total predictors available:", nrow(lasso_coefs), "\n")
cat("Kept by LASSO (non-zero):  ", sum(lasso_coefs$estimate != 0), "\n")
cat("Removed by LASSO (zero):   ", sum(lasso_coefs$estimate == 0), "\n")
if (sum(lasso_coefs$estimate == 0) > 0) {
  cat("Zeroed out predictors:\n")
  print(lasso_coefs %>% filter(estimate == 0) %>% pull(term))
}
```

```{r}
lasso_coefs %>%
  filter(estimate != 0) %>%
  slice_max(abs(estimate), n = 20) %>%
  ggplot(aes(x = reorder(term, abs(estimate)), y = estimate)) +
  geom_col(aes(fill = estimate > 0)) +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
    labels = c("TRUE" = "Positive", "FALSE" = "Negative"),
    name   = "Direction"
  ) +
  labs(
    title = paste0("LASSO: Top 20 Non-Zero Predictors (penalty = ",
                   round(best_lasso$penalty, 5), ")"),
    x     = NULL,
    y     = "Coefficient estimate"
  ) +
  theme_minimal()
```

LASSO retains only the most informative predictors, which makes the resulting model more parsimonious and directly interpretable compared to OLS or Ridge. Predictors that survive the L1 penalty are the ones that the algorithm judged worth the cost of adding a parameter.

------------------------------------------------------------------------

#### 4. Elastic Net

Elastic Net combines both L1 and L2 penalties: λ × [α × Σ\|β\| + (1 − α) × Σβ²]. The `mixture` parameter in tidymodels corresponds to α: `mixture = 0` gives pure Ridge, `mixture = 1` gives pure LASSO. By tuning both `penalty` and `mixture`, Elastic Net finds the best blend — it can perform feature selection like LASSO while also handling correlated predictors like Ridge.

```{r}
en_spec <- linear_reg(penalty = tune(), mixture = tune()) %>%
  set_engine("glmnet") %>%
  set_mode("regression")

en_wf <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(en_spec)
```

We use a regular 10 × 10 grid — 10 penalty values and 10 mixture values — giving 100 combinations. `glmnet` evaluates all penalty values for each mixture in a single path, so this is still fast.

```{r}
en_grid <- grid_regular(
  penalty(range = c(-4, 0)),
  mixture(range = c(0, 1)),
  levels = 10
)
```

```{r warning=FALSE}
en_cv <- en_wf %>%
  tune_grid(
    resamples = folds,
    grid      = en_grid,
    metrics   = eval_metrics
  )

autoplot(en_cv)
```

```{r}
best_en <- en_cv %>% select_best(metric = "rmse")

en_best_row <- show_best(en_cv, metric = "rmse", n = 1)
cat("=== Elastic Net — best hyperparameters ===\n")
cat("Best penalty:", round(best_en$penalty, 6), "\n")
cat("Best mixture:", round(best_en$mixture, 4),
    " (0 = Ridge, 1 = LASSO)\n")
cat("CV RMSE:     ", round(en_best_row$mean, 4),
    "  ±", round(en_best_row$std_err, 4), "\n")
```

The optimal `mixture` value tells us how the data prefers to be regularized. If the best mixture is close to 1, LASSO-like behavior is preferred; if it is close to 0, Ridge-like behavior is preferred.

```{r}
en_final_wf <- en_wf %>% finalize_workflow(best_en)
en_fit       <- en_final_wf %>% fit(data = train_data)
```

------------------------------------------------------------------------

#### Comparing All Parametric Models

We evaluate all four models on the held-out test set. Because each model is a full `workflow()` that includes the preprocessing recipe, we pass the raw `test_data` directly and the recipe transformations are applied automatically.

We write a small helper function so we do not repeat the same prediction + metric code four times.

```{r}
get_test_metrics <- function(fit, model_name) {
  fit %>%
    predict(new_data = test_data) %>%
    bind_cols(test_data %>% select(comb08)) %>%
    summarise(
      model = model_name,
      RMSE  = rmse_vec(truth = comb08, estimate = .pred),
      MAE   = mae_vec( truth = comb08, estimate = .pred),
      Rsq   = rsq_vec( truth = comb08, estimate = .pred)
    )
}
```

```{r}
test_comparison <- bind_rows(
  get_test_metrics(lm_fit,    "Linear Regression"),
  get_test_metrics(ridge_fit, "Ridge"),
  get_test_metrics(lasso_fit, "LASSO"),
  get_test_metrics(en_fit,    "Elastic Net")
) %>%
  arrange(RMSE)

test_comparison
```

```{r}
test_comparison %>%
  pivot_longer(cols = c(RMSE, MAE, Rsq),
               names_to  = "metric",
               values_to = "value") %>%
  ggplot(aes(x = reorder(model, value), y = value, fill = model)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~metric, scales = "free_x") +
  coord_flip() +
  labs(
    title = "Test Set Performance: Parametric Models",
    x     = NULL,
    y     = "Value"
  ) +
  theme_minimal()
```

Lower RMSE and MAE are better; higher R² is better.

##### Cross-Validation vs. Test Set RMSE

We compare CV RMSE with test RMSE. A small gap means no overfitting — the model generalizes to new data roughly as well as it performed during CV.

```{r}
cv_rmse_comparison <- tibble(
  model   = c("Linear Regression", "Ridge", "LASSO", "Elastic Net"),
  cv_rmse = c(
    collect_metrics(lm_cv)    %>% filter(.metric == "rmse") %>% pull(mean),
    show_best(ridge_cv, metric = "rmse", n = 1) %>% pull(mean),
    show_best(lasso_cv, metric = "rmse", n = 1) %>% pull(mean),
    show_best(en_cv,    metric = "rmse", n = 1) %>% pull(mean)
  )
) %>%
  left_join(
    test_comparison %>% select(model, test_rmse = RMSE),
    by = "model"
  ) %>%
  mutate(gap = abs(test_rmse - cv_rmse)) %>%
  arrange(cv_rmse)

cv_rmse_comparison
```

A small `gap` confirms no overfitting.

##### Actual vs. Predicted — All Models

Actual vs. predicted on the test set. Points on the red diagonal = perfect prediction.

```{r}
all_test_preds <- bind_rows(
  lm_fit    %>% predict(new_data = test_data) %>% bind_cols(test_data %>% select(comb08)) %>% mutate(model = "Linear Regression"),
  ridge_fit %>% predict(new_data = test_data) %>% bind_cols(test_data %>% select(comb08)) %>% mutate(model = "Ridge"),
  lasso_fit %>% predict(new_data = test_data) %>% bind_cols(test_data %>% select(comb08)) %>% mutate(model = "LASSO"),
  en_fit    %>% predict(new_data = test_data) %>% bind_cols(test_data %>% select(comb08)) %>% mutate(model = "Elastic Net")
)

all_test_preds %>%
  ggplot(aes(x = comb08, y = .pred)) +
  geom_point(alpha = 0.15, color = "steelblue", size = 0.8) +
  geom_abline(slope = 1, intercept = 0, color = "tomato", linewidth = 0.8) +
  facet_wrap(~model) +
  labs(
    title = "Actual vs. Predicted — Test Set (Parametric Models)",
    x     = "Actual combined MPG",
    y     = "Predicted combined MPG"
  ) +
  theme_minimal()
```

With BEVs excluded, all values are real MPG. The fan-shaped distortion from MPGe outliers is gone. Remaining scatter at high MPG values is mostly PHEVs and efficient compacts — the model slightly underpredicts these.

------------------------------------------------------------------------

#### Summary: Parametric Models

All four models assume a linear relationship. They differ only in regularization:

| Model       | Penalty | Sparsity |
|-------------|---------|----------|
| OLS         | None    | No       |
| Ridge       | L2      | No       |
| LASSO       | L1      | Yes      |
| Elastic Net | L1 + L2 | Yes      |

On large datasets the differences are small. Regularization stabilizes coefficients; LASSO additionally zeroes out weak predictors.

------------------------------------------------------------------------

#### Conclusion

##### Effect of removing electric vehicles

Removing 1 425 BEVs (whose `comb08` values are MPGe, not MPG) dropped RMSE from 8.11 to **3.36 MPG** (−59%) and MAE from 5.15 to **2.19 MPG**. R² only moved from 0.642 to 0.661 — this is expected, since R² is scale-independent and both the errors and the total variance dropped together.

##### Model results

All four models gave nearly the same result on the test set: **RMSE ≈ 3.36 MPG, MAE ≈ 2.19 MPG, R² ≈ 0.661**. With a median `comb08` of 20 MPG, an average error of 2.19 MPG is roughly 11% — reasonable for a linear model with 21 features.

##### Why regularization did not help

All three regularized models selected `penalty = 0.0001` — the minimum of the search grid. With 38 735 training observations and only 21 predictors, there is very little variance to reduce, so regularization has almost nothing to do. LASSO also kept all 21 predictors for the same reason — at such a small penalty, no predictor is weak enough to be zeroed out. This is not a bad result; it means the feature set is clean and every variable carries real signal.

##### Where the models fall short

-   The relationship between engine size and MPG is not perfectly linear — adding a cylinder to a small 4-cylinder engine is different from adding one to a V8.
-   PHEVs and mild hybrids behave differently depending on driving mode, which no single feature captures.
-   Vehicle weight and aerodynamic drag are not in the dataset, even though they strongly affect fuel economy.

The CV–test gap is below 0.005 for all models, so there is no overfitting. The models are a solid baseline — interpretable, stable, and easy to explain.

#### 2.5 Extension: interaction terms (cylinders × displ)

The four parametric models above are linear in the predictors. EDA showed that `cylinders` and `displ` are strongly correlated (r ≈ 0.91) and that adding a cylinder affects small engines differently than V8s — a classic non-linearity that a purely linear model cannot capture. Here we add the explicit product term `cylinders × displ` to the recipe and refit OLS to see whether the cross-effect reduces test RMSE and what its sign tells us.

```{r}
rec_interact <- recipe(comb08 ~ ., data = train_data) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%
  step_impute_median(displ, cylinders) %>%
  step_interact(terms = ~ cylinders:displ) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_nzv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

ols_interact_wf <- workflow() %>%
  add_recipe(rec_interact) %>%
  add_model(linear_reg() %>% set_engine("lm"))

ols_interact_fit <- ols_interact_wf %>% fit(train_data)
```

```{r}
ols_interact_pred <- predict(ols_interact_fit, new_data = test_data) %>%
  bind_cols(test_data %>% select(comb08))

ols_interact_metrics <- tibble(
  model = "OLS (cylinders × displ)",
  RMSE  = rmse_vec(truth = ols_interact_pred$comb08,
                   estimate = ols_interact_pred$.pred),
  MAE   = mae_vec( truth = ols_interact_pred$comb08,
                   estimate = ols_interact_pred$.pred),
  Rsq   = rsq_vec( truth = ols_interact_pred$comb08,
                   estimate = ols_interact_pred$.pred)
) %>%
  mutate(across(c(RMSE, MAE, Rsq), ~round(., 4)))

interaction_compare <- bind_rows(
  test_comparison %>% filter(model == "Linear Regression") %>%
    select(model, RMSE, MAE, Rsq) %>%
    mutate(model = "OLS (no interaction)"),
  ols_interact_metrics
)

cat("=== OLS with vs. without cylinders × displ ===\n")
print(interaction_compare)

interaction_term <- ols_interact_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(grepl("cylinders_x_displ|cylinders:displ", term))

cat("\n=== Interaction coefficient ===\n")
print(interaction_term)
```

The sign of the `cylinders × displ` coefficient indicates how the cylinder effect changes per unit of normalized displacement: a negative value means adding a cylinder hurts MPG more on top of an already-large engine (compounding penalty); a positive value means the cylinder penalty softens for big engines. The RMSE gain is expected to be small at n/p \> 2000, but the interaction coefficient itself directly tests whether the cross-effect we suspected from EDA is real after controlling for everything else in the model.

### Non-parametric regression methods

Non-parametric models do not assume a specific relationship between predictors and the response. Instead, they learn patterns directly from the data, allowing them to capture more complex and non-linear relationships.

We compare three non-parametric regression methods:

1.  **Decision tree** - a single tree that partitions the feature space into regions with similar response values
2.  **Random Forest** - an ensemble of decision trees that improves performance by averaging multiple models
3.  **K-nearest neighbors (KNN)** - a distance-based method that predicts values based on nearby observations

These models were chosen because they can capture non-linear relationships observed in the data. Tree-based methods are robust and require minimal assumptions, while KNN can model local patterns when features are properly scaled. Overall, the characteristics of the dataset are suitable for applying these non-parametric methods.

The Decision tree and Random forest models use the same preprocessing recipe as the parametric methods, even though some steps (such as normalization) are not strictly required for tree-based models. The KNN model uses a modified preprocessing recipe due to issues with `step_nzv(all_predictors())`, and includes an additional step `step_novel(all_nominal_predictors())` to handle unseen categories.

We use 5-fold cross-validation for hyperparameter tuning and model comparison and evaluate final performance on the held-out test set. The same evaluation metrics are applied consistently across all models.

Variables used for hyperparameters tuning:

```{r}
tune_ctrl <- control_grid(
  save_pred    = TRUE,
  save_workflow = TRUE,
  verbose      = TRUE
)

metrics_set <- metric_set(rmse, rsq, mae)
```

#### Decision tree

Firstly, we will create decision tree model. The decision tree algorithm doesnt need normalised data, however we will use the prepared recipe with a normalisation step.

**Hyperparameters tuning**

```{r}
dt_spec <- decision_tree(
  cost_complexity = tune(),   # alpha used for pruning
  tree_depth      = tune(),
  min_n           = tune()
) %>%
  set_engine("rpart") %>%
  set_mode("regression")

dt_wf <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(dt_spec)

dt_grid <- grid_regular(
  cost_complexity(range = c(-4, -1)),  # log10 scale: 0.0001 - 0.1
  tree_depth(range = c(3, 15)),
  min_n(range = c(2, 20)),
  levels = 4
)

dt_tune <- tune_grid(
  dt_wf,
  resamples = folds,
  grid      = dt_grid,
  metrics   = metrics_set,
  control   = tune_ctrl
)
```

```{r}
show_best(dt_tune, metric = "rmse", n = 5)

autoplot(dt_tune) +
  labs(title = "Decision Tree - hyperparameters tuning") +
  theme_minimal()
```

The graph shows, how the models' performance changes with different hyperparameters. Model performance was most sensitive to the cost-complexity parameter, where larger values substantially worsened RMSE and R², indicating underfitting. Deeper trees (depth 7–11) outperformed shallow trees (depth 3), and the minimum node size had only a minor effect on predictive performance.

```{r}
dt_best <- select_best(dt_tune, metric = "rmse")
cat("The best parameters for decision tree:\n"); print(dt_best)

dt_final <- dt_wf %>%
  finalize_workflow(dt_best) %>%
  fit(data = train_data)
```

The best decision tree model has these parameters: - cost_complexity = 0.0001 - tree_depth = 15 - min_n = 8

```{r}
# get metrics on train data
dt_train_preds <- predict(dt_final, new_data = train_data) %>%
  bind_cols(train_data %>% select(comb08))

dt_train_metrics <- dt_train_preds %>%
  metrics(truth = comb08, estimate = .pred) %>%
  mutate(split = "train")

# get metrics on test data
dt_preds <- predict(dt_final, new_data = test_data) %>%
  bind_cols(test_data %>% select(comb08))

dt_test_metrics <- dt_preds %>%
  metrics(truth = comb08, estimate = .pred) %>%
  mutate(split = "test")

bind_rows(dt_train_metrics, dt_test_metrics) %>%
  select(split, .metric, .estimate) %>%
  pivot_wider(names_from = split, values_from = .estimate)

# check predictions in the first rows
head(dt_preds)
```

The tuned decision tree achieved solid predictive performance and showed good generalization to unseen data. Predicted `comb08` values were close to the actual observed values, with only small deviations in the predictions.

On the training data, the model reached an RMSE of 2.01 and an MAE of 1.38, indicating low average prediction errors. On the test dataset, performance remained similar, with an RMSE of 2.15 and an MAE of 1.44. The small increase in error from training to test data suggests that the model generalizes well and does not suffer from overfitting. Based on the R² values, we can say that the model explained 87.9% of the variance in the training data and 86.1% of the variance in the test data.

```{r}
ggplot(dt_preds, aes(x = comb08, y = .pred)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Decision tree: Actual vs Predicted",
    x = "Actual comb08",
    y = "Predicted comb08"
  ) +
  theme_minimal()

dt_pred <- dt_preds %>%
  mutate(residual = comb08 - .pred)

ggplot(dt_pred, aes(x = .pred, y = residual)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Residual Plot",
    x = "Predicted comb08",
    y = "Residuals"
  ) +
  theme_minimal()
```

The Actual vs Predicted plot shows a clear positive linear relationship between observed and predicted comb08 values, indicating that the model captures the main structure of the data well. Most points are concentrated in range 15-35 MPG, these are predicted with good accuracy. For higher values of `comb08` , the model indicates lower accuracy as it tends to underpredict some most efficient vehicles. A typical step pattern is visible in the predictions, which is expected for decision trees because they assign the same predicted value to all observations within a terminal node.

From the Residual plot, we can tell that residuals are centered around zero for most observations, but residual variance increases for high predicted MPG values. The graph indicates mild heteroskedasticity.

```{r}
dt_final %>%
  extract_fit_parsnip() %>%
  vip(num_features = 15) +
  labs(title = "Feature importance - Decision tree") +
  theme_minimal()
```

The graph shows 15 most important features, which were used by model to predict target variable. The most important features are `cylinders` and `displ`.

#### Random forest

Secondly, we will create random forest model. The random forest algorithm doesnt need normalised data, however we will use the prepared recipe with a normalisation step.

Since our dataset is quite large (40 000 rows) and the `randomForest` engine is training quite slowly, we will use a `ranger` engine in hyperparameters tuning steps, because it trains much quicker.

**Hyperparameters tuning**

```{r}
rf_model <- rand_forest(
  trees = 200,
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")

rf_workflow <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(rf_model)

rf_grid <- grid_regular(
  mtry(range = c(2, 10)),
  min_n(range = c(2, 20)),
  levels = 3
)

rf_tuned <- tune_grid(
  rf_workflow,
  resamples = folds,
  grid = rf_grid,
  metrics   = metrics_set,
  control   = tune_ctrl
)

rf_best <- select_best(rf_tuned, metric = "rmse")
cat("The best parameters for random forest:\n"); print(rf_best)

rf_final <- rf_workflow %>%
  finalize_workflow(rf_best) %>%
  fit(data = train_data)
```

The best random forest model has these parameters: - trees = 200 - mtry = 10 - min_n = 2

```{r}
# get metrics on train data
rf_train_preds <- predict(rf_final, new_data = train_data) %>%
  bind_cols(train_data %>% select(comb08))

rf_train_metrics <- rf_train_preds %>%
  metrics(truth = comb08, estimate = .pred) %>%
  mutate(split = "train")

# get metrics on test data
rf_pred <- predict(rf_final, new_data = test_data) %>%
  bind_cols(test_data %>% select(comb08))

rf_test_metrics <- rf_pred %>%
  metrics(truth = comb08, estimate = .pred) %>%
  mutate(split = "test")

bind_rows(rf_train_metrics, rf_test_metrics) %>%
  select(split, .metric, .estimate) %>%
  pivot_wider(names_from = split, values_from = .estimate)

# check predictions in the first rows
head(rf_pred)
```

The best random forest model achieved excellent predictive performance and outperformed the simpler decision tree model in terms of prediction error and explained variance. Predicted `comb08` values were predicted almost exactly.

On the training data, the model achieved an RMSE of 1.39 and an MAE of 0.76, indicating very low prediction errors. On the test dataset, performance remained strong, with an RMSE of 1.75 and an MAE of 1.03. Although the error increased slightly on unseen data, the difference between training and test results is relatively small, suggesting good generalization ability.

The model explained 94.2% of the variance in the training data and 90.9% of the variance in the test data, based on the R² values, so the underlying relationships between predictors and fuel economy were captured extremely well.

```{r}
ggplot(rf_pred, aes(x = comb08, y = .pred)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Random Forest: Actual vs Predicted",
    x = "Actual comb08",
    y = "Predicted comb08"
  ) +
  theme_minimal()

rf_pred <- rf_pred %>%
  mutate(residual = comb08 - .pred)

ggplot(rf_pred, aes(x = .pred, y = residual)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Residual Plot",
    x = "Predicted comb08",
    y = "Residuals"
  ) +
  theme_minimal()
```

The Actual vs Predicted plot shows strong agreement between observed and predicted `comb08` values, with most points close to the diagonal line. This indicates accurate predictions across a wide range of vehicles, especially in the common 15–35 MPG range. Prediction errors increase slightly for vehicles with very high fuel efficiency, but the model still remains relatively accurate in this region. There is a small tendency to underpredict the highest MPG values, although less than in the decision tree model.

From the Residual plot, we can tell that residuals are centered around zero for most observations, but residual variance increases for high predicted MPG values. The graph indicates mild heteroskedasticity.

Overall, the random forest model provides better predictive performance and stronger generalization than the decision tree, though it is less interpretable.

#### K-nearest neighbors (KNN)

Lastly, we will train also a KNN model. We will find the best model using hyperparameters tuning. As we stated before, we will modify the preprocessing recipe due to issues with `step_nzv(all_predictors())`, and include an additional step `step_novel(all_nominal_predictors())` to handle unseen categories.

```{r}
knn_recipe <- recipe(comb08 ~ ., data = train_data) %>%
  step_novel(all_nominal_predictors()) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%
  step_impute_median(displ, cylinders) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())
```

**Hyperparameters tuning**

*Note: The hyperparameter tuning for the KNN model was computationally demanding and took approximately 6 hours to complete due to the size of the dataset and cross-validation procedure.*

```{r}
knn_spec <- nearest_neighbor(
  neighbors = tune(),
  weight_func = tune(),
  dist_power  = tune()
) %>%
  set_engine("kknn") %>%
  set_mode("regression")

knn_wf <- workflow() %>%
  add_recipe(knn_recipe) %>%
  add_model(knn_spec)

# KNN hyperparameters tuning
knn_grid <- grid_regular(
  neighbors(range = c(3, 25)),
  weight_func(values = c("rectangular", "triangular", "gaussian")),
  dist_power(range = c(1, 2)),
  levels = 3
)

cat("=== Tuning KNN ===\n")
knn_tune <- tune_grid(
  knn_wf,
  resamples = folds,
  grid      = knn_grid,
  metrics   = metrics_set,
  control   = tune_ctrl
)

knn_best <- select_best(knn_tune, metric = "rmse")
cat("The best parameters for KNN:\n"); print(knn_best)


knn_final <- knn_wf %>%
  finalize_workflow(knn_best) %>%
  fit(data = train_data)
```

The best KNN model is a model with parameters: - neighbors = 3 - weight_func = gaussian - dist_power = 1

```{r}
# get metrics on train data
knn_train_preds <- predict(knn_final, new_data = train_data) %>%
  bind_cols(train_data %>% select(comb08))

knn_train_metrics <- knn_train_preds %>%
  metrics(truth = comb08, estimate = .pred) %>%
  mutate(split = "train")

# get metrics on test data
knn_pred <- predict(knn_final, new_data = test_data) %>%
  bind_cols(test_data %>% select(comb08))

knn_test_metrics <- knn_pred %>%
  metrics(truth = comb08, estimate = .pred) %>%
  mutate(split = "test")

bind_rows(knn_train_metrics, knn_test_metrics) %>%
  select(split, .metric, .estimate) %>%
  pivot_wider(names_from = split, values_from = .estimate)

# check predictions in the first rows
head(knn_pred)
```

The KNN model achieved strong predictive performance on both training and test data. Predicted `comb08` values were again very close to the actual observed values.

On the training data, the model reached an RMSE of 1.12 and an MAE of 0.60, indicating very low prediction errors. On the test dataset, performance remained strong, with an RMSE of 1.69 and an MAE of 0.98. The moderate increase in error from training to test data suggests good generalization with mild overfitting.

The model explained 96.2% of the variance in the training data and 91.4% of the variance in the test data, based on the R² values. This indicates that the KNN model captured the relationships between predictors and fuel economy extremely well.

```{r}
ggplot(knn_pred, aes(x = comb08, y = .pred)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "KNN: Actual vs Predicted",
    x = "Actual comb08",
    y = "Predicted comb08"
  ) +
  theme_minimal()

knn_pred <- knn_pred %>%
  mutate(residual = comb08 - .pred)

ggplot(knn_pred, aes(x = .pred, y = residual)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Residual Plot",
    x = "Predicted comb08",
    y = "Residuals"
  ) +
  theme_minimal()
```

The actual vs. predicted plot as well as the residuals plot is very similar to random forests plots. The model has strong predictive performance for standard vehicles, however it tends to underestimate more efficient vehicles with higher MPG.

## Scenario 3: Feature Selection

We have already removed redundant, leaking and uninformative features during preprocessing. Feature selection extends the analysis by determining which of the remaining 22 features truly contribute to predicting the target variable and which can be removed without a significant decrease in performance. We compare one algorithmic method (forward selection) and two embedded methods (LASSO and Elastic Net), then evaluate how many features each retains and whether the reduced feature set preserves predictive performance compared to models trained in previous tasks.

Before applying any formal feature selection method, we first examined feature importance scores from Decision tree and a Random forest models trained on the full feature set. This will provide an initial indication of which predictors carry the most predictive signal.

```{r}
dt_importance <- dt_final %>%
  extract_fit_parsnip() %>%
  vip(num_features = 22) +
  labs(title = "Feature importance - Decision tree") +
  theme_minimal()

rf_importance <- rf_final %>%
  extract_fit_parsnip() %>%
  vip(num_features = 22) +
  labs(title = "Feature importance - Random forest") +
  theme_minimal()

dt_importance + rf_importance + plot_layout(ncol = 2, nrow = 1) 
```

Both models agree strongly on the top features: `displ` (engine displacement) and `cylinders` consistently rank as the two most important predictors by a large margin, suggesting that engine size is the primary driver of fuel efficiency. Features `drive_Front.Wheel.Drive` and `year` also appear prominently in both models, indicating that drivetrain type and vehicle age contribute meaningfully to predictions. In contrast, manufacturer-level dummy variables such as `make_GMC`, `make_BMW`, `make_Dodge`, and `make_Chevrolet` show near-zero importance in both models, suggesting they add little predictive value.

Overall, the importance rankings are largely consistent across both methods, which increases confidence in the signal and motivates retaining the top features while being selective about low-importance categorical dummies in subsequent selection steps. This gives us a clear starting point for feature selection.

### Algorithmic method

#### Forward stepwise selection

Forward stepwise selection is an algorithmic feature selection method that incrementally builds a regression model. Starting from a null model with no predictors, it evaluates all candidate variables at each step and adds the one that most improves model fit, measured by AIC (Akaike Information Criterion). The process continues until no remaining variable reduces the AIC further. Among the three algorithmic approaches (forward, backward and mixed selection) forward selection was chosen because it is a starting point when there is no strong prior knowledge about which predictors are relevant. Backward selection requires fitting a full model first, which can be unstable when some predictors are correlated, as seen in our EDA where `displ` and `cylinders` showed high multicollinearity. Mixed selection combines both directions but adds unnecessary complexity.

```{r}
null_model <- lm(comb08 ~ 1, data = train_processed)
full_model <- lm(comb08 ~ ., data = train_processed)

cat("\nNull model AIC:", AIC(null_model), "\n")
cat("Full model AIC:", AIC(full_model), "\n")
```

The null model, containing only an intercept and no predictors, achieved an AIC of 245612.9. The full model using all available predictors had AIC of 202102.7, which is a substantial improvement indicating that the predictors collectively carry meaningful information about fuel efficiency.

```{r}
cat("\n=== Forward stepwise selection ===\n")
forward_model <- stepAIC(
  null_model,
  scope     = list(lower = null_model, upper = full_model),
  direction = "forward",
  trace     = 1
)
```

The forward selection algorithm started from the null model and added predictors one at a time, retaining each variable only if it reduced the AIC. The first variable added was `displ`, which produced the largest single AIC drop from 135 685 to 109 690, confirming it as the strongest individual predictor of fuel efficiency. Subsequent steps added `year`, `drive_Front.Wheel.Drive`, `fuel_group_Gasoline`, and further variables with progressively reduced AIC. The algorithm terminated after 21 steps when no remaining variable could further reduce the AIC.

```{r}
cat("\n=== Result of forward selection ===\n")
forward_summary <- summary(forward_model)
print(forward_summary)
```

The final model retained 21 out of the available predictors. The model explains 67.5% of variance in `comb08` (R² = 0.675), with all 21 selected variables being statistically significant at the 0.001 level. The strongest negative predictors are `displ` with estimate −2.25 and `has_discrete_gears` with estimate −1.72, while the largest positive predictors are `drive_Front.Wheel.Drive` with estimate 1.46 and `year` with estimate 0.95.

```{r}
selected_vars_forward <- names(coef(forward_model))[-1]  # without intercept
cat("\nNumber of selected variables:", length(selected_vars_forward), "\n")
cat("Selected variables:\n")
print(selected_vars_forward)

all_vars_forward <- names(train_processed %>% 
                        select(-comb08))
removed_vars_forward <- setdiff(all_vars_forward, selected_vars_forward)
cat("\nRemoved variables:\n")
print(removed_vars_forward)
```

The model retained 21 predictors and removed the `VClass_Two.Seaters` variable.

```{r}
forward_train_preds <- predict(forward_model, newdata = train_processed)

forward_train_metrics <- tibble(
  truth    = train_processed$comb08,
  estimate = forward_train_preds
) %>%
  metrics(truth = truth, estimate = estimate) %>%
  mutate(split = "train")

forward_preds <- predict(forward_model, newdata = test_processed)

forward_test_metrics <- tibble(
  truth    = test_processed$comb08,
  estimate = forward_preds
) %>%
  metrics(truth = truth, estimate = estimate) %>%
  mutate(split = "test")

bind_rows(forward_train_metrics, forward_test_metrics) %>%
  select(split, .metric, .estimate) %>%
  pivot_wider(names_from = split, values_from = .estimate)
```

On the test set the model achieved RMSE = 3.28, MAE = 2.16, and R² = 0.678. Comparing train and test metrics shows practically no difference, indicating that the model generalizes well with no signs of overfitting. However, compared to the random forest model (RMSE = 2.56, R² = 0.964), the linear forward selection model is considerably weaker, suggesting that the relationship between predictors and fuel efficiency contains substantial non-linear structure that a linear model cannot capture.

```{r}
tibble(actual = test_processed$comb08,
       predicted = forward_preds) %>%
  ggplot(aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Forward selection: Actual vs. Predicted",
    x     = "Actual comb08",
    y     = "Predicted comb08"
  ) +
  theme_minimal()

tibble(predicted = forward_preds,
       residual  = test_processed$comb08 - forward_preds) %>%
  ggplot(aes(x = predicted, y = residual)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Residual Plot",
    x     = "Predicted comb08",
    y     = "Residuals"
  ) +
  theme_minimal()

aic_progress <- as.data.frame(forward_model$anova) %>%
  rownames_to_column("step") %>%
  mutate(step_num = row_number())

ggplot(aic_progress, aes(x = step_num, y = AIC)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  labs(
    title = "Forward selection - AIC decrease during selection",
    x     = "Number of added variables",
    y     = "AIC"
  ) +
  theme_minimal()
```

The Actual vs Predicted plot shows that the forward selection model captures the general positive relationship between actual and predicted `comb08`, especially for common values around 10–30 MPG. However, the model underpredicts highly fuel-efficient vehicles. This suggests the linear model cannot fully capture more complex relationships in the data.

The Residual plot supports this conclusion. Residuals are not randomly scattered around zero and their spread increases for higher predicted values, indicating heteroscedasticity. Several large positive residuals show cases where actual fuel efficiency is much higher than predicted.

The third graph illustrates the decrease of AIC during forward selection. A sharp decrease in AIC is visible during the first few steps, especially between steps 1 and 5. This suggests that these are likely the strongest predictors of `comb08`. After approximately steps 6 to 10, the decline in AIC becomes much smaller. Each additional predictor still improves the model slightly, but the gains are modest compared with the first selected features. From around steps 12 to 15 onward, the curve becomes almost flat. This means that adding more predictors provides little additional information.

#### Summary

Forward stepwise selection identified 21 statistically significant predictors of `comb08` out of all available candidates. The most influential variables added in the early steps were `displ`, `year`, `drive_Front.Wheel.Drive`, and `fuel_group_Gasoline`, which together account for the largest share of explained variance. The final model achieved R² = 0.678 and RMSE = 3.28 on the test set, with no signs of overfitting as train and test metrics were nearly identical. Compared to the non-parametric models trained earlier, forward selection performs considerably worse - Random Forest achieved RMSE = 2.56 and R² = 0.964 on the same data. This gap highlights the limitation of linear feature selection methods when the underlying relationships are non-linear. However, forward selection offers a clear advantage in interpretability, because each coefficient has a direct meaning and statistical significance can be assessed for every predictor, which is not possible with Random Forest or KNN.

### Embedded methods

Embedded methods integrate feature selection into model training itself — the algorithm simultaneously learns which features matter and what their coefficients are. This differs from filter methods (which score features independently of any model) and wrapper methods (which train a model for each candidate subset).

Two models from Scenario 2 are embedded methods by design:

-   **LASSO** (L1 penalty): the L1 geometry causes the optimization to land exactly at zero for weak predictors. The regularization path records which features enter or leave the model as λ grows.
-   **Elastic Net** (L1 + L2): inherits LASSO's sparsity while distributing weight more evenly among correlated predictors like `displ` and `cylinders`.

OLS has no regularization — no selection. Ridge (L2 only) shrinks coefficients toward zero but never reaches exactly zero — not an embedded selector.

Because both models were already fitted in Scenario 2, no new training is needed. We inspect their embedded selection behavior directly.

#### 3.1 Which Scenario 2 models qualify as embedded?

We classify the four Scenario 2 models to justify why LASSO and Elastic Net are selected for Scenario 3.

```{r}
model_classification <- tibble(
  model        = c("Linear Regression", "Ridge", "LASSO", "Elastic Net"),
  penalty_type = c("None", "L2 only", "L1", "L1 + L2"),
  zeros_coefs  = c("No", "No", "Yes", "Yes"),
  embedded     = c("No", "No", "YES", "YES")
)

cat("=== Scenario 2 models — embedded method classification ===\n")
print(model_classification)
cat("\nSelected for Scenario 3: LASSO and Elastic Net\n")
cat("(best-performing embedded-eligible models from Scenario 2 — test RMSE ≈ 3.36 MPG)\n")
```

#### 3.2 LASSO — Regularization Path

The regularization path shows how many features the model keeps as the penalty λ grows. At λ → 0 all features are retained (same as OLS). As λ increases, the L1 geometry forces progressively more coefficients to exactly zero. The CV-optimal λ is marked — this is where Scenario 2's cross-validation stopped.

```{r}
lasso_glmnet  <- lasso_fit %>% extract_fit_engine()
opt_log_lasso <- log10(best_lasso$penalty)

path_lasso <- tibble(
  log_lambda = log10(lasso_glmnet$lambda),
  n_selected = lasso_glmnet$df   # non-zero coefficients at each lambda
)

cat("=== LASSO — regularization path summary ===\n")
cat("Lambda range: [", round(min(lasso_glmnet$lambda), 6), ",",
                       round(max(lasso_glmnet$lambda), 3), "]\n")
cat("Features at λ_min (≈ OLS):   ",
    path_lasso$n_selected[which.min(path_lasso$log_lambda)], "\n")
cat("Features at λ_opt (CV best): ",
    path_lasso$n_selected[which.min(abs(path_lasso$log_lambda - opt_log_lasso))], "\n")
cat("Features at λ_max (max reg): ",
    path_lasso$n_selected[which.max(path_lasso$log_lambda)], "\n")

ggplot(path_lasso, aes(x = log_lambda, y = n_selected)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_vline(xintercept = opt_log_lasso, color = "tomato", lty = 2) +
  annotate("text",
           x     = opt_log_lasso + 0.15,
           y     = max(path_lasso$n_selected) * 0.95,
           label = paste0("λ_opt = ", round(best_lasso$penalty, 5)),
           color = "tomato", hjust = 0, size = 3.5) +
  labs(
    title    = "LASSO — number of selected features along the regularization path",
    subtitle = "Red dashed line = CV-optimal λ",
    x        = "log₁₀(λ)",
    y        = "Number of non-zero coefficients"
  ) +
  theme_minimal()
```

The path shows what LASSO does as an embedded selector: at low λ all features are kept; as λ grows, the weakest predictors drop out one by one. The CV-optimal λ sits at the left end of the path — meaning the 5-fold CV found that keeping all (or nearly all) features produces the lowest prediction error on this dataset.

#### 3.3 LASSO — Selected Features at Optimal λ

These are the predictors LASSO kept at the CV-optimal penalty. A zero coefficient means the model decided that feature adds no useful information given the others.

```{r fig.height=6}
lasso_coefs_s3 <- lasso_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(term != "(Intercept)")

n_kept_lasso   <- sum(lasso_coefs_s3$estimate != 0)
n_zeroed_lasso <- sum(lasso_coefs_s3$estimate == 0)

cat("=== LASSO — embedded feature selection result ===\n")
cat("Total features available:    ", nrow(lasso_coefs_s3), "\n")
cat("Features selected (non-zero):", n_kept_lasso, "\n")
cat("Features removed (zeroed):   ", n_zeroed_lasso, "\n")

if (n_zeroed_lasso > 0) {
  cat("\nZeroed predictors:\n")
  print(lasso_coefs_s3 %>% filter(estimate == 0) %>% pull(term))
}

lasso_coefs_s3 %>%
  filter(estimate != 0) %>%
  slice_max(abs(estimate), n = 20) %>%
  ggplot(aes(x = reorder(term, abs(estimate)), y = estimate,
             fill = estimate > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
    labels = c("TRUE" = "Positive", "FALSE" = "Negative"),
    name   = "Direction"
  ) +
  labs(
    title    = paste0("LASSO — selected features at λ = ",
                      round(best_lasso$penalty, 5)),
    subtitle = paste0(n_kept_lasso, " features retained, ",
                      n_zeroed_lasso, " removed"),
    x        = NULL,
    y        = "Coefficient estimate"
  ) +
  theme_minimal()
```

#### 3.4 Elastic Net — Regularization Path

Elastic Net mixes L1 and L2 penalties. The optimal mixture was found by CV in Scenario 2 (jointly tuning both `penalty` and `mixture`). We show its path using the same approach as LASSO.

```{r}
en_glmnet  <- en_fit %>% extract_fit_engine()
opt_log_en <- log10(best_en$penalty)

path_en <- tibble(
  log_lambda = log10(en_glmnet$lambda),
  n_selected = en_glmnet$df
)

cat("=== Elastic Net — regularization path summary ===\n")
cat("Optimal penalty (λ):", round(best_en$penalty, 6), "\n")
cat("Optimal mixture (α):", round(best_en$mixture, 4),
    " (0 = Ridge, 1 = LASSO)\n")
cat("Features at λ_min (≈ OLS):   ",
    path_en$n_selected[which.min(path_en$log_lambda)], "\n")
cat("Features at λ_opt (CV best): ",
    path_en$n_selected[which.min(abs(path_en$log_lambda - opt_log_en))], "\n")
cat("Features at λ_max (max reg): ",
    path_en$n_selected[which.max(path_en$log_lambda)], "\n")

ggplot(path_en, aes(x = log_lambda, y = n_selected)) +
  geom_line(color = "#27AE60", linewidth = 0.8) +
  geom_vline(xintercept = opt_log_en, color = "tomato", lty = 2) +
  annotate("text",
           x     = opt_log_en + 0.15,
           y     = max(path_en$n_selected) * 0.95,
           label = paste0("λ_opt = ", round(best_en$penalty, 5)),
           color = "tomato", hjust = 0, size = 3.5) +
  labs(
    title    = "Elastic Net — number of selected features along the regularization path",
    subtitle = paste0("Red dashed line = CV-optimal λ  |  α = ",
                      round(best_en$mixture, 3), "  (0 = Ridge, 1 = LASSO)"),
    x        = "log₁₀(λ)",
    y        = "Number of non-zero coefficients"
  ) +
  theme_minimal()
```

#### 3.5 Elastic Net — Selected Features at Optimal λ

```{r fig.height=6}
en_coefs_s3 <- en_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(term != "(Intercept)")

n_kept_en   <- sum(en_coefs_s3$estimate != 0)
n_zeroed_en <- sum(en_coefs_s3$estimate == 0)

cat("=== Elastic Net — embedded feature selection result ===\n")
cat("Total features available:    ", nrow(en_coefs_s3), "\n")
cat("Features selected (non-zero):", n_kept_en, "\n")
cat("Features removed (zeroed):   ", n_zeroed_en, "\n")

if (n_zeroed_en > 0) {
  cat("\nZeroed predictors:\n")
  print(en_coefs_s3 %>% filter(estimate == 0) %>% pull(term))
}

en_coefs_s3 %>%
  filter(estimate != 0) %>%
  slice_max(abs(estimate), n = 20) %>%
  ggplot(aes(x = reorder(term, abs(estimate)), y = estimate,
             fill = estimate > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
    labels = c("TRUE" = "Positive", "FALSE" = "Negative"),
    name   = "Direction"
  ) +
  labs(
    title    = paste0("Elastic Net — selected features at λ = ",
                      round(best_en$penalty, 5)),
    subtitle = paste0(n_kept_en, " features retained, ",
                      n_zeroed_en, " removed"),
    x        = NULL,
    y        = "Coefficient estimate"
  ) +
  theme_minimal()
```

#### 3.6 Comparing LASSO vs. Elastic Net Feature Selection

We compare which features each model kept and where they agree. Features retained by both are the most robustly selected — they carry enough signal to survive both pure L1 and blended L1+L2 penalization.

```{r}
selection_comparison <- lasso_coefs_s3 %>%
  select(term, lasso = estimate) %>%
  left_join(
    en_coefs_s3 %>% select(term, en = estimate),
    by = "term"
  ) %>%
  mutate(
    lasso_selected = lasso != 0,
    en_selected    = en    != 0,
    agreement      = case_when(
      lasso_selected & en_selected   ~ "Both selected",
      lasso_selected & !en_selected  ~ "LASSO only",
      !lasso_selected & en_selected  ~ "EN only",
      TRUE                           ~ "Both zeroed"
    )
  )

cat("=== LASSO vs. Elastic Net — selection agreement ===\n")
print(table(selection_comparison$agreement))

cat("\n--- Features selected by BOTH models (top 10 by |LASSO coef|) ---\n")
selection_comparison %>%
  filter(agreement == "Both selected") %>%
  arrange(desc(abs(lasso))) %>%
  head(10) %>%
  mutate(lasso = round(lasso, 4), en = round(en, 4)) %>%
  print()

selection_comparison %>%
  filter(lasso_selected | en_selected) %>%
  slice_max(abs(lasso) + abs(en), n = 20) %>%
  pivot_longer(cols = c(lasso, en), names_to = "model", values_to = "coef") %>%
  ggplot(aes(x = reorder(term, abs(coef)), y = coef, fill = model)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_manual(
    values = c("lasso" = "steelblue", "en" = "#27AE60"),
    labels = c("lasso" = "LASSO", "en" = "Elastic Net")
  ) +
  labs(
    title    = "LASSO vs. Elastic Net — coefficient comparison",
    subtitle = "Top 20 features selected by at least one model",
    x        = NULL,
    y        = "Coefficient estimate",
    fill     = "Model"
  ) +
  theme_minimal()
```

#### 3.7 Test Set Performance

We evaluate both models on the held-out test set, showing prediction accuracy alongside the number of features used — the embedded selection framing.

```{r}
lasso_test_s3 <- lasso_fit %>%
  predict(new_data = test_data) %>%
  bind_cols(test_data %>% select(comb08))

en_test_s3 <- en_fit %>%
  predict(new_data = test_data) %>%
  bind_cols(test_data %>% select(comb08))

s3_comparison <- tibble(
  model      = c("LASSO", "Elastic Net"),
  n_features = c(n_kept_lasso, n_kept_en),
  RMSE = c(
    rmse_vec(truth = lasso_test_s3$comb08, estimate = lasso_test_s3$.pred),
    rmse_vec(truth = en_test_s3$comb08,   estimate = en_test_s3$.pred)
  ),
  MAE = c(
    mae_vec( truth = lasso_test_s3$comb08, estimate = lasso_test_s3$.pred),
    mae_vec( truth = en_test_s3$comb08,   estimate = en_test_s3$.pred)
  ),
  Rsq = c(
    rsq_vec( truth = lasso_test_s3$comb08, estimate = lasso_test_s3$.pred),
    rsq_vec( truth = en_test_s3$comb08,   estimate = en_test_s3$.pred)
  )
) %>%
  mutate(across(c(RMSE, MAE, Rsq), ~round(., 4)))

cat("=== Scenario 3 — Test set performance ===\n")
print(s3_comparison)

cat("\n=== OLS baseline (all", ncol(train_processed) - 1, "features) ===\n")
test_comparison %>%
  filter(model == "Linear Regression") %>%
  mutate(n_features = ncol(train_processed) - 1) %>%
  select(model, n_features, RMSE, MAE, Rsq) %>%
  print()
```

#### 3.8 Extension: extended penalty grid (sparsity vs. accuracy)

The original Scenario 2 grid (`penalty(range = c(-4, 0))`) topped out at λ = 1, and CV picked λ = 1e-4 — the smallest value — which zeroed no features. The regularization path plot in §3.2 hinted at what would happen at higher λ, but we never measured the prediction error along that path. Here we re-tune LASSO over a much wider grid (10⁻⁴ to 10¹) and read off, for each λ, both the number of selected features and the test RMSE — the actual sparsity-vs-accuracy tradeoff.

```{r warning=FALSE}
lasso_grid_wide <- grid_regular(penalty(range = c(-4, 1)), levels = 25)

lasso_cv_wide <- lasso_wf %>%
  tune_grid(
    resamples = folds,
    grid      = lasso_grid_wide,
    metrics   = eval_metrics
  )

cv_rmse_wide <- collect_metrics(lasso_cv_wide) %>%
  filter(.metric == "rmse") %>%
  select(penalty, cv_rmse = mean)
```

```{r warning=FALSE}
sparsity_path <- map_dfr(lasso_grid_wide$penalty, function(lam) {
  fit_l <- lasso_wf %>%
    finalize_workflow(tibble(penalty = lam)) %>%
    fit(train_data)

  betas <- fit_l %>%
    extract_fit_parsnip() %>%
    tidy() %>%
    filter(term != "(Intercept)")

  pred_l <- predict(fit_l, new_data = test_data) %>%
    bind_cols(test_data %>% select(comb08))

  tibble(
    penalty    = lam,
    log_lambda = log10(lam),
    n_selected = sum(betas$estimate != 0),
    test_rmse  = rmse_vec(truth = pred_l$comb08, estimate = pred_l$.pred)
  )
}) %>%
  left_join(cv_rmse_wide, by = "penalty")

cat("=== LASSO — sparsity vs. accuracy across λ ===\n")
print(sparsity_path %>%
        mutate(across(c(cv_rmse, test_rmse, log_lambda), ~round(., 4))))
```

```{r}
ggplot(sparsity_path, aes(x = log_lambda)) +
  geom_line(aes(y = test_rmse, color = "Test RMSE"), linewidth = 0.9) +
  geom_line(aes(y = cv_rmse,   color = "CV RMSE"),   linewidth = 0.9, lty = 2) +
  geom_vline(xintercept = log10(best_lasso$penalty),
             color = "tomato", lty = 2) +
  scale_color_manual(values = c("Test RMSE" = "steelblue", "CV RMSE" = "#27AE60")) +
  labs(
    title    = "LASSO — RMSE vs. log₁₀(λ)  (extended grid)",
    subtitle = "Red dashed line = original CV-optimal λ",
    x        = "log₁₀(λ)",
    y        = "RMSE (MPG)",
    color    = NULL
  ) +
  theme_minimal()

ggplot(sparsity_path, aes(x = n_selected, y = test_rmse)) +
  geom_line(color = "steelblue", linewidth = 0.9) +
  geom_point(color = "steelblue", size = 2) +
  labs(
    title    = "LASSO — test RMSE vs. number of selected features",
    subtitle = "Each point is one λ on the extended grid",
    x        = "Number of non-zero coefficients",
    y        = "Test RMSE (MPG)"
  ) +
  theme_minimal()
```

```{r}
n_total <- sparsity_path$n_selected[which.min(sparsity_path$log_lambda)]

representative <- sparsity_path %>%
  arrange(log_lambda) %>%
  filter(
    penalty == best_lasso$penalty |
      n_selected == n_total |
      n_selected <= 1 |
      abs(n_selected - round(n_total / 2)) ==
        min(abs(n_selected - round(n_total / 2))) |
      abs(n_selected - 10) == min(abs(n_selected - 10))
  ) %>%
  distinct(n_selected, .keep_all = TRUE) %>%
  arrange(desc(n_selected)) %>%
  mutate(across(c(cv_rmse, test_rmse, penalty), ~round(., 5)))

cat("=== Representative sparsity levels — what does shrinkage cost? ===\n")
print(representative %>%
        select(penalty, log_lambda, n_selected, cv_rmse, test_rmse))
```

The plot shows where the elbow lies: down to a moderate number of features the test RMSE barely moves, then rises sharply once λ becomes large enough to zero out genuinely informative predictors. This is the answer the original narrow-grid Scenario 3 could not give — embedded selection on this data could discard a meaningful fraction of features at almost no cost in accuracy, until it crosses the threshold where the engine-size signal itself starts being shrunk.

------------------------------------------------------------------------

### Conclusion

#### Embedded selection vs. Scenario 2 analysis

Scenario 2 used LASSO and Elastic Net as prediction models. Scenario 3 looks at the same models from a different angle: which features did they select, and what does the regularization path reveal about the feature space? The path plot is the embedded selector's "work log" — it shows which predictors are robust enough to survive increasing penalization, even if the CV-optimal λ happens to keep most features.

#### Regularization path insight

The CV-optimal λ sits near the far-left end of both paths — close to zero, meaning almost no features are removed. This is consistent with Scenario 2's finding: with 38 735 training observations and only \~60 dummy-encoded predictors, there is very little variance to reduce. Every predictor carries enough signal that the cross-validation penalty is not large enough to zero it out. The path still demonstrates the embedded selection mechanism: at higher λ values, features do drop out in order of their contribution — the path just happens to stop before that region is reached.

#### Which features were consistently selected

Features retained by both LASSO and Elastic Net at the CV-optimal λ are the most robustly embedded-selected predictors. The strongest predictors — `displ`, `cylinders`, `fuel_group_Electricity`, `drive` variants — appear with large coefficients in both models. Where the models differ slightly in coefficient magnitudes, the L2 component of Elastic Net distributes weight more evenly among correlated features (particularly `displ` and `cylinders` with r ≈ 0.91 in the EDA).

#### Performance vs. number of features

Both embedded models produce identical or near-identical RMSE to OLS with all features. This confirms that even where LASSO zeroes a predictor, the removed feature was genuinely redundant — the prediction does not suffer. Embedded selection here acts as a validator: it endorses the full feature set rather than pruning it, which is itself a meaningful result given the manual feature engineering done in preprocessing.

#### Suitability

LASSO is the simpler embedded selector — clean, interpretable, sparse when the data supports it. Elastic Net is a better fit for this dataset because several predictors are correlated (`displ`/`cylinders`, multiple `drive` dummies) — the L2 component stabilizes their joint selection. The CV-tuned mixture value shows how much L2 the data preferred. The same fundamental limitations apply as in Scenario 2: non-linear engine-size effects, PHEV driving-mode behavior, and missing weight/drag features are not addressable by changing the regularization strategy.

------------------------------------------------------------------------

## Scenario 5: Dimensionality Reduction

This scenario was not the primary focus of our analysis, as the dataset is not highly multidimensional. After preprocessing, the data contains 23 features, which does not typically require dimensionality reduction techniques. Nevertheless, we included it to explore dimensionality reduction techniques such as PCA and to assess their effect on model performance, interpretability and information loss.

### Principal Component Analysis (PCA)

PCA is a dimensionality reduction technique that transforms the original feature space into a new set of variables called principal components. These components are linear combinations of the original features and are constructed in such a way that each successive component captures the maximum remaining variance in the data while being orthogonal to the previous ones.

As a result, PCA rotates and reprojects the original feature space into a new coordinate system, where the first few components capture most of the variability in the data. This allows the dataset to be represented with fewer dimensions while preserving as much information as possible.

```{r}
preprocess_recipe <- recipe(comb08 ~ ., data = train_data) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%
  step_impute_median(displ, cylinders) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_nzv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

prep_preprocess_recipe <- prep(preprocess_recipe, training = train_data)
train_preprocessed <- bake(prep_preprocess_recipe, new_data = NULL)
```

```{r}
pca_input <- train_preprocessed %>%
  select(-comb08)   # remove target

cat("Size of PCA input:", nrow(pca_input), "×", ncol(pca_input), "\n")
```

```{r}
pca_result <- prcomp(pca_input, center = FALSE, scale. = FALSE)
# center=FALSE a scale.=FALSE because train_preprocessed is already normalised
```

```{r}
explained_var <- summary(pca_result)$importance
cat("\nFirst 10 components:\n")
print(explained_var[, 1:10])
```

The PCA was performed on the preprocessed training data, which had already been normalized and dummy-encoded. The first principal component explains approximately 13.6% of the total variance, while the second accounts for 11,97%. The explained variance decreases gradually across subsequent components, with no single dominant component, what suggests that the information in the dataset is moderately distributed across multiple dimensions rather than being concentrated in a few directions. The first 10 components together capture about 68.9% of the total variance.

```{r}
cum_var <- as.data.frame(t(explained_var)) %>%
  rownames_to_column("PC") %>%
  setNames(c("PC", "std_dev", "prop_var", "cum_var")) %>%
  mutate(PC_num = as.integer(str_extract(PC, "\\d+")))
```

```{r}
for (threshold in c(0.80, 0.90, 0.95)) {
  n <- min(which(cum_var$cum_var >= threshold))
  cat(sprintf("Number of PC for %.0f%% of variance: %d\n", threshold * 100, n))
}
```

To reach 80%, 90%, and 95% of cumulative variance, 13, 16, and 18 components are required respectively. The relatively high number of components needed to explain a substantial portion of variance is consistent with the high-dimensional nature of the dataset after dummy encoding, where many categorical variables were expanded into binary columns.

```{r}
p1 <- cum_var %>%
  filter(PC_num <= 30) %>%
  ggplot(aes(x = PC_num, y = prop_var)) +
  geom_col(fill = "steelblue") +
  geom_line(aes(group = 1), color = "darkblue", linewidth = 0.8) +
  geom_point(color = "darkblue", size = 2) +
  scale_x_continuous(breaks = 1:30) +
  labs(
    title = "Scree Plot - explained variance ratio",
    x     = "Principal component",
    y     = "Variance ratio"
  ) +
  theme_minimal()

p2 <- cum_var %>%
  filter(PC_num <= 30) %>%
  ggplot(aes(x = PC_num, y = cum_var)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_hline(yintercept = c(0.80, 0.90, 0.95),
             linetype = "dashed",
             color = c("orange", "red", "darkred")) +
  annotate("text", x = 22, y = 0.81, label = "80%", color = "orange",    size = 3) +
  annotate("text", x = 22, y = 0.91, label = "90%", color = "red",       size = 3) +
  annotate("text", x = 22, y = 0.96, label = "95%", color = "darkred",   size = 3) +
  scale_x_continuous(breaks = seq(0, 23, 5)) +
  labs(
    title = "Cummulative explained variability",
    x     = "Number of principal components",
    y     = "Cummulative variance ratio"
  ) +
  theme_minimal()

p1 + p2
```

The scree plot shows that the first principal component explains the largest share of variance (\~14%), followed by a notable drop to the second component (\~12%). After the third component, the decline becomes more gradual and the curve flattens, forming a characteristic elbow shape. This indicates that no single component strongly dominates, and the remaining variance is spread relatively evenly across the subsequent components.

The cumulative variance plot confirms this pattern. The curve rises steadily without any sharp jumps, which is typical for datasets with many moderately correlated features after dummy encoding. The three reference thresholds show that 80% of variance is reached at 13 components, 90% at 16 components, and 95% at 18 components.

```{r}
n_components <- min(which(cum_var$cum_var >= 0.95))
cat("\nWe will use", n_components, "components (95% of variance)\n")
```

For subsequent modeling, 18 principal components were retained, capturing 95% of the total variance. This threshold was chosen as a balance between dimensionality reduction and information retention - discarding the remaining 5% removes noise and redundancy while keeping the majority of the signal present in the original features. The reduction was small but still useful given the relatively small original feature space after preprocessing.

**Discussion on explainability and information loss** After applying PCA, the original features are replaced by principal components (PC1, PC2,...), which are combinations of multiple variables. Since these components do not have a direct or intuitive interpretation, model explainability decreases. It becomes difficult to attribute predictions to specific original features or to understand how individual variables influence the outcome.

In addition, dimensionality reduction inevitably leads to some loss of information. While the first few components capture most of the variance, the remaining variability contained in the discarded components is lost. This may remove noise, but it can also eliminate subtle patterns that could be useful for prediction, potentially reducing model performance.

**Random forest model with PCA** Now, we create a recipe, which will be used for comparison of models.

```{r}
pca_recipe <- recipe(comb08 ~ ., data = train_data) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_impute_mode(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_nzv(all_predictors()) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_pca(all_numeric_predictors(), num_comp = n_components)
```

Now we train a random forest model using preprocessed data with pca_recipe. Although the KNN model achieved slightly better predictive performance in the previous task, the random forest model is more robust and less sensitive to overfitting.

```{r}
pca_rf_model <- rand_forest(
  trees = 200,
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")

pca_rf_workflow <- workflow() %>%
  add_recipe(pca_recipe) %>%
  add_model(pca_rf_model)

pca_rf_grid <- grid_regular(
  mtry(range = c(2, 10)),
  min_n(range = c(2, 20)),
  levels = 3
)

pca_rf_tuned <- tune_grid(
  pca_rf_workflow,
  resamples = folds,
  grid = pca_rf_grid,
  metrics   = metrics_set,
  control   = tune_ctrl
)

pca_rf_best <- select_best(pca_rf_tuned, metric = "rmse")
cat("The best parameters for random forest with PCA:\n"); print(pca_rf_best)

pca_rf_final <- pca_rf_workflow %>%
  finalize_workflow(pca_rf_best) %>%
  fit(data = train_data)
```

The best random forest model with PCA has these parameters: - trees = 200 - mtry = 10 - min_n = 11

```{r}
# get metrics on train data
pca_rf_train_preds <- predict(pca_rf_final, new_data = train_data) %>%
  bind_cols(train_data %>% select(comb08))

pca_rf_train_metrics <- pca_rf_train_preds %>%
  metrics(truth = comb08, estimate = .pred) %>%
  mutate(split = "train")

# get metrics on test data
pca_rf_pred <- predict(pca_rf_final, new_data = test_data) %>%
  bind_cols(test_data %>% select(comb08))

pca_rf_test_metrics <- pca_rf_pred %>%
  metrics(truth = comb08, estimate = .pred) %>%
  mutate(split = "test")

bind_rows(pca_rf_train_metrics, pca_rf_test_metrics) %>%
  select(split, .metric, .estimate) %>%
  pivot_wider(names_from = split, values_from = .estimate)

# check predictions in the first rows
head(pca_rf_pred)
```

The best random forest model with PCA shows good overall performance, but it is slightly less accurate than models trained on the original features.

On the training set, the model achieved an RMSE of 1.36, MAE of 0.76, and R² of 0.944, indicating a very strong fit and that it explains about 94% of the variance in `comb08`. However, performance decreases on the test set, where the model reaches an RMSE of 1.97, MAE of 1.16, and R² of 0.885. This gap suggests mild overfitting, although part of the decrease in test performance may also be attributed to information loss introduced by PCA dimensionality reduction.

The sample predictions show that the model still captures general trends in fuel consumption well, but with slightly larger errors compared to models without PCA.

Overall, the model remains reliable and captures the main structure of the data, but the PCA transformation likely removed some fine-grained information that was useful for prediction. As a result, predictive performance is slightly reduced compared to models trained on the original feature space.

```{r}
ggplot(pca_rf_pred, aes(x = comb08, y = .pred)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Random Forest with PCA: Actual vs Predicted",
    x = "Actual comb08",
    y = "Predicted comb08"
  ) +
  theme_minimal()

pca_rf_pred <- pca_rf_pred %>%
  mutate(residual = comb08 - .pred)

ggplot(pca_rf_pred, aes(x = .pred, y = residual)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Residual Plot",
    x = "Predicted comb08",
    y = "Residuals"
  ) +
  theme_minimal()
```

The Actual vs. Predicted plot shows that the model tracks the diagonal reasonably well in the lower range (10–30 MPG), where most observations are concentrated. However, as actual values increase beyond 40, predictions increasingly underestimate the true values.

The Residual plot reveals a clear funnel shape: residuals are tightly clustered near zero for low predicted values but spread substantially as predictions increase. This heteroscedasticity suggests the model struggles with higher fuel efficiency vehicles, likely because they are underrepresented in the training data after PCA transformation. The residuals are roughly centered around zero overall, indicating no systematic bias in the mid-range, but the upper tail remains problematic.

Overall, the model performs adequately for typical vehicles but loses precision for fuel-efficient outliers.

### Partial Least Squares (PLS)

PLS (Partial Least Squares) is a supervised dimensionality reduction method for regression. It finds latent components — like PCA axes, but guided by the target — that maximally **correlate with `comb08`**, not just capture variance in X. We reuse the same preprocessed data as Scenario 2, so the comparison is direct: same features, same split, different method.

#### 5.1 Fitting PLS with Cross-Validation

We fit PLS with up to 15 components and use built-in 5-fold CV to find the optimal number of components. `scale = FALSE` because `train_processed` is already normalized by `step_normalize()` in the Scenario 2 recipe.

```{r}
pls_model <- plsr(
  comb08 ~ .,
  data       = train_processed,
  ncomp      = 15,
  validation = "CV",
  segments   = 5,     # 5-fold CV, matching Scenario 2
  scale      = FALSE  # already step_normalize()'d in recipe
)

# CV RMSEP per ncomp — index 1 = 0-component (intercept-only) model
rmsep_cv <- RMSEP(pls_model)$val["CV", 1, ]

# Elbow selection: fewest components whose RMSEP is within 1% of the global minimum.
# which.min() would pick the global minimum (often an overfit plateau); this
# approach favours parsimony while still being close to optimal.
min_rmsep <- min(rmsep_cv)
opt_ncomp <- which(rmsep_cv <= min_rmsep * 1.01)[1] - 1

cat("=== PLS — CV RMSEP by number of components ===\n")
for (k in 0:min(10, length(rmsep_cv) - 1)) {
  marker <- if (k == opt_ncomp) "  <-- selected" else ""
  cat(sprintf("  ncomp = %2d  |  CV RMSEP = %.4f%s\n", k, rmsep_cv[k + 1], marker))
}
cat("\nOptimal number of components:", opt_ncomp, "\n")
```

```{r}
plot(RMSEP(pls_model),
     main = "PLS — CV RMSEP by number of components",
     xlab = "Number of components",
     ylab = "CV RMSEP (MPG)")
abline(v = opt_ncomp, col = "tomato", lty = 2)
```

The elbow shows where adding more components stops improving CV accuracy. Components past the elbow add complexity without reducing error.

#### 5.2 Explained Variance per Component

PLS components are ordered by how much covariance with `comb08` they capture. `explvar()` gives variance explained in X; `R2()` gives R² for y — the more important metric for a regression task.

```{r}
x_var <- explvar(pls_model)
y_r2  <- R2(pls_model, estimate = "CV")$val["CV", 1, -1]  # drop intercept (index 1)

cat("=== PLS — Explained variance per component ===\n")
# x_var is incremental (% of X variance added by each component) — cumulate it.
# y_r2 from R2() is already cumulative (R² when using k components total) — display directly.
cat(sprintf("%-6s  %9s  %10s  %12s\n",
    "Comp", "X var %", "Cum X %", "R²(y) CV"))
cumx <- 0
for (k in seq_len(min(opt_ncomp + 3, length(x_var)))) {
  cumx   <- cumx + x_var[k]
  marker <- if (k == opt_ncomp) "  <--" else ""
  cat(sprintf("%-6d  %9.2f  %10.2f  %12.4f%s\n",
      k, x_var[k], cumx, y_r2[k], marker))
}
```

Comp1 captures the dominant engine-size axis — the same linear combination that drove LD1 in LDA. Subsequent components pick up smaller, residual patterns.

#### 5.3 Dimensionality Reduction — Projection onto First Two Components

We project all training vehicles onto Comp1 and Comp2. This is the dimensionality reduction result: \~60 normalized features collapse into 2 coordinates. Color shows actual MPG — a clean gradient confirms Comp1 captures the efficiency axis.

```{r}
scores_df <- tibble(
  Comp1  = scores(pls_model)[, 1],
  Comp2  = scores(pls_model)[, 2],
  comb08 = train_data$comb08
)

cat("=== PLS — Score ranges (training data) ===\n")
cat("Comp1:", round(range(scores_df$Comp1), 3), "\n")
cat("Comp2:", round(range(scores_df$Comp2), 3), "\n")

ggplot(scores_df, aes(x = Comp1, y = Comp2, color = comb08)) +
  geom_point(alpha = 0.3, size = 0.6) +
  scale_color_viridis_c(option = "plasma", name = "MPG") +
  labs(
    title    = "PLS — projection of training data onto first two components",
    subtitle = "Color = actual combined MPG (purple = low, yellow = high)",
    x        = "PLS Component 1",
    y        = "PLS Component 2"
  ) +
  theme_minimal()
```

#### 5.4 Component Loadings

Loadings show which predictors drive Comp1. Negative loadings push toward lower MPG; positive loadings push toward higher MPG.

```{r fig.height=6}
loadings_df <- as_tibble(loadings(pls_model)[, 1, drop = FALSE],
                          rownames = "predictor") %>%
  rename(Comp1 = 2)

loadings_df %>%
  slice_max(abs(Comp1), n = 15) %>%
  ggplot(aes(x = reorder(predictor, Comp1), y = Comp1, fill = Comp1 > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#27AE60", "FALSE" = "#E74C3C"), guide = "none") +
  labs(
    title    = "PLS — loadings of Component 1",
    subtitle = "Top 15 predictors — green = higher MPG, red = lower MPG",
    x        = NULL,
    y        = "Comp1 loading"
  ) +
  theme_minimal()

cat("=== PLS — Top 5 predictors for Comp1 (absolute loading) ===\n")
loadings_df %>%
  slice_max(abs(Comp1), n = 5) %>%
  mutate(Comp1 = round(Comp1, 4)) %>%
  print()
```

#### 5.5 Test Set Evaluation

We predict on the held-out test set using `opt_ncomp` components and compare directly with the four Scenario 2 models — same test set, same metrics.

```{r}
pls_pred <- predict(pls_model, newdata = test_processed, ncomp = opt_ncomp)[, , 1]

pls_metrics <- tibble(
  truth = test_data$comb08,
  pred  = as.numeric(pls_pred)
)

pls_rmse <- rmse_vec(truth = pls_metrics$truth, estimate = pls_metrics$pred)
pls_mae  <- mae_vec( truth = pls_metrics$truth, estimate = pls_metrics$pred)
pls_rsq  <- rsq_vec( truth = pls_metrics$truth, estimate = pls_metrics$pred)

cat("=== PLS — Test set metrics (ncomp =", opt_ncomp, ") ===\n")
cat("RMSE:", round(pls_rmse, 4), "\n")
cat("MAE: ", round(pls_mae,  4), "\n")
cat("R²:  ", round(pls_rsq,  4), "\n")

cat("\n=== All models: PLS vs. Scenario 2 ===\n")
bind_rows(
  test_comparison,
  tibble(model = paste0("PLS (", opt_ncomp, " comp)"),
         RMSE  = pls_rmse,
         MAE   = pls_mae,
         Rsq   = pls_rsq)
) %>%
  arrange(RMSE) %>%
  print()
```

```{r}
ggplot(pls_metrics, aes(x = truth, y = pred)) +
  geom_point(alpha = 0.15, color = "steelblue", size = 0.8) +
  geom_abline(slope = 1, intercept = 0, color = "tomato", linewidth = 0.8) +
  labs(
    title = "PLS — Actual vs. Predicted (test set)",
    x     = "Actual combined MPG",
    y     = "Predicted combined MPG"
  ) +
  theme_minimal()
```

#### 5.6 Extension: PCR (unsupervised components + OLS)

PLS chooses components that maximize covariance with `comb08` — a *supervised* projection. PCA chooses components that maximize variance of the predictors alone, ignoring the target — *unsupervised*. PCR is OLS fitted on the top-k PCA components. Comparing PLS at its optimal k = 3 with PCR at the same k tells us whether the supervision actually helped on this dataset, or whether the data's principal axes already happen to align with MPG.

```{r warning=FALSE}
fit_pcr_k <- function(k) {
  rec_pcr_k <- recipe(comb08 ~ ., data = train_data) %>%
    step_unknown(all_nominal_predictors()) %>%
    step_other(all_nominal_predictors(), threshold = 0.01) %>%
    step_impute_median(displ, cylinders) %>%
    step_dummy(all_nominal_predictors()) %>%
    step_zv(all_predictors()) %>%
    step_nzv(all_predictors()) %>%
    step_normalize(all_numeric_predictors()) %>%
    step_pca(all_numeric_predictors(), num_comp = k)

  pcr_wf_k <- workflow() %>%
    add_recipe(rec_pcr_k) %>%
    add_model(linear_reg() %>% set_engine("lm"))

  fit_k  <- pcr_wf_k %>% fit(train_data)
  pred_k <- predict(fit_k, new_data = test_data) %>%
    bind_cols(test_data %>% select(comb08))

  tibble(
    k    = k,
    RMSE = rmse_vec(truth = pred_k$comb08, estimate = pred_k$.pred),
    MAE  = mae_vec( truth = pred_k$comb08, estimate = pred_k$.pred),
    Rsq  = rsq_vec( truth = pred_k$comb08, estimate = pred_k$.pred)
  )
}

pcr_results <- map_dfr(1:15, fit_pcr_k)

cat("=== PCR — test RMSE per number of components ===\n")
print(pcr_results %>% mutate(across(c(RMSE, MAE, Rsq), ~round(., 4))))
```

```{r}
pls_per_k <- tibble(
  k    = 1:15,
  RMSE = sapply(1:15, function(k) {
    p <- predict(pls_model, newdata = test_processed, ncomp = k)[, , 1]
    rmse_vec(truth = test_processed$comb08, estimate = as.numeric(p))
  })
)

ggplot() +
  geom_line(data = pcr_results, aes(x = k, y = RMSE, color = "PCR (PCA + OLS)"),
            linewidth = 0.9) +
  geom_point(data = pcr_results, aes(x = k, y = RMSE, color = "PCR (PCA + OLS)"),
             size = 2) +
  geom_line(data = pls_per_k, aes(x = k, y = RMSE, color = "PLS"),
            linewidth = 0.9) +
  geom_point(data = pls_per_k, aes(x = k, y = RMSE, color = "PLS"),
             size = 2) +
  geom_vline(xintercept = 3, color = "tomato", lty = 2) +
  scale_color_manual(values = c("PCR (PCA + OLS)" = "#27AE60", "PLS" = "steelblue")) +
  labs(
    title    = "PLS vs. PCR — test RMSE by number of components",
    subtitle = "Red dashed line = k = 3 (PLS elbow)",
    x        = "Number of components",
    y        = "Test RMSE",
    color    = "Method"
  ) +
  theme_minimal()
```

```{r}
pcr_k3 <- pcr_results %>% filter(k == 3)
pcr_k15 <- pcr_results %>% filter(k == 15)

pls_pcr_table <- tibble(
  model        = c("PLS", "PCR", "PCR", "OLS (full)"),
  n_components = c("3", "3", "15", "≈ 60 (all features)"),
  RMSE         = c(round(pls_rmse, 4),
                   round(pcr_k3$RMSE, 4),
                   round(pcr_k15$RMSE, 4),
                   test_comparison %>% filter(model == "Linear Regression") %>% pull(RMSE)),
  Rsq          = c(round(pls_rsq, 4),
                   round(pcr_k3$Rsq, 4),
                   round(pcr_k15$Rsq, 4),
                   test_comparison %>% filter(model == "Linear Regression") %>% pull(Rsq))
)

cat("=== PLS vs. PCR vs. OLS — test set comparison ===\n")
print(pls_pcr_table)
```

If PCR at k = 3 matches PLS at k = 3, the dataset's principal axes of variance already point in the direction of `comb08` — supervision adds nothing because the unsupervised projection happens to be aimed correctly. If PCR is meaningfully worse at k = 3 but catches up by k = 15, supervision genuinely helped: PLS finds the MPG-relevant direction faster than PCA does.

------------------------------------------------------------------------

#### Conclusion





*Save trained models for display in Shiny app*

```{r}
saveRDS(dt_final,     "models/dt_final.rds")
saveRDS(rf_final,     "models/rf_final.rds")
saveRDS(knn_final,    "models/knn_final.rds")
saveRDS(pca_rf_final, "models/pca_rf_final.rds")

saveRDS(cor_matrix, "models/cor_matrix.rds")

```
