---
title: "Vehicles dataset"
author: "Dominika Lukašáková, Lukáš Lovás"
date: "2026-04-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In this project, we work with a dataset of vehicles that contains various technical and categorical features describing their characteristics, such as engine specifications and fuel type. The target variable of interest is combined fuel economy (`comb08`), measured in miles per gallon (MPG). Understanding and accurately predicting fuel efficiency is important for evaluating vehicle performance and environmental impact.

We formulate the following hypothesis:

**Fuel economy can be accurately predicted based on vehicle characteristics, such as engine specifications, transmission type, vehicle size and other attributes.**

We test this hypothesis by applying a range of regression models, evaluating their predictive performance, and comparing their ability to capture patterns in the data. To investigate this problem, we focus on **Scenario 2** and **Scenario 3**. We also include Scenario 5, although it is not a primary focus, as the dataset is not highly multidimensional (22 features and 1 target variable after preprocessing).

# EDA

Vehicles dataset\
Source: <https://fueleconomy.gov/feg/ws/index.shtml>

Based on the column descriptions, we could potentially predict fuel consumption (MPG).

-   **MPG (Miles per gallon)** – a measure of fuel economy in automobiles, higher number= higher efficiency\
-   **Target variable:** `comb08`

```{r}
install.packages(c("MASS", "tidyverse", "ggcorrplot", "patchwork", "rsample", "recipes",
                     "stringr", "forcats", "tidymodels", "broom", "glmnet", "pls", "leaps",
                     "rpart", "ranger", "kknn", "vip", "parsnip", "factoextra"))
library(MASS)
library(tidyverse)
library(ggcorrplot)
library(patchwork)
library(rsample)
library(recipes)
library(stringr)
library(forcats)
library(tidymodels)
library(broom)
library(glmnet)
library(pls)
library(leaps)
library(rpart)
library(ranger)
library(kknn)
library(vip)
library(parsnip)
library(factoextra)

rm(list = ls())

# MASS also exports select(); force the dplyr version used throughout this report.
select <- dplyr::select

data <- read_csv("data/vehicles.csv")
data
```

## Basic Structure

First, we check dataset dimensions and data types.

```{r}
dim(data)
```

The dataset has 49,846 rows and 84 columns.

```{r}
glimpse(data)
```

```{r}
data %>%
  summarise(across(everything(), class))
```

The dataset contains columns of type: character, double, logical. Columns createdOn and modifiedOn are dates stored as strings. The dataset has some missing values.

## Missing Values

```{r}
sum(is.na(data))
```

The dataset contains 538 466 missing values.

# Identify columns with missing values

```{r}
missing_summary <- data %>%
  summarise(across(everything(), ~mean(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_pct") %>%
  arrange(desc(missing_pct))

missing_summary %>%
  print(n = 84)
```

Columns with more than 60% missing values:

-   `c240bDscr`, `c240Dscr`, `sCharger`, `rangeA`, `fuelType2`, `guzzler`, `evMotor`,
-   `atvType`, `tCharger`, `trans_dscr`, `startStop`, `mfrCode`

We will not use these columns and remove them later.

-   Column `eng_dscr` has \~37% missing values
-   Columns `cylinders`, `displ`, `drive`, `trany`, `fuelType1` contain only a small amount of missing data (up to \~3%), so they could potentially be imputed.

## Unique Values

Next we identify columns with unique values.

```{r}
data %>%
  summarise(across(everything(), n_distinct)) %>%
  pivot_longer(everything(), 
               names_to = "variable", 
               values_to = "n_unique") %>%
  arrange(desc(n_unique)) %>%
  print(n = 84)
```

From these we filter out non-informative columns. We remove variables with nearly all unique values, as they act as identifiers and do not provide generalizable patterns for prediction. Also variables with very low variability (binary or near-constant features) were excluded, as they provide minimal information for explaining variation in the target variable.

```{r}
data %>%
  summarise(across(everything(), n_distinct)) %>%
  pivot_longer(everything(), 
               names_to = "variable", 
               values_to = "n_unique") %>%
  arrange(desc(n_unique)) %>%
  filter(!(n_unique >= 0.9 * nrow(data))) %>%
  filter(!(n_unique <= 2)) %>%
  print(n = 84)
```

-   We should also remove `model` column, because its extremely high number of unique values makes it impractical for modeling, and `engId,` as it also represents an identifier.
-   The dataset contains a large number of variables with high cardinality, indicating a high-dimensional feature space that requires careful feature selection.

## Numerical Variables

First, we check the distribution of numerical variables.

```{r}
num_data <- data %>% 
  select(where(is.numeric)) %>% 
  select(-id, -engId)

summary(num_data)
```

-   Many columns have median = 0 because most vehicles do not have that feature, for example `highwayE` applies only to electric vehicles. So, these variables might be informative for specific subsets, but not for the general fuel consumption prediction.

-   Some columns have min and median = -1, which is unusual and likely represents missing values. The columns are: `co2`, `co2A`, `feScore`, `ghgScore`, `ghgScoreA`.

-   Although column `youSaveSpend` does not have issues in terms of distribution, its definition reveals that it is directly derived from fuel consumption and fuel cost variables. As a result, it contains information closely related to the target variable and could introduce data leakage and artificially inflating model performance.

For further analysis we will remove: - columns with median = 0 or -1 - column `youSaveSpend`

```{r}
num_data <- num_data %>%
  select(where(~ {
    med <- median(., na.rm = TRUE)
    !(med == 0 | med == -1)
  })) %>%
  select(-youSaveSpend)

summary(num_data)
```

-   We are left with 11 numerical columns.

-   The similarity between mean and median values suggests that most variables are not heavily skewed, although there might be some outliers.

```{r}
num_data %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~variable, scales = "free") +
  theme_minimal()
# A warning is displayed because of NA in columns cylinders and displ
```

Most variables show right-skewed distributions, with the majority of observations with lower values and with a long tail toward higher values.

A few variables, such as `barrels08` and `co2TailpipeGpm` appear approximately normally distributed, though still with slight positive skew. In contrast, `cylinders` is a discrete variable, with clear peaks at common engine configurations - 4, 6, and 8 cylinders.

The variable `year` shows a relatively even but slightly increasing distribution toward more recent years, indicating a higher representation of newer vehicles in the dataset.

```{r}
num_data %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "value") %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()
# A warning is displayed because of NA in columns cylinders and displ
```

-   The variable `fuelCost08` shows a wide IQR and numerous outliers.
-   Column `co2TailpipeGpm` has a relatively compact central distribution but still includes several higher outliers.
-   The `year`column has a narrow spread from 1984 to 2026, indicating that most observations are concentrated within a similar time range, so there are no outliers.

We will plot another boxplot without columns `year`, `fuelCost08`, and `co2TailpipeGpm` due to differences in scale:

```{r}
num_data %>%
  select(-year, -fuelCost08, -co2TailpipeGpm) %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "value") %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()
# A warning is displayed because of NA in columns cylinders and displ
```

-   All columns have a significant amount of outsiders on the right side.

The boxplots reveal that most variables are right-skewed, as indicated by longer upper whiskers and a large number of high-value outliers. This pattern is especially clear for fuel economy variables (`city08`, `highway08`, `comb08`, `UCity`, `UHighway`), where a small number of highly efficient vehicles extend the upper range.

Variables such as `barrels08` and `displ` exhibit moderate right skewness with some extreme values. In contrast, `cylinders` is a discrete variable, making its boxplot less informative and leading to apparent outliers due to limited possible values.

Overall, the boxplots highlight skewness and the presence of outliers.

## Correlation Analysis

```{r}
cor_matrix <- num_data %>%
  select(where(is.numeric)) %>%
  cor(use = "pairwise.complete.obs")

cor_matrix
ggcorrplot(cor_matrix)
```

-   Target `comb08` is highly positively correlated with:

    -   `city08`, `highway08`, `UCity`, `UHighway` (up to \~0.996), so the usage of these columns in a model would cause data leakage

-   Target `comb08` is strongly negatively correlated with:

    -   `barrels08`, `co2TailpipeGpm`, `cylinders`, `displ`, `fuelCost08` (up to \~ -0.7)
    -   however we know from the columns description that columns `barrels08`, `co2TailpipeGpm` and `fuelCost08` are derived from fuel consumption, so they would also cause data leakage
    -   `cylinders` and `displ` are suitable predictors but they are highly correlated with each other (\~0.91)

So, for the model, we will keep only column **`displ`**, because it has fewer missing values and a stronger correlation with target (\~ -0.71), and column `year`, because it has a moderate positive correlation (\~0.32) with target and has almost no correlation with `displ`.

## Categorical Variables

```{r}
cat_data <- data %>% 
  select(where(is.character))

ncol(cat_data)

names(cat_data)
```

The dataset has 23 categorical variables.

```{r}
cat_overview <- cat_data %>%
  summarise(across(everything(), ~n_distinct(.))) %>%
  pivot_longer(everything(),
               names_to = "variable",
               values_to = "n_levels")

cat_overview
```

Based on the number of unique values and the description of each column, we will further explore only `fuelType`, `drive`, `VClass` and `trany` columns, because they have rational amount of categories and are related to fuel efficiency.

```{r}
data %>%
  select(fuelType, drive, VClass, trany) %>%
  pivot_longer(everything(),
               names_to = "variable",
               values_to = "category") %>%
  filter(!is.na(category)) %>%
  group_by(variable, category) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(variable) %>%
  slice_max(n, n = 10) %>%
  ggplot(aes(x = reorder(category, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  facet_wrap(~variable, scales = "free") +
  theme_minimal()
```

The `fuelType` variable shows a highly imbalanced distribution, with a few dominant categories (`Regular` and `Premium`) accounting for most observations, while several alternative fuel types appear only rarely. We will group rare categories of the fuelType variable into an "Other" class.

The `drive` variable is moderately imbalanced. `Front-Wheel Drive` and `Rear-Wheel Drive` represent the majority of observations, while several four-wheel-drive variants and two-wheel-drive vehicles occur much less frequently. The presence of multiple related four-wheel-drive labels indicates potential redundancy in category definitions, so merging semantically similar categories could reduce noise.

The `trany` variable contains many transmission categories, with a clear dominance of `Automatic 4-speed` and `Manual 5-speed` types.

The VClass has 3 dominant categories - `Compact Cars`, `Midsize Cars`, and `Subcompact Cars`. The other catogories appear less frequently. Since vehicle class captures structural differences in size, it is expected to be an informative predictor and should be retained in the modeling stage.

```{r}
ggplot(data, aes(x = fuelType, y = comb08)) +
  geom_boxplot() +
  coord_flip()
```

The boxplot shows how different fuel type influences fuel consumption. We can see that the electricity type has higher values then gasoline types.

```{r}
data %>%
  filter(!is.na(drive)) %>%
  ggplot(aes(x = drive, y = comb08)) +
  geom_boxplot() +
  coord_flip()
```

Drive type clearly affects fuel efficiency. `Front-Wheel Drive` vehicles tend to have higher combined MPG, while `4-Wheel or All-Wheel Drive` has smaller values.

```{r}
data %>%
  filter(!is.na(VClass)) %>%
  mutate(VClass = fct_lump(VClass, n = 10)) %>%
  ggplot(aes(x = VClass, y = comb08)) +
  geom_boxplot() +
  coord_flip()
```

Vehicle class has a clear relationship with fuel efficiency. Smaller vehicles (`compact` and `subcompact cars`) tend to achieve higher MPG, while pickups, SUVs, and larger vehicles have smaller MPG values, so they are less fuel efficient.

```{r}
data %>%
  filter(!is.na(trany)) %>%
  mutate(trany = fct_lump(trany, n = 10)) %>%
  ggplot(aes(x = trany, y = comb08)) +
  geom_boxplot() +
  coord_flip()
```

Transmission type strongly differentiates vehicle efficiency. `Automatic (A1)` shows the highest MPG values, while other categories have smaller values.

## Tagret variable

```{r}
summary(data$comb08)
```

The target variable `comb08` (combined fuel efficiency) has strongly right-skewed distribution. Most vehicles have between 15 and 30 MPG, with a median value of 20 MPG. However, a long right tail is observed, extending up to 146 MPG.

```{r}
hist <- ggplot(data, aes(x = comb08)) +
  geom_histogram(bins = 40, fill = "steelblue") +
  theme_minimal()

box <- ggplot(data, aes(y = comb08)) +
  geom_boxplot(fill = "steelblue") +
  coord_flip() +
  theme_minimal()

hist + box + plot_layout(ncol = 1, nrow = 2) 
```

Based on the histogram, we can say that the distribution is not normal. The boxplot confirms the presence of numerous high-value outliers. However, these observations are not errors but represent a distinct technological group of vehicles with significantly higher efficiency.

## Preprocessing and Train/Test Split

### Column Groups to Remove

#### Columns with Too Many Missing Values (`cols_missing`)

These variables have more than 60% missing values and cannot be reliably used in a general model.

```{r}
cols_missing <- c(
  "c240bDscr", "c240Dscr", "sCharger", "rangeA", "fuelType2",
  "guzzler", "evMotor", "atvType", "tCharger", "trans_dscr",
  "startStop", "mfrCode"
)
```

#### Identifier Columns (`cols_ids`)

These variables are identifiers or overly specific strings. Using them would encourage the model to memorize specific vehicles or models.

```{r}
cols_ids <- c("id", "engId", "model")
```

#### Data Leakage Columns (`cols_leakage`)

These variables are either directly equivalent to `comb08` or derived from the same fuel consumption, emissions, or cost calculations. Including them would incorrectly inflate model performance.

```{r}
cols_leakage <- c(
  "city08", "highway08", "UCity", "UHighway",
  "barrels08", "co2TailpipeGpm", "fuelCost08", "youSaveSpend",
  "comb08U", "city08U", "highway08U",
  "combA08", "cityA08", "highwayA08",
  "combA08U", "cityA08U", "highwayA08U",
  "co2", "co2A", "co2TailpipeAGpm",
  "feScore", "ghgScore", "ghgScoreA",
  "phevCity", "phevHwy", "phevComb",
  "combE", "cityE", "highwayE",
  "combinedCD", "combinedUF",
  "cityCD", "cityUF",
  "highwayCD", "highwayUF"
)
```

#### Sparse or Invalid Columns

Many numeric columns have median = 0 or contain encoded missing values (−1). These are typically relevant only for electric or hybrid vehicles. Including them in the general regression dataset would introduce large numbers of zeros, encoded missing values, or derived information.

```{r}
cols_sparse_or_invalid <- c(
  "barrelsA08", "charge120", "charge240", "charge240b",
  "city08U", "cityA08", "cityA08U", "cityCD", "cityE", "cityUF",
  "co2", "co2A", "co2TailpipeAGpm",
  "comb08U", "combA08", "combA08U", "combE",
  "combinedCD", "combinedUF",
  "feScore", "fuelCostA08", "ghgScore", "ghgScoreA",
  "highway08U", "highwayA08", "highwayA08U",
  "highwayCD", "highwayE", "highwayUF",
  "hlv", "hpv", "lv2", "lv4", "pv2", "pv4",
  "range", "rangeCity", "rangeCityA", "rangeHwy", "rangeHwyA",
  "UCityA", "UHighwayA",
  "phevCity", "phevHwy", "phevComb"
)
```

#### Unused Columns

-   `eng_dscr` has \~37% missing values and is a free-text engine description.
-   `createdOn` and `modifiedOn` are metadata with no predictive value.
-   `baseModel` would support memorization of specific model lines rather than generalization.
-   `fuelType` and `fuelType1` are replaced by the new `fuel_group` variable.

```{r}
cols_unused <- c("createdOn", "modifiedOn", "eng_dscr", "baseModel")
```

#### Combine All Removal Lists

```{r}
cols_to_remove <- unique(c(cols_missing, cols_ids, cols_leakage,
                            cols_sparse_or_invalid, cols_unused))

cat("Total columns scheduled for removal:", length(cols_to_remove), "\n")
```

### Electric Vehicles

Electric vehicles do not have a combustion engine and therefore have no `displ` or `cylinders`. Their missing values are structurally meaningful — they signal the absence of an engine.

-   `is_electric`: TRUE if the vehicle runs on electricity only.
-   `has_combustion_engine`: TRUE if the vehicle has a combustion engine.

For EVs, `displ` and `cylinders` are set to 0 — meaning zero displacement, not a statistical fill-in.

```{r}
data_fe <- data %>%
  mutate(
    is_electric = (fuelType == "Electricity") | (fuelType1 == "Electricity"),
    is_electric = replace_na(is_electric, FALSE),
    has_combustion_engine = !is_electric,
    displ     = if_else(is_electric, 0, displ),
    cylinders = if_else(is_electric, 0, cylinders)
  )

cat("Electric vehicles:", sum(data_fe$is_electric), "\n")
cat("Combustion engine vehicles:", sum(data_fe$has_combustion_engine), "\n")
```

### Categorical Variable Simplification

#### `fuel_group` — Fuel Type Grouping

We create a new variable `fuel_group` that distributes all fuel categories into six groups that simplify using the `fuel_type` variable. `fuelType` variable contains many rare categories that don't appear often enough. If we used these categories separatelly, the model could become more unstable and less accurate.

```{r}
data_fe <- data_fe %>%
  mutate(
    fuel_group = case_when(
      fuelType %in% c("Regular", "Premium", "Midgrade")       ~ "Gasoline",
      fuelType == "Diesel"                                     ~ "Diesel",
      fuelType == "Electricity"                                ~ "Electricity",
      fuelType %in% c(
        "Premium and Electricity",
        "Regular Gas and Electricity",
        "Premium Gas or Electricity",
        "Regular Gas or Electricity",
        "Electricity and Hydrogen"
      )                                                        ~ "Hybrid/Electric mix",
      fuelType %in% c("Gasoline or E85", "Premium or E85")    ~ "Flex-fuel/E85",
      fuelType %in% c(
        "CNG", "Hydrogen",
        "Gasoline or natural gas",
        "Gasoline or propane"
      )                                                        ~ "Alternative fuel",
      TRUE                                                     ~ "Other"
    )
  )

table(data_fe$fuel_group)
```

#### `drive` — Reduce 4WD/AWD Variants

We identified multiple overlapping 4-wheel and all-wheel drive labels. We group them into three interpretable groups.

```{r}
data_fe <- data_fe %>%
  mutate(drive = case_when(
    str_detect(drive, "4-Wheel|All-Wheel") ~ "AWD/4WD",
    str_detect(drive, "Front-Wheel")       ~ "Front-Wheel Drive",
    str_detect(drive, "Rear-Wheel")        ~ "Rear-Wheel Drive",
    is.na(drive)                           ~ "Unknown",
    TRUE                                   ~ "Other"
  ))

table(data_fe$drive)
```

#### `VClass` — Keep Top 10 Categories

`VClass` has many levels. We leave the 10 most frequent and collapse the rest into "Other".

```{r}
data_fe <- data_fe %>%
  mutate(VClass = fct_lump(as.factor(VClass), n = 10, other_level = "Other"))

table(data_fe$VClass)
```

#### `trany` — Extract Important Information Into More Interpretable Features

Instead of keeping the raw high-cardinality `trany` string, we extract four meaningful features:

-   `transmission_type`: "Automatic", "Manual", or "Other"
-   `is_cvt`: TRUE if the transmission uses "continuously variable gear ratios" - no set number of gears
-   `n_gears`: number of gears; 0 means "not applicable" (CVT or non-standard transmission)
-   `has_discrete_gears`: TRUE if the number of gear steps was successfully extracted from `trany`

```{r}
data_fe <- data_fe %>%
  mutate(
    transmission_type = case_when(
      str_detect(trany, regex("^Automatic", ignore_case = TRUE)) ~ "Automatic",
      str_detect(trany, regex("^Manual",    ignore_case = TRUE)) ~ "Manual",
      TRUE ~ "Other"
    ),
    is_cvt  = str_detect(trany, regex("variable gear ratios", ignore_case = TRUE)),
    is_cvt  = replace_na(is_cvt, FALSE),
    n_gears = as.integer(str_match(trany, "(\\d+)-spd")[, 2]),
    has_discrete_gears = as.integer(!is.na(n_gears)),
    n_gears = replace_na(n_gears, 0)
  )

cat("Transmission type distribution:\n")
print(table(data_fe$transmission_type))
cat("\nCVT vehicles:", sum(data_fe$is_cvt), "\n")
cat("\nn_gears summary:\n")
print(summary(data_fe$n_gears))
cat("\nhas_discrete_gears distribution:\n")
print(table(data_fe$has_discrete_gears))
```

### Build Final Modeling Dataset

We apply all column removals. Pure battery electric vehicles are excluded from the modeling dataset before the train/test split. The target variable `comb08` represents Miles Per Gallon (MPG) for combustion-engine vehicles, but for Electric Vehicles the MPG is converted to MPGe — an artificial conversion. That does not reflect actual fuel consumption, which means including these vehicles that have two structurally different aspects in a single regression target would only harm the prediction.

```{r}
n_ev <- sum(data_fe$is_electric)

vehicles_model <- data_fe %>%
  filter(!is_electric) %>%
  select(-any_of(cols_to_remove)) %>%
  select(-any_of(c("trany", "fuelType", "fuelType1", "is_electric", "has_combustion_engine")))

cat("Electric vehicles excluded:", n_ev, "\n")
cat("Remaining vehicles:        ", nrow(vehicles_model), "\n")
```

`cylinders` is kept as a candidate predictor even though its strong correlation with `displ` (\~0.91) but both `cylinders` and `displ` should be significant enough to be kept in the final dataset because we think they will be strong predictors based on real life assumptions: `displ` reflects total engine displacement while `cylinders` captures engine architecture. From theory, both of these directly influence MPG.

### Train/Test Split

We split the data into 80% training and 20% test. The test set is only used at the very end to evaluate final model performance. We stratify on `comb08` so both sets have a similar distribution of MPG values.

```{r}
set.seed(123)

data_split <- initial_split(vehicles_model, prop = 0.8, strata = comb08)

train_data <- training(data_split)
test_data  <- testing(data_split)

cat("Train:", nrow(train_data), "×", ncol(train_data), "\n")
cat("Test: ", nrow(test_data),  "×", ncol(test_data),  "\n")
```

### Check Train/Test Distribution

```{r}
cat("Train comb08:\n"); print(summary(train_data$comb08))
cat("Test  comb08:\n"); print(summary(test_data$comb08))
```

```{r}
bind_rows(
  train_data %>% select(comb08) %>% mutate(split = "train"),
  test_data  %>% select(comb08) %>% mutate(split = "test")
) %>%
  ggplot(aes(x = comb08, fill = split)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("train" = "steelblue", "test" = "tomato")) +
  labs(
    title = "Distribution of comb08: Train vs. Test",
    x     = "Combined MPG (comb08)",
    y     = "Density"
  ) +
  theme_minimal()
```

The density curves confirm that stratification produces very similar `comb08` distributions in both sets.

### Cross-Validation Folds

All models require hyperparameter tuning, which involves comparing performance across different parameter values. We use 5-fold cross-validation for this: the training data is split into 5 folds, the model trains on 4 and is evaluated on the remaining one, combinating through all folds. This way we can estimate results without using the test set and risking data leakage.

```{r}
folds <- vfold_cv(train_data, v = 5, strata = comb08)

cat("Number of CV folds:", nrow(folds), "\n")
print(folds)
```

### Preprocessing Recipe

We define preprocessing as a `recipe` to prevent data leakage — all transformations (normalization, imputation, dummy encoding) are learned only from the training data and then applied to the test set using those saved values. This also works correctly inside cross-validation, where each fold fits its own preprocessing independently.

-   `step_novel()` — handles new factor levels that may appear in data
-   `step_unknown()` — handles missing values in nominal predictors
-   `step_other()` — collapses any residual rare categories
-   `step_impute_median(displ, cylinders)` — fills the numeric NAs in engine parameters for electric car rows; `n_gears` is intentionally set to 0 in feature engineering and does not need imputation
-   `step_dummy()` — one-hot encodes all nominal predictors; required by linear regression methods
-   `step_zv()` / `step_nzv()` — removes zero- and near-zero-variance columns, consistent with EDA findings on low-variability variables
-   `step_normalize()` — standardizes all numeric predictors; required for many models

```{r}
model_recipe <- recipe(comb08 ~ ., data = train_data) %>%
  step_novel(all_nominal_predictors()) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%
  step_impute_median(displ, cylinders) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_nzv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

model_recipe
```

### Fit and Apply Recipe

The recipe is fitted on `train_data`. The same fitted statistics (medians, modes, dummy levels) are then applied to both sets.

```{r}
prep_recipe <- prep(model_recipe, training = train_data)

train_processed <- bake(prep_recipe, new_data = NULL)
test_processed  <- bake(prep_recipe, new_data = test_data)
```

### Verify Preprocessing Output

```{r}
cat("train_processed:", nrow(train_processed), "×", ncol(train_processed), "\n")
cat("test_processed: ", nrow(test_processed),  "×", ncol(test_processed),  "\n")

cat("\nMissing values — train_processed:", sum(is.na(train_processed)), "\n")
cat("Missing values — test_processed: ", sum(is.na(test_processed)),  "\n")

cat("\ncomb08 in train_processed:", "comb08" %in% names(train_processed), "\n")
cat("comb08 in test_processed: ", "comb08" %in% names(test_processed),  "\n")

n_predictors <- ncol(train_processed) - 1
cat("\nNumber of predictors after dummy encoding:", n_predictors, "\n")
```

Both processed datasets contain no missing values in predictors. 

## Scenario 2: Parametric vs. Non-Parametric Models

### Parametric models

Parametric models assume that the relationship between predictors and the response follows a specific linear combination of weighted features.

We compare four parametric regression methods:

1.  **Linear Regression (OLS)** — the classic baseline with no regularization
2.  **Ridge Regression** — OLS plus an L2 penalty that shrinks all coefficients toward zero
3.  **LASSO** — OLS plus an L1 penalty that can shrink some coefficients to exactly zero, selecting features
4.  **Elastic Net** — a blend of Ridge and LASSO penalties, combining both approaches

 We use the 5-fold cross-validation folds for hyperparameter tuning and model comparison, and evaluate final performance on the test set.

We define the three evaluation metrics once and reuse them for every model.

```{r}
eval_metrics <- metric_set(rmse, rsq, mae)
```

------------------------------------------------------------------------

#### 1. Linear Regression (OLS)

Ordinary least squares minimizes the sum of squared residuals and imposes no constraint on coefficient size. It is our baseline model — any benefit from regularization is an improvement.

```{r}
lm_spec <- linear_reg() %>%
  set_engine("lm") %>%
  set_mode("regression")

lm_wf <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(lm_spec)
```

We wrap the recipe and model in a `workflow` so preprocessing is applied automatically during cross-validation and when predicting on new data.

```{r warning=FALSE}
lm_cv <- lm_wf %>%
  fit_resamples(
    resamples = folds,
    metrics   = eval_metrics,
    control   = control_resamples(save_pred = TRUE)
  )

lm_cv_metrics <- collect_metrics(lm_cv)
cat("=== OLS — 5-fold CV ===\n")
cat("RMSE:", round(lm_cv_metrics %>% filter(.metric == "rmse") %>% pull(mean), 4),
    "  ±", round(lm_cv_metrics %>% filter(.metric == "rmse") %>% pull(std_err), 4), "\n")
cat("MAE: ", round(lm_cv_metrics %>% filter(.metric == "mae")  %>% pull(mean), 4), "\n")
cat("R²:  ", round(lm_cv_metrics %>% filter(.metric == "rsq")  %>% pull(mean), 4), "\n")
```

Now we fit on the full training set to inspect which predictors drive the predictions.

```{r}
lm_fit <- lm_wf %>% fit(data = train_data)

lm_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(term != "(Intercept)") %>%
  slice_max(abs(estimate), n = 20) %>%
  ggplot(aes(x = reorder(term, abs(estimate)), y = estimate)) +
  geom_col(aes(fill = estimate > 0)) +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
    labels = c("TRUE" = "Positive", "FALSE" = "Negative"),
    name   = "Direction"
  ) +
  labs(
    title = "OLS: Top 20 Predictors by Absolute Coefficient",
    x     = NULL,
    y     = "Coefficient estimate"
  ) +
  theme_minimal()
```

Blue bars indicate predictors that increase `comb08`, red bars indicate predictors that decrease it.

------------------------------------------------------------------------

#### 2. Ridge Regression

Ridge regression adds an L2 penalty. This forces the optimization to keep coefficients small, reducing variance at the cost of some bias. It should effective when predictors are correlated — e.g `displ` and `cylinders` - we saw this in our EDA.

The regularization strength λ is controlled by `penalty`. We tune it using cross-validation.

```{r}
ridge_spec <- linear_reg(penalty = tune(), mixture = 0) %>%
  set_engine("glmnet") %>%
  set_mode("regression")

ridge_wf <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(ridge_spec)
```


```{r}
ridge_grid <- grid_regular(penalty(range = c(-4, 2)), levels = 50)
```

Now we define 50 penalty thresholds ranging from 0.0001 to 100.

```{r warning=FALSE}
ridge_cv <- ridge_wf %>%
  tune_grid(
    resamples = folds,
    grid      = ridge_grid,
    metrics   = eval_metrics
  )

```

At very low penalty, Ridge behaves like OLS. As penalty increases, coefficients shrink and the model has higher bias but less variance. The optimal result is where RMSE on the folds is minimized.

```{r}
best_ridge <- ridge_cv %>% select_best(metric = "rmse")

ridge_best_row <- show_best(ridge_cv, metric = "rmse", n = 1)
cat("=== Ridge — best hyperparameter ===\n")
cat("Best penalty:", round(best_ridge$penalty, 6), "\n")
cat("CV RMSE:     ", round(ridge_best_row$mean, 4),
    "  ±", round(ridge_best_row$std_err, 4), "\n")
```

```{r}
ridge_final_wf <- ridge_wf %>% finalize_workflow(best_ridge)
ridge_fit       <- ridge_final_wf %>% fit(data = train_data)
```

```{r}
ridge_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(term != "(Intercept)") %>%
  slice_max(abs(estimate), n = 20) %>%
  ggplot(aes(x = reorder(term, abs(estimate)), y = estimate)) +
  geom_col(aes(fill = estimate > 0)) +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
    labels = c("TRUE" = "Positive", "FALSE" = "Negative"),
    name   = "Direction"
  ) +
  labs(
    title = paste0("Ridge: Top 20 Predictors (penalty = ",
                   round(best_ridge$penalty, 5), ")"),
    x     = NULL,
    y     = "Coefficient estimate"
  ) +
  theme_minimal()
```

Compared to OLS, Ridge produces smaller coefficient weights across all predictors. The ordering and direction of the most important features generally remains the same, but correlated predictors share their weight more evenly.

------------------------------------------------------------------------

#### 3. LASSO Regression

LASSO (Least Absolute Shrinkage and Selection Operator) uses an L1 penalty. The key property of L1 is that the regularization can scale the coefficients to zero. LASSO performs simultaneous regression and feature selection: it automatically identifies the most informative predictors and removes the rest.

```{r}
lasso_spec <- linear_reg(penalty = tune(), mixture = 1) %>%
  set_engine("glmnet") %>%
  set_mode("regression")

lasso_wf <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(lasso_spec)
```

Giving LASSO the same penalty range as Ridge would only harm the prediction, because scaling too many predictors to 0 could introduce underfitting.

```{r}
lasso_grid <- grid_regular(penalty(range = c(-4, 0)), levels = 50)
```

```{r warning=FALSE}
lasso_cv <- lasso_wf %>%
  tune_grid(
    resamples = folds,
    grid      = lasso_grid,
    metrics   = eval_metrics
  )

```

```{r}
best_lasso <- lasso_cv %>% select_best(metric = "rmse")

lasso_best_row <- show_best(lasso_cv, metric = "rmse", n = 1)
cat("=== LASSO — best hyperparameter ===\n")
cat("Best penalty:", round(best_lasso$penalty, 6), "\n")
cat("CV RMSE:     ", round(lasso_best_row$mean, 4),
    "  ±", round(lasso_best_row$std_err, 4), "\n")
```

```{r}
lasso_final_wf <- lasso_wf %>% finalize_workflow(best_lasso)
lasso_fit       <- lasso_final_wf %>% fit(data = train_data)
```

```{r}
lasso_coefs <- lasso_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(term != "(Intercept)")

cat("=== LASSO — sparsity ===\n")
cat("Total predictors available:", nrow(lasso_coefs), "\n")
cat("Kept by LASSO:  ", sum(lasso_coefs$estimate != 0), "\n")
cat("Removed by LASSO:   ", sum(lasso_coefs$estimate == 0), "\n")
if (sum(lasso_coefs$estimate == 0) > 0) {
  cat("Zeroed out predictors:\n")
  print(lasso_coefs %>% filter(estimate == 0) %>% pull(term))
}
```

```{r}
lasso_coefs %>%
  filter(estimate != 0) %>%
  slice_max(abs(estimate), n = 20) %>%
  ggplot(aes(x = reorder(term, abs(estimate)), y = estimate)) +
  geom_col(aes(fill = estimate > 0)) +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
    labels = c("TRUE" = "Positive", "FALSE" = "Negative"),
    name   = "Direction"
  ) +
  labs(
    title = paste0("LASSO: Top 20 Non-Zero Predictors (penalty = ",
                   round(best_lasso$penalty, 5), ")"),
    x     = NULL,
    y     = "Coefficient estimate"
  ) +
  theme_minimal()
```

Predictors that survive the L1 penalty are the ones that the algorithm judged worth the cost of adding a parameter.

------------------------------------------------------------------------

#### 4. Elastic Net

Elastic Net combines both L1 and L2 penalties. The `mixture` parameter corresponds to α: `mixture = 0` gives pure Ridge, `mixture = 1` gives pure LASSO. By tuning both `penalty` and `mixture`, Elastic Net finds the best combination — it can perform feature selection like LASSO while also handling correlated predictors like Ridge.

```{r}
en_spec <- linear_reg(penalty = tune(), mixture = tune()) %>%
  set_engine("glmnet") %>%
  set_mode("regression")

en_wf <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(en_spec)
```

We use a regular 10 × 10 grid — 10 penalty values and 10 mixture values — giving 100 combinations. This should leave enough space to choose the best combination without limiting one either penalty or mixture.
```{r}
en_grid <- grid_regular(
  penalty(range = c(-4, 0)),
  mixture(range = c(0, 1)),
  levels = 10
)
```

```{r warning=FALSE}
en_cv <- en_wf %>%
  tune_grid(
    resamples = folds,
    grid      = en_grid,
    metrics   = eval_metrics
  )

```

```{r}
best_en <- en_cv %>% select_best(metric = "rmse")

en_best_row <- show_best(en_cv, metric = "rmse", n = 1)
cat("=== Elastic Net — best hyperparameters ===\n")
cat("Best penalty:", round(best_en$penalty, 6), "\n")
cat("Best mixture:", round(best_en$mixture, 4),
    " (0 = Ridge, 1 = LASSO)\n")
cat("CV RMSE:     ", round(en_best_row$mean, 4),
    "  ±", round(en_best_row$std_err, 4), "\n")
```

The optimal `mixture` value tells us how the data should to be regularized. If the best mixture is close to 1, LASSO-like behavior is preferred; if it is close to 0, Ridge-like behavior is preferred.

```{r}
en_final_wf <- en_wf %>% finalize_workflow(best_en)
en_fit       <- en_final_wf %>% fit(data = train_data)
```

------------------------------------------------------------------------

#### Comparing All Parametric Models

We evaluate all four models on the held-out test set.

```{r}
get_test_metrics <- function(fit, model_name) {
  fit %>%
    predict(new_data = test_data) %>%
    bind_cols(test_data %>% select(comb08)) %>%
    summarise(
      model = model_name,
      RMSE  = rmse_vec(truth = comb08, estimate = .pred),
      MAE   = mae_vec( truth = comb08, estimate = .pred),
      Rsq   = rsq_vec( truth = comb08, estimate = .pred)
    )
}
```

```{r}
test_comparison <- bind_rows(
  get_test_metrics(lm_fit,    "Linear Regression"),
  get_test_metrics(ridge_fit, "Ridge"),
  get_test_metrics(lasso_fit, "LASSO"),
  get_test_metrics(en_fit,    "Elastic Net")
) %>%
  arrange(RMSE)

test_comparison
```


##### Cross-Validation vs. Test Set RMSE

We compare CV RMSE with test RMSE. A small gap means no overfitting — the model generalizes to new data roughly as well as it performed during CV.

```{r}
cv_rmse_comparison <- tibble(
  model   = c("Linear Regression", "Ridge", "LASSO", "Elastic Net"),
  cv_rmse = c(
    collect_metrics(lm_cv)    %>% filter(.metric == "rmse") %>% pull(mean),
    show_best(ridge_cv, metric = "rmse", n = 1) %>% pull(mean),
    show_best(lasso_cv, metric = "rmse", n = 1) %>% pull(mean),
    show_best(en_cv,    metric = "rmse", n = 1) %>% pull(mean)
  )
) %>%
  left_join(
    test_comparison %>% select(model, test_rmse = RMSE),
    by = "model"
  ) %>%
  mutate(gap = abs(test_rmse - cv_rmse)) %>%
  arrange(cv_rmse)

cv_rmse_comparison
```

##### Actual vs. Predicted — All Models

Actual vs. predicted on the test set.

```{r}
all_test_preds <- bind_rows(
  lm_fit    %>% predict(new_data = test_data) %>% bind_cols(test_data %>% select(comb08)) %>% mutate(model = "Linear Regression"),
  ridge_fit %>% predict(new_data = test_data) %>% bind_cols(test_data %>% select(comb08)) %>% mutate(model = "Ridge"),
  lasso_fit %>% predict(new_data = test_data) %>% bind_cols(test_data %>% select(comb08)) %>% mutate(model = "LASSO"),
  en_fit    %>% predict(new_data = test_data) %>% bind_cols(test_data %>% select(comb08)) %>% mutate(model = "Elastic Net")
)

all_test_preds %>%
  ggplot(aes(x = comb08, y = .pred)) +
  geom_point(alpha = 0.15, color = "steelblue", size = 0.8) +
  geom_abline(slope = 1, intercept = 0, color = "tomato", linewidth = 0.8) +
  facet_wrap(~model) +
  labs(
    title = "Actual vs. Predicted — Test Set (Parametric Models)",
    x     = "Actual combined MPG",
    y     = "Predicted combined MPG"
  ) +
  theme_minimal()
```

Remaining scattered observations at high MPG values is mostly hybrid vehicles or vehicle models made with efficiency in mind — the model slightly underpredicts these.

------------------------------------------------------------------------

#### Conclusion

##### Model results

All four models gave nearly the same result on the test set: **RMSE ≈ 3.28 MPG, MAE ≈ 2.16 MPG, R² ≈ 0.677**. With a median `comb08` of 20 MPG, an average error of 2.16 MPG is roughly 11%.

##### Why regularization did not help

All three regularized models selected `penalty = 0.0001` — the minimum of the penalty value set. At such a small penalty, no predictor is weak enough to be zeroed out. This means the feature set is clean and every variable carries real signal.

The baseline linear models assume that each predictor contributes independently to MPG. This is likely too restrictive for our vehicle data. The effect of engine displacement may depend on cylinder count, vehicle year, or transmission structure, because the same increase in engine size can have different consequences for different engine designs. Interaction terms could allow the model to consider these relationships.

The CV–test gap is below 0.005 for all models, so there is no overfitting. The models are a decent baseline.

#### Scenario 2 Extension: Attempt to improve Parametric models with interaction terms

Since Ridge, LASSO and Elastic Net did not improve meaningfully over the OLS baseline, we extend OLS with four interpretable interaction terms. These interactions test whether engine-size effects depend on engine architecture, vehicle age and transmission structure.

-   `cylinders:displ` checks whether the MPG effect of engine displacement changes with engine architecture.
-   `displ:year` checks whether the effect of engine size changed over time, because newer engines can use more efficient technology.
-   `cylinders:year` checks whether the MPG penalty of additional cylinders differs across vehicle generations.
-   `displ:has_discrete_gears` checks whether the effect of engine size depends on transmission structure.
```{r}
rec_interact <- recipe(comb08 ~ ., data = train_data) %>%
  step_novel(all_nominal_predictors()) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%
  step_impute_median(displ, cylinders) %>%
  step_interact(terms = ~
    cylinders:displ +
    displ:year +
    cylinders:year +
    displ:has_discrete_gears
  ) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_nzv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

ols_interact_wf <- workflow() %>%
  add_recipe(rec_interact) %>%
  add_model(linear_reg() %>% set_engine("lm"))

ols_interact_fit <- ols_interact_wf %>% fit(train_data)
```

```{r}
ols_interact_pred <- predict(ols_interact_fit, new_data = test_data) %>%
  bind_cols(test_data %>% select(comb08))

ols_interact_metrics <- tibble(
  model = "OLS (4 interactions)",
  RMSE  = rmse_vec(truth = ols_interact_pred$comb08,
                   estimate = ols_interact_pred$.pred),
  MAE   = mae_vec( truth = ols_interact_pred$comb08,
                   estimate = ols_interact_pred$.pred),
  Rsq   = rsq_vec( truth = ols_interact_pred$comb08,
                   estimate = ols_interact_pred$.pred)
) %>%
  mutate(across(c(RMSE, MAE, Rsq), ~round(., 4)))

interaction_compare <- bind_rows(
  test_comparison %>% filter(model == "Linear Regression") %>%
    select(model, RMSE, MAE, Rsq) %>%
    mutate(model = "OLS (no interactions)"),
  ols_interact_metrics
)

cat("=== OLS with vs. without 4 interaction terms ===\n")
print(interaction_compare)

interaction_terms <- ols_interact_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(grepl("cylinders_x_displ|displ_x_year|cylinders_x_year|displ_x_has_discrete_gears", term))

cat("\n=== Interaction coefficients ===\n")
print(interaction_terms)
```

Adding the four interaction terms improved the OLS test performance from RMSE = 3.28 to RMSE = 3.13 and increased R² from 0.677 to 0.706. This means that the main limitation of the original linear model is its purely additive structure. The baseline OLS model assumes that the effect of each predictor is constant regardless of the values of the other predictors.

The strongest interaction is `cylinders:displ`, which is highly significant. This confirms that the MPG penalty of displacement depends on the number of cylinders.

 The interaction `displ:has_discrete_gears` is also significant, suggesting that the effect of engine size differs between vehicles with discrete gears and vehicles with CVT or other non-discrete transmissions.
 
  The `displ:year` interaction is significant as well, which indicates that the relationship between engine size and MPG changed over time, likely because newer engine technologies improved efficiency.

The only weak interaction is `cylinders:year`, whose p-value is not significant.

This result shows non-parametric regression that are more flexible should perform a lot better because they can learn non-linear and interaction effects automatically instead of requiring us to specify them.



### Non-parametric regression methods

Non-parametric models do not assume a specific relationship between predictors and the response. Instead, they learn patterns directly from the data, allowing them to capture more complex and non-linear relationships.

We compare three non-parametric regression methods:

1.  **Decision tree** - a single tree that partitions the feature space into regions with similar response values
2.  **Random Forest** - an ensemble of decision trees that improves performance by averaging multiple models
3.  **K-nearest neighbors (KNN)** - a distance-based method that predicts values based on nearby observations

These models were chosen because they can capture non-linear relationships observed in the data. Tree-based methods are robust and require minimal assumptions, while KNN can model local patterns when features are properly scaled. Overall, the characteristics of the dataset are suitable for applying these non-parametric methods.

The Decision tree and Random forest models use the same preprocessing recipe as the parametric methods, even though some steps (such as normalization) are not strictly required for tree-based models. The KNN model uses a modified preprocessing recipe due to issues with `step_nzv(all_predictors())`, and includes an additional step `step_novel(all_nominal_predictors())` to handle unseen categories.

We use 5-fold cross-validation for hyperparameter tuning and model comparison and evaluate final performance on the held-out test set. The same evaluation metrics are applied consistently across all models.

Variables used for hyperparameters tuning:

```{r}
tune_ctrl <- control_grid(
  save_pred    = TRUE,
  save_workflow = TRUE,
  verbose      = TRUE
)
```

#### Decision tree

Firstly, we will create decision tree model. The decision tree algorithm doesnt need normalised data, however we will use the prepared recipe with a normalisation step.

**Hyperparameters tuning**

```{r}
dt_spec <- decision_tree(
  cost_complexity = tune(),   # alpha used for pruning
  tree_depth      = tune(),
  min_n           = tune()
) %>%
  set_engine("rpart") %>%
  set_mode("regression")

dt_wf <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(dt_spec)

dt_grid <- grid_regular(
  cost_complexity(range = c(-4, -1)),  # log10 scale: 0.0001 - 0.1
  tree_depth(range = c(3, 15)),
  min_n(range = c(2, 20)),
  levels = 4
)

dt_tune <- tune_grid(
  dt_wf,
  resamples = folds,
  grid      = dt_grid,
  metrics   = eval_metrics,
  control   = tune_ctrl
)
```

```{r}
show_best(dt_tune, metric = "rmse", n = 5)

autoplot(dt_tune) +
  labs(title = "Decision Tree - hyperparameters tuning") +
  theme_minimal()
```

The graph shows, how the models' performance changes with different hyperparameters. Model performance was most sensitive to the cost-complexity parameter, where larger values substantially worsened RMSE and R², indicating underfitting. Deeper trees (depth 7–11) outperformed shallow trees (depth 3), and the minimum node size had only a minor effect on predictive performance.

```{r}
dt_best <- select_best(dt_tune, metric = "rmse")
cat("The best parameters for decision tree:\n"); print(dt_best)

dt_final <- dt_wf %>%
  finalize_workflow(dt_best) %>%
  fit(data = train_data)
```

The best decision tree model has these parameters: - cost_complexity = 0.0001 - tree_depth = 15 - min_n = 8

```{r}
# get metrics on train data
dt_train_preds <- predict(dt_final, new_data = train_data) %>%
  bind_cols(train_data %>% select(comb08))

dt_train_metrics <- dt_train_preds %>%
  metrics(truth = comb08, estimate = .pred) %>%
  mutate(split = "train")

# get metrics on test data
dt_preds <- predict(dt_final, new_data = test_data) %>%
  bind_cols(test_data %>% select(comb08))

dt_test_metrics <- dt_preds %>%
  metrics(truth = comb08, estimate = .pred) %>%
  mutate(split = "test")

bind_rows(dt_train_metrics, dt_test_metrics) %>%
  select(split, .metric, .estimate) %>%
  pivot_wider(names_from = split, values_from = .estimate)

# check predictions in the first rows
head(dt_preds)
```

The tuned decision tree achieved solid predictive performance and showed good generalization to unseen data. Predicted `comb08` values were close to the actual observed values, with only small deviations in the predictions.

On the training data, the model reached an RMSE of 2.01 and an MAE of 1.38, indicating low average prediction errors. On the test dataset, performance remained similar, with an RMSE of 2.15 and an MAE of 1.44. The small increase in error from training to test data suggests that the model generalizes well and does not suffer from overfitting. Based on the R² values, we can say that the model explained 87.9% of the variance in the training data and 86.1% of the variance in the test data.

```{r}
ggplot(dt_preds, aes(x = comb08, y = .pred)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Decision tree: Actual vs Predicted",
    x = "Actual comb08",
    y = "Predicted comb08"
  ) +
  theme_minimal()

dt_pred <- dt_preds %>%
  mutate(residual = comb08 - .pred)

ggplot(dt_pred, aes(x = .pred, y = residual)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Residual Plot",
    x = "Predicted comb08",
    y = "Residuals"
  ) +
  theme_minimal()
```

The Actual vs Predicted plot shows a clear positive linear relationship between observed and predicted comb08 values, indicating that the model captures the main structure of the data well. Most points are concentrated in range 15-35 MPG, these are predicted with good accuracy. For higher values of `comb08` , the model indicates lower accuracy as it tends to underpredict some most efficient vehicles. A typical step pattern is visible in the predictions, which is expected for decision trees because they assign the same predicted value to all observations within a terminal node.

From the Residual plot, we can tell that residuals are centered around zero for most observations, but residual variance increases for high predicted MPG values. The graph indicates mild heteroskedasticity.

```{r}
dt_final %>%
  extract_fit_parsnip() %>%
  vip(num_features = 15) +
  labs(title = "Feature importance - Decision tree") +
  theme_minimal()
```

The graph shows 15 most important features, which were used by model to predict target variable. The most important features are `cylinders` and `displ`.

#### Random forest

Secondly, we will create random forest model. The random forest algorithm doesnt need normalised data, however we will use the prepared recipe with a normalisation step.

Since our dataset is quite large (40 000 rows) and the `randomForest` engine is training quite slowly, we will use a `ranger` engine in hyperparameters tuning steps, because it trains much quicker.

**Hyperparameters tuning**

```{r}
rf_model <- rand_forest(
  trees = 200,
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")

rf_workflow <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(rf_model)

rf_grid <- grid_regular(
  mtry(range = c(2, 10)),
  min_n(range = c(2, 20)),
  levels = 3
)

rf_tuned <- tune_grid(
  rf_workflow,
  resamples = folds,
  grid = rf_grid,
  metrics   = eval_metrics,
  control   = tune_ctrl
)

rf_best <- select_best(rf_tuned, metric = "rmse")
cat("The best parameters for random forest:\n"); print(rf_best)

rf_final <- rf_workflow %>%
  finalize_workflow(rf_best) %>%
  fit(data = train_data)
```

The best random forest model has these parameters: - trees = 200 - mtry = 10 - min_n = 2

```{r}
# get metrics on train data
rf_train_preds <- predict(rf_final, new_data = train_data) %>%
  bind_cols(train_data %>% select(comb08))

rf_train_metrics <- rf_train_preds %>%
  metrics(truth = comb08, estimate = .pred) %>%
  mutate(split = "train")

# get metrics on test data
rf_pred <- predict(rf_final, new_data = test_data) %>%
  bind_cols(test_data %>% select(comb08))

rf_test_metrics <- rf_pred %>%
  metrics(truth = comb08, estimate = .pred) %>%
  mutate(split = "test")

bind_rows(rf_train_metrics, rf_test_metrics) %>%
  select(split, .metric, .estimate) %>%
  pivot_wider(names_from = split, values_from = .estimate)

# check predictions in the first rows
head(rf_pred)
```

The best random forest model achieved excellent predictive performance and outperformed the simpler decision tree model in terms of prediction error and explained variance. Predicted `comb08` values were predicted almost exactly.

On the training data, the model achieved an RMSE of 1.39 and an MAE of 0.76, indicating very low prediction errors. On the test dataset, performance remained strong, with an RMSE of 1.75 and an MAE of 1.03. Although the error increased slightly on unseen data, the difference between training and test results is relatively small, suggesting good generalization ability.

The model explained 94.2% of the variance in the training data and 90.9% of the variance in the test data, based on the R² values, so the underlying relationships between predictors and fuel economy were captured extremely well.

```{r}
ggplot(rf_pred, aes(x = comb08, y = .pred)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Random Forest: Actual vs Predicted",
    x = "Actual comb08",
    y = "Predicted comb08"
  ) +
  theme_minimal()

rf_pred <- rf_pred %>%
  mutate(residual = comb08 - .pred)

ggplot(rf_pred, aes(x = .pred, y = residual)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Residual Plot",
    x = "Predicted comb08",
    y = "Residuals"
  ) +
  theme_minimal()
```

The Actual vs Predicted plot shows strong agreement between observed and predicted `comb08` values, with most points close to the diagonal line. This indicates accurate predictions across a wide range of vehicles, especially in the common 15–35 MPG range. Prediction errors increase slightly for vehicles with very high fuel efficiency, but the model still remains relatively accurate in this region. There is a small tendency to underpredict the highest MPG values, although less than in the decision tree model.

From the Residual plot, we can tell that residuals are centered around zero for most observations, but residual variance increases for high predicted MPG values. The graph indicates mild heteroskedasticity.

Overall, the random forest model provides better predictive performance and stronger generalization than the decision tree, though it is less interpretable.

#### K-nearest neighbors (KNN)

Lastly, we will train also a KNN model. We will find the best model using hyperparameters tuning. As we stated before, we will modify the preprocessing recipe due to issues with `step_nzv(all_predictors())`, and include an additional step `step_novel(all_nominal_predictors())` to handle unseen categories.

```{r}
knn_recipe <- recipe(comb08 ~ ., data = train_data) %>%
  step_novel(all_nominal_predictors()) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%
  step_impute_median(displ, cylinders) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())
```

**Hyperparameters tuning**

*Note: The hyperparameter tuning for the KNN model was computationally demanding and took approximately 6 hours to complete due to the size of the dataset and cross-validation procedure.*

```{r}
knn_spec <- nearest_neighbor(
  neighbors = tune(),
  weight_func = tune(),
  dist_power  = tune()
) %>%
  set_engine("kknn") %>%
  set_mode("regression")

knn_wf <- workflow() %>%
  add_recipe(knn_recipe) %>%
  add_model(knn_spec)

# KNN hyperparameters tuning
knn_grid <- grid_regular(
  neighbors(range = c(3, 25)),
  weight_func(values = c("rectangular", "triangular", "gaussian")),
  dist_power(range = c(1, 2)),
  levels = 3
)

cat("=== Tuning KNN ===\n")
knn_tune <- tune_grid(
  knn_wf,
  resamples = folds,
  grid      = knn_grid,
  metrics   = eval_metrics,
  control   = tune_ctrl
)

knn_best <- select_best(knn_tune, metric = "rmse")
cat("The best parameters for KNN:\n"); print(knn_best)


knn_final <- knn_wf %>%
  finalize_workflow(knn_best) %>%
  fit(data = train_data)
```

The best KNN model is a model with parameters: - neighbors = 3 - weight_func = gaussian - dist_power = 1

```{r}
# get metrics on train data
knn_train_preds <- predict(knn_final, new_data = train_data) %>%
  bind_cols(train_data %>% select(comb08))

knn_train_metrics <- knn_train_preds %>%
  metrics(truth = comb08, estimate = .pred) %>%
  mutate(split = "train")

# get metrics on test data
knn_pred <- predict(knn_final, new_data = test_data) %>%
  bind_cols(test_data %>% select(comb08))

knn_test_metrics <- knn_pred %>%
  metrics(truth = comb08, estimate = .pred) %>%
  mutate(split = "test")

bind_rows(knn_train_metrics, knn_test_metrics) %>%
  select(split, .metric, .estimate) %>%
  pivot_wider(names_from = split, values_from = .estimate)

# check predictions in the first rows
head(knn_pred)
```

The KNN model achieved strong predictive performance on both training and test data. Predicted `comb08` values were again very close to the actual observed values.

On the training data, the model reached an RMSE of 1.12 and an MAE of 0.60, indicating very low prediction errors. On the test dataset, performance remained strong, with an RMSE of 1.69 and an MAE of 0.98. The moderate increase in error from training to test data suggests good generalization with mild overfitting.

The model explained 96.2% of the variance in the training data and 91.4% of the variance in the test data, based on the R² values. This indicates that the KNN model captured the relationships between predictors and fuel economy extremely well.

```{r}
ggplot(knn_pred, aes(x = comb08, y = .pred)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "KNN: Actual vs Predicted",
    x = "Actual comb08",
    y = "Predicted comb08"
  ) +
  theme_minimal()

knn_pred <- knn_pred %>%
  mutate(residual = comb08 - .pred)

ggplot(knn_pred, aes(x = .pred, y = residual)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Residual Plot",
    x = "Predicted comb08",
    y = "Residuals"
  ) +
  theme_minimal()
```

The actual vs. predicted plot as well as the residuals plot is very similar to random forests plots. The model has strong predictive performance for standard vehicles, however it tends to underestimate more efficient vehicles with higher MPG.

### Conclusion

The parametric models provide a decent interpretable baseline. All achieved very similar performance (test RMSE around 3.28 MPG). Regularization did not improve prediction because the dataset has many observations relative to the number of predictors and the manually engineered feature set is already cleaned from preprocessing. The OLS with interaction terms improved the result to RMSE around 3.13, hinting that some relationships are not meant to be explained by a simple addition. Non-parametric model results later confirm this.

The non-parametric models performed substantially better, because they were able to capture local patterns and non-linear relationships between engine size, drivetrain, year and MPG. KNN achieved the strongest test performance among the tested models, while Random forest offered a better balance between accuracy, robustness and feature importance interpretation. The improvement over parametric models confirms that fuel efficiency is not fully explained by a simple linear relationship.

The trade-off is interpretability. Parametric models are generally easier to explain because each coefficient has a direct meaning. Among non-parametric models, simpler methods such as Decision trees remain interpretable, as their structure can be visualized and understood as a set of decision rules. However, more complex models like Random forest and KNN are harder to interpret due to their aggregated or instance-based nature. Parametric models are also more reproducible as they produce stable results across runs and are less sensitive to random variation in the data. In contrast, non-parametric models such as Random forest depend on randomness and hyperparameter tuning, which can lead to slightly different results across runs.

## Scenario 3: Feature Selection

We have already removed redundant, leaking and uninformative features during preprocessing. Feature selection extends the analysis by determining which of the remaining 22 features truly contribute to predicting the target variable and which can be removed without a significant decrease in performance. We compare one algorithmic method (forward selection) and two embedded methods (LASSO and Elastic Net), then evaluate how many features each retains and whether the reduced feature set preserves predictive performance compared to models trained in previous tasks.

Before applying any formal feature selection method, we first examined feature importance scores from Decision tree and a Random forest models trained on the full feature set. This will provide an initial indication of which predictors carry the most predictive signal.

```{r}
dt_importance <- dt_final %>%
  extract_fit_parsnip() %>%
  vip(num_features = 22) +
  labs(title = "Feature importance - Decision tree") +
  theme_minimal()

rf_importance <- rf_final %>%
  extract_fit_parsnip() %>%
  vip(num_features = 22) +
  labs(title = "Feature importance - Random forest") +
  theme_minimal()

dt_importance + rf_importance + plot_layout(ncol = 2, nrow = 1) 
```

Both models agree strongly on the top features: `displ` (engine displacement) and `cylinders` consistently rank as the two most important predictors by a large margin, suggesting that engine size is the primary driver of fuel efficiency. Features `drive_Front.Wheel.Drive` and `year` also appear prominently in both models, indicating that drivetrain type and vehicle age contribute meaningfully to predictions. In contrast, manufacturer-level dummy variables such as `make_GMC`, `make_BMW`, `make_Dodge`, and `make_Chevrolet` show near-zero importance in both models, suggesting they add little predictive value.

Overall, the importance rankings are largely consistent across both methods, which increases confidence in the signal and motivates retaining the top features while being selective about low-importance categorical dummies in subsequent selection steps. This gives us a clear starting point for feature selection.

### Algorithmic method

#### Forward stepwise selection

Forward stepwise selection is an algorithmic feature selection method that incrementally builds a regression model. Starting from a null model with no predictors, it evaluates all candidate variables at each step and adds the one that most improves model fit, measured by AIC (Akaike Information Criterion). The process continues until no remaining variable reduces the AIC further. Among the three algorithmic approaches (forward, backward and mixed selection) forward selection was chosen because it is a starting point when there is no strong prior knowledge about which predictors are relevant. Backward selection requires fitting a full model first, which can be unstable when some predictors are correlated, as seen in our EDA where `displ` and `cylinders` showed high multicollinearity. Mixed selection combines both directions but adds unnecessary complexity.

```{r}
null_model <- lm(comb08 ~ 1, data = train_processed)
full_model <- lm(comb08 ~ ., data = train_processed)

cat("\nNull model AIC:", AIC(null_model), "\n")
cat("Full model AIC:", AIC(full_model), "\n")
```

The null model, containing only an intercept and no predictors, achieved an AIC of 245612.9. The full model using all available predictors had AIC of 202102.7, which is a substantial improvement indicating that the predictors collectively carry meaningful information about fuel efficiency.

```{r}
cat("\n=== Forward stepwise selection ===\n")
forward_model <- stepAIC(
  null_model,
  scope     = list(lower = null_model, upper = full_model),
  direction = "forward",
  trace     = 1
)
```

The forward selection algorithm started from the null model and added predictors one at a time, retaining each variable only if it reduced the AIC. The first variable added was `displ`, which produced the largest single AIC drop from 135 685 to 109 690, confirming it as the strongest individual predictor of fuel efficiency. Subsequent steps added `year`, `drive_Front.Wheel.Drive`, `fuel_group_Gasoline`, and further variables with progressively reduced AIC. The algorithm terminated after 21 steps when no remaining variable could further reduce the AIC.

```{r}
cat("\n=== Result of forward selection ===\n")
forward_summary <- summary(forward_model)
print(forward_summary)
```

The final model retained 21 out of the available predictors. The model explains 67.5% of variance in `comb08` (R² = 0.675), with all 21 selected variables being statistically significant at the 0.001 level. The strongest negative predictors are `displ` with estimate −2.25 and `has_discrete_gears` with estimate −1.72, while the largest positive predictors are `drive_Front.Wheel.Drive` with estimate 1.46 and `year` with estimate 0.95.

```{r}
selected_vars_forward <- names(coef(forward_model))[-1]  # without intercept
cat("\nNumber of selected variables:", length(selected_vars_forward), "\n")
cat("Selected variables:\n")
print(selected_vars_forward)

all_vars_forward <- names(train_processed %>% 
                        select(-comb08))
removed_vars_forward <- setdiff(all_vars_forward, selected_vars_forward)
cat("\nRemoved variables:\n")
print(removed_vars_forward)
```

The model retained 21 predictors and removed the `VClass_Two.Seaters` variable.

```{r}
forward_train_preds <- predict(forward_model, newdata = train_processed)

forward_train_metrics <- tibble(
  truth    = train_processed$comb08,
  estimate = forward_train_preds
) %>%
  metrics(truth = truth, estimate = estimate) %>%
  mutate(split = "train")

forward_preds <- predict(forward_model, newdata = test_processed)

forward_test_metrics <- tibble(
  truth    = test_processed$comb08,
  estimate = forward_preds
) %>%
  metrics(truth = truth, estimate = estimate) %>%
  mutate(split = "test")

bind_rows(forward_train_metrics, forward_test_metrics) %>%
  select(split, .metric, .estimate) %>%
  pivot_wider(names_from = split, values_from = .estimate)
```

On the test set the model achieved RMSE = 3.28, MAE = 2.16, and R² = 0.677. Comparing train and test metrics shows practically no difference, indicating that the model generalizes well with no signs of overfitting. However, compared to the random forest model (RMSE = 2.56, R² = 0.964), the linear forward selection model is considerably weaker, suggesting that the relationship between predictors and fuel efficiency contains substantial non-linear structure that a linear model cannot capture.

```{r}
tibble(actual = test_processed$comb08,
       predicted = forward_preds) %>%
  ggplot(aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Forward selection: Actual vs. Predicted",
    x     = "Actual comb08",
    y     = "Predicted comb08"
  ) +
  theme_minimal()

tibble(predicted = forward_preds,
       residual  = test_processed$comb08 - forward_preds) %>%
  ggplot(aes(x = predicted, y = residual)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Residual Plot",
    x     = "Predicted comb08",
    y     = "Residuals"
  ) +
  theme_minimal()

aic_progress <- as.data.frame(forward_model$anova) %>%
  rownames_to_column("step") %>%
  mutate(step_num = row_number())

ggplot(aic_progress, aes(x = step_num, y = AIC)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  labs(
    title = "Forward selection - AIC decrease during selection",
    x     = "Number of added variables",
    y     = "AIC"
  ) +
  theme_minimal()
```

The Actual vs Predicted plot shows that the forward selection model captures the general positive relationship between actual and predicted `comb08`, especially for common values around 10–30 MPG. However, the model underpredicts highly fuel-efficient vehicles. This suggests the linear model cannot fully capture more complex relationships in the data.

The Residual plot supports this conclusion. Residuals are not randomly scattered around zero and their spread increases for higher predicted values, indicating heteroscedasticity. Several large positive residuals show cases where actual fuel efficiency is much higher than predicted.

The third graph illustrates the decrease of AIC during forward selection. A sharp decrease in AIC is visible during the first few steps, especially between steps 1 and 5. This suggests that these are likely the strongest predictors of `comb08`. After approximately steps 6 to 10, the decline in AIC becomes much smaller. Each additional predictor still improves the model slightly, but the gains are modest compared with the first selected features. From around steps 12 to 15 onward, the curve becomes almost flat. This means that adding more predictors provides little additional information.

#### Summary

Forward stepwise selection identified 21 statistically significant predictors of `comb08` out of all available candidates. The most influential variables added in the early steps were `displ`, `year`, `drive_Front.Wheel.Drive`, and `fuel_group_Gasoline`, which together account for the largest share of explained variance. The final model achieved R² = 0.677 and RMSE = 3.28 on the test set, with no signs of overfitting as train and test metrics were nearly identical. Compared to the non-parametric models trained earlier, forward selection performs considerably worse - Random Forest achieved RMSE = 2.56 and R² = 0.964 on the same data. This gap highlights the limitation of linear feature selection methods when the underlying relationships are non-linear. However, forward selection offers a clear advantage in interpretability, because each coefficient has a direct meaning and statistical significance can be assessed for every predictor, which is not possible with Random Forest or KNN.

### Embedded methods

Embedded methods integrate feature selection into model training itself — the algorithm simultaneously learns which features matter and what their coefficients are.

Two models from Scenario 2 are embedded methods by design:

-   **LASSO** (L1 penalty)
-   **Elastic Net** (L1 + L2) 

Because both models were already fitted in Scenario 2, we can reuse them.

#### LASSO — Regularization Path

The regularization path shows how many features the model keeps as the penalty λ grows. At λ → 0 all features are retained (same as OLS). As λ increases, the L1 geometry forces progressively more coefficients to exactly zero. The CV-optimal λ is marked — this is where Scenario 2's cross-validation stopped.

```{r}
lasso_glmnet  <- lasso_fit %>% extract_fit_engine()
opt_log_lasso <- log10(best_lasso$penalty)

path_lasso <- tibble(
  log_lambda = log10(lasso_glmnet$lambda),
  n_selected = lasso_glmnet$df   # non-zero coefficients at each lambda
)

path_lasso <- bind_rows(
  tibble(log_lambda = c(opt_log_lasso, min(path_lasso$log_lambda) - 0.01),
         n_selected = max(path_lasso$n_selected)),
  path_lasso
)

cat("=== LASSO — regularization path summary ===\n")
cat("Lambda range: [", round(min(lasso_glmnet$lambda), 6), ",",
                       round(max(lasso_glmnet$lambda), 3), "]\n")
cat("Features at λ_min (≈ OLS):   ",
    path_lasso$n_selected[which.min(path_lasso$log_lambda)], "\n")
cat("Features at λ_opt (CV best): ",
    path_lasso$n_selected[which.min(abs(path_lasso$log_lambda - opt_log_lasso))], "\n")
cat("Features at λ_max (max reg): ",
    path_lasso$n_selected[which.max(path_lasso$log_lambda)], "\n")

ggplot(path_lasso, aes(x = log_lambda, y = n_selected)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_vline(xintercept = opt_log_lasso, color = "tomato", lty = 2) +
  annotate("text",
           x     = opt_log_lasso + 0.15,
           y     = max(path_lasso$n_selected) * 0.95,
           label = paste0("λ_opt = ", round(best_lasso$penalty, 5)),
           color = "tomato", hjust = 0, size = 3.5) +
  labs(
    title    = "LASSO — number of selected features along the regularization path",
    subtitle = "Red dashed line = CV-optimal λ",
    x        = "log₁₀(λ)",
    y        = "Number of non-zero coefficients"
  ) +
  theme_minimal()
```

The CV-optimal λ is at the left end of the path — meaning the 5-fold CV found that keeping all (or nearly all) features produces the lowest prediction error on this dataset.

#### LASSO — Selected Features at Optimal λ

These are the predictors LASSO kept at the CV-optimal penalty. A zero coefficient means the model decided that feature adds no useful information given the others.

```{r fig.height=6}
lasso_coefs_s3 <- lasso_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(term != "(Intercept)")

n_kept_lasso   <- sum(lasso_coefs_s3$estimate != 0)
n_zeroed_lasso <- sum(lasso_coefs_s3$estimate == 0)

cat("=== LASSO — embedded feature selection result ===\n")
cat("Total features available:    ", nrow(lasso_coefs_s3), "\n")
cat("Features selected (non-zero):", n_kept_lasso, "\n")
cat("Features removed (zeroed):   ", n_zeroed_lasso, "\n")

if (n_zeroed_lasso > 0) {
  cat("\nZeroed predictors:\n")
  print(lasso_coefs_s3 %>% filter(estimate == 0) %>% pull(term))
}

lasso_coefs_s3 %>%
  filter(estimate != 0) %>%
  slice_max(abs(estimate), n = 20) %>%
  ggplot(aes(x = reorder(term, abs(estimate)), y = estimate,
             fill = estimate > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
    labels = c("TRUE" = "Positive", "FALSE" = "Negative"),
    name   = "Direction"
  ) +
  labs(
    title    = paste0("LASSO — selected features at λ = ",
                      round(best_lasso$penalty, 5)),
    subtitle = paste0(n_kept_lasso, " features retained, ",
                      n_zeroed_lasso, " removed"),
    x        = NULL,
    y        = "Coefficient estimate"
  ) +
  theme_minimal()
```

#### 3.4 Elastic Net — Regularization Path

Elastic Net mixes L1 and L2 penalties. The optimal mixture was found by CV in Scenario 2.

```{r}
en_glmnet  <- en_fit %>% extract_fit_engine()
opt_log_en <- log10(best_en$penalty)

path_en <- tibble(
  log_lambda = log10(en_glmnet$lambda),
  n_selected = en_glmnet$df
)

ggplot(path_en, aes(x = log_lambda, y = n_selected)) +
  geom_line(color = "#27AE60", linewidth = 0.8) +
  geom_vline(xintercept = opt_log_en, color = "tomato", lty = 2) +
  annotate("text",
           x     = opt_log_en + 0.15,
           y     = max(path_en$n_selected) * 0.95,
           label = paste0("λ_opt = ", round(best_en$penalty, 5)),
           color = "tomato", hjust = 0, size = 3.5) +
  labs(
    title    = "Elastic Net — regularization path",
    subtitle = paste0("Red dashed line = CV-optimal λ  |  α = ",
                      round(best_en$mixture, 3), "  (0 = Ridge, 1 = LASSO)"),
    x        = "log₁₀(λ)",
    y        = "Number of non-zero coefficients"
  ) +
  theme_minimal()
```

#### 3.5 Elastic Net — Selected Features at Optimal λ

```{r fig.height=6}
en_coefs_s3 <- en_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(term != "(Intercept)")

n_kept_en   <- sum(en_coefs_s3$estimate != 0)
n_zeroed_en <- sum(en_coefs_s3$estimate == 0)

cat("=== Elastic Net — embedded feature selection result ===\n")
cat("Total features available:    ", nrow(en_coefs_s3), "\n")
cat("Features selected (non-zero):", n_kept_en, "\n")
cat("Features removed (zeroed):   ", n_zeroed_en, "\n")

if (n_zeroed_en > 0) {
  cat("\nZeroed predictors:\n")
  print(en_coefs_s3 %>% filter(estimate == 0) %>% pull(term))
}

en_coefs_s3 %>%
  filter(estimate != 0) %>%
  slice_max(abs(estimate), n = 20) %>%
  ggplot(aes(x = reorder(term, abs(estimate)), y = estimate,
             fill = estimate > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
    labels = c("TRUE" = "Positive", "FALSE" = "Negative"),
    name   = "Direction"
  ) +
  labs(
    title    = paste0("Elastic Net — selected features at λ = ",
                      round(best_en$penalty, 5)),
    subtitle = paste0(n_kept_en, " features retained, ",
                      n_zeroed_en, " removed"),
    x        = NULL,
    y        = "Coefficient estimate"
  ) +
  theme_minimal()
```

#### 3.6 Comparing LASSO vs. Elastic Net Feature Selection

We compare which features each model kept and where they agree. Features retained by both carry enough signal to survive both pure L1 and combined L1+L2 penalization.

```{r}
selection_comparison <- lasso_coefs_s3 %>%
  select(term, lasso = estimate) %>%
  left_join(
    en_coefs_s3 %>% select(term, en = estimate),
    by = "term"
  ) %>%
  mutate(
    lasso_selected = lasso != 0,
    en_selected    = en    != 0,
    agreement      = case_when(
      lasso_selected & en_selected   ~ "Both selected",
      lasso_selected & !en_selected  ~ "LASSO only",
      !lasso_selected & en_selected  ~ "EN only",
      TRUE                           ~ "Both zeroed"
    )
  )

cat("=== LASSO vs. Elastic Net — selection agreement ===\n")
print(table(selection_comparison$agreement))

cat("\n--- Features selected by BOTH models ---\n")
selection_comparison %>%
  filter(agreement == "Both selected") %>%
  arrange(desc(abs(lasso))) %>%
  head(10) %>%
  mutate(lasso = round(lasso, 4), en = round(en, 4)) %>%
  print()

selection_comparison %>%
  filter(lasso_selected | en_selected) %>%
  slice_max(abs(lasso) + abs(en), n = 20) %>%
  pivot_longer(cols = c(lasso, en), names_to = "model", values_to = "coef") %>%
  ggplot(aes(x = reorder(term, abs(coef)), y = coef, fill = model)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_manual(
    values = c("lasso" = "steelblue", "en" = "#27AE60"),
    labels = c("lasso" = "LASSO", "en" = "Elastic Net")
  ) +
  labs(
    title    = "LASSO vs. Elastic Net — coefficient comparison",
    subtitle = "Top 20 features selected by at least one model",
    x        = NULL,
    y        = "Coefficient estimate",
    fill     = "Model"
  ) +
  theme_minimal()
```

#### 3.7 Test Set Performance

We evaluate both models on the test set, showing prediction accuracy alongside the number of features used — the embedded selection.

```{r}
lasso_test_s3 <- lasso_fit %>%
  predict(new_data = test_data) %>%
  bind_cols(test_data %>% select(comb08))

en_test_s3 <- en_fit %>%
  predict(new_data = test_data) %>%
  bind_cols(test_data %>% select(comb08))

s3_comparison <- tibble(
  model      = c("LASSO", "Elastic Net"),
  n_features = c(n_kept_lasso, n_kept_en),
  RMSE = c(
    rmse_vec(truth = lasso_test_s3$comb08, estimate = lasso_test_s3$.pred),
    rmse_vec(truth = en_test_s3$comb08,   estimate = en_test_s3$.pred)
  ),
  MAE = c(
    mae_vec( truth = lasso_test_s3$comb08, estimate = lasso_test_s3$.pred),
    mae_vec( truth = en_test_s3$comb08,   estimate = en_test_s3$.pred)
  ),
  Rsq = c(
    rsq_vec( truth = lasso_test_s3$comb08, estimate = lasso_test_s3$.pred),
    rsq_vec( truth = en_test_s3$comb08,   estimate = en_test_s3$.pred)
  )
) %>%
  mutate(across(c(RMSE, MAE, Rsq), ~round(., 4)))

cat("=== Scenario 3 — Test set performance ===\n")
print(s3_comparison)

cat("\n=== OLS baseline (all", ncol(train_processed) - 1, "features) ===\n")
test_comparison %>%
  filter(model == "Linear Regression") %>%
  mutate(n_features = ncol(train_processed) - 1) %>%
  select(model, n_features, RMSE, MAE, Rsq) %>%
  print()
```

#### 3.8 Extended penalty range

The CV-optimal LASSO penalty was the smallest tested value (`λ = 1e-4`), so the final predictive model kept all 22 predictors. The fitted path still shows what would happen under stronger penalization: as `λ` increases, coefficients are gradually shrunk and eventually set to zero.

```{r}
path_summary <- tibble(
  lambda     = lasso_glmnet$lambda,
  log_lambda = log10(lambda),
  n_selected = lasso_glmnet$df
) %>%
  slice(c(1, round(n() / 2), n())) %>%
  mutate(lambda = round(lambda, 5),
         log_lambda = round(log_lambda, 3))

cat("=== LASSO path: representative sparsity levels ===\n")
print(path_summary)
```

This shows that LASSO is would be able of feature selection, but the cross-validated prediction model did not need sparsity. Stronger penalties would simplify the model, but they would also move it away from the penalty value selected for best predictions.

------------------------------------------------------------------------

### Conclusion

#### Embedded selection vs. Scenario 2 analysis

Scenario 2 used LASSO and Elastic Net as prediction models. Scenario 3 looks at the same models from a different angle: which features did they select, and what does the regularization path reveal about the feature space? The path plot is the embedded selector's "work log" — it shows which predictors are robust enough to survive increasing penalization, even if the CV-optimal λ happens to keep most features.

#### Algorithmic vs. embedded selection

Forward stepwise selection is an algorithmic wrapper method. It repeatedly fits candidate models and adds the predictor that improves AIC the most. This produces a clear sequence of decisions and standard regression output, including p-values for the selected variables. In our case it retained 21 of the 22 available predictors and removed only `VClass_Two.Seaters`, so the algorithmic method mostly confirmed that the preprocessing step had already removed the weakest information.

LASSO and Elastic Net are embedded methods - feature selection happens inside the model fitting process through the penalty term. At the CV-optimal penalty both retained all 22 predictors, meaning that the best predictive model did not benefit from removing variables. The regularization path still shows that under stronger penalization predictors would gradually drop out, but that would move the model away from the penalty value selected for best predictive performance.

Overall, the results are quite consistent across methods, with the algorithmic method producing better results. The most important predictors, such as `displ`, `cylinders`, `drive_Front.Wheel.Drive`, `fuel_group_Gasoline`, `year` and `has_discrete_gears` appear in all models and have a stable influence on the prediction. On the other hand, weaker variables, such as some manufacturer or vehicle class dummies, do not contribute much and are the first to be removed or shrink when the model is simplified.

------------------------------------------------------------------------

## Scenario 5: Dimensionality Reduction

This scenario was not the primary focus of our analysis, as the dataset is not highly multidimensional. After preprocessing, the data contains 22 features and 1 target variable, which does not typically require dimensionality reduction techniques. Nevertheless, we included it to explore dimensionality reduction techniques such as PCA and to assess their effect on model performance, interpretability and information loss.

### Principal Component Analysis (PCA)

PCA is a dimensionality reduction technique that transforms the original feature space into a new set of variables called principal components. These components are linear combinations of the original features and are constructed in such a way that each successive component captures the maximum remaining variance in the data while being orthogonal to the previous ones.

As a result, PCA rotates and reprojects the original feature space into a new coordinate system, where the first few components capture most of the variability in the data. This allows the dataset to be represented with fewer dimensions while preserving as much information as possible.

```{r}
preprocess_recipe <- recipe(comb08 ~ ., data = train_data) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%
  step_impute_median(displ, cylinders) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_nzv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

prep_preprocess_recipe <- prep(preprocess_recipe, training = train_data)
train_preprocessed <- bake(prep_preprocess_recipe, new_data = NULL)
```

```{r}
pca_input <- train_preprocessed %>%
  select(-comb08)   # remove target

cat("Size of PCA input:", nrow(pca_input), "×", ncol(pca_input), "\n")
```

```{r}
pca_result <- prcomp(pca_input, center = FALSE, scale. = FALSE)
# center=FALSE a scale.=FALSE because train_preprocessed is already normalised
```

```{r}
explained_var <- summary(pca_result)$importance
cat("\nFirst 10 components:\n")
print(explained_var[, 1:10])
```

The PCA was performed on the preprocessed training data, which had already been normalized and dummy-encoded. The first principal component explains approximately 13.6% of the total variance, while the second accounts for 11,97%. The explained variance decreases gradually across subsequent components, with no single dominant component, what suggests that the information in the dataset is moderately distributed across multiple dimensions rather than being concentrated in a few directions. The first 10 components together capture about 68.9% of the total variance.

```{r}
cum_var <- as.data.frame(t(explained_var)) %>%
  rownames_to_column("PC") %>%
  setNames(c("PC", "std_dev", "prop_var", "cum_var")) %>%
  mutate(PC_num = as.integer(str_extract(PC, "\\d+")))
```

```{r}
for (threshold in c(0.80, 0.90, 0.95)) {
  n <- min(which(cum_var$cum_var >= threshold))
  cat(sprintf("Number of PC for %.0f%% of variance: %d\n", threshold * 100, n))
}
```

To reach 80%, 90%, and 95% of cumulative variance, 13, 16, and 18 components are required respectively. The relatively high number of components needed to explain a substantial portion of variance is consistent with the high-dimensional nature of the dataset after dummy encoding, where many categorical variables were expanded into binary columns.

```{r}
p1 <- cum_var %>%
  filter(PC_num <= 30) %>%
  ggplot(aes(x = PC_num, y = prop_var)) +
  geom_col(fill = "steelblue") +
  geom_line(aes(group = 1), color = "darkblue", linewidth = 0.8) +
  geom_point(color = "darkblue", size = 2) +
  scale_x_continuous(breaks = 1:30) +
  labs(
    title = "Scree Plot - explained variance ratio",
    x     = "Principal component",
    y     = "Variance ratio"
  ) +
  theme_minimal()

p2 <- cum_var %>%
  filter(PC_num <= 30) %>%
  ggplot(aes(x = PC_num, y = cum_var)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_hline(yintercept = c(0.80, 0.90, 0.95),
             linetype = "dashed",
             color = c("orange", "red", "darkred")) +
  annotate("text", x = 21, y = 0.81, label = "80%", color = "orange",    size = 3) +
  annotate("text", x = 21, y = 0.91, label = "90%", color = "red",       size = 3) +
  annotate("text", x = 2, y = 0.96, label = "95%", color = "darkred",   size = 3) +
  scale_x_continuous(breaks = seq(0, 22, 5)) +
  labs(
    title = "Cummulative explained variability",
    x     = "Number of principal components",
    y     = "Cummulative variance ratio"
  ) +
  theme_minimal()

p1 + p2
```

The scree plot shows that the first principal component explains the largest share of variance (\~14%), followed by a notable drop to the second component (\~12%). After the third component, the decline becomes more gradual and the curve flattens, forming a characteristic elbow shape. This indicates that no single component strongly dominates, and the remaining variance is spread relatively evenly across the subsequent components.

The cumulative variance plot confirms this pattern. The curve rises steadily without any sharp jumps, which is typical for datasets with many moderately correlated features after dummy encoding. The three reference thresholds show that 80% of variance is reached at 13 components, 90% at 16 components, and 95% at 18 components.

```{r}
n_components <- min(which(cum_var$cum_var >= 0.95))
cat("\nWe will use", n_components, "components (95% of variance)\n")
```

For subsequent modeling, 18 principal components were retained, capturing 95% of the total variance. This threshold was chosen as a balance between dimensionality reduction and information retention - discarding the remaining 5% removes noise and redundancy while keeping the majority of the signal present in the original features. The reduction was small but still useful given the relatively small original feature space after preprocessing.

**Discussion on explainability and information loss** After applying PCA, the original features are replaced by principal components (PC1, PC2,...), which are combinations of multiple variables. Since these components do not have a direct or intuitive interpretation, model explainability decreases. It becomes difficult to attribute predictions to specific original features or to understand how individual variables influence the outcome.

In addition, dimensionality reduction inevitably leads to some loss of information. While the first few components capture most of the variance, the remaining variability contained in the discarded components is lost. This may remove noise, but it can also eliminate subtle patterns that could be useful for prediction, potentially reducing model performance.

**Random forest model with PCA** Now, we create a recipe, which will be used for comparison of models.

```{r}
pca_recipe <- recipe(comb08 ~ ., data = train_data) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_impute_mode(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_nzv(all_predictors()) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_pca(all_numeric_predictors(), num_comp = n_components)
```

Now we train a random forest model using preprocessed data with pca_recipe. Although the KNN model achieved slightly better predictive performance in the previous task, the random forest model is more robust and less sensitive to overfitting.

```{r}
pca_rf_model <- rand_forest(
  trees = 200,
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")

pca_rf_workflow <- workflow() %>%
  add_recipe(pca_recipe) %>%
  add_model(pca_rf_model)

pca_rf_grid <- grid_regular(
  mtry(range = c(2, 10)),
  min_n(range = c(2, 20)),
  levels = 3
)

pca_rf_tuned <- tune_grid(
  pca_rf_workflow,
  resamples = folds,
  grid = pca_rf_grid,
  metrics   = eval_metrics,
  control   = tune_ctrl
)

pca_rf_best <- select_best(pca_rf_tuned, metric = "rmse")
cat("The best parameters for random forest with PCA:\n"); print(pca_rf_best)

pca_rf_final <- pca_rf_workflow %>%
  finalize_workflow(pca_rf_best) %>%
  fit(data = train_data)
```

The best random forest model with PCA has these parameters: - trees = 200 - mtry = 10 - min_n = 11

```{r}
# get metrics on train data
pca_rf_train_preds <- predict(pca_rf_final, new_data = train_data) %>%
  bind_cols(train_data %>% select(comb08))

pca_rf_train_metrics <- pca_rf_train_preds %>%
  metrics(truth = comb08, estimate = .pred) %>%
  mutate(split = "train")

# get metrics on test data
pca_rf_pred <- predict(pca_rf_final, new_data = test_data) %>%
  bind_cols(test_data %>% select(comb08))

pca_rf_test_metrics <- pca_rf_pred %>%
  metrics(truth = comb08, estimate = .pred) %>%
  mutate(split = "test")

bind_rows(pca_rf_train_metrics, pca_rf_test_metrics) %>%
  select(split, .metric, .estimate) %>%
  pivot_wider(names_from = split, values_from = .estimate)

# check predictions in the first rows
head(pca_rf_pred)
```

The best random forest model with PCA shows good overall performance, but it is slightly less accurate than models trained on the original features.

On the training set, the model achieved an RMSE of 1.36, MAE of 0.76, and R² of 0.944, indicating a very strong fit and that it explains about 94% of the variance in `comb08`. However, performance decreases on the test set, where the model reaches an RMSE of 1.97, MAE of 1.16, and R² of 0.885. This gap suggests mild overfitting, although part of the decrease in test performance may also be attributed to information loss introduced by PCA dimensionality reduction.

The sample predictions show that the model still captures general trends in fuel consumption well, but with slightly larger errors compared to models without PCA.

Overall, the model remains reliable and captures the main structure of the data, but the PCA transformation likely removed some fine-grained information that was useful for prediction. As a result, predictive performance is slightly reduced compared to models trained on the original feature space.

```{r}
ggplot(pca_rf_pred, aes(x = comb08, y = .pred)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Random Forest with PCA: Actual vs Predicted",
    x = "Actual comb08",
    y = "Predicted comb08"
  ) +
  theme_minimal()

pca_rf_pred <- pca_rf_pred %>%
  mutate(residual = comb08 - .pred)

ggplot(pca_rf_pred, aes(x = .pred, y = residual)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Residual Plot",
    x = "Predicted comb08",
    y = "Residuals"
  ) +
  theme_minimal()
```

The Actual vs. Predicted plot shows that the model tracks the diagonal reasonably well in the lower range (10–30 MPG), where most observations are concentrated. However, as actual values increase beyond 40, predictions increasingly underestimate the true values.

The Residual plot reveals a clear funnel shape: residuals are tightly clustered near zero for low predicted values but spread substantially as predictions increase. This heteroscedasticity suggests the model struggles with higher fuel efficiency vehicles, likely because they are underrepresented in the training data after PCA transformation. The residuals are roughly centered around zero overall, indicating no systematic bias in the mid-range, but the upper tail remains problematic.

Overall, the model performs adequately for typical vehicles but loses precision for fuel-efficient outliers.

### Partial Least Squares (PLS)

PLS (Partial Least Squares) is a supervised dimensionality reduction method for regression. It finds latent components — like PCA axes, but guided by the target — that maximally **correlate with `comb08`**.

#### Cross-Validation

We fit PLS with up to 15 components and use 5-fold CV to find the optimal number of components. The optimal number of components is selected using a simple rule: we choose the smallest number of components whose RMSE from cross-validation is within 1% of the best observed RMSE.
```{r}
pls_model <- plsr(
    comb08 ~ .,
    data       = train_processed,
    ncomp      = 15,
    validation = "CV",
    segments   = 5,
    scale      = FALSE
  )

rmsep_cv <- RMSEP(pls_model)$val["CV", 1, ]

min_rmsep <- min(rmsep_cv)
opt_ncomp <- which(rmsep_cv <= min_rmsep * 1.01)[1] - 1
cat("\nOptimal number of components:", opt_ncomp, "\n")
```

```{r}
plot(RMSEP(pls_model),
     main = "PLS — CV RMSEP by number of components",
     xlab = "Number of components",
     ylab = "CV RMSEP (MPG)")
abline(v = opt_ncomp, col = "tomato", lty = 2)
```

Components past the "elbow" add complexity without reducing error.


#### Explained Variance per Component

PLS components are ordered by how much covariance with `comb08` they capture.

```{r}
predictor_variance_by_component <- explvar(pls_model)
cv_rsq_by_component <- R2(pls_model, estimate = "CV")$val["CV", 1, -1]

cat("=== PLS — Explained variance per component ===\n")
cat(sprintf("%-6s  %9s  %10s  %12s\n",
    "Comp", "X var %", "Cum X %", "R²(y) CV"))

cumulative_predictor_variance <- 0

for (component in seq_len(min(opt_ncomp + 3, length(predictor_variance_by_component)))) {
  cumulative_predictor_variance <- cumulative_predictor_variance +
    predictor_variance_by_component[component]
  selected_marker <- if (component == opt_ncomp) "  <--" else ""

  cat(sprintf("%-6d  %9.2f  %10.2f  %12.4f%s\n",
      component,
      predictor_variance_by_component[component],
      cumulative_predictor_variance,
      cv_rsq_by_component[component],
      selected_marker))
}
```

Comp1 captures the dominant engine-size axis, mainly showing the difference between larger combustion engines with more efficient drivetrain and gearing patterns. Subsequent components pick up smaller, residual patterns.

#### Dimensionality Reduction — Projection onto First Two Components

We project all training vehicles onto Comp1 and Comp2. This is the dimensionality reduction result: 22 normalized predictors collapse into 2 coordinates. Color shows actual MPG — a clean gradient confirms Comp1 captures the efficiency axis.

```{r}
scores_df <- tibble(
  Comp1  = scores(pls_model)[, 1],
  Comp2  = scores(pls_model)[, 2],
  comb08 = train_data$comb08
)

cat("=== PLS — Score ranges (training data) ===\n")
cat("Comp1:", round(range(scores_df$Comp1), 3), "\n")
cat("Comp2:", round(range(scores_df$Comp2), 3), "\n")

ggplot(scores_df, aes(x = Comp1, y = Comp2, color = comb08)) +
  geom_point(alpha = 0.3, size = 0.6) +
  scale_color_viridis_c(option = "plasma", name = "MPG") +
  labs(
    title    = "PLS — training data on first two components",
    subtitle = "Color = combined MPG (purple = low, yellow = high)",
    x        = "PLS Component 1",
    y        = "PLS Component 2"
  ) +
  theme_minimal()
```

#### Component Loadings

Loadings show which predictors drive Comp1.

```{r fig.height=6}
loadings_df <- as_tibble(loadings(pls_model)[, 1, drop = FALSE],
                          rownames = "predictor") %>%
  rename(Comp1 = 2)

loadings_df %>%
  slice_max(abs(Comp1), n = 15) %>%
  ggplot(aes(x = reorder(predictor, Comp1), y = Comp1, fill = Comp1 > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#27AE60", "FALSE" = "#E74C3C"), guide = "none") +
  labs(
    title    = "PLS — loadings of Component 1",
    subtitle = "Top 15 predictors — green = higher MPG, red = lower MPG",
    x        = NULL,
    y        = "Comp1 loading"
  ) +
  theme_minimal()
```

#### Test Set Evaluation

We predict on the test set using `opt_ncomp` components and compare directly with the four Scenario 2 models — same test set, same metrics.

```{r}
pls_pred <- predict(pls_model, newdata = test_processed, ncomp = opt_ncomp)[, , 1]

pls_metrics <- tibble(
  truth = test_data$comb08,
  pred  = as.numeric(pls_pred)
)

pls_rmse <- rmse_vec(truth = pls_metrics$truth, estimate = pls_metrics$pred)
pls_mae  <- mae_vec( truth = pls_metrics$truth, estimate = pls_metrics$pred)
pls_rsq  <- rsq_vec( truth = pls_metrics$truth, estimate = pls_metrics$pred)

cat("=== PLS — Test set metrics (ncomp =", opt_ncomp, ") ===\n")
cat("RMSE:", round(pls_rmse, 4), "\n")
cat("MAE: ", round(pls_mae,  4), "\n")
cat("R²:  ", round(pls_rsq,  4), "\n")

cat("\n=== All models: PLS vs. Scenario 2 ===\n")
bind_rows(
  test_comparison,
  tibble(model = paste0("PLS (", opt_ncomp, " comp)"),
         RMSE  = pls_rmse,
         MAE   = pls_mae,
         Rsq   = pls_rsq)
) %>%
  arrange(RMSE) %>%
  print()
```

```{r}
ggplot(pls_metrics, aes(x = truth, y = pred)) +
  geom_point(alpha = 0.15, color = "steelblue", size = 0.8) +
  geom_abline(slope = 1, intercept = 0, color = "tomato", linewidth = 0.8) +
  labs(
    title = "PLS — Actual vs. Predicted (test set)",
    x     = "Actual combined MPG",
    y     = "Predicted combined MPG"
  ) +
  theme_minimal()
```

The PLS model selected 3 components using the "elbow" rule. These 3 components explain only 22.91% of the predictor variance, but already reach a cross-validated R² of 0.6703, which shows that PLS focuses on the part of the feature space that is most relevant for predicting `comb08`, not necessarily on explaining all predictor variance.

On the test set, PLS achieved RMSE = 3.3005, MAE = 2.1785 and R² = 0.6737. This is very close to the parametric models from Scenario 2, but slightly worse than the full linear regression, LASSO and Elastic Net. The top loadings of the first component are `displ`, `cylinders`, `drive_Front.Wheel.Drive`, `drive_Rear.Wheel.Drive` and `has_discrete_gears`, which confirms that the main PLS component captures the same engine-size and drivetrain structure identified throughout the analysis.
 
Overall, PLS provides an interpretable regression model, reducing 22 predictors to 3 supervised components with only a small loss in performance. However, it does not outperform the original feature-space models.

------------------------------------------------------------------------

#### PCA VS PLS

PCA and PLS both reduce the dimensionality of the feature space, but they optimize different goals. PCA is unsupervised: it keeps directions with the largest variance in the predictors, regardless of whether that variance helps predict `comb08`. PLS is supervised: it builds components that explain covariance with the target.

The PCA-based model used Random forest after the PCA transformation. It achieved strong performance, but still performed worse than Random forest trained on the original feature space. This suggests that PCA removed some detailed information. PCA also reduces explainability because it replaces original variables with principal components that combine multiple predictors. So its hard to say how a specific original feature affects MPG.

PLS compressed its base OLS model to 3 supervised components and achieved decent results, but weaker than the non-parametric models. An advantage is the interpretability of the  structure: the loadings show that Component 1 is mainly driven by `displ`, `cylinders`, drivetrain and gearing variables. The information loss is visible in the worse prediction performance, but the component structure remains easier to connect back to the regression target than PCA components.

Overall, PCA is useful for reducing the number of features, but it makes the model harder to interpret. PLS is more targeted for regression and more compact, but seems to behave closer to a linear supervised model and cannot match the best non-linear models. In this dataset, dimensionality reduction is not essential because we only have 22 predictors after preprocessing.

## Final Conclusion

The project started with a large vehicle fuel economy dataset and transformed it into a cleaner modeling dataset focused on combustion and hybrid vehicles. We removed variables with leakage, many missing values, identifiers, sparse technical fields and electric-only MPGe observations. We then engineered predictors such as fuel group, drivetrain group, transmission type, number of gears and discrete-gearing indicators. The final modeling data contained 22 predictors and the target variable `comb08`.

In Scenario 2, we compared parametric and non-parametric regression models. The linear parametric models were stable and interpretable, but their performance was limited by the linear form. Ridge, LASSO and Elastic Net did not improve over OLS because the dataset was large and the feature set was already compact. Adding four domain-driven interaction terms improved the OLS baseline, confirming that non-additive methods show a big impact. The non-parametric models performed best overall, especially Random forest and KNN, because they captured non-linear structure that linear models could not.

In Scenario 3, we studied feature selection. Forward stepwise selection removed only one predictor and embedded methods kept all features at the CV-optimal penalty. This means the preprocessing stage had already removed most uninformative variables and the remaining predictors carried meaningful signal. The embedded regularization path still showed how features would be removed under stronger penalization, but would introduce highly increased error once important engine-size information was lost.

In Scenario 5, we compared PCA and PLS as dimensionality reduction approaches. PCA reduced the feature space but introduced information loss and weaker explainability because principal components are mixtures of original predictors. PLS used supervised components and gave a compact regression model with interpretable loadings, but its predictive performance stayed close to the linear models rather than the stronger non-parametric models. Since the final dataset is not highly dimensional, dimensionality reduction is not necessary for the best prediction.

Overall, the main result is clear - fuel efficiency is strongly driven by engine size, cylinder count, drivetrain, transmission structure, fuel group and vehicle year. Linear models explain these effects transparently, but non-parametric models predict better.

*Save trained models for display in Shiny app*

```{r}
dir.create("models", showWarnings = FALSE)

saveRDS(ols_interact_fit, "models/ols_interact_fit.rds")

saveRDS(dt_final,     "models/dt_final.rds")
saveRDS(rf_final,     "models/rf_final.rds")
saveRDS(knn_final,    "models/knn_final.rds")

saveRDS(forward_model, "models/forward_model.rds")
saveRDS(en_fit,        "models/en_fit.rds")

saveRDS(pca_rf_final, "models/pca_rf_final.rds")
saveRDS(list(model = pls_model, ncomp = opt_ncomp), "models/pls_final.rds")

saveRDS(cor_matrix, "models/cor_matrix.rds")
```
