---
title: "Vehicles dataset"
author: "Dominika Lukašáková, Lukáš Lovás"
date: "2026-04-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("D:/Škola/Ing/2.semester/OZNAL_projekt")
```

## EDA

Vehicles dataset\
Source: <https://fueleconomy.gov/feg/ws/index.shtml>

Based on the column descriptions, we could potentially predict fuel consumption (MPG).

-   **MPG (Miles per gallon)** – a measure of fuel economy in automobiles, higher number= higher efficiency\
-   **Target variable:** `comb08`

```{r}
library(tidyverse)
library(ggcorrplot)
library(patchwork)

rm(list = ls())
data <- read_csv("data/vehicles.csv")
data
View(data)
```

## Basic Structure

First, we check dataset dimensions and data types.

```{r}
dim(data)
```

The dataset has 49,846 rows and 84 columns.

```{r}
glimpse(data)
```

```{r}
data %>%
  summarise(across(everything(), class))
```

The dataset contains columns of type: character, double, logical. Columns createdOn and modifiedOn are dates stored as strings. The dataset has some missing values.

## Missing Values

```{r}
sum(is.na(data))
```

The dataset contains 538 466 missing values.

# Identify columns with missing values

```{r}
missing_summary <- data %>%
  summarise(across(everything(), ~mean(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_pct") %>%
  arrange(desc(missing_pct))

missing_summary %>%
  print(n = 84)
```

Columns with more than 60% missing values:

-   `c240bDscr`, `c240Dscr`, `sCharger`, `rangeA`, `fuelType2`, `guzzler`, `evMotor`,
-   `atvType`, `tCharger`, `trans_dscr`, `startStop`, `mfrCode`

We will not use these columns and remove them later.

-   Column `eng_dscr` has \~37% missing values
-   Columns `cylinders`, `displ`, `drive`, `trany`, `fuelType1` contain only a small amount of missing data (up to \~3%), so they could potentially be imputed.

## Unique Values

Next we identify columns with unique values.

```{r}
data %>%
  summarise(across(everything(), n_distinct)) %>%
  pivot_longer(everything(), 
               names_to = "variable", 
               values_to = "n_unique") %>%
  arrange(desc(n_unique)) %>%
  print(n = 84)
```

From these we filter out non-informative columns. We remove variables with nearly all unique values, as they act as identifiers and do not provide generalizable patterns for prediction. Also variables with very low variability (binary or near-constant features) were excluded, as they provide minimal information for explaining variation in the target variable.

```{r}
data %>%
  summarise(across(everything(), n_distinct)) %>%
  pivot_longer(everything(), 
               names_to = "variable", 
               values_to = "n_unique") %>%
  arrange(desc(n_unique)) %>%
  filter(!(n_unique >= 0.9 * nrow(data))) %>%
  filter(!(n_unique <= 2)) %>%
  print(n = 84)
```

-   We should also remove `model` column, because its extremely high number of unique values makes it impractical for modeling, and `engId,` as it also represents an identifier.
-   The dataset contains a large number of variables with high cardinality, indicating a high-dimensional feature space that requires careful feature selection.

## Numerical Variables

First, we check the distribution of numerical variables.

```{r}
num_data <- data %>% 
  select(where(is.numeric)) %>% 
  select(-id, -engId)

summary(num_data)
```

-   Many columns have median = 0 because most vehicles do not have that feature, for example `highwayE` applies only to electric vehicles. So, these variables might be informative for specific subsets, but not for the general fuel consumption prediction.

-   Some columns have min and median = -1, which is unusual and likely represents missing values. The columns are: `co2`, `co2A`, `feScore`, `ghgScore`, `ghgScoreA`.

-   Although column `youSaveSpend` does not have issues in terms of distribution, its definition reveals that it is directly derived from fuel consumption and fuel cost variables. As a result, it contains information closely related to the target variable and could introduce data leakage and artificially inflating model performance.

For further analysis we will remove: - columns with median = 0 or -1 - column `youSaveSpend`

```{r}
num_data <- num_data %>%
  select(where(~ {
    med <- median(., na.rm = TRUE)
    !(med == 0 | med == -1)
  })) %>%
  select(-youSaveSpend)

summary(num_data)
```

-   We are left with 11 numerical columns.

-   The similarity between mean and median values suggests that most variables are not heavily skewed, although there might be some outliers.

```{r}
num_data %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~variable, scales = "free") +
  theme_minimal()
# A warning is displayed because of NA in columns cylinders and displ
```

Most variables show right-skewed distributions, with the majority of observations with lower values and with a long tail toward higher values.

A few variables, such as `barrels08` and `co2TailpipeGpm` appear approximately normally distributed, though still with slight positive skew. In contrast, `cylinders` is a discrete variable, with clear peaks at common engine configurations - 4, 6, and 8 cylinders.

The variable `year` shows a relatively even but slightly increasing distribution toward more recent years, indicating a higher representation of newer vehicles in the dataset.

```{r}
num_data %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "value") %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()
# A warning is displayed because of NA in columns cylinders and displ
```

-   The variable `fuelCost08` shows a wide IQR and numerous outliers.
-   Column `co2TailpipeGpm` has a relatively compact central distribution but still includes several higher outliers.
-   The `year`column has a narrow spread from 1984 to 2026, indicating that most observations are concentrated within a similar time range, so there are no outliers.

We will plot another boxplot without columns `year`, `fuelCost08`, and `co2TailpipeGpm` due to differences in scale:

```{r}
num_data %>%
  select(-year, -fuelCost08, -co2TailpipeGpm) %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "value") %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()
# A warning is displayed because of NA in columns cylinders and displ
```

-   All columns have a significant amount of outsiders on the right side.

The boxplots reveal that most variables are right-skewed, as indicated by longer upper whiskers and a large number of high-value outliers. This pattern is especially clear for fuel economy variables (`city08`, `highway08`, `comb08`, `UCity`, `UHighway`), where a small number of highly efficient vehicles extend the upper range.

Variables such as `barrels08` and `displ` exhibit moderate right skewness with some extreme values. In contrast, `cylinders` is a discrete variable, making its boxplot less informative and leading to apparent outliers due to limited possible values.

Overall, the boxplots highlight skewness and the presence of outliers.

## Correlation Analysis

```{r}
cor_matrix <- num_data %>%
  select(where(is.numeric)) %>%
  cor(use = "pairwise.complete.obs")

cor_matrix
ggcorrplot(cor_matrix)
```

-   Target `comb08` is highly positively correlated with:

    -   `city08`, `highway08`, `UCity`, `UHighway` (up to \~0.996), so the usage of these columns in a model would cause data leakage

-   Target `comb08` is strongly negatively correlated with:

    -   `barrels08`, `co2TailpipeGpm`, `cylinders`, `displ`, `fuelCost08` (up to \~ -0.7)
    -   however we know from the columns description that columns `barrels08`, `co2TailpipeGpm` and `fuelCost08` are derived from fuel consumption, so they would also cause data leakage
    -   `cylinders` and `displ` are suitable predictors but they are highly correlated with each other (\~0.91)

So, for the model, we will keep only column **`displ`**, because it has fewer missing values and a stronger correlation with target (\~ -0.71), and column `year`, because it has a moderate positive correlation (\~0.32) with target and has almost no correlation with `displ`.

## Categorical Variables

```{r}
cat_data <- data %>% 
  select(where(is.character))

ncol(cat_data)

names(cat_data)
```

The dataset has 23 categorical variables.

```{r}
cat_overview <- cat_data %>%
  summarise(across(everything(), ~n_distinct(.))) %>%
  pivot_longer(everything(),
               names_to = "variable",
               values_to = "n_levels")

cat_overview
```

Based on the number of unique values and the description of each column, we will further explore only `fuelType`, `drive`, `VClass` and `trany` columns, because they have rational amount of categories and are related to fuel efficiency.

```{r}
data %>%
  select(fuelType, drive, VClass, trany) %>%
  pivot_longer(everything(),
               names_to = "variable",
               values_to = "category") %>%
  filter(!is.na(category)) %>%
  group_by(variable, category) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(variable) %>%
  slice_max(n, n = 10) %>%
  ggplot(aes(x = reorder(category, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  facet_wrap(~variable, scales = "free") +
  theme_minimal()
```

The `fuelType` variable shows a highly imbalanced distribution, with a few dominant categories (`Regular` and `Premium`) accounting for most observations, while several alternative fuel types appear only rarely. We will group rare categories of the fuelType variable into an "Other" class.

The `drive` variable is moderately imbalanced. `Front-Wheel Drive` and `Rear-Wheel Drive` represent the majority of observations, while several four-wheel-drive variants and two-wheel-drive vehicles occur much less frequently. The presence of multiple related four-wheel-drive labels indicates potential redundancy in category definitions, so merging semantically similar categories could reduce noise.

The `trany` variable contains many transmission categories, with a clear dominance of `Automatic 4-speed` and `Manual 5-speed` types.

The VClass has 3 dominant categories - `Compact Cars`, `Midsize Cars`, and `Subcompact Cars`. The other catogories appear less frequently. Since vehicle class captures structural differences in size, it is expected to be an informative predictor and should be retained in the modeling stage.

```{r}
ggplot(data, aes(x = fuelType, y = comb08)) +
  geom_boxplot() +
  coord_flip()
```

The boxplot shows how different fuel type influences fuel consumption. We can see that the electricity type has higher values then gasoline types.

```{r}
data %>%
  filter(!is.na(drive)) %>%
  ggplot(aes(x = drive, y = comb08)) +
  geom_boxplot() +
  coord_flip()
```

Drive type clearly affects fuel efficiency. `Front-Wheel Drive` vehicles tend to have higher combined MPG, while `4-Wheel or All-Wheel Drive` has smaller values.

```{r}
data %>%
  filter(!is.na(VClass)) %>%
  mutate(VClass = fct_lump(VClass, n = 10)) %>%
  ggplot(aes(x = VClass, y = comb08)) +
  geom_boxplot() +
  coord_flip()
```

Vehicle class has a clear relationship with fuel efficiency. Smaller vehicles (`compact` and `subcompact cars`) tend to achieve higher MPG, while pickups, SUVs, and larger vehicles have smaller MPG values, so they are less fuel efficient.

```{r}
data %>%
  filter(!is.na(trany)) %>%
  mutate(trany = fct_lump(trany, n = 10)) %>%
  ggplot(aes(x = trany, y = comb08)) +
  geom_boxplot() +
  coord_flip()
```

Transmission type strongly differentiates vehicle efficiency. `Automatic (A1)` shows the highest MPG values, while other categories have smaller values.

## Tagret variable

```{r}
summary(data$comb08)
```

The target variable `comb08` (combined fuel efficiency) has strongly right-skewed distribution. Most vehicles have between 15 and 30 MPG, with a median value of 20 MPG. However, a long right tail is observed, extending up to 146 MPG.

```{r}
hist <- ggplot(data, aes(x = comb08)) +
  geom_histogram(bins = 40, fill = "steelblue") +
  theme_minimal()

box <- ggplot(data, aes(y = comb08)) +
  geom_boxplot(fill = "steelblue") +
  coord_flip() +
  theme_minimal()

hist + box + plot_layout(ncol = 1, nrow = 2) 
```

Based on the histogram, we can say that the distribution is not normal. The boxplot confirms the presence of numerous high-value outliers. However, these observations are not errors but represent a distinct technological group of vehicles with significantly higher efficiency.

## Preprocessing and Train/Test Split

```{r message=FALSE, warning=FALSE}
library(rsample)
library(recipes)
library(stringr)
library(forcats)
```

### Column Groups to Remove

We define explicit groups of problematic columns, then combine them and use `any_of()` so the code does not fail if a column does not exist.

#### Columns with Too Many Missing Values (`cols_missing`)

These variables have more than 60% missing values and cannot be reliably used in a general model.

```{r}
cols_missing <- c(
  "c240bDscr", "c240Dscr", "sCharger", "rangeA", "fuelType2",
  "guzzler", "evMotor", "atvType", "tCharger", "trans_dscr",
  "startStop", "mfrCode"
)
```

#### Identifier Columns (`cols_ids`)

These variables are identifiers or overly specific strings. Using them would encourage the model to memorize specific vehicles or model lines instead of learning generalizable patterns.

```{r}
cols_ids <- c("id", "engId", "model")
```

#### Data Leakage Columns (`cols_leakage`)

These variables are either directly equivalent to `comb08` or derived from the same fuel consumption, emissions, or cost calculations. Including them would incorrectly inflate model performance.

```{r}
cols_leakage <- c(
  "city08", "highway08", "UCity", "UHighway",
  "barrels08", "co2TailpipeGpm", "fuelCost08", "youSaveSpend",
  "comb08U", "city08U", "highway08U",
  "combA08", "cityA08", "highwayA08",
  "combA08U", "cityA08U", "highwayA08U",
  "co2", "co2A", "co2TailpipeAGpm",
  "feScore", "ghgScore", "ghgScoreA",
  "phevCity", "phevHwy", "phevComb",
  "combE", "cityE", "highwayE",
  "combinedCD", "combinedUF",
  "cityCD", "cityUF",
  "highwayCD", "highwayUF"
)
```

#### Sparse or Invalid Columns

As identified in the EDA, many numeric columns have median = 0 or contain encoded missing values (−1). These are typically relevant only for electric or hybrid vehicles. Including them in the general regression dataset would introduce large numbers of zeros, encoded missing values, or derived information. They could be useful for targeted analyses of EV or PHEV subsets, but are excluded from the baseline model.

```{r}
cols_sparse_or_invalid <- c(
  "barrelsA08", "charge120", "charge240", "charge240b",
  "city08U", "cityA08", "cityA08U", "cityCD", "cityE", "cityUF",
  "co2", "co2A", "co2TailpipeAGpm",
  "comb08U", "combA08", "combA08U", "combE",
  "combinedCD", "combinedUF",
  "feScore", "fuelCostA08", "ghgScore", "ghgScoreA",
  "highway08U", "highwayA08", "highwayA08U",
  "highwayCD", "highwayE", "highwayUF",
  "hlv", "hpv", "lv2", "lv4", "pv2", "pv4",
  "range", "rangeCity", "rangeCityA", "rangeHwy", "rangeHwyA",
  "UCityA", "UHighwayA",
  "phevCity", "phevHwy", "phevComb"
)
```

#### Unused Columns

- `eng_dscr` has ~37% missing values and is a free-text engine description — not suitable for standard regression without NLP.
- `createdOn` and `modifiedOn` are administrative metadata with no predictive value.
- `baseModel` has high cardinality and would support memorization of specific model lines rather than generalization.
- `fuelType` and `fuelType1` are replaced by the new `fuel_group` variable and removed from `vehicles_model` via an explicit `select()`. The raw columns remain intact in the original `data` object.

```{r}
cols_unused <- c("createdOn", "modifiedOn", "eng_dscr", "baseModel")
```

#### Combine All Removal Lists

```{r}
cols_to_remove <- unique(c(cols_missing, cols_ids, cols_leakage,
                            cols_sparse_or_invalid, cols_unused))

cat("Total columns scheduled for removal:", length(cols_to_remove), "\n")
```

### Feature Engineering: Electric Vehicles

Electric vehicles do not have a combustion engine and therefore have no `displ` or `cylinders`. Their missing values are structurally meaningful — they signal the absence of an engine, not random data loss. We capture this before any imputation.

- `is_electric`: TRUE if the vehicle runs on electricity only.
- `has_combustion_engine`: TRUE if the vehicle has a combustion engine.

For EVs, `displ` and `cylinders` are set to 0 — meaning zero displacement, not a statistical fill-in. Remaining missing values in non-EV rows will be handled later by the recipe.

```{r}
data_fe <- data %>%
  mutate(
    is_electric = (fuelType == "Electricity") | (fuelType1 == "Electricity"),
    is_electric = replace_na(is_electric, FALSE),
    has_combustion_engine = !is_electric,
    displ     = if_else(is_electric, 0, displ),
    cylinders = if_else(is_electric, 0, cylinders)
  )

cat("Electric vehicles:", sum(data_fe$is_electric), "\n")
cat("Combustion engine vehicles:", sum(data_fe$has_combustion_engine), "\n")
```

### Categorical Variable Simplification

All transformations below use deterministic rules (not statistics learned from data), so they can safely be applied before the train/test split.

#### `fuel_group` — Structured Fuel Type Grouping

Instead of lumping rare levels of `fuelType`, we create a new interpretable variable `fuel_group` that consolidates all fuel categories into six meaningful groups. The raw `fuelType` and `fuelType1` columns remain unchanged in the original `data` object. For modeling, only `fuel_group` is used as a predictor — `fuelType` and `fuelType1` are removed from `vehicles_model` to avoid duplicating the same information across multiple columns.

```{r}
data_fe <- data_fe %>%
  mutate(
    fuel_group = case_when(
      fuelType %in% c("Regular", "Premium", "Midgrade")       ~ "Gasoline",
      fuelType == "Diesel"                                     ~ "Diesel",
      fuelType == "Electricity"                                ~ "Electricity",
      fuelType %in% c(
        "Premium and Electricity",
        "Regular Gas and Electricity",
        "Premium Gas or Electricity",
        "Regular Gas or Electricity",
        "Electricity and Hydrogen"
      )                                                        ~ "Hybrid/Electric mix",
      fuelType %in% c("Gasoline or E85", "Premium or E85")    ~ "Flex-fuel/E85",
      fuelType %in% c(
        "CNG", "Hydrogen",
        "Gasoline or natural gas",
        "Gasoline or propane"
      )                                                        ~ "Alternative fuel",
      TRUE                                                     ~ "Other"
    )
  )

table(data_fe$fuel_group)
```

#### `drive` — Consolidate 4WD/AWD Variants

The EDA identified multiple overlapping 4-wheel and all-wheel drive labels. We consolidate them into three interpretable groups.

```{r}
data_fe <- data_fe %>%
  mutate(drive = case_when(
    str_detect(drive, "4-Wheel|All-Wheel") ~ "AWD/4WD",
    str_detect(drive, "Front-Wheel")       ~ "Front-Wheel Drive",
    str_detect(drive, "Rear-Wheel")        ~ "Rear-Wheel Drive",
    is.na(drive)                           ~ "Unknown",
    TRUE                                   ~ "Other"
  ))

table(data_fe$drive)
```

#### `VClass` — Keep Top 10 Categories

`VClass` has many levels. We retain the 10 most frequent and collapse the rest into "Other".

```{r}
data_fe <- data_fe %>%
  mutate(VClass = fct_lump(as.factor(VClass), n = 10, other_level = "Other"))

table(data_fe$VClass)
```

#### `trany` — Extract Structured Transmission Features

Instead of keeping the raw high-cardinality `trany` string, we extract four interpretable features:

- `transmission_type`: "Automatic", "Manual", or "Other"
- `is_cvt`: TRUE if the transmission uses continuously variable gear ratios
- `n_gears`: number of speeds extracted from text; 0 means "not applicable" (CVT or non-standard transmission)
- `has_discrete_gears`: TRUE if the number of gear steps was successfully extracted from `trany`

`has_discrete_gears` and `n_gears` are designed to be read together: `has_discrete_gears = FALSE` means a CVT or other transmission without a discrete gear count — this is not randomly missing data, but a structural property of the transmission. Setting `n_gears = 0` via `replace_na` provides a numeric sentinel that is safe to use in models without median imputation.

The raw `trany` column is then removed.

```{r}
data_fe <- data_fe %>%
  mutate(
    transmission_type = case_when(
      str_detect(trany, regex("^Automatic", ignore_case = TRUE)) ~ "Automatic",
      str_detect(trany, regex("^Manual",    ignore_case = TRUE)) ~ "Manual",
      TRUE ~ "Other"
    ),
    is_cvt  = str_detect(trany, regex("variable gear ratios", ignore_case = TRUE)),
    is_cvt  = replace_na(is_cvt, FALSE),
    n_gears = as.integer(str_match(trany, "(\\d+)-spd")[, 2]),
    has_discrete_gears = !is.na(n_gears),
    n_gears = replace_na(n_gears, 0)
  )

cat("Transmission type distribution:\n")
print(table(data_fe$transmission_type))
cat("\nCVT vehicles:", sum(data_fe$is_cvt), "\n")
cat("\nn_gears summary:\n")
print(summary(data_fe$n_gears))
cat("\nhas_discrete_gears distribution:\n")
print(table(data_fe$has_discrete_gears))
```

### Build Final Modeling Dataset

We apply all column removals. The raw `trany` is dropped because it is fully replaced by `transmission_type`, `is_cvt`, and `n_gears`. The raw `fuelType` and `fuelType1` are dropped because their information is captured by `fuel_group`.

Pure battery electric vehicles (BEVs) are excluded from the modeling dataset before the train/test split. The target variable `comb08` represents Miles Per Gallon (MPG) for combustion-engine vehicles, but for BEVs the EPA reports MPGe — an artificial conversion (33.7 kWh = 1 gallon equivalent) that does not reflect actual fuel consumption. Including BEVs would mix two structurally different quantities in a single regression target. Moreover, the features that drive MPGe — battery capacity, motor efficiency, regenerative braking — are not present in the dataset after removing high-missingness columns such as `evMotor`. The `is_electric` and `has_combustion_engine` flags are also dropped after filtering, as they become zero-variance constants.

```{r}
n_ev <- sum(data_fe$is_electric)

vehicles_model <- data_fe %>%
  filter(!is_electric) %>%
  select(-any_of(cols_to_remove)) %>%
  select(-any_of(c("trany", "fuelType", "fuelType1", "is_electric", "has_combustion_engine")))

cat("Electric vehicles excluded:", n_ev, "\n")
cat("Remaining vehicles:        ", nrow(vehicles_model), "\n")
cat("Dimensions of vehicles_model:", nrow(vehicles_model), "×", ncol(vehicles_model), "\n")
cat("\nColumns retained:\n")
print(names(vehicles_model))
```

`cylinders` is kept as a candidate predictor despite its strong correlation with `displ` (~0.91). We expect multicollinearity, but the final decision on whether to drop it is deferred to feature selection in a later scenario.

### Sanity Checks

```{r}
cat("Duplicate rows in vehicles_model:", sum(duplicated(vehicles_model)), "\n\n")

cat("summary(comb08):\n")
print(summary(vehicles_model$comb08))

cat("\nYear range:", range(vehicles_model$year, na.rm = TRUE), "\n")

cat("\ncomb08 <= 0:", sum(vehicles_model$comb08 <= 0, na.rm = TRUE), "\n")
cat("displ < 0:",    sum(vehicles_model$displ    < 0, na.rm = TRUE), "\n")
cat("cylinders < 0:", sum(vehicles_model$cylinders < 0, na.rm = TRUE), "\n")

missing_before <- colSums(is.na(vehicles_model))
cat("\nColumns with missing values (before recipe):\n")
print(missing_before[missing_before > 0])
cat("Total missing values:", sum(is.na(vehicles_model)), "\n")
```

After excluding BEVs, the remaining `comb08` values reflect actual MPG from combustion-engine and hybrid vehicles. Extreme values at the high end (diesel hybrids, efficient compact cars) are retained as valid observations.

### Train/Test Split

We use stratified splitting on `comb08` to ensure both sets reflect similar proportions of common and highly efficient vehicles. Because `comb08` is numeric, `rsample` bins it internally before stratifying.

```{r}
set.seed(123)

data_split <- initial_split(vehicles_model, prop = 0.8, strata = comb08)

train_data <- training(data_split)
test_data  <- testing(data_split)

cat("Train:", nrow(train_data), "×", ncol(train_data), "\n")
cat("Test: ", nrow(test_data),  "×", ncol(test_data),  "\n")
```

### Check Train/Test Distribution

```{r}
cat("Train comb08:\n"); print(summary(train_data$comb08))
cat("Test  comb08:\n"); print(summary(test_data$comb08))
```

```{r}
bind_rows(
  train_data %>% select(comb08) %>% mutate(split = "train"),
  test_data  %>% select(comb08) %>% mutate(split = "test")
) %>%
  ggplot(aes(x = comb08, fill = split)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("train" = "steelblue", "test" = "tomato")) +
  labs(
    title = "Distribution of comb08: Train vs. Test",
    x     = "Combined MPG (comb08)",
    y     = "Density"
  ) +
  theme_minimal()
```

The density curves confirm that stratification produces very similar `comb08` distributions in both sets.

### Cross-Validation Folds

We prepare 5-fold cross-validation on the training set, stratified by `comb08`. These folds will be used later for hyperparameter tuning and comparing models.

```{r}
folds <- vfold_cv(train_data, v = 5, strata = comb08)

cat("Number of CV folds:", nrow(folds), "\n")
print(folds)
```

### Preprocessing Recipe

The recipe is defined on `train_data` and fitted only on the training set to prevent leakage of test statistics into training:

- `step_unknown()` — handles unseen factor levels that may appear in the test set
- `step_other()` — collapses any residual rare categories below 1%
- `step_impute_median(displ, cylinders)` — fills the small number of numeric NAs in engine parameters for non-BEV rows; `n_gears` is intentionally set to 0 in feature engineering and does not need imputation
- `step_dummy()` — one-hot encodes all nominal predictors; required by linear regression methods
- `step_zv()` / `step_nzv()` — removes zero- and near-zero-variance columns, consistent with EDA findings on low-variability variables
- `step_normalize()` — standardizes all numeric predictors; required for ridge, lasso, elastic net, kNN, and PCA/PCR/PLS

```{r}
model_recipe <- recipe(comb08 ~ ., data = train_data) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%
  step_impute_median(displ, cylinders) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_nzv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

model_recipe
```

### Fit and Apply Recipe

The recipe is fitted exclusively on `train_data`. The same fitted statistics (medians, modes, dummy levels) are then applied to both sets.

```{r}
prep_recipe <- prep(model_recipe, training = train_data)

train_processed <- bake(prep_recipe, new_data = NULL)
test_processed  <- bake(prep_recipe, new_data = test_data)
```

### Verify Preprocessing Output

```{r}
cat("train_processed:", nrow(train_processed), "×", ncol(train_processed), "\n")
cat("test_processed: ", nrow(test_processed),  "×", ncol(test_processed),  "\n")

cat("\nMissing values — train_processed:", sum(is.na(train_processed)), "\n")
cat("Missing values — test_processed: ", sum(is.na(test_processed)),  "\n")

cat("\ncomb08 in train_processed:", "comb08" %in% names(train_processed), "\n")
cat("comb08 in test_processed: ", "comb08" %in% names(test_processed),  "\n")

n_predictors <- ncol(train_processed) - 1
cat("\nNumber of predictors after dummy encoding:", n_predictors, "\n")

cat("\nFirst 10 column names:\n")
print(head(names(train_processed), 10))
```

Both processed datasets contain no missing values in predictors, and `comb08` is retained as the regression target. The objects `vehicles_model`, `train_data`, `test_data`, `folds`, `model_recipe`, `prep_recipe`, `train_processed`, and `test_processed` are ready for regression scenarios 2, 3, and 5.

## Scenario 2: Parametric vs. Non-Parametric Models — Parametric Part

Parametric models assume that the relationship between predictors and the response follows a specific, fixed mathematical form — in regression, a linear combination of weighted features.

We compare four parametric regression methods:

1. **Linear Regression (OLS)** — the classical least-squares baseline, no regularization
2. **Ridge Regression** — OLS plus an L2 penalty that shrinks all coefficients toward zero
3. **LASSO** — OLS plus an L1 penalty that can shrink some coefficients to exactly zero, selecting features
4. **Elastic Net** — a blend of Ridge and LASSO penalties, combining both approaches

All four methods share the same preprocessing recipe defined in the previous section. We use the 5-fold cross-validation folds for hyperparameter tuning and model comparison, and evaluate final performance on the held-out test set.

```{r message=FALSE, warning=FALSE}
library(tidymodels)
library(broom)
library(glmnet)
```

We define the three evaluation metrics once and reuse them across every model.

```{r}
eval_metrics <- metric_set(rmse, rsq, mae)
```

---

### 1. Linear Regression (OLS)

Ordinary least squares minimizes the sum of squared residuals and imposes no constraint on coefficient size. It is our reference point — any benefit from regularization will appear as an improvement over this baseline.

```{r}
lm_spec <- linear_reg() %>%
  set_engine("lm") %>%
  set_mode("regression")

lm_wf <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(lm_spec)
```

`fit_resamples()` runs the complete preprocessing + fitting pipeline independently on each of the 5 folds, so no information from the validation fold leaks into training.

```{r warning=FALSE}
set.seed(123)

lm_cv <- lm_wf %>%
  fit_resamples(
    resamples = folds,
    metrics   = eval_metrics,
    control   = control_resamples(save_pred = TRUE)
  )

lm_cv_metrics <- collect_metrics(lm_cv)
cat("=== OLS — 5-fold CV ===\n")
cat("RMSE:", round(lm_cv_metrics %>% filter(.metric == "rmse") %>% pull(mean), 4),
    "  ±", round(lm_cv_metrics %>% filter(.metric == "rmse") %>% pull(std_err), 4), "\n")
cat("MAE: ", round(lm_cv_metrics %>% filter(.metric == "mae")  %>% pull(mean), 4), "\n")
cat("R²:  ", round(lm_cv_metrics %>% filter(.metric == "rsq")  %>% pull(mean), 4), "\n")
```

Now we fit on the full training set to inspect which predictors drive the predictions.

```{r}
lm_fit <- lm_wf %>% fit(data = train_data)

lm_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(term != "(Intercept)") %>%
  slice_max(abs(estimate), n = 20) %>%
  ggplot(aes(x = reorder(term, abs(estimate)), y = estimate)) +
  geom_col(aes(fill = estimate > 0)) +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
    labels = c("TRUE" = "Positive", "FALSE" = "Negative"),
    name   = "Direction"
  ) +
  labs(
    title = "OLS: Top 20 Predictors by Absolute Coefficient",
    x     = NULL,
    y     = "Coefficient estimate"
  ) +
  theme_minimal()
```

Blue bars indicate predictors that increase `comb08` (e.g. electric drivetrain), red bars indicate predictors that decrease it (e.g. larger engine displacement).

---

### 2. Ridge Regression

Ridge regression adds an L2 penalty — λ × Σβ² — to the OLS loss. This forces the optimiser to keep coefficients small, reducing variance at the cost of some bias. Ridge never removes a predictor entirely: all coefficients are shrunk toward zero, but remain non-zero. It is especially effective when predictors are correlated — a situation we identified between `displ` and `cylinders` in the EDA.

The regularization strength λ is controlled by `penalty`. We tune it over a log-spaced grid.

```{r}
ridge_spec <- linear_reg(penalty = tune(), mixture = 0) %>%
  set_engine("glmnet") %>%
  set_mode("regression")

ridge_wf <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(ridge_spec)
```

`glmnet` fits the entire regularization path in a single pass, so evaluating 50 penalty values has almost no extra cost compared to evaluating one.

```{r}
ridge_grid <- grid_regular(penalty(range = c(-4, 2)), levels = 50)
```

```{r warning=FALSE}
set.seed(123)

ridge_cv <- ridge_wf %>%
  tune_grid(
    resamples = folds,
    grid      = ridge_grid,
    metrics   = eval_metrics
  )

autoplot(ridge_cv)
```

At very low penalty (left end), Ridge behaves like OLS. As penalty increases, coefficients shrink and the model becomes more biased but less variable. The optimal point is where RMSE on the validation folds is minimized.

```{r}
best_ridge <- ridge_cv %>% select_best(metric = "rmse")

ridge_best_row <- show_best(ridge_cv, metric = "rmse", n = 1)
cat("=== Ridge — best hyperparameter ===\n")
cat("Best penalty:", round(best_ridge$penalty, 6), "\n")
cat("CV RMSE:     ", round(ridge_best_row$mean, 4),
    "  ±", round(ridge_best_row$std_err, 4), "\n")
```

```{r}
ridge_final_wf <- ridge_wf %>% finalize_workflow(best_ridge)
ridge_fit       <- ridge_final_wf %>% fit(data = train_data)
```

```{r}
ridge_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(term != "(Intercept)") %>%
  slice_max(abs(estimate), n = 20) %>%
  ggplot(aes(x = reorder(term, abs(estimate)), y = estimate)) +
  geom_col(aes(fill = estimate > 0)) +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
    labels = c("TRUE" = "Positive", "FALSE" = "Negative"),
    name   = "Direction"
  ) +
  labs(
    title = paste0("Ridge: Top 20 Predictors (penalty = ",
                   round(best_ridge$penalty, 5), ")"),
    x     = NULL,
    y     = "Coefficient estimate"
  ) +
  theme_minimal()
```

Compared to OLS, Ridge produces smaller coefficient magnitudes across all predictors. The ordering and direction of the most important features generally remains the same, but correlated predictors share their weight more evenly.

---

### 3. LASSO Regression

LASSO (Least Absolute Shrinkage and Selection Operator) uses an L1 penalty — λ × Σ|β|. The key geometric property of L1 is that the regularized solution often lands exactly at zero for many coefficients. LASSO therefore performs simultaneous regression and feature selection: it automatically identifies the most informative predictors and removes the rest.

```{r}
lasso_spec <- linear_reg(penalty = tune(), mixture = 1) %>%
  set_engine("glmnet") %>%
  set_mode("regression")

lasso_wf <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(lasso_spec)
```

LASSO typically needs smaller penalties than Ridge to achieve similar shrinkage, so we search a narrower range.

```{r}
lasso_grid <- grid_regular(penalty(range = c(-4, 0)), levels = 50)
```

```{r warning=FALSE}
set.seed(123)

lasso_cv <- lasso_wf %>%
  tune_grid(
    resamples = folds,
    grid      = lasso_grid,
    metrics   = eval_metrics
  )

autoplot(lasso_cv)
```

```{r}
best_lasso <- lasso_cv %>% select_best(metric = "rmse")

lasso_best_row <- show_best(lasso_cv, metric = "rmse", n = 1)
cat("=== LASSO — best hyperparameter ===\n")
cat("Best penalty:", round(best_lasso$penalty, 6), "\n")
cat("CV RMSE:     ", round(lasso_best_row$mean, 4),
    "  ±", round(lasso_best_row$std_err, 4), "\n")
```

```{r}
lasso_final_wf <- lasso_wf %>% finalize_workflow(best_lasso)
lasso_fit       <- lasso_final_wf %>% fit(data = train_data)
```

One of LASSO's main advantages is sparsity. Let's see how many predictors it retained.

```{r}
lasso_coefs <- lasso_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(term != "(Intercept)")

cat("=== LASSO — sparsity ===\n")
cat("Total predictors available:", nrow(lasso_coefs), "\n")
cat("Kept by LASSO (non-zero):  ", sum(lasso_coefs$estimate != 0), "\n")
cat("Removed by LASSO (zero):   ", sum(lasso_coefs$estimate == 0), "\n")
if (sum(lasso_coefs$estimate == 0) > 0) {
  cat("Zeroed out predictors:\n")
  print(lasso_coefs %>% filter(estimate == 0) %>% pull(term))
}
```

```{r}
lasso_coefs %>%
  filter(estimate != 0) %>%
  slice_max(abs(estimate), n = 20) %>%
  ggplot(aes(x = reorder(term, abs(estimate)), y = estimate)) +
  geom_col(aes(fill = estimate > 0)) +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
    labels = c("TRUE" = "Positive", "FALSE" = "Negative"),
    name   = "Direction"
  ) +
  labs(
    title = paste0("LASSO: Top 20 Non-Zero Predictors (penalty = ",
                   round(best_lasso$penalty, 5), ")"),
    x     = NULL,
    y     = "Coefficient estimate"
  ) +
  theme_minimal()
```

LASSO retains only the most informative predictors, which makes the resulting model more parsimonious and directly interpretable compared to OLS or Ridge. Predictors that survive the L1 penalty are the ones that the algorithm judged worth the cost of adding a parameter.

---

### 4. Elastic Net

Elastic Net combines both L1 and L2 penalties: λ × [α × Σ|β| + (1 − α) × Σβ²]. The `mixture` parameter in tidymodels corresponds to α:  `mixture = 0` gives pure Ridge, `mixture = 1` gives pure LASSO. By tuning both `penalty` and `mixture`, Elastic Net finds the best blend — it can perform feature selection like LASSO while also handling correlated predictors like Ridge.

```{r}
en_spec <- linear_reg(penalty = tune(), mixture = tune()) %>%
  set_engine("glmnet") %>%
  set_mode("regression")

en_wf <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(en_spec)
```

We use a regular 10 × 10 grid — 10 penalty values and 10 mixture values — giving 100 combinations. `glmnet` evaluates all penalty values for each mixture in a single path, so this is still fast.

```{r}
en_grid <- grid_regular(
  penalty(range = c(-4, 0)),
  mixture(range = c(0, 1)),
  levels = 10
)
```

```{r warning=FALSE}
set.seed(123)

en_cv <- en_wf %>%
  tune_grid(
    resamples = folds,
    grid      = en_grid,
    metrics   = eval_metrics
  )

autoplot(en_cv)
```

```{r}
best_en <- en_cv %>% select_best(metric = "rmse")

en_best_row <- show_best(en_cv, metric = "rmse", n = 1)
cat("=== Elastic Net — best hyperparameters ===\n")
cat("Best penalty:", round(best_en$penalty, 6), "\n")
cat("Best mixture:", round(best_en$mixture, 4),
    " (0 = Ridge, 1 = LASSO)\n")
cat("CV RMSE:     ", round(en_best_row$mean, 4),
    "  ±", round(en_best_row$std_err, 4), "\n")
```

The optimal `mixture` value tells us how the data prefers to be regularized. If the best mixture is close to 1, LASSO-like behavior is preferred; if it is close to 0, Ridge-like behavior is preferred.

```{r}
en_final_wf <- en_wf %>% finalize_workflow(best_en)
en_fit       <- en_final_wf %>% fit(data = train_data)
```

---

### Comparing All Parametric Models

We evaluate all four models on the held-out test set. Because each model is a full `workflow()` that includes the preprocessing recipe, we pass the raw `test_data` directly and the recipe transformations are applied automatically.

We write a small helper function so we do not repeat the same prediction + metric code four times.

```{r}
get_test_metrics <- function(fit, model_name) {
  fit %>%
    predict(new_data = test_data) %>%
    bind_cols(test_data %>% select(comb08)) %>%
    summarise(
      model = model_name,
      RMSE  = rmse_vec(truth = comb08, estimate = .pred),
      MAE   = mae_vec( truth = comb08, estimate = .pred),
      Rsq   = rsq_vec( truth = comb08, estimate = .pred)
    )
}
```

```{r}
test_comparison <- bind_rows(
  get_test_metrics(lm_fit,    "Linear Regression"),
  get_test_metrics(ridge_fit, "Ridge"),
  get_test_metrics(lasso_fit, "LASSO"),
  get_test_metrics(en_fit,    "Elastic Net")
) %>%
  arrange(RMSE)

test_comparison
```

```{r}
test_comparison %>%
  pivot_longer(cols = c(RMSE, MAE, Rsq),
               names_to  = "metric",
               values_to = "value") %>%
  ggplot(aes(x = reorder(model, value), y = value, fill = model)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~metric, scales = "free_x") +
  coord_flip() +
  labs(
    title = "Test Set Performance: Parametric Models",
    x     = NULL,
    y     = "Value"
  ) +
  theme_minimal()
```

For RMSE and MAE, lower is better. For R², higher is better. The bar chart lets us compare the four models at a glance across all three metrics simultaneously.

#### Cross-Validation vs. Test Set RMSE

As a sanity check, we compare the best CV RMSE achieved during tuning with the final test RMSE. A close match confirms that our cross-validation estimate was reliable and there is no significant overfitting.

```{r}
cv_rmse_comparison <- tibble(
  model   = c("Linear Regression", "Ridge", "LASSO", "Elastic Net"),
  cv_rmse = c(
    collect_metrics(lm_cv)    %>% filter(.metric == "rmse") %>% pull(mean),
    show_best(ridge_cv, metric = "rmse", n = 1) %>% pull(mean),
    show_best(lasso_cv, metric = "rmse", n = 1) %>% pull(mean),
    show_best(en_cv,    metric = "rmse", n = 1) %>% pull(mean)
  )
) %>%
  left_join(
    test_comparison %>% select(model, test_rmse = RMSE),
    by = "model"
  ) %>%
  mutate(gap = abs(test_rmse - cv_rmse)) %>%
  arrange(cv_rmse)

cv_rmse_comparison
```

A small `gap` value confirms that each model generalizes to new data about as well as it performed during cross-validation, with no significant overfit to the training set.

#### Actual vs. Predicted — All Models

Finally, we plot actual versus predicted values for all four models on the test set. Points along the diagonal red line indicate perfect predictions.

```{r}
all_test_preds <- bind_rows(
  lm_fit    %>% predict(new_data = test_data) %>% bind_cols(test_data %>% select(comb08)) %>% mutate(model = "Linear Regression"),
  ridge_fit %>% predict(new_data = test_data) %>% bind_cols(test_data %>% select(comb08)) %>% mutate(model = "Ridge"),
  lasso_fit %>% predict(new_data = test_data) %>% bind_cols(test_data %>% select(comb08)) %>% mutate(model = "LASSO"),
  en_fit    %>% predict(new_data = test_data) %>% bind_cols(test_data %>% select(comb08)) %>% mutate(model = "Elastic Net")
)

all_test_preds %>%
  ggplot(aes(x = comb08, y = .pred)) +
  geom_point(alpha = 0.15, color = "steelblue", size = 0.8) +
  geom_abline(slope = 1, intercept = 0, color = "tomato", linewidth = 0.8) +
  facet_wrap(~model) +
  labs(
    title = "Actual vs. Predicted — Test Set (Parametric Models)",
    x     = "Actual combined MPG",
    y     = "Predicted combined MPG"
  ) +
  theme_minimal()
```

With battery electric vehicles excluded, all observations in this plot share the same physical interpretation of `comb08` — actual fuel consumption in MPG. Points along the diagonal indicate accurate predictions. Any remaining scatter at higher MPG values reflects plug-in hybrids and highly efficient compacts where the linear model may still be under-fitted, but the fan-shaped distortion caused by MPGe outliers is no longer present.

---

### Summary: Parametric Models

All four parametric models share the same fundamental assumption: the relationship between predictors and `comb08` is linear. Their differences lie entirely in regularization:

| Model | Penalty | Sparsity | Main strength |
|---|---|---|---|
| OLS | None | No | Full interpretability, baseline |
| Ridge | L2 (sum of β²) | No | Handles correlated predictors |
| LASSO | L1 (sum of \|β\|) | Yes | Automatic feature selection |
| Elastic Net | L1 + L2 | Yes | Adaptive blend of Ridge and LASSO |

On a large dataset like ours (~48 000 rows), the performance differences between OLS and the regularized models tend to be modest — there is enough data that variance is already well-controlled. The greater advantage of Ridge, LASSO, and Elastic Net is **reproducibility**: the penalty stabilizes coefficient estimates, making the model less sensitive to random variation in the training sample. LASSO additionally enhances **explainability** by producing a sparse model that a practitioner can read through quickly without being distracted by dozens of near-zero coefficients.

---

### Conclusion: Are Parametric Models Suitable for This Problem?

#### Impact of excluding battery electric vehicles

Before any model comparison, it is worth quantifying the effect of the BEV exclusion decision. The original mixed dataset (49 846 vehicles) contained 1 425 battery electric vehicles whose `comb08` values represent MPGe — a unit-converted proxy incompatible with the MPG values of combustion vehicles. Removing them reduced the dataset to 48 421 observations and had a substantial impact on model accuracy: RMSE dropped from 8.11 MPG to **3.36 MPG** (a 59% reduction) and MAE from 5.15 to **2.19 MPG**, while R² improved only modestly from 0.642 to 0.661. The small R² change is expected — R² is scale-independent and measures the proportion of explained variance, which remains similar once the target distribution is homogeneous. The large absolute improvement in RMSE confirms that the BEV segment was the dominant source of prediction error, not the model itself.

#### Overall performance

All four models achieved nearly identical results on the held-out test set: **RMSE ≈ 3.36 MPG, MAE ≈ 2.19 MPG, R² ≈ 0.661**. In context, the median `comb08` in this dataset is 20 MPG, so the average absolute prediction error of 2.19 MPG represents roughly 11% of the median — a reasonable level of accuracy for a linear model using only 21 features. The range of the target variable after BEV exclusion is 7–74 MPG, with the bulk of observations concentrated between 15 and 30 MPG, which is where the model performs best.

#### Why regularization did not help

The cross-validation tuning selected `penalty = 0.0001` (the smallest value in the search grid) for Ridge, LASSO, and Elastic Net alike. This is not a failure of the tuning procedure — it is a meaningful result. With 38 735 training observations and only 21 predictors after dummy encoding, the ordinary least squares estimator already operates in a regime where variance is very low. Regularization is most beneficial when the number of predictors is large relative to the number of observations, or when predictors are strongly collinear. Neither condition dominates here: the recipe collapsed rare factor levels, removed near-zero-variance columns, and normalized all numeric predictors, leaving a well-conditioned design matrix.

The near-zero optimal penalty also explains why LASSO retained all 21 predictors. Variable selection via LASSO is a threshold phenomenon — coefficients are set to exactly zero only when the penalty exceeds a level specific to each predictor's contribution. At `penalty = 1e-4`, every predictor carries enough signal that removing it would increase prediction error. This is a positive diagnostic: it means the preprocessing produced a lean, non-redundant feature set where every column earns its place.

The Elastic Net selected `mixture = 0.222`, indicating a slight preference for Ridge-like behavior over LASSO. Combined with a near-zero penalty, this further confirms that the dataset does not require aggressive shrinkage.

#### Why the models are still a reasonable choice

Despite the uniformity of results across all four methods, parametric models serve several legitimate purposes in this setting:

1. **Interpretability.** The coefficient plots reveal that `cylinders`, fuel type (gasoline vs. alternative), `year`, and `displ` are the dominant drivers of fuel efficiency — in directions fully consistent with domain knowledge. More powerful black-box methods may improve RMSE by a small margin, but at the cost of being unable to explain *why* a vehicle receives a specific MPG rating. For regulatory, engineering, or audit use cases, interpretability is a hard requirement.

2. **Generalization.** The CV–test RMSE gap is below 0.005 for all four models, confirming essentially zero overfitting. The cross-validation estimates were reliable predictors of held-out performance. A model that generalizes this cleanly requires minimal post-deployment monitoring.

3. **Appropriate baseline.** Before applying more flexible methods, it is standard practice to establish what a well-tuned linear model can achieve. RMSE ≈ 3.36 MPG and R² ≈ 0.661 set the benchmark any non-parametric method must surpass to justify its added complexity.

#### Where parametric models fall short

The core limitation is the linearity assumption. Fuel efficiency is shaped by interactions and non-linearities that a linear model cannot capture without explicit feature engineering:

- The effect of `cylinders` on MPG is not constant — the marginal cost of adding one cylinder differs between a 4-cylinder and a V8 engine.
- Plug-in hybrids (PHEVs) and mild hybrids introduce efficiency characteristics that depend on the interplay between electric and combustion components, not captured by individual feature coefficients.
- Manufacturer-specific engineering decisions (turbocharging, variable valve timing, cylinder deactivation) create within-brand patterns that a global linear model averages away.

These limitations appear as residual scatter in the actual vs. predicted plots, particularly above 40 MPG where highly efficient compacts and hybrids cluster. Unlike the previous runs, this scatter no longer shows a systematic fan shape driven by MPGe outliers — what remains is genuine model uncertainty about non-linear efficiency relationships.

#### Final assessment

Parametric models are well-suited to this problem as an **interpretable, stable, and computationally efficient solution for combustion-engine vehicle fuel efficiency prediction**. After appropriate exclusion of battery electric vehicles, they achieve RMSE ≈ 3.36 MPG (≈11% of the median), explain 66% of the variance in fuel economy using 21 features, and generalize to unseen data with negligible loss. Coefficients align with physical intuition and are stable across all four model variants. The primary remaining limitation is the inability to capture non-linear interactions between engine parameters — a structural constraint of the linear model family that motivates the use of non-parametric methods as a comparison point.
