---
title: "Vehicles dataset"
author: "Dominika Lukašáková, Lukáš Lovás"
date: "2026-04-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#setwd("D:/Škola/Ing/2.semester/OZNAL_projekt")
setwd("C:/school/oznal/OZNAL_Project")
```

## EDA

Vehicles dataset\
Source: <https://fueleconomy.gov/feg/ws/index.shtml>

Based on the column descriptions, we could potentially predict fuel consumption (MPG).

-   **MPG (Miles per gallon)** – a measure of fuel economy in automobiles, higher number= higher efficiency\
-   **Target variable:** `comb08`

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(ggcorrplot)
library(patchwork)
```

```{r}
rm(list = ls())
data <- read_csv("data/vehicles.csv")
data
View(data)
```

## Basic Structure

First, we check dataset dimensions and data types.

```{r}
dim(data)
```

The dataset has 49,846 rows and 84 columns.

```{r}
glimpse(data)
```

```{r}
data %>%
  summarise(across(everything(), class))
```

The dataset contains columns of type: character, double, logical. Columns createdOn and modifiedOn are dates stored as strings. The dataset has some missing values.

## Missing Values

```{r}
sum(is.na(data))
```

The dataset contains 538 466 missing values.

# Identify columns with missing values

```{r}
missing_summary <- data %>%
  summarise(across(everything(), ~mean(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_pct") %>%
  arrange(desc(missing_pct))

missing_summary %>%
  print(n = 84)
```

Columns with more than 60% missing values:

-   `c240bDscr`, `c240Dscr`, `sCharger`, `rangeA`, `fuelType2`, `guzzler`, `evMotor`,
-   `atvType`, `tCharger`, `trans_dscr`, `startStop`, `mfrCode`

We will not use these columns and remove them later.

-   Column `eng_dscr` has \~37% missing values
-   Columns `cylinders`, `displ`, `drive`, `trany`, `fuelType1` contain only a small amount of missing data (up to \~3%), so they could potentially be imputed.

## Unique Values

Next we identify columns with unique values.

```{r}
data %>%
  summarise(across(everything(), n_distinct)) %>%
  pivot_longer(everything(), 
               names_to = "variable", 
               values_to = "n_unique") %>%
  arrange(desc(n_unique)) %>%
  print(n = 84)
```

From these we filter out non-informative columns. We remove variables with nearly all unique values, as they act as identifiers and do not provide generalizable patterns for prediction. Also variables with very low variability (binary or near-constant features) were excluded, as they provide minimal information for explaining variation in the target variable.

```{r}
data %>%
  summarise(across(everything(), n_distinct)) %>%
  pivot_longer(everything(), 
               names_to = "variable", 
               values_to = "n_unique") %>%
  arrange(desc(n_unique)) %>%
  filter(!(n_unique >= 0.9 * nrow(data))) %>%
  filter(!(n_unique <= 2)) %>%
  print(n = 84)
```

-   We should also remove `model` column, because its extremely high number of unique values makes it impractical for modeling, and `engId,` as it also represents an identifier.
-   The dataset contains a large number of variables with high cardinality, indicating a high-dimensional feature space that requires careful feature selection.

## Numerical Variables

First, we check the distribution of numerical variables.

```{r}
num_data <- data %>% 
  select(where(is.numeric)) %>% 
  select(-id, -engId)

summary(num_data)
```

-   Many columns have median = 0 because most vehicles do not have that feature, for example `highwayE` applies only to electric vehicles. So, these variables might be informative for specific subsets, but not for the general fuel consumption prediction.

-   Some columns have min and median = -1, which is unusual and likely represents missing values. The columns are: `co2`, `co2A`, `feScore`, `ghgScore`, `ghgScoreA`.

-   Although column `youSaveSpend` does not have issues in terms of distribution, its definition reveals that it is directly derived from fuel consumption and fuel cost variables. As a result, it contains information closely related to the target variable and could introduce data leakage and artificially inflating model performance.

For further analysis we will remove: - columns with median = 0 or -1 - column `youSaveSpend`

```{r}
num_data <- num_data %>%
  select(where(~ {
    med <- median(., na.rm = TRUE)
    !(med == 0 | med == -1)
  })) %>%
  select(-youSaveSpend)

summary(num_data)
```

-   We are left with 11 numerical columns.

-   The similarity between mean and median values suggests that most variables are not heavily skewed, although there might be some outliers.

```{r}
num_data %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~variable, scales = "free") +
  theme_minimal()
# A warning is displayed because of NA in columns cylinders and displ
```

Most variables show right-skewed distributions, with the majority of observations with lower values and with a long tail toward higher values.

A few variables, such as `barrels08` and `co2TailpipeGpm` appear approximately normally distributed, though still with slight positive skew. In contrast, `cylinders` is a discrete variable, with clear peaks at common engine configurations - 4, 6, and 8 cylinders.

The variable `year` shows a relatively even but slightly increasing distribution toward more recent years, indicating a higher representation of newer vehicles in the dataset.

```{r}
num_data %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "value") %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()
# A warning is displayed because of NA in columns cylinders and displ
```

-   The variable `fuelCost08` shows a wide IQR and numerous outliers.
-   Column `co2TailpipeGpm` has a relatively compact central distribution but still includes several higher outliers.
-   The `year`column has a narrow spread from 1984 to 2026, indicating that most observations are concentrated within a similar time range, so there are no outliers.

We will plot another boxplot without columns `year`, `fuelCost08`, and `co2TailpipeGpm` due to differences in scale:

```{r}
num_data %>%
  select(-year, -fuelCost08, -co2TailpipeGpm) %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "value") %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()
# A warning is displayed because of NA in columns cylinders and displ
```

-   All columns have a significant amount of outsiders on the right side.

The boxplots reveal that most variables are right-skewed, as indicated by longer upper whiskers and a large number of high-value outliers. This pattern is especially clear for fuel economy variables (`city08`, `highway08`, `comb08`, `UCity`, `UHighway`), where a small number of highly efficient vehicles extend the upper range.

Variables such as `barrels08` and `displ` exhibit moderate right skewness with some extreme values. In contrast, `cylinders` is a discrete variable, making its boxplot less informative and leading to apparent outliers due to limited possible values.

Overall, the boxplots highlight skewness and the presence of outliers.

## Correlation Analysis

```{r}
cor_matrix <- num_data %>%
  select(where(is.numeric)) %>%
  cor(use = "pairwise.complete.obs")

cor_matrix
ggcorrplot(cor_matrix)
```

-   Target `comb08` is highly positively correlated with:

    -   `city08`, `highway08`, `UCity`, `UHighway` (up to \~0.996), so the usage of these columns in a model would cause data leakage

-   Target `comb08` is strongly negatively correlated with:

    -   `barrels08`, `co2TailpipeGpm`, `cylinders`, `displ`, `fuelCost08` (up to \~ -0.7)
    -   however we know from the columns description that columns `barrels08`, `co2TailpipeGpm` and `fuelCost08` are derived from fuel consumption, so they would also cause data leakage
    -   `cylinders` and `displ` are suitable predictors but they are highly correlated with each other (\~0.91)

So, for the model, we will keep only column **`displ`**, because it has fewer missing values and a stronger correlation with target (\~ -0.71), and column `year`, because it has a moderate positive correlation (\~0.32) with target and has almost no correlation with `displ`.

## Categorical Variables

```{r}
cat_data <- data %>% 
  select(where(is.character))

ncol(cat_data)

names(cat_data)
```

The dataset has 23 categorical variables.

```{r}
cat_overview <- cat_data %>%
  summarise(across(everything(), ~n_distinct(.))) %>%
  pivot_longer(everything(),
               names_to = "variable",
               values_to = "n_levels")

cat_overview
```

Based on the number of unique values and the description of each column, we will further explore only `fuelType`, `drive`, `VClass` and `trany` columns, because they have rational amount of categories and are related to fuel efficiency.

```{r}
data %>%
  select(fuelType, drive, VClass, trany) %>%
  pivot_longer(everything(),
               names_to = "variable",
               values_to = "category") %>%
  filter(!is.na(category)) %>%
  group_by(variable, category) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(variable) %>%
  slice_max(n, n = 10) %>%
  ggplot(aes(x = reorder(category, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  facet_wrap(~variable, scales = "free") +
  theme_minimal()
```

The `fuelType` variable shows a highly imbalanced distribution, with a few dominant categories (`Regular` and `Premium`) accounting for most observations, while several alternative fuel types appear only rarely. We will group rare categories of the fuelType variable into an "Other" class.

The `drive` variable is moderately imbalanced. `Front-Wheel Drive` and `Rear-Wheel Drive` represent the majority of observations, while several four-wheel-drive variants and two-wheel-drive vehicles occur much less frequently. The presence of multiple related four-wheel-drive labels indicates potential redundancy in category definitions, so merging semantically similar categories could reduce noise.

The `trany` variable contains many transmission categories, with a clear dominance of `Automatic 4-speed` and `Manual 5-speed` types.

The VClass has 3 dominant categories - `Compact Cars`, `Midsize Cars`, and `Subcompact Cars`. The other catogories appear less frequently. Since vehicle class captures structural differences in size, it is expected to be an informative predictor and should be retained in the modeling stage.

```{r}
ggplot(data, aes(x = fuelType, y = comb08)) +
  geom_boxplot() +
  coord_flip()
```

The boxplot shows how different fuel type influences fuel consumption. We can see that the electricity type has higher values then gasoline types.

```{r}
data %>%
  filter(!is.na(drive)) %>%
  ggplot(aes(x = drive, y = comb08)) +
  geom_boxplot() +
  coord_flip()
```

Drive type clearly affects fuel efficiency. `Front-Wheel Drive` vehicles tend to have higher combined MPG, while `4-Wheel or All-Wheel Drive` has smaller values.

```{r}
data %>%
  filter(!is.na(VClass)) %>%
  mutate(VClass = fct_lump(VClass, n = 10)) %>%
  ggplot(aes(x = VClass, y = comb08)) +
  geom_boxplot() +
  coord_flip()
```

Vehicle class has a clear relationship with fuel efficiency. Smaller vehicles (`compact` and `subcompact cars`) tend to achieve higher MPG, while pickups, SUVs, and larger vehicles have smaller MPG values, so they are less fuel efficient.

```{r}
data %>%
  filter(!is.na(trany)) %>%
  mutate(trany = fct_lump(trany, n = 10)) %>%
  ggplot(aes(x = trany, y = comb08)) +
  geom_boxplot() +
  coord_flip()
```

Transmission type strongly differentiates vehicle efficiency. `Automatic (A1)` shows the highest MPG values, while other categories have smaller values.

## Tagret variable

```{r}
summary(data$comb08)
```

The target variable `comb08` (combined fuel efficiency) has strongly right-skewed distribution. Most vehicles have between 15 and 30 MPG, with a median value of 20 MPG. However, a long right tail is observed, extending up to 146 MPG.

```{r}
hist <- ggplot(data, aes(x = comb08)) +
  geom_histogram(bins = 40, fill = "steelblue") +
  theme_minimal()

box <- ggplot(data, aes(y = comb08)) +
  geom_boxplot(fill = "steelblue") +
  coord_flip() +
  theme_minimal()

hist + box + plot_layout(ncol = 1, nrow = 2) 
```

Based on the histogram, we can say that the distribution is not normal. The boxplot confirms the presence of numerous high-value outliers. However, these observations are not errors but represent a distinct technological group of vehicles with significantly higher efficiency.

## Preprocessing and Train/Test Split

We now prepare the data for regression modeling. This section covers column selection, feature engineering, categorical simplification, sanity checks, train/test split, cross-validation folds, and a preprocessing recipe. No models are trained here.

```{r message=FALSE, warning=FALSE}
library(rsample)
library(recipes)
library(stringr)
library(forcats)
```

### Column Groups to Remove

We define explicit groups of problematic columns, then combine them and use `any_of()` so the code does not fail if a column does not exist.

#### Columns with Too Many Missing Values (`cols_missing`)

These variables have more than 60% missing values and cannot be reliably used in a general model.

```{r}
cols_missing <- c(
  "c240bDscr", "c240Dscr", "sCharger", "rangeA", "fuelType2",
  "guzzler", "evMotor", "atvType", "tCharger", "trans_dscr",
  "startStop", "mfrCode"
)
```

#### Identifier Columns (`cols_ids`)

These variables are identifiers or overly specific strings. Using them would encourage the model to memorize specific vehicles or models.

```{r}
cols_ids <- c("id", "engId", "model")
```

#### Data Leakage Columns (`cols_leakage`)

These variables are either directly equivalent to `comb08` or derived from the same fuel consumption, emissions, or cost calculations. Including them would incorrectly inflate model performance.

```{r}
cols_leakage <- c(
  "city08", "highway08", "UCity", "UHighway",
  "barrels08", "co2TailpipeGpm", "fuelCost08", "youSaveSpend",
  "comb08U", "city08U", "highway08U",
  "combA08", "cityA08", "highwayA08",
  "combA08U", "cityA08U", "highwayA08U",
  "co2", "co2A", "co2TailpipeAGpm",
  "feScore", "ghgScore", "ghgScoreA",
  "phevCity", "phevHwy", "phevComb",
  "combE", "cityE", "highwayE",
  "combinedCD", "combinedUF",
  "cityCD", "cityUF",
  "highwayCD", "highwayUF"
)
```

#### Sparse or Invalid Columns

As identified in the EDA, many numeric columns have median = 0 or contain encoded missing values (−1). These are typically relevant only for electric or hybrid vehicles. Including them in the general regression dataset would introduce large numbers of zeros, encoded missing values, or derived information.

```{r}
cols_sparse_or_invalid <- c(
  "barrelsA08", "charge120", "charge240", "charge240b",
  "city08U", "cityA08", "cityA08U", "cityCD", "cityE", "cityUF",
  "co2", "co2A", "co2TailpipeAGpm",
  "comb08U", "combA08", "combA08U", "combE",
  "combinedCD", "combinedUF",
  "feScore", "fuelCostA08", "ghgScore", "ghgScoreA",
  "highway08U", "highwayA08", "highwayA08U",
  "highwayCD", "highwayE", "highwayUF",
  "hlv", "hpv", "lv2", "lv4", "pv2", "pv4",
  "range", "rangeCity", "rangeCityA", "rangeHwy", "rangeHwyA",
  "UCityA", "UHighwayA",
  "phevCity", "phevHwy", "phevComb"
)
```

#### Unused Columns

-   `eng_dscr` has \~37% missing values and is a free-text engine description.
-   `createdOn` and `modifiedOn` are metadata with no predictive value.
-   `baseModel` would support memorization of specific model lines rather than generalization.
-   `fuelType` and `fuelType1` are replaced by the new `fuel_group` variable.

```{r}
cols_unused <- c("createdOn", "modifiedOn", "eng_dscr", "baseModel")
```

#### Combine All Removal Lists

```{r}
cols_to_remove <- unique(c(cols_missing, cols_ids, cols_leakage,
                            cols_sparse_or_invalid, cols_unused))

cat("Total columns scheduled for removal:", length(cols_to_remove), "\n")
```

### Feature Engineering: Electric Vehicles

Electric vehicles do not have a combustion engine and therefore have no `displ` or `cylinders`. Their missing values are structurally meaningful — they signal the absence of an engine, not random data loss. We capture this before any imputation.

-   `is_electric`: TRUE if the vehicle runs on electricity only.
-   `has_combustion_engine`: TRUE if the vehicle has a combustion engine.

For EVs, `displ` and `cylinders` are set to 0 — meaning zero displacement, not a statistical fill-in. Remaining missing values in non-EV rows will be handled later by the recipe.

```{r}
data_fe <- data %>%
  mutate(
    is_electric = (fuelType == "Electricity") | (fuelType1 == "Electricity"),
    is_electric = replace_na(is_electric, FALSE),
    has_combustion_engine = !is_electric,
    displ     = if_else(is_electric, 0, displ),
    cylinders = if_else(is_electric, 0, cylinders)
  )

cat("Electric vehicles:", sum(data_fe$is_electric), "\n")
cat("Combustion engine vehicles:", sum(data_fe$has_combustion_engine), "\n")
```

### Categorical Variable Simplification

All transformations below use deterministic rules (not statistics learned from data), so they can safely be applied before the train/test split.

#### `fuel_group` — Structured Fuel Type Grouping

Instead of lumping rare levels of `fuelType`, we create a new interpretable variable `fuel_group` that consolidates all fuel categories into six meaningful groups. The raw `fuelType` and `fuelType1` columns remain unchanged in the original `data` object. For modeling, only `fuel_group` is used as a predictor — `fuelType` and `fuelType1` are removed from `vehicles_model` to avoid duplicating the same information across multiple columns.

```{r}
data_fe <- data_fe %>%
  mutate(
    fuel_group = case_when(
      fuelType %in% c("Regular", "Premium", "Midgrade")       ~ "Gasoline",
      fuelType == "Diesel"                                     ~ "Diesel",
      fuelType == "Electricity"                                ~ "Electricity",
      fuelType %in% c(
        "Premium and Electricity",
        "Regular Gas and Electricity",
        "Premium Gas or Electricity",
        "Regular Gas or Electricity",
        "Electricity and Hydrogen"
      )                                                        ~ "Hybrid/Electric mix",
      fuelType %in% c("Gasoline or E85", "Premium or E85")    ~ "Flex-fuel/E85",
      fuelType %in% c(
        "CNG", "Hydrogen",
        "Gasoline or natural gas",
        "Gasoline or propane"
      )                                                        ~ "Alternative fuel",
      TRUE                                                     ~ "Other"
    )
  )

table(data_fe$fuel_group)
```

#### `drive` — Consolidate 4WD/AWD Variants

The EDA identified multiple overlapping 4-wheel and all-wheel drive labels. We consolidate them into three interpretable groups.

```{r}
data_fe <- data_fe %>%
  mutate(drive = case_when(
    str_detect(drive, "4-Wheel|All-Wheel") ~ "AWD/4WD",
    str_detect(drive, "Front-Wheel")       ~ "Front-Wheel Drive",
    str_detect(drive, "Rear-Wheel")        ~ "Rear-Wheel Drive",
    is.na(drive)                           ~ "Unknown",
    TRUE                                   ~ "Other"
  ))

table(data_fe$drive)
```

#### `VClass` — Keep Top 10 Categories

`VClass` has many levels. We retain the 10 most frequent and collapse the rest into "Other".

```{r}
data_fe <- data_fe %>%
  mutate(VClass = fct_lump(as.factor(VClass), n = 10, other_level = "Other"))

table(data_fe$VClass)
```

#### `trany` — Extract Structured Transmission Features

Instead of keeping the raw high-cardinality `trany` string, we extract four interpretable features:

-   `transmission_type`: "Automatic", "Manual", or "Other"
-   `is_cvt`: TRUE if the transmission uses continuously variable gear ratios
- `n_gears`: number of speeds extracted from text; 0 means "not applicable" (CVT or non-standard transmission)
- `has_discrete_gears`: TRUE if the number of gear steps was successfully extracted from `trany`

`has_discrete_gears` and `n_gears` are designed to be read together: `has_discrete_gears = FALSE` means a CVT or other transmission without a discrete gear count — this is not randomly missing data, but a structural property of the transmission. Setting `n_gears = 0` via `replace_na` provides a numeric sentinel that is safe to use in models without median imputation.

The raw `trany` column is then removed.

```{r}
data_fe <- data_fe %>%
  mutate(
    transmission_type = case_when(
      str_detect(trany, regex("^Automatic", ignore_case = TRUE)) ~ "Automatic",
      str_detect(trany, regex("^Manual",    ignore_case = TRUE)) ~ "Manual",
      TRUE ~ "Other"
    ),
    is_cvt  = str_detect(trany, regex("variable gear ratios", ignore_case = TRUE)),
    is_cvt  = replace_na(is_cvt, FALSE),
    n_gears = as.integer(str_match(trany, "(\\d+)-spd")[, 2]),
    has_discrete_gears = as.integer(!is.na(n_gears)),
    n_gears = replace_na(n_gears, 0)
  )

cat("Transmission type distribution:\n")
print(table(data_fe$transmission_type))
cat("\nCVT vehicles:", sum(data_fe$is_cvt), "\n")
cat("\nn_gears summary:\n")
print(summary(data_fe$n_gears))
cat("\nhas_discrete_gears distribution:\n")
print(table(data_fe$has_discrete_gears))
```

### Build Final Modeling Dataset

We apply all column removals. The raw `trany` is dropped because it is fully replaced by `transmission_type`, `is_cvt`, and `n_gears`. The raw `fuelType` and `fuelType1` are dropped because their information is captured by `fuel_group`.

Pure battery electric vehicles (BEVs) are excluded from the modeling dataset before the train/test split. The target variable `comb08` represents Miles Per Gallon (MPG) for combustion-engine vehicles, but for BEVs the EPA reports MPGe — an artificial conversion (33.7 kWh = 1 gallon equivalent) that does not reflect actual fuel consumption. Including BEVs would mix two structurally different quantities in a single regression target. Moreover, the features that drive MPGe — battery capacity, motor efficiency, regenerative braking — are not present in the dataset after removing high-missingness columns such as `evMotor`. The `is_electric` and `has_combustion_engine` flags are also dropped after filtering, as they become zero-variance constants.
```{r}
n_ev <- sum(data_fe$is_electric)
vehicles_model <- data_fe %>%
  filter(!is_electric) %>%
  select(-any_of(cols_to_remove)) %>%
  select(-any_of(c("trany", "fuelType", "fuelType1", "is_electric", "has_combustion_engine")))
cat("Electric vehicles excluded:", n_ev, "\n")
cat("Remaining vehicles:        ", nrow(vehicles_model), "\n")
cat("Dimensions of vehicles_model:", nrow(vehicles_model), "×", ncol(vehicles_model), "\n")
cat("\nColumns retained:\n")
print(names(vehicles_model))
```

`cylinders` is kept as a candidate predictor despite its strong correlation with `displ` (\~0.91). We expect multicollinearity, but the final decision on whether to drop it is deferred to feature selection in a later scenario.

### Sanity Checks

```{r}
cat("Duplicate rows in vehicles_model:", sum(duplicated(vehicles_model)), "\n\n")

cat("summary(comb08):\n")
print(summary(vehicles_model$comb08))

cat("\nYear range:", range(vehicles_model$year, na.rm = TRUE), "\n")

cat("\ncomb08 <= 0:", sum(vehicles_model$comb08 <= 0, na.rm = TRUE), "\n")
cat("displ < 0:",    sum(vehicles_model$displ    < 0, na.rm = TRUE), "\n")
cat("cylinders < 0:", sum(vehicles_model$cylinders < 0, na.rm = TRUE), "\n")

missing_before <- colSums(is.na(vehicles_model))
cat("\nColumns with missing values (before recipe):\n")
print(missing_before[missing_before > 0])
cat("Total missing values:", sum(is.na(vehicles_model)), "\n")
```

After excluding BEVs, the remaining `comb08` values reflect actual MPG from combustion-engine and hybrid vehicles. Extreme values at the high end (diesel hybrids, efficient compact cars) are retained as valid observations.
### Train/Test Split

We split the data into 80% training and 20% test. The test set is only used at the very end to evaluate final model performance — we do not touch it during training or tuning. We stratify on `comb08` so both sets have a similar distribution of MPG values.

```{r}
set.seed(123)

data_split <- initial_split(vehicles_model, prop = 0.8, strata = comb08)

train_data <- training(data_split)
test_data  <- testing(data_split)

cat("Train:", nrow(train_data), "×", ncol(train_data), "\n")
cat("Test: ", nrow(test_data),  "×", ncol(test_data),  "\n")
```

### Check Train/Test Distribution

```{r}
cat("Train comb08:\n"); print(summary(train_data$comb08))
cat("Test  comb08:\n"); print(summary(test_data$comb08))
```

```{r}
bind_rows(
  train_data %>% select(comb08) %>% mutate(split = "train"),
  test_data  %>% select(comb08) %>% mutate(split = "test")
) %>%
  ggplot(aes(x = comb08, fill = split)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("train" = "steelblue", "test" = "tomato")) +
  labs(
    title = "Distribution of comb08: Train vs. Test",
    x     = "Combined MPG (comb08)",
    y     = "Density"
  ) +
  theme_minimal()
```

The density curves confirm that stratification produces very similar `comb08` distributions in both sets.

### Cross-Validation Folds

We prepare 5-fold cross-validation on the training set, stratified by `comb08`. These folds will be used later for hyperparameter tuning and comparing models.

```{r}
folds <- vfold_cv(train_data, v = 5, strata = comb08)

cat("Number of CV folds:", nrow(folds), "\n")
print(folds)
```

### Preprocessing Recipe

We define preprocessing as a `recipe` to prevent data leakage — all transformations (normalization, imputation, dummy encoding) are learned only from the training data and then applied to the test set using those saved values. This also works correctly inside cross-validation, where each fold fits its own preprocessing independently.

-   `step_unknown()` — handles unseen factor levels that may appear in the test set
-   `step_other()` — collapses any residual rare categories below 1%
-   `step_impute_median(displ, cylinders)` — fills the small number of numeric NAs in engine parameters for non-BEV rows; `n_gears` is intentionally set to 0 in feature engineering and does not need imputation
-   `step_dummy()` — one-hot encodes all nominal predictors; required by linear regression methods
-   `step_zv()` / `step_nzv()` — removes zero- and near-zero-variance columns, consistent with EDA findings on low-variability variables
-   `step_normalize()` — standardizes all numeric predictors; required for ridge, lasso, elastic net, kNN, and PCA/PCR/PLS

```{r}
model_recipe <- recipe(comb08 ~ ., data = train_data) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%
  step_impute_median(displ, cylinders) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_nzv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

model_recipe
```

### Fit and Apply Recipe

The recipe is fitted exclusively on `train_data`. The same fitted statistics (medians, modes, dummy levels) are then applied to both sets.

```{r}
prep_recipe <- prep(model_recipe, training = train_data)

train_processed <- bake(prep_recipe, new_data = NULL)
test_processed  <- bake(prep_recipe, new_data = test_data)
```

### Verify Preprocessing Output

```{r}
cat("train_processed:", nrow(train_processed), "×", ncol(train_processed), "\n")
cat("test_processed: ", nrow(test_processed),  "×", ncol(test_processed),  "\n")

cat("\nMissing values — train_processed:", sum(is.na(train_processed)), "\n")
cat("Missing values — test_processed: ", sum(is.na(test_processed)),  "\n")

cat("\ncomb08 in train_processed:", "comb08" %in% names(train_processed), "\n")
cat("comb08 in test_processed: ", "comb08" %in% names(test_processed),  "\n")

n_predictors <- ncol(train_processed) - 1
cat("\nNumber of predictors after dummy encoding:", n_predictors, "\n")

cat("\nFirst 10 column names:\n")
print(head(names(train_processed), 10))
```

Both processed datasets contain no missing values in predictors, and `comb08` is retained as the regression target. The objects `vehicles_model`, `train_data`, `test_data`, `folds`, `model_recipe`, `prep_recipe`, `train_processed`, and `test_processed` are ready for regression scenarios 2, 3, and 5.

## Non-parametric regression methods

```{r message=FALSE, warning=FALSE}
library(tidymodels)
library(rpart)       #for Decision tree
library(rpart.plot)  # decision tree visualisation
library(ranger)      # for Random forest
library(randomForest) # for Random forest
library(kknn)        # for KNN
library(vip)
library(parsnip)
```

Variables used for hyperparameters tuning:
```{r}
tune_ctrl <- control_grid(
  save_pred    = TRUE,
  save_workflow = TRUE,
  verbose      = TRUE
)

metrics_set <- metric_set(rmse, rsq, mae)
```

### Decision tree

Firstly, we will create decision tree model. The decision tree algorithm doesnt need normalised data, however we will use the prepared recipe with a normalisation step.

Hyperparameters tuning
```{r}
dt_spec <- decision_tree(
  cost_complexity = tune(),   # alpha used for pruning
  tree_depth      = tune(),
  min_n           = tune()
) %>%
  set_engine("rpart") %>%
  set_mode("regression")

dt_wf <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(dt_spec)

dt_grid <- grid_regular(
  cost_complexity(range = c(-4, -1)),  # log10 scale: 0.0001 - 0.1
  tree_depth(range = c(3, 15)),
  min_n(range = c(2, 20)),
  levels = 4
)

dt_tune <- tune_grid(
  dt_wf,
  resamples = folds,
  grid      = dt_grid,
  metrics   = metrics_set,
  control   = tune_ctrl
)
```

```{r}
show_best(dt_tune, metric = "rmse", n = 5)

autoplot(dt_tune) +
  labs(title = "Decision Tree - hyperparameters tuning") +
  theme_minimal()
```
The graph shows, how the models' performance changes with different hyperparameters. Model performance was most sensitive to the cost-complexity parameter, where larger values substantially worsened RMSE and R², indicating underfitting. Deeper trees (depth 7–11) outperformed shallow trees (depth 3), and the minimum node size had only a minor effect on predictive performance.

```{r}
dt_best <- select_best(dt_tune, metric = "rmse")
cat("The best parameters for decision tree:\n"); print(dt_best)

dt_final <- dt_wf %>%
  finalize_workflow(dt_best) %>%
  fit(data = train_data)
```
The best decision tree model has these parameters:
-   cost_complexity = 0.0001
-   tree_depth = 15
-   min_n = 8

```{r}
# get metrics on train data
dt_train_preds <- predict(dt_final, new_data = train_data) %>%
  bind_cols(train_data %>% select(comb08))

dt_train_metrics <- dt_train_preds %>%
  metrics(truth = comb08, estimate = .pred) %>%
  mutate(split = "train")

# get metrics on test data
dt_preds <- predict(dt_final, new_data = test_data) %>%
  bind_cols(test_data %>% select(comb08))

dt_test_metrics <- dt_preds %>%
  metrics(truth = comb08, estimate = .pred) %>%
  mutate(split = "test")

bind_rows(dt_train_metrics, dt_test_metrics) %>%
  select(split, .metric, .estimate) %>%
  pivot_wider(names_from = split, values_from = .estimate)

# check predictions in the first rows
head(dt_preds)
```
The tuned decision tree achieved solid predictive performance and showed good generalization to unseen data. Predicted `comb08` values were close to the actual observed values, with only small deviations in the predictions.

On the training data, the model reached an RMSE of 2.01 and an MAE of 1.38, indicating low average prediction errors. On the test dataset, performance remained similar, with an RMSE of 2.15 and an MAE of 1.44. The small increase in error from training to test data suggests that the model generalizes well and does not suffer from overfitting.
Based on the R² values, we can say that the model explained 87.9% of the variance in the training data and 86.1% of the variance in the test data.

```{r}
ggplot(dt_preds, aes(x = comb08, y = .pred)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Decision tree: Actual vs Predicted",
    x = "Actual comb08",
    y = "Predicted comb08"
  ) +
  theme_minimal()

dt_pred <- dt_preds %>%
  mutate(residual = comb08 - .pred)

ggplot(dt_pred, aes(x = .pred, y = residual)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Residual Plot",
    x = "Predicted comb08",
    y = "Residuals"
  ) +
  theme_minimal()
```
The Actual vs Predicted plot shows a clear positive linear relationship between observed and predicted comb08 values, indicating that the model captures the main structure of the data well. Most points are concentrated in range 15-35 MPG, these are predicted with good accuracy. For higher values of `comb08` , the model indicates lower accuracy as it tends to underpredict some most efficient vehicles.
A typical step pattern is visible in the predictions, which is expected for decision trees because they assign the same predicted value to all observations within a terminal node.

From the Residual plot, we can tell that residuals are centered around zero for most observations, but residual variance increases for high predicted MPG values. The graph indicates mild heteroskedasticity.

```{r}
dt_final %>%
  extract_fit_parsnip() %>%
  vip(num_features = 15) +
  labs(title = "Feature importance - Decision Tree") +
  theme_minimal()
```
The graph shows 15 most important features, which were used by model to predict target variable. The most important features are `cylinders` and `displ`.

### Random forest

Secondly, we will create random forest model. The random forest algorithm doesnt need normalised data, however we will use the prepared recipe with a normalisation step.

Since our dataset is quite large (40 000 rows) and the `randomForest` engine is training quite slowly, we will use a `ranger` engine in hyperparameters tuning steps, because it trains much quicker.

Hyperparameters tuning
```{r}
rf_model <- rand_forest(
  trees = 200,
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")

rf_workflow <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(rf_model)

rf_grid <- grid_regular(
  mtry(range = c(2, 10)),
  min_n(range = c(2, 20)),
  levels = 3
)

rf_tuned <- tune_grid(
  rf_workflow,
  resamples = folds,
  grid = rf_grid,
  metrics   = metrics_set,
  control   = tune_ctrl
)

rf_best <- select_best(rf_tuned, metric = "rmse")
cat("The best parameters for random forest:\n"); print(rf_best)

rf_final <- rf_workflow %>%
  finalize_workflow(rf_best) %>%
  fit(data = train_data)
```

The best random forest model has these parameters:
-   trees = 200
-   mtry = 10
-   min_n = 2

```{r}
# get metrics on train data
rf_train_preds <- predict(rf_final, new_data = train_data) %>%
  bind_cols(train_data %>% select(comb08))

rf_train_metrics <- rf_train_preds %>%
  metrics(truth = comb08, estimate = .pred) %>%
  mutate(split = "train")

# get metrics on test data
rf_pred <- predict(rf_final, new_data = test_data) %>%
  bind_cols(test_data %>% select(comb08))

rf_test_metrics <- rf_pred %>%
  metrics(truth = comb08, estimate = .pred) %>%
  mutate(split = "test")

bind_rows(rf_train_metrics, rf_test_metrics) %>%
  select(split, .metric, .estimate) %>%
  pivot_wider(names_from = split, values_from = .estimate)

# check predictions in the first rows
head(rf_pred)
```
The best random forest model achieved excellent predictive performance and outperformed the simpler decision tree model in terms of prediction error and explained variance. Predicted `comb08` values were predicted almost exactly.

On the training data, the model achieved an RMSE of 1.39 and an MAE of 0.76, indicating very low prediction errors. On the test dataset, performance remained strong, with an RMSE of 1.75 and an MAE of 1.03. Although the error increased slightly on unseen data, the difference between training and test results is relatively small, suggesting good generalization ability.

The model explained 94.2% of the variance in the training data and 90.9% of the variance in the test data, based on the R² values, so the underlying relationships between predictors and fuel economy were captured extremely well.

```{r}
ggplot(rf_pred, aes(x = comb08, y = .pred)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Random Forest: Actual vs Predicted",
    x = "Actual comb08",
    y = "Predicted comb08"
  ) +
  theme_minimal()

rf_pred <- rf_pred %>%
  mutate(residual = comb08 - .pred)

ggplot(rf_pred, aes(x = .pred, y = residual)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Residual Plot",
    x = "Predicted comb08",
    y = "Residuals"
  ) +
  theme_minimal()
```
The Actual vs Predicted plot shows strong agreement between observed and predicted `comb08` values, with most points close to the diagonal line. This indicates accurate predictions across a wide range of vehicles, especially in the common 15–35 MPG range.
Prediction errors increase slightly for vehicles with very high fuel efficiency, but the model still remains relatively accurate in this region. There is a small tendency to underpredict the highest MPG values, although less than in the decision tree model.

From the Residual plot, we can tell that residuals are centered around zero for most observations, but residual variance increases for high predicted MPG values. The graph indicates mild heteroskedasticity.

Overall, the random forest model provides better predictive performance and stronger generalization than the decision tree, though it is less interpretable.

### KNN
Lastly, we will train also a KNN model. We will find the best model using hyperparameters tuning.

```{r}
knn_recipe <- recipe(comb08 ~ ., data = train_data) %>%
  step_novel(all_nominal_predictors()) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%
  step_impute_median(displ, cylinders) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())
```


Hyperparameters tuning
```{r}
knn_spec <- nearest_neighbor(
  neighbors = tune(),
  weight_func = tune(),
  dist_power  = tune()
) %>%
  set_engine("kknn") %>%
  set_mode("regression")

knn_wf <- workflow() %>%
  add_recipe(knn_recipe) %>%
  add_model(knn_spec)

# KNN hyperparameters tuning
knn_grid <- grid_regular(
  neighbors(range = c(3, 25)),
  weight_func(values = c("rectangular", "triangular", "gaussian")),
  dist_power(range = c(1, 2)),
  levels = 3
)

cat("=== Tuning KNN ===\n")
knn_tune <- tune_grid(
  knn_wf,
  resamples = folds,
  grid      = knn_grid,
  metrics   = metrics_set,
  control   = tune_ctrl
)

knn_best <- select_best(knn_tune, metric = "rmse")
cat("The best parameters for KNN:\n"); print(knn_best)


knn_final <- knn_wf %>%
  finalize_workflow(knn_best) %>%
  fit(data = train_data)
```
The best KNN model is a model with parameters: 
-   neighbors = 3
-   weight_func = gaussian
-   dist_power = 1

```{r}
# get metrics on train data
knn_train_preds <- predict(knn_final, new_data = train_data) %>%
  bind_cols(train_data %>% select(comb08))

knn_train_metrics <- knn_train_preds %>%
  metrics(truth = comb08, estimate = .pred) %>%
  mutate(split = "train")

# get metrics on test data
knn_pred <- predict(knn_final, new_data = test_data) %>%
  bind_cols(test_data %>% select(comb08))

knn_test_metrics <- knn_pred %>%
  metrics(truth = comb08, estimate = .pred) %>%
  mutate(split = "test")

bind_rows(knn_train_metrics, knn_test_metrics) %>%
  select(split, .metric, .estimate) %>%
  pivot_wider(names_from = split, values_from = .estimate)

# check predictions in the first rows
head(knn_pred)
```

The KNN model achieved strong predictive performance on both training and test data. Predicted `comb08` values were again very close to the actual observed values.

On the training data, the model reached an RMSE of 1.12 and an MAE of 0.60, indicating very low prediction errors. On the test dataset, performance remained strong, with an RMSE of 1.69 and an MAE of 0.98. The moderate increase in error from training to test data suggests good generalization with mild overfitting.

The model explained 96.2% of the variance in the training data and 91.4% of the variance in the test data, based on the R² values. This indicates that the KNN model captured the relationships between predictors and fuel economy extremely well.

```{r}
ggplot(knn_pred, aes(x = comb08, y = .pred)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "KNN: Actual vs Predicted",
    x = "Actual comb08",
    y = "Predicted comb08"
  ) +
  theme_minimal()

knn_pred <- knn_pred %>%
  mutate(residual = comb08 - .pred)

ggplot(knn_pred, aes(x = .pred, y = residual)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Residual Plot",
    x = "Predicted comb08",
    y = "Residuals"
  ) +
  theme_minimal()
```
The actual vs. predicted plot as well as the residuals plot is very similar to random forests plots. The model has strong predictive performance for standard vehicles, however it tends to underestimate more efficient vehicles with higher MPG.



## PCA - Principal Component Analysis

```{r message=FALSE, warning=FALSE}
library(factoextra)
```

```{r}
preprocess_recipe <- recipe(comb08 ~ ., data = train_data) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%
  step_impute_median(displ, cylinders) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_nzv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

prep_preprocess_recipe <- prep(preprocess_recipe, training = train_data)
train_preprocessed <- bake(prep_preprocess_recipe, new_data = NULL)
```

```{r}
pca_input <- train_preprocessed %>%
  select(-comb08)   # remove target

cat("Size of PCA input:", nrow(pca_input), "×", ncol(pca_input), "\n")
```

```{r}
pca_result <- prcomp(pca_input, center = FALSE, scale. = FALSE)
# center=FALSE a scale.=FALSE because train_preprocessed is already normalised
```

```{r}
explained_var <- summary(pca_result)$importance
cat("\nFirst 10 components:\n")
print(explained_var[, 1:10])
```
The PCA was performed on the preprocessed training data, which had already been normalized and dummy-encoded. The first principal component explains approximately 13.6% of the total variance, while the second accounts for 11,97%. The explained variance decreases gradually across subsequent components, with no single dominant component, what suggests that the information in the dataset is moderately distributed across multiple dimensions rather than being concentrated in a few directions. 
The first 10 components together capture about 68.9% of the total variance.

```{r}
cum_var <- as.data.frame(t(explained_var)) %>%
  rownames_to_column("PC") %>%
  setNames(c("PC", "std_dev", "prop_var", "cum_var")) %>%
  mutate(PC_num = as.integer(str_extract(PC, "\\d+")))
```

```{r}
for (threshold in c(0.80, 0.90, 0.95)) {
  n <- min(which(cum_var$cum_var >= threshold))
  cat(sprintf("Number of PC for %.0f%% of variance: %d\n", threshold * 100, n))
}
```
To reach 80%, 90%, and 95% of cumulative variance, 13, 16, and 18 components are required respectively. The relatively high number of components needed to explain a substantial portion of variance is consistent with the high-dimensional nature of the dataset after dummy encoding, where many categorical variables were expanded into binary columns.

```{r}
p1 <- cum_var %>%
  filter(PC_num <= 30) %>%
  ggplot(aes(x = PC_num, y = prop_var)) +
  geom_col(fill = "steelblue") +
  geom_line(aes(group = 1), color = "darkblue", linewidth = 0.8) +
  geom_point(color = "darkblue", size = 2) +
  scale_x_continuous(breaks = 1:30) +
  labs(
    title = "Scree Plot - explained variance ratio",
    x     = "Principal component",
    y     = "Variance ratio"
  ) +
  theme_minimal()

p2 <- cum_var %>%
  filter(PC_num <= 30) %>%
  ggplot(aes(x = PC_num, y = cum_var)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_hline(yintercept = c(0.80, 0.90, 0.95),
             linetype = "dashed",
             color = c("orange", "red", "darkred")) +
  annotate("text", x = 22, y = 0.81, label = "80%", color = "orange",    size = 3) +
  annotate("text", x = 22, y = 0.91, label = "90%", color = "red",       size = 3) +
  annotate("text", x = 22, y = 0.96, label = "95%", color = "darkred",   size = 3) +
  scale_x_continuous(breaks = seq(0, 23, 5)) +
  labs(
    title = "Cummulative explained variability",
    x     = "Number of principal components",
    y     = "Cummulative variance ratio"
  ) +
  theme_minimal()

p1 + p2
```
The scree plot shows that the first principal component explains the largest share of variance (~14%), followed by a notable drop to the second component (~12%). After the third component, the decline becomes more gradual and the curve flattens, forming a characteristic elbow shape. This indicates that no single component strongly dominates, and the remaining variance is spread relatively evenly across the subsequent components.

The cumulative variance plot confirms this pattern. The curve rises steadily without any sharp jumps, which is typical for datasets with many moderately correlated features after dummy encoding. The three reference thresholds show that 80% of variance is reached at 13 components, 90% at 16 components, and 95% at 18 components.

```{r}
n_components <- min(which(cum_var$cum_var >= 0.95))
cat("\nWe will use", n_components, "components (95% of variance)\n")
```
For subsequent modeling, 18 principal components were retained, capturing 95% of the total variance. This threshold was chosen as a balance between dimensionality reduction and information retention - discarding the remaining 5% removes noise and redundancy while keeping the majority of the signal present in the original features.
The reduction was small but still useful given the relatively small original feature space after preprocessing.

Discussion on explainability: After PCA there are features PC1, PC2... which do not have a direct interpretation, so explainability decreases.


Now, we create a recipe, which will be used for comparison of models.
```{r}
pca_recipe <- recipe(comb08 ~ ., data = train_data) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_impute_mode(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_nzv(all_predictors()) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_pca(all_numeric_predictors(), num_comp = n_components)
```

Now we train a random forest model using preprocessed data with pca_recipe. 
Although the KNN model achieved slightly better predictive performance in the previous task, the random forest model is more robust and less sensitive to overfitting.
```{r}
pca_rf_model <- rand_forest(
  trees = 200,
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")

pca_rf_workflow <- workflow() %>%
  add_recipe(pca_recipe) %>%
  add_model(pca_rf_model)

pca_rf_grid <- grid_regular(
  mtry(range = c(2, 10)),
  min_n(range = c(2, 20)),
  levels = 3
)

pca_rf_tuned <- tune_grid(
  pca_rf_workflow,
  resamples = folds,
  grid = pca_rf_grid,
  metrics   = metrics_set,
  control   = tune_ctrl
)

pca_rf_best <- select_best(pca_rf_tuned, metric = "rmse")
cat("The best parameters for random forest with PCA:\n"); print(pca_rf_best)

pca_rf_final <- pca_rf_workflow %>%
  finalize_workflow(pca_rf_best) %>%
  fit(data = train_data)
```

The best random forest model with PCA has these parameters:
-   trees = 200
-   mtry = 10
-   min_n = 11

```{r}
# get metrics on train data
pca_rf_train_preds <- predict(pca_rf_final, new_data = train_data) %>%
  bind_cols(train_data %>% select(comb08))

pca_rf_train_metrics <- pca_rf_train_preds %>%
  metrics(truth = comb08, estimate = .pred) %>%
  mutate(split = "train")

# get metrics on test data
pca_rf_pred <- predict(pca_rf_final, new_data = test_data) %>%
  bind_cols(test_data %>% select(comb08))

pca_rf_test_metrics <- pca_rf_pred %>%
  metrics(truth = comb08, estimate = .pred) %>%
  mutate(split = "test")

bind_rows(pca_rf_train_metrics, pca_rf_test_metrics) %>%
  select(split, .metric, .estimate) %>%
  pivot_wider(names_from = split, values_from = .estimate)

# check predictions in the first rows
head(pca_rf_pred)
```
The best random forest model with PCA shows good overall performance, but it is slightly less accurate than models trained on the original features.

On the training set, the model achieved an RMSE of 1.36, MAE of 0.76, and R² of 0.944, indicating a very strong fit and that it explains about 94% of the variance in `comb08`. However, performance decreases on the test set, where the model reaches an RMSE of 1.97, MAE of 1.16, and R² of 0.885. This gap suggests mild overfitting, although part of the decrease in test performance may also be attributed to information loss introduced by PCA dimensionality reduction.

The sample predictions show that the model still captures general trends in fuel consumption well, but with slightly larger errors compared to models without PCA.

Overall, the model remains reliable and captures the main structure of the data, but the PCA transformation likely removed some fine-grained information that was useful for prediction. As a result, predictive performance is slightly reduced compared to models trained on the original feature space.

```{r}
ggplot(pca_rf_pred, aes(x = comb08, y = .pred)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Random Forest with PCA: Actual vs Predicted",
    x = "Actual comb08",
    y = "Predicted comb08"
  ) +
  theme_minimal()

pca_rf_pred <- pca_rf_pred %>%
  mutate(residual = comb08 - .pred)

ggplot(pca_rf_pred, aes(x = .pred, y = residual)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Residual Plot",
    x = "Predicted comb08",
    y = "Residuals"
  ) +
  theme_minimal()
```
The Actual vs. Predicted plot shows that the model tracks the diagonal reasonably well in the lower range (10–30 MPG), where most observations are concentrated. However, as actual values increase beyond 40, predictions increasingly underestimate the true values.

The Residual plot reveals a clear funnel shape: residuals are tightly clustered near zero for low predicted values but spread substantially as predictions increase. This heteroscedasticity suggests the model struggles with higher fuel efficiency vehicles, likely because they are underrepresented in the training data after PCA transformation. The residuals are roughly centered around zero overall, indicating no systematic bias in the mid-range, but the upper tail remains problematic.

Overall, the model performs adequately for typical vehicles but loses precision for fuel-efficient outliers.

```{r}

```

```{r}

```