---
title: "Vehicles dataset"
author: "Dominika Lukašáková, Lukáš Lovás"
date: "2026-04-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("D:/Škola/Ing/2.semester/OZNAL_projekt")
```

## EDA

Vehicles dataset\
Source: <https://fueleconomy.gov/feg/ws/index.shtml>

Based on the column descriptions, we could potentially predict fuel consumption (MPG).

-   **MPG (Miles per gallon)** – a measure of fuel economy in automobiles, higher number= higher efficiency\
-   **Target variable:** `comb08`

```{r}
library(tidyverse)
library(ggcorrplot)
library(patchwork)

rm(list = ls())
data <- read_csv("data/vehicles.csv")
data
View(data)
```

## Basic Structure

First, we check dataset dimensions and data types.

```{r}
dim(data)
```

The dataset has 49,846 rows and 84 columns.

```{r}
glimpse(data)
```

```{r}
data %>%
  summarise(across(everything(), class))
```

The dataset contains columns of type: character, double, logical. Columns createdOn and modifiedOn are dates stored as strings. The dataset has some missing values.

## Missing Values

```{r}
sum(is.na(data))
```

The dataset contains 538 466 missing values.

# Identify columns with missing values

```{r}
missing_summary <- data %>%
  summarise(across(everything(), ~mean(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_pct") %>%
  arrange(desc(missing_pct))

missing_summary %>%
  print(n = 84)
```

Columns with more than 60% missing values:

-   `c240bDscr`, `c240Dscr`, `sCharger`, `rangeA`, `fuelType2`, `guzzler`, `evMotor`,
-   `atvType`, `tCharger`, `trans_dscr`, `startStop`, `mfrCode`

We will not use these columns and remove them later.

-   Column `eng_dscr` has \~37% missing values
-   Columns `cylinders`, `displ`, `drive`, `trany`, `fuelType1` contain only a small amount of missing data (up to \~3%), so they could potentially be imputed.

## Unique Values

Next we identify columns with unique values.

```{r}
data %>%
  summarise(across(everything(), n_distinct)) %>%
  pivot_longer(everything(), 
               names_to = "variable", 
               values_to = "n_unique") %>%
  arrange(desc(n_unique)) %>%
  print(n = 84)
```

From these we filter out non-informative columns. We remove variables with nearly all unique values, as they act as identifiers and do not provide generalizable patterns for prediction. Also variables with very low variability (binary or near-constant features) were excluded, as they provide minimal information for explaining variation in the target variable.

```{r}
data %>%
  summarise(across(everything(), n_distinct)) %>%
  pivot_longer(everything(), 
               names_to = "variable", 
               values_to = "n_unique") %>%
  arrange(desc(n_unique)) %>%
  filter(!(n_unique >= 0.9 * nrow(data))) %>%
  filter(!(n_unique <= 2)) %>%
  print(n = 84)
```

-   We should also remove `model` column, because its extremely high number of unique values makes it impractical for modeling, and `engId,` as it also represents an identifier.
-   The dataset contains a large number of variables with high cardinality, indicating a high-dimensional feature space that requires careful feature selection.

## Numerical Variables

First, we check the distribution of numerical variables.

```{r}
num_data <- data %>% 
  select(where(is.numeric)) %>% 
  select(-id, -engId)

summary(num_data)
```

-   Many columns have median = 0 because most vehicles do not have that feature, for example `highwayE` applies only to electric vehicles. So, these variables might be informative for specific subsets, but not for the general fuel consumption prediction.

-   Some columns have min and median = -1, which is unusual and likely represents missing values. The columns are: `co2`, `co2A`, `feScore`, `ghgScore`, `ghgScoreA`.

-   Although column `youSaveSpend` does not have issues in terms of distribution, its definition reveals that it is directly derived from fuel consumption and fuel cost variables. As a result, it contains information closely related to the target variable and could introduce data leakage and artificially inflating model performance.

For further analysis we will remove: - columns with median = 0 or -1 - column `youSaveSpend`

```{r}
num_data <- num_data %>%
  select(where(~ {
    med <- median(., na.rm = TRUE)
    !(med == 0 | med == -1)
  })) %>%
  select(-youSaveSpend)

summary(num_data)
```

-   We are left with 11 numerical columns.

-   The similarity between mean and median values suggests that most variables are not heavily skewed, although there might be some outliers.

```{r}
num_data %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~variable, scales = "free") +
  theme_minimal()
# A warning is displayed because of NA in columns cylinders and displ
```

Most variables show right-skewed distributions, with the majority of observations with lower values and with a long tail toward higher values.

A few variables, such as `barrels08` and `co2TailpipeGpm` appear approximately normally distributed, though still with slight positive skew. In contrast, `cylinders` is a discrete variable, with clear peaks at common engine configurations - 4, 6, and 8 cylinders.

The variable `year` shows a relatively even but slightly increasing distribution toward more recent years, indicating a higher representation of newer vehicles in the dataset.

```{r}
num_data %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "value") %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()
# A warning is displayed because of NA in columns cylinders and displ
```

-   The variable `fuelCost08` shows a wide IQR and numerous outliers.
-   Column `co2TailpipeGpm` has a relatively compact central distribution but still includes several higher outliers.
-   The `year`column has a narrow spread from 1984 to 2026, indicating that most observations are concentrated within a similar time range, so there are no outliers.

We will plot another boxplot without columns `year`, `fuelCost08`, and `co2TailpipeGpm` due to differences in scale:

```{r}
num_data %>%
  select(-year, -fuelCost08, -co2TailpipeGpm) %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "value") %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()
# A warning is displayed because of NA in columns cylinders and displ
```

-   All columns have a significant amount of outsiders on the right side.

The boxplots reveal that most variables are right-skewed, as indicated by longer upper whiskers and a large number of high-value outliers. This pattern is especially clear for fuel economy variables (`city08`, `highway08`, `comb08`, `UCity`, `UHighway`), where a small number of highly efficient vehicles extend the upper range.

Variables such as `barrels08` and `displ` exhibit moderate right skewness with some extreme values. In contrast, `cylinders` is a discrete variable, making its boxplot less informative and leading to apparent outliers due to limited possible values.

Overall, the boxplots highlight skewness and the presence of outliers.

## Correlation Analysis

```{r}
cor_matrix <- num_data %>%
  select(where(is.numeric)) %>%
  cor(use = "pairwise.complete.obs")

cor_matrix
ggcorrplot(cor_matrix)
```

-   Target `comb08` is highly positively correlated with:

    -   `city08`, `highway08`, `UCity`, `UHighway` (up to \~0.996), so the usage of these columns in a model would cause data leakage

-   Target `comb08` is strongly negatively correlated with:

    -   `barrels08`, `co2TailpipeGpm`, `cylinders`, `displ`, `fuelCost08` (up to \~ -0.7)
    -   however we know from the columns description that columns `barrels08`, `co2TailpipeGpm` and `fuelCost08` are derived from fuel consumption, so they would also cause data leakage
    -   `cylinders` and `displ` are suitable predictors but they are highly correlated with each other (\~0.91)

So, for the model, we will keep only column **`displ`**, because it has fewer missing values and a stronger correlation with target (\~ -0.71), and column `year`, because it has a moderate positive correlation (\~0.32) with target and has almost no correlation with `displ`.

## Categorical Variables

```{r}
cat_data <- data %>% 
  select(where(is.character))

ncol(cat_data)

names(cat_data)
```

The dataset has 23 categorical variables.

```{r}
cat_overview <- cat_data %>%
  summarise(across(everything(), ~n_distinct(.))) %>%
  pivot_longer(everything(),
               names_to = "variable",
               values_to = "n_levels")

cat_overview
```

Based on the number of unique values and the description of each column, we will further explore only `fuelType`, `drive`, `VClass` and `trany` columns, because they have rational amount of categories and are related to fuel efficiency.

```{r}
data %>%
  select(fuelType, drive, VClass, trany) %>%
  pivot_longer(everything(),
               names_to = "variable",
               values_to = "category") %>%
  filter(!is.na(category)) %>%
  group_by(variable, category) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(variable) %>%
  slice_max(n, n = 10) %>%
  ggplot(aes(x = reorder(category, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  facet_wrap(~variable, scales = "free") +
  theme_minimal()
```

The `fuelType` variable shows a highly imbalanced distribution, with a few dominant categories (`Regular` and `Premium`) accounting for most observations, while several alternative fuel types appear only rarely. We will group rare categories of the fuelType variable into an "Other" class.

The `drive` variable is moderately imbalanced. `Front-Wheel Drive` and `Rear-Wheel Drive` represent the majority of observations, while several four-wheel-drive variants and two-wheel-drive vehicles occur much less frequently. The presence of multiple related four-wheel-drive labels indicates potential redundancy in category definitions, so merging semantically similar categories could reduce noise.

The `trany` variable contains many transmission categories, with a clear dominance of `Automatic 4-speed` and `Manual 5-speed` types.

The VClass has 3 dominant categories - `Compact Cars`, `Midsize Cars`, and `Subcompact Cars`. The other catogories appear less frequently. Since vehicle class captures structural differences in size, it is expected to be an informative predictor and should be retained in the modeling stage.

```{r}
ggplot(data, aes(x = fuelType, y = comb08)) +
  geom_boxplot() +
  coord_flip()
```

The boxplot shows how different fuel type influences fuel consumption. We can see that the electricity type has higher values then gasoline types.

```{r}
data %>%
  filter(!is.na(drive)) %>%
  ggplot(aes(x = drive, y = comb08)) +
  geom_boxplot() +
  coord_flip()
```

Drive type clearly affects fuel efficiency. `Front-Wheel Drive` vehicles tend to have higher combined MPG, while `4-Wheel or All-Wheel Drive` has smaller values.

```{r}
data %>%
  filter(!is.na(VClass)) %>%
  mutate(VClass = fct_lump(VClass, n = 10)) %>%
  ggplot(aes(x = VClass, y = comb08)) +
  geom_boxplot() +
  coord_flip()
```

Vehicle class has a clear relationship with fuel efficiency. Smaller vehicles (`compact` and `subcompact cars`) tend to achieve higher MPG, while pickups, SUVs, and larger vehicles have smaller MPG values, so they are less fuel efficient.

```{r}
data %>%
  filter(!is.na(trany)) %>%
  mutate(trany = fct_lump(trany, n = 10)) %>%
  ggplot(aes(x = trany, y = comb08)) +
  geom_boxplot() +
  coord_flip()
```

Transmission type strongly differentiates vehicle efficiency. `Automatic (A1)` shows the highest MPG values, while other categories have smaller values.

## Tagret variable

```{r}
summary(data$comb08)
```

The target variable `comb08` (combined fuel efficiency) has strongly right-skewed distribution. Most vehicles have between 15 and 30 MPG, with a median value of 20 MPG. However, a long right tail is observed, extending up to 146 MPG.

```{r}
hist <- ggplot(data, aes(x = comb08)) +
  geom_histogram(bins = 40, fill = "steelblue") +
  theme_minimal()

box <- ggplot(data, aes(y = comb08)) +
  geom_boxplot(fill = "steelblue") +
  coord_flip() +
  theme_minimal()

hist + box + plot_layout(ncol = 1, nrow = 2) 
```

Based on the histogram, we can say that the distribution is not normal. The boxplot confirms the presence of numerous high-value outliers. However, these observations are not errors but represent a distinct technological group of vehicles with significantly higher efficiency.

## Preprocessing and Train/Test Split

```{r message=FALSE, warning=FALSE}
library(rsample)
library(recipes)
library(stringr)
library(forcats)
```

### Column Groups to Remove

#### Columns with Too Many Missing Values (`cols_missing`)

These variables have more than 60% missing values and cannot be reliably used in a general model.

```{r}
cols_missing <- c(
  "c240bDscr", "c240Dscr", "sCharger", "rangeA", "fuelType2",
  "guzzler", "evMotor", "atvType", "tCharger", "trans_dscr",
  "startStop", "mfrCode"
)
```

#### Identifier Columns (`cols_ids`)

These variables are identifiers or overly specific strings. Using them would encourage the model to memorize specific vehicles or models.

```{r}
cols_ids <- c("id", "engId", "model")
```

#### Data Leakage Columns (`cols_leakage`)

These variables are either directly equivalent to `comb08` or derived from the same fuel consumption, emissions, or cost calculations. Including them would incorrectly inflate model performance.

```{r}
cols_leakage <- c(
  "city08", "highway08", "UCity", "UHighway",
  "barrels08", "co2TailpipeGpm", "fuelCost08", "youSaveSpend",
  "comb08U", "city08U", "highway08U",
  "combA08", "cityA08", "highwayA08",
  "combA08U", "cityA08U", "highwayA08U",
  "co2", "co2A", "co2TailpipeAGpm",
  "feScore", "ghgScore", "ghgScoreA",
  "phevCity", "phevHwy", "phevComb",
  "combE", "cityE", "highwayE",
  "combinedCD", "combinedUF",
  "cityCD", "cityUF",
  "highwayCD", "highwayUF"
)
```

#### Sparse or Invalid Columns

As identified in the EDA, many numeric columns have median = 0 or contain encoded missing values (−1). These are typically relevant only for electric or hybrid vehicles. Including them in the general regression dataset would introduce large numbers of zeros, encoded missing values, or derived information.

```{r}
cols_sparse_or_invalid <- c(
  "barrelsA08", "charge120", "charge240", "charge240b",
  "city08U", "cityA08", "cityA08U", "cityCD", "cityE", "cityUF",
  "co2", "co2A", "co2TailpipeAGpm",
  "comb08U", "combA08", "combA08U", "combE",
  "combinedCD", "combinedUF",
  "feScore", "fuelCostA08", "ghgScore", "ghgScoreA",
  "highway08U", "highwayA08", "highwayA08U",
  "highwayCD", "highwayE", "highwayUF",
  "hlv", "hpv", "lv2", "lv4", "pv2", "pv4",
  "range", "rangeCity", "rangeCityA", "rangeHwy", "rangeHwyA",
  "UCityA", "UHighwayA",
  "phevCity", "phevHwy", "phevComb"
)
```

#### Unused Columns

- `eng_dscr` has ~37% missing values and is a free-text engine description.
- `createdOn` and `modifiedOn` are metadata with no predictive value.
- `baseModel` would support memorization of specific model lines rather than generalization.
- `fuelType` and `fuelType1` are replaced by the new `fuel_group` variable.

```{r}
cols_unused <- c("createdOn", "modifiedOn", "eng_dscr", "baseModel")
```

#### Combine All Removal Lists

```{r}
cols_to_remove <- unique(c(cols_missing, cols_ids, cols_leakage,
                            cols_sparse_or_invalid, cols_unused))

cat("Total columns scheduled for removal:", length(cols_to_remove), "\n")
```

### Feature Engineering: Electric Vehicles

Electric vehicles do not have a combustion engine and therefore have no `displ` or `cylinders`. Their missing values are structurally meaningful — they signal the absence of an engine, not random data loss. We capture this before any imputation.

- `is_electric`: TRUE if the vehicle runs on electricity only.
- `has_combustion_engine`: TRUE if the vehicle has a combustion engine.

For EVs, `displ` and `cylinders` are set to 0 — meaning zero displacement, not a statistical fill-in. Remaining missing values in non-EV rows will be handled later by the recipe.

```{r}
data_fe <- data %>%
  mutate(
    is_electric = (fuelType == "Electricity") | (fuelType1 == "Electricity"),
    is_electric = replace_na(is_electric, FALSE),
    has_combustion_engine = !is_electric,
    displ     = if_else(is_electric, 0, displ),
    cylinders = if_else(is_electric, 0, cylinders)
  )

cat("Electric vehicles:", sum(data_fe$is_electric), "\n")
cat("Combustion engine vehicles:", sum(data_fe$has_combustion_engine), "\n")
```

### Categorical Variable Simplification

All transformations below use deterministic rules (not statistics learned from data), so they can safely be applied before the train/test split.

#### `fuel_group` — Structured Fuel Type Grouping

Instead of lumping rare levels of `fuelType`, we create a new interpretable variable `fuel_group` that consolidates all fuel categories into six meaningful groups. The raw `fuelType` and `fuelType1` columns remain unchanged in the original `data` object. For modeling, only `fuel_group` is used as a predictor — `fuelType` and `fuelType1` are removed from `vehicles_model` to avoid duplicating the same information across multiple columns.

```{r}
data_fe <- data_fe %>%
  mutate(
    fuel_group = case_when(
      fuelType %in% c("Regular", "Premium", "Midgrade")       ~ "Gasoline",
      fuelType == "Diesel"                                     ~ "Diesel",
      fuelType == "Electricity"                                ~ "Electricity",
      fuelType %in% c(
        "Premium and Electricity",
        "Regular Gas and Electricity",
        "Premium Gas or Electricity",
        "Regular Gas or Electricity",
        "Electricity and Hydrogen"
      )                                                        ~ "Hybrid/Electric mix",
      fuelType %in% c("Gasoline or E85", "Premium or E85")    ~ "Flex-fuel/E85",
      fuelType %in% c(
        "CNG", "Hydrogen",
        "Gasoline or natural gas",
        "Gasoline or propane"
      )                                                        ~ "Alternative fuel",
      TRUE                                                     ~ "Other"
    )
  )

table(data_fe$fuel_group)
```

#### `drive` — Consolidate 4WD/AWD Variants

The EDA identified multiple overlapping 4-wheel and all-wheel drive labels. We consolidate them into three interpretable groups.

```{r}
data_fe <- data_fe %>%
  mutate(drive = case_when(
    str_detect(drive, "4-Wheel|All-Wheel") ~ "AWD/4WD",
    str_detect(drive, "Front-Wheel")       ~ "Front-Wheel Drive",
    str_detect(drive, "Rear-Wheel")        ~ "Rear-Wheel Drive",
    is.na(drive)                           ~ "Unknown",
    TRUE                                   ~ "Other"
  ))

table(data_fe$drive)
```

#### `VClass` — Keep Top 10 Categories

`VClass` has many levels. We retain the 10 most frequent and collapse the rest into "Other".

```{r}
data_fe <- data_fe %>%
  mutate(VClass = fct_lump(as.factor(VClass), n = 10, other_level = "Other"))

table(data_fe$VClass)
```

#### `trany` — Extract Structured Transmission Features

Instead of keeping the raw high-cardinality `trany` string, we extract four interpretable features:

- `transmission_type`: "Automatic", "Manual", or "Other"
- `is_cvt`: TRUE if the transmission uses continuously variable gear ratios
- `n_gears`: number of speeds extracted from text; 0 means "not applicable" (CVT or non-standard transmission)
- `has_discrete_gears`: TRUE if the number of gear steps was successfully extracted from `trany`

`has_discrete_gears` and `n_gears` are designed to be read together: `has_discrete_gears = FALSE` means a CVT or other transmission without a discrete gear count — this is not randomly missing data, but a structural property of the transmission. Setting `n_gears = 0` via `replace_na` provides a numeric sentinel that is safe to use in models without median imputation.

The raw `trany` column is then removed.

```{r}
data_fe <- data_fe %>%
  mutate(
    transmission_type = case_when(
      str_detect(trany, regex("^Automatic", ignore_case = TRUE)) ~ "Automatic",
      str_detect(trany, regex("^Manual",    ignore_case = TRUE)) ~ "Manual",
      TRUE ~ "Other"
    ),
    is_cvt  = str_detect(trany, regex("variable gear ratios", ignore_case = TRUE)),
    is_cvt  = replace_na(is_cvt, FALSE),
    n_gears = as.integer(str_match(trany, "(\\d+)-spd")[, 2]),
    has_discrete_gears = as.integer(!is.na(n_gears)),
    n_gears = replace_na(n_gears, 0)
  )

cat("Transmission type distribution:\n")
print(table(data_fe$transmission_type))
cat("\nCVT vehicles:", sum(data_fe$is_cvt), "\n")
cat("\nn_gears summary:\n")
print(summary(data_fe$n_gears))
cat("\nhas_discrete_gears distribution:\n")
print(table(data_fe$has_discrete_gears))
```

### Build Final Modeling Dataset

We apply all column removals. The raw `trany` is dropped because it is fully replaced by `transmission_type`, `is_cvt`, and `n_gears`. The raw `fuelType` and `fuelType1` are dropped because their information is captured by `fuel_group`.

Pure battery electric vehicles (BEVs) are excluded from the modeling dataset before the train/test split. The target variable `comb08` represents Miles Per Gallon (MPG) for combustion-engine vehicles, but for BEVs the EPA reports MPGe — an artificial conversion (33.7 kWh = 1 gallon equivalent) that does not reflect actual fuel consumption. Including BEVs would mix two structurally different quantities in a single regression target. Moreover, the features that drive MPGe — battery capacity, motor efficiency, regenerative braking — are not present in the dataset after removing high-missingness columns such as `evMotor`. The `is_electric` and `has_combustion_engine` flags are also dropped after filtering, as they become zero-variance constants.

```{r}
n_ev <- sum(data_fe$is_electric)

vehicles_model <- data_fe %>%
  filter(!is_electric) %>%
  select(-any_of(cols_to_remove)) %>%
  select(-any_of(c("trany", "fuelType", "fuelType1", "is_electric", "has_combustion_engine")))

cat("Electric vehicles excluded:", n_ev, "\n")
cat("Remaining vehicles:        ", nrow(vehicles_model), "\n")
cat("Dimensions of vehicles_model:", nrow(vehicles_model), "×", ncol(vehicles_model), "\n")
cat("\nColumns retained:\n")
print(names(vehicles_model))
```

`cylinders` is kept as a candidate predictor despite its strong correlation with `displ` (~0.91). Multicollinearity between the two was expected, but LASSO (Scenario 2) selected `penalty = 1e-4` — too small to zero any coefficient — and retained both predictors. With n/p > 2 000, OLS variance is low enough that the multicollinearity does not destabilize estimates. Both `cylinders` and `displ` also appear as the two strongest PLS Component 1 loadings (Scenario 5), confirming they each carry independent signal: `displ` reflects total engine displacement while `cylinders` captures engine architecture (e.g. a turbocharged 4-cylinder can have large displacement but few cylinders). Dropping either variable based solely on the correlation would discard real predictive information.

### Sanity Checks

```{r}
cat("Duplicate rows in vehicles_model:", sum(duplicated(vehicles_model)), "\n\n")

cat("summary(comb08):\n")
print(summary(vehicles_model$comb08))

cat("\nYear range:", range(vehicles_model$year, na.rm = TRUE), "\n")

cat("\ncomb08 <= 0:", sum(vehicles_model$comb08 <= 0, na.rm = TRUE), "\n")
cat("displ < 0:",    sum(vehicles_model$displ    < 0, na.rm = TRUE), "\n")
cat("cylinders < 0:", sum(vehicles_model$cylinders < 0, na.rm = TRUE), "\n")

missing_before <- colSums(is.na(vehicles_model))
cat("\nColumns with missing values (before recipe):\n")
print(missing_before[missing_before > 0])
cat("Total missing values:", sum(is.na(vehicles_model)), "\n")
```

After excluding BEVs, the remaining `comb08` values reflect actual MPG from combustion-engine and hybrid vehicles. Extreme values at the high end (diesel hybrids, efficient compact cars) are retained as valid observations.

### Train/Test Split

We split the data into 80% training and 20% test. The test set is only used at the very end to evaluate final model performance — we do not touch it during training or tuning. We stratify on `comb08` so both sets have a similar distribution of MPG values.

```{r}
set.seed(123)

data_split <- initial_split(vehicles_model, prop = 0.8, strata = comb08)

train_data <- training(data_split)
test_data  <- testing(data_split)

cat("Train:", nrow(train_data), "×", ncol(train_data), "\n")
cat("Test: ", nrow(test_data),  "×", ncol(test_data),  "\n")
```

### Check Train/Test Distribution

```{r}
cat("Train comb08:\n"); print(summary(train_data$comb08))
cat("Test  comb08:\n"); print(summary(test_data$comb08))
```

```{r}
bind_rows(
  train_data %>% select(comb08) %>% mutate(split = "train"),
  test_data  %>% select(comb08) %>% mutate(split = "test")
) %>%
  ggplot(aes(x = comb08, fill = split)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("train" = "steelblue", "test" = "tomato")) +
  labs(
    title = "Distribution of comb08: Train vs. Test",
    x     = "Combined MPG (comb08)",
    y     = "Density"
  ) +
  theme_minimal()
```

The density curves confirm that stratification produces very similar `comb08` distributions in both sets.

### Cross-Validation Folds

Ridge, LASSO, and Elastic Net need a `penalty` hyperparameter tuned by comparing performance across different values. We use 5-fold cross-validation for this: the training data is split into 5 folds, the model trains on 4 and is evaluated on the remaining one, rotating through all folds. This gives a reliable performance estimate without using the test set.

```{r}
folds <- vfold_cv(train_data, v = 5, strata = comb08)

cat("Number of CV folds:", nrow(folds), "\n")
print(folds)
```

### Preprocessing Recipe

We define preprocessing as a `recipe` to prevent data leakage — all transformations (normalization, imputation, dummy encoding) are learned only from the training data and then applied to the test set using those saved values. This also works correctly inside cross-validation, where each fold fits its own preprocessing independently.

- `step_unknown()` — handles unseen factor levels that may appear in the test set
- `step_other()` — collapses any residual rare categories below 1%
- `step_impute_median(displ, cylinders)` — fills the small number of numeric NAs in engine parameters for non-BEV rows; `n_gears` is intentionally set to 0 in feature engineering and does not need imputation
- `step_dummy()` — one-hot encodes all nominal predictors; required by linear regression methods
- `step_zv()` / `step_nzv()` — removes zero- and near-zero-variance columns, consistent with EDA findings on low-variability variables
- `step_normalize()` — standardizes all numeric predictors; required for ridge, lasso, elastic net, kNN, and PCA/PCR/PLS

```{r}
model_recipe <- recipe(comb08 ~ ., data = train_data) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%
  step_impute_median(displ, cylinders) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_nzv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

model_recipe
```

### Fit and Apply Recipe

The recipe is fitted exclusively on `train_data`. The same fitted statistics (medians, modes, dummy levels) are then applied to both sets.

```{r}
prep_recipe <- prep(model_recipe, training = train_data)

train_processed <- bake(prep_recipe, new_data = NULL)
test_processed  <- bake(prep_recipe, new_data = test_data)
```

### Verify Preprocessing Output

```{r}
cat("train_processed:", nrow(train_processed), "×", ncol(train_processed), "\n")
cat("test_processed: ", nrow(test_processed),  "×", ncol(test_processed),  "\n")

cat("\nMissing values — train_processed:", sum(is.na(train_processed)), "\n")
cat("Missing values — test_processed: ", sum(is.na(test_processed)),  "\n")

cat("\ncomb08 in train_processed:", "comb08" %in% names(train_processed), "\n")
cat("comb08 in test_processed: ", "comb08" %in% names(test_processed),  "\n")

n_predictors <- ncol(train_processed) - 1
cat("\nNumber of predictors after dummy encoding:", n_predictors, "\n")

cat("\nFirst 10 column names:\n")
print(head(names(train_processed), 10))
```

Both processed datasets contain no missing values in predictors, and `comb08` is retained as the regression target. The objects `vehicles_model`, `train_data`, `test_data`, `folds`, `model_recipe`, `prep_recipe`, `train_processed`, and `test_processed` are ready for regression scenarios 2, 3, and 5.

## Scenario 2: Parametric vs. Non-Parametric Models — Parametric Part

Parametric models assume that the relationship between predictors and the response follows a specific, fixed mathematical form — in regression, a linear combination of weighted features.

We compare four parametric regression methods:

1. **Linear Regression (OLS)** — the classical least-squares baseline, no regularization
2. **Ridge Regression** — OLS plus an L2 penalty that shrinks all coefficients toward zero
3. **LASSO** — OLS plus an L1 penalty that can shrink some coefficients to exactly zero, selecting features
4. **Elastic Net** — a blend of Ridge and LASSO penalties, combining both approaches

All four methods share the same preprocessing recipe defined in the previous section. We use the 5-fold cross-validation folds for hyperparameter tuning and model comparison, and evaluate final performance on the held-out test set.

```{r message=FALSE, warning=FALSE}
library(tidymodels)
library(broom)
library(glmnet)
```

We define the three evaluation metrics once and reuse them across every model.

```{r}
eval_metrics <- metric_set(rmse, rsq, mae)
```

---

### 1. Linear Regression (OLS)

Ordinary least squares minimizes the sum of squared residuals and imposes no constraint on coefficient size. It is our reference point — any benefit from regularization will appear as an improvement over this baseline.

```{r}
lm_spec <- linear_reg() %>%
  set_engine("lm") %>%
  set_mode("regression")

lm_wf <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(lm_spec)
```

We wrap the recipe and model in a `workflow` so preprocessing is applied automatically during cross-validation and when predicting on new data.

```{r warning=FALSE}
set.seed(123)

lm_cv <- lm_wf %>%
  fit_resamples(
    resamples = folds,
    metrics   = eval_metrics,
    control   = control_resamples(save_pred = TRUE)
  )

lm_cv_metrics <- collect_metrics(lm_cv)
cat("=== OLS — 5-fold CV ===\n")
cat("RMSE:", round(lm_cv_metrics %>% filter(.metric == "rmse") %>% pull(mean), 4),
    "  ±", round(lm_cv_metrics %>% filter(.metric == "rmse") %>% pull(std_err), 4), "\n")
cat("MAE: ", round(lm_cv_metrics %>% filter(.metric == "mae")  %>% pull(mean), 4), "\n")
cat("R²:  ", round(lm_cv_metrics %>% filter(.metric == "rsq")  %>% pull(mean), 4), "\n")
```

Now we fit on the full training set to inspect which predictors drive the predictions.

```{r}
lm_fit <- lm_wf %>% fit(data = train_data)

lm_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(term != "(Intercept)") %>%
  slice_max(abs(estimate), n = 20) %>%
  ggplot(aes(x = reorder(term, abs(estimate)), y = estimate)) +
  geom_col(aes(fill = estimate > 0)) +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
    labels = c("TRUE" = "Positive", "FALSE" = "Negative"),
    name   = "Direction"
  ) +
  labs(
    title = "OLS: Top 20 Predictors by Absolute Coefficient",
    x     = NULL,
    y     = "Coefficient estimate"
  ) +
  theme_minimal()
```

Blue bars indicate predictors that increase `comb08` (e.g. electric drivetrain), red bars indicate predictors that decrease it (e.g. larger engine displacement).

---

### 2. Ridge Regression

Ridge regression adds an L2 penalty — λ × Σβ² — to the OLS loss. This forces the optimiser to keep coefficients small, reducing variance at the cost of some bias. Ridge never removes a predictor entirely: all coefficients are shrunk toward zero, but remain non-zero. It is especially effective when predictors are correlated — a situation we identified between `displ` and `cylinders` in the EDA.

The regularization strength λ is controlled by `penalty`. We tune it over a log-spaced grid.

```{r}
ridge_spec <- linear_reg(penalty = tune(), mixture = 0) %>%
  set_engine("glmnet") %>%
  set_mode("regression")

ridge_wf <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(ridge_spec)
```

`glmnet` fits the entire regularization path in a single pass, so evaluating 50 penalty values has almost no extra cost compared to evaluating one.

```{r}
ridge_grid <- grid_regular(penalty(range = c(-4, 2)), levels = 50)
```

```{r warning=FALSE}
set.seed(123)

ridge_cv <- ridge_wf %>%
  tune_grid(
    resamples = folds,
    grid      = ridge_grid,
    metrics   = eval_metrics
  )

autoplot(ridge_cv)
```

At very low penalty (left end), Ridge behaves like OLS. As penalty increases, coefficients shrink and the model becomes more biased but less variable. The optimal point is where RMSE on the validation folds is minimized.

```{r}
best_ridge <- ridge_cv %>% select_best(metric = "rmse")

ridge_best_row <- show_best(ridge_cv, metric = "rmse", n = 1)
cat("=== Ridge — best hyperparameter ===\n")
cat("Best penalty:", round(best_ridge$penalty, 6), "\n")
cat("CV RMSE:     ", round(ridge_best_row$mean, 4),
    "  ±", round(ridge_best_row$std_err, 4), "\n")
```

```{r}
ridge_final_wf <- ridge_wf %>% finalize_workflow(best_ridge)
ridge_fit       <- ridge_final_wf %>% fit(data = train_data)
```

```{r}
ridge_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(term != "(Intercept)") %>%
  slice_max(abs(estimate), n = 20) %>%
  ggplot(aes(x = reorder(term, abs(estimate)), y = estimate)) +
  geom_col(aes(fill = estimate > 0)) +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
    labels = c("TRUE" = "Positive", "FALSE" = "Negative"),
    name   = "Direction"
  ) +
  labs(
    title = paste0("Ridge: Top 20 Predictors (penalty = ",
                   round(best_ridge$penalty, 5), ")"),
    x     = NULL,
    y     = "Coefficient estimate"
  ) +
  theme_minimal()
```

Compared to OLS, Ridge produces smaller coefficient magnitudes across all predictors. The ordering and direction of the most important features generally remains the same, but correlated predictors share their weight more evenly.

---

### 3. LASSO Regression

LASSO (Least Absolute Shrinkage and Selection Operator) uses an L1 penalty — λ × Σ|β|. The key geometric property of L1 is that the regularized solution often lands exactly at zero for many coefficients. LASSO therefore performs simultaneous regression and feature selection: it automatically identifies the most informative predictors and removes the rest.

```{r}
lasso_spec <- linear_reg(penalty = tune(), mixture = 1) %>%
  set_engine("glmnet") %>%
  set_mode("regression")

lasso_wf <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(lasso_spec)
```

LASSO typically needs smaller penalties than Ridge to achieve similar shrinkage, so we search a narrower range.

```{r}
lasso_grid <- grid_regular(penalty(range = c(-4, 0)), levels = 50)
```

```{r warning=FALSE}
set.seed(123)

lasso_cv <- lasso_wf %>%
  tune_grid(
    resamples = folds,
    grid      = lasso_grid,
    metrics   = eval_metrics
  )

autoplot(lasso_cv)
```

```{r}
best_lasso <- lasso_cv %>% select_best(metric = "rmse")

lasso_best_row <- show_best(lasso_cv, metric = "rmse", n = 1)
cat("=== LASSO — best hyperparameter ===\n")
cat("Best penalty:", round(best_lasso$penalty, 6), "\n")
cat("CV RMSE:     ", round(lasso_best_row$mean, 4),
    "  ±", round(lasso_best_row$std_err, 4), "\n")
```

```{r}
lasso_final_wf <- lasso_wf %>% finalize_workflow(best_lasso)
lasso_fit       <- lasso_final_wf %>% fit(data = train_data)
```

One of LASSO's main advantages is sparsity. Let's see how many predictors it retained.

```{r}
lasso_coefs <- lasso_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(term != "(Intercept)")

cat("=== LASSO — sparsity ===\n")
cat("Total predictors available:", nrow(lasso_coefs), "\n")
cat("Kept by LASSO (non-zero):  ", sum(lasso_coefs$estimate != 0), "\n")
cat("Removed by LASSO (zero):   ", sum(lasso_coefs$estimate == 0), "\n")
if (sum(lasso_coefs$estimate == 0) > 0) {
  cat("Zeroed out predictors:\n")
  print(lasso_coefs %>% filter(estimate == 0) %>% pull(term))
}
```

```{r}
lasso_coefs %>%
  filter(estimate != 0) %>%
  slice_max(abs(estimate), n = 20) %>%
  ggplot(aes(x = reorder(term, abs(estimate)), y = estimate)) +
  geom_col(aes(fill = estimate > 0)) +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
    labels = c("TRUE" = "Positive", "FALSE" = "Negative"),
    name   = "Direction"
  ) +
  labs(
    title = paste0("LASSO: Top 20 Non-Zero Predictors (penalty = ",
                   round(best_lasso$penalty, 5), ")"),
    x     = NULL,
    y     = "Coefficient estimate"
  ) +
  theme_minimal()
```

LASSO retains only the most informative predictors, which makes the resulting model more parsimonious and directly interpretable compared to OLS or Ridge. Predictors that survive the L1 penalty are the ones that the algorithm judged worth the cost of adding a parameter.

---

### 4. Elastic Net

Elastic Net combines both L1 and L2 penalties: λ × [α × Σ|β| + (1 − α) × Σβ²]. The `mixture` parameter in tidymodels corresponds to α:  `mixture = 0` gives pure Ridge, `mixture = 1` gives pure LASSO. By tuning both `penalty` and `mixture`, Elastic Net finds the best blend — it can perform feature selection like LASSO while also handling correlated predictors like Ridge.

```{r}
en_spec <- linear_reg(penalty = tune(), mixture = tune()) %>%
  set_engine("glmnet") %>%
  set_mode("regression")

en_wf <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(en_spec)
```

We use a regular 10 × 10 grid — 10 penalty values and 10 mixture values — giving 100 combinations. `glmnet` evaluates all penalty values for each mixture in a single path, so this is still fast.

```{r}
en_grid <- grid_regular(
  penalty(range = c(-4, 0)),
  mixture(range = c(0, 1)),
  levels = 10
)
```

```{r warning=FALSE}
set.seed(123)

en_cv <- en_wf %>%
  tune_grid(
    resamples = folds,
    grid      = en_grid,
    metrics   = eval_metrics
  )

autoplot(en_cv)
```

```{r}
best_en <- en_cv %>% select_best(metric = "rmse")

en_best_row <- show_best(en_cv, metric = "rmse", n = 1)
cat("=== Elastic Net — best hyperparameters ===\n")
cat("Best penalty:", round(best_en$penalty, 6), "\n")
cat("Best mixture:", round(best_en$mixture, 4),
    " (0 = Ridge, 1 = LASSO)\n")
cat("CV RMSE:     ", round(en_best_row$mean, 4),
    "  ±", round(en_best_row$std_err, 4), "\n")
```

The optimal `mixture` value tells us how the data prefers to be regularized. If the best mixture is close to 1, LASSO-like behavior is preferred; if it is close to 0, Ridge-like behavior is preferred.

```{r}
en_final_wf <- en_wf %>% finalize_workflow(best_en)
en_fit       <- en_final_wf %>% fit(data = train_data)
```

---

### Comparing All Parametric Models

We evaluate all four models on the held-out test set. Because each model is a full `workflow()` that includes the preprocessing recipe, we pass the raw `test_data` directly and the recipe transformations are applied automatically.

We write a small helper function so we do not repeat the same prediction + metric code four times.

```{r}
get_test_metrics <- function(fit, model_name) {
  fit %>%
    predict(new_data = test_data) %>%
    bind_cols(test_data %>% select(comb08)) %>%
    summarise(
      model = model_name,
      RMSE  = rmse_vec(truth = comb08, estimate = .pred),
      MAE   = mae_vec( truth = comb08, estimate = .pred),
      Rsq   = rsq_vec( truth = comb08, estimate = .pred)
    )
}
```

```{r}
test_comparison <- bind_rows(
  get_test_metrics(lm_fit,    "Linear Regression"),
  get_test_metrics(ridge_fit, "Ridge"),
  get_test_metrics(lasso_fit, "LASSO"),
  get_test_metrics(en_fit,    "Elastic Net")
) %>%
  arrange(RMSE)

test_comparison
```

```{r}
test_comparison %>%
  pivot_longer(cols = c(RMSE, MAE, Rsq),
               names_to  = "metric",
               values_to = "value") %>%
  ggplot(aes(x = reorder(model, value), y = value, fill = model)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~metric, scales = "free_x") +
  coord_flip() +
  labs(
    title = "Test Set Performance: Parametric Models",
    x     = NULL,
    y     = "Value"
  ) +
  theme_minimal()
```

Lower RMSE and MAE are better; higher R² is better.

#### Cross-Validation vs. Test Set RMSE

We compare CV RMSE with test RMSE. A small gap means no overfitting — the model generalizes to new data roughly as well as it performed during CV.

```{r}
cv_rmse_comparison <- tibble(
  model   = c("Linear Regression", "Ridge", "LASSO", "Elastic Net"),
  cv_rmse = c(
    collect_metrics(lm_cv)    %>% filter(.metric == "rmse") %>% pull(mean),
    show_best(ridge_cv, metric = "rmse", n = 1) %>% pull(mean),
    show_best(lasso_cv, metric = "rmse", n = 1) %>% pull(mean),
    show_best(en_cv,    metric = "rmse", n = 1) %>% pull(mean)
  )
) %>%
  left_join(
    test_comparison %>% select(model, test_rmse = RMSE),
    by = "model"
  ) %>%
  mutate(gap = abs(test_rmse - cv_rmse)) %>%
  arrange(cv_rmse)

cv_rmse_comparison
```

A small `gap` confirms no overfitting.

#### Actual vs. Predicted — All Models

Actual vs. predicted on the test set. Points on the red diagonal = perfect prediction.

```{r}
all_test_preds <- bind_rows(
  lm_fit    %>% predict(new_data = test_data) %>% bind_cols(test_data %>% select(comb08)) %>% mutate(model = "Linear Regression"),
  ridge_fit %>% predict(new_data = test_data) %>% bind_cols(test_data %>% select(comb08)) %>% mutate(model = "Ridge"),
  lasso_fit %>% predict(new_data = test_data) %>% bind_cols(test_data %>% select(comb08)) %>% mutate(model = "LASSO"),
  en_fit    %>% predict(new_data = test_data) %>% bind_cols(test_data %>% select(comb08)) %>% mutate(model = "Elastic Net")
)

all_test_preds %>%
  ggplot(aes(x = comb08, y = .pred)) +
  geom_point(alpha = 0.15, color = "steelblue", size = 0.8) +
  geom_abline(slope = 1, intercept = 0, color = "tomato", linewidth = 0.8) +
  facet_wrap(~model) +
  labs(
    title = "Actual vs. Predicted — Test Set (Parametric Models)",
    x     = "Actual combined MPG",
    y     = "Predicted combined MPG"
  ) +
  theme_minimal()
```

With BEVs excluded, all values are real MPG. The fan-shaped distortion from MPGe outliers is gone. Remaining scatter at high MPG values is mostly PHEVs and efficient compacts — the model slightly underpredicts these.

---

### Summary: Parametric Models

All four models assume a linear relationship. They differ only in regularization:

| Model | Penalty | Sparsity |
|---|---|---|
| OLS | None | No |
| Ridge | L2 | No |
| LASSO | L1 | Yes |
| Elastic Net | L1 + L2 | Yes |

On large datasets the differences are small. Regularization stabilizes coefficients; LASSO additionally zeroes out weak predictors.

---

### Conclusion

#### Effect of removing electric vehicles

Removing 1 425 BEVs (whose `comb08` values are MPGe, not MPG) dropped RMSE from 8.11 to **3.36 MPG** (−59%) and MAE from 5.15 to **2.19 MPG**. R² only moved from 0.642 to 0.661 — this is expected, since R² is scale-independent and both the errors and the total variance dropped together.

#### Model results

All four models gave nearly the same result on the test set: **RMSE ≈ 3.36 MPG, MAE ≈ 2.19 MPG, R² ≈ 0.661**. With a median `comb08` of 20 MPG, an average error of 2.19 MPG is roughly 11% — reasonable for a linear model with 21 features.

#### Why regularization did not help

All three regularized models selected `penalty = 0.0001` — the minimum of the search grid. With 38 735 training observations and only 21 predictors, there is very little variance to reduce, so regularization has almost nothing to do. LASSO also kept all 21 predictors for the same reason — at such a small penalty, no predictor is weak enough to be zeroed out. This is not a bad result; it means the feature set is clean and every variable carries real signal.

#### Where the models fall short

- The relationship between engine size and MPG is not perfectly linear — adding a cylinder to a small 4-cylinder engine is different from adding one to a V8.
- PHEVs and mild hybrids behave differently depending on driving mode, which no single feature captures.
- Vehicle weight and aerodynamic drag are not in the dataset, even though they strongly affect fuel economy.

The CV–test gap is below 0.005 for all models, so there is no overfitting. The models are a solid baseline — interpretable, stable, and easy to explain.

### 2.5 Extension: interaction terms (cylinders × displ)

The four parametric models above are linear in the predictors. EDA showed that `cylinders` and `displ` are strongly correlated (r ≈ 0.91) and that adding a cylinder affects small engines differently than V8s — a classic non-linearity that a purely linear model cannot capture. Here we add the explicit product term `cylinders × displ` to the recipe and refit OLS to see whether the cross-effect reduces test RMSE and what its sign tells us.

```{r}
rec_interact <- recipe(comb08 ~ ., data = train_data) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%
  step_impute_median(displ, cylinders) %>%
  step_interact(terms = ~ cylinders:displ) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_nzv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

ols_interact_wf <- workflow() %>%
  add_recipe(rec_interact) %>%
  add_model(linear_reg() %>% set_engine("lm"))

ols_interact_fit <- ols_interact_wf %>% fit(train_data)
```

```{r}
ols_interact_pred <- predict(ols_interact_fit, new_data = test_data) %>%
  bind_cols(test_data %>% select(comb08))

ols_interact_metrics <- tibble(
  model = "OLS (cylinders × displ)",
  RMSE  = rmse_vec(truth = ols_interact_pred$comb08,
                   estimate = ols_interact_pred$.pred),
  MAE   = mae_vec( truth = ols_interact_pred$comb08,
                   estimate = ols_interact_pred$.pred),
  Rsq   = rsq_vec( truth = ols_interact_pred$comb08,
                   estimate = ols_interact_pred$.pred)
) %>%
  mutate(across(c(RMSE, MAE, Rsq), ~round(., 4)))

interaction_compare <- bind_rows(
  test_comparison %>% filter(model == "Linear Regression") %>%
    select(model, RMSE, MAE, Rsq) %>%
    mutate(model = "OLS (no interaction)"),
  ols_interact_metrics
)

cat("=== OLS with vs. without cylinders × displ ===\n")
print(interaction_compare)

interaction_term <- ols_interact_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(grepl("cylinders_x_displ|cylinders:displ", term))

cat("\n=== Interaction coefficient ===\n")
print(interaction_term)
```

The sign of the `cylinders × displ` coefficient indicates how the cylinder effect changes per unit of normalized displacement: a negative value means adding a cylinder hurts MPG more on top of an already-large engine (compounding penalty); a positive value means the cylinder penalty softens for big engines. The RMSE gain is expected to be small at n/p > 2000, but the interaction coefficient itself directly tests whether the cross-effect we suspected from EDA is real after controlling for everything else in the model.

---

## Scenario 3: Embedded Feature Selection (LASSO and Elastic Net)

Embedded methods integrate feature selection into model training itself — the algorithm simultaneously learns which features matter and what their coefficients are. This differs from filter methods (which score features independently of any model) and wrapper methods (which train a model for each candidate subset).

Two models from Scenario 2 are embedded methods by design:

- **LASSO** (L1 penalty): the L1 geometry causes the optimization to land exactly at zero for weak predictors. The regularization path records which features enter or leave the model as λ grows.
- **Elastic Net** (L1 + L2): inherits LASSO's sparsity while distributing weight more evenly among correlated predictors like `displ` and `cylinders`.

OLS has no regularization — no selection. Ridge (L2 only) shrinks coefficients toward zero but never reaches exactly zero — not an embedded selector.

Because both models were already fitted in Scenario 2, no new training is needed. We inspect their embedded selection behavior directly.

### 3.1 Which Scenario 2 models qualify as embedded?

We classify the four Scenario 2 models to justify why LASSO and Elastic Net are selected for Scenario 3.

```{r}
model_classification <- tibble(
  model        = c("Linear Regression", "Ridge", "LASSO", "Elastic Net"),
  penalty_type = c("None", "L2 only", "L1", "L1 + L2"),
  zeros_coefs  = c("No", "No", "Yes", "Yes"),
  embedded     = c("No", "No", "YES", "YES")
)

cat("=== Scenario 2 models — embedded method classification ===\n")
print(model_classification)
cat("\nSelected for Scenario 3: LASSO and Elastic Net\n")
cat("(best-performing embedded-eligible models from Scenario 2 — test RMSE ≈ 3.36 MPG)\n")
```

### 3.2 LASSO — Regularization Path

The regularization path shows how many features the model keeps as the penalty λ grows. At λ → 0 all features are retained (same as OLS). As λ increases, the L1 geometry forces progressively more coefficients to exactly zero. The CV-optimal λ is marked — this is where Scenario 2's cross-validation stopped.

```{r}
lasso_glmnet  <- lasso_fit %>% extract_fit_engine()
opt_log_lasso <- log10(best_lasso$penalty)

path_lasso <- tibble(
  log_lambda = log10(lasso_glmnet$lambda),
  n_selected = lasso_glmnet$df   # non-zero coefficients at each lambda
)

cat("=== LASSO — regularization path summary ===\n")
cat("Lambda range: [", round(min(lasso_glmnet$lambda), 6), ",",
                       round(max(lasso_glmnet$lambda), 3), "]\n")
cat("Features at λ_min (≈ OLS):   ",
    path_lasso$n_selected[which.min(path_lasso$log_lambda)], "\n")
cat("Features at λ_opt (CV best): ",
    path_lasso$n_selected[which.min(abs(path_lasso$log_lambda - opt_log_lasso))], "\n")
cat("Features at λ_max (max reg): ",
    path_lasso$n_selected[which.max(path_lasso$log_lambda)], "\n")

ggplot(path_lasso, aes(x = log_lambda, y = n_selected)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_vline(xintercept = opt_log_lasso, color = "tomato", lty = 2) +
  annotate("text",
           x     = opt_log_lasso + 0.15,
           y     = max(path_lasso$n_selected) * 0.95,
           label = paste0("λ_opt = ", round(best_lasso$penalty, 5)),
           color = "tomato", hjust = 0, size = 3.5) +
  labs(
    title    = "LASSO — number of selected features along the regularization path",
    subtitle = "Red dashed line = CV-optimal λ",
    x        = "log₁₀(λ)",
    y        = "Number of non-zero coefficients"
  ) +
  theme_minimal()
```

The path shows what LASSO does as an embedded selector: at low λ all features are kept; as λ grows, the weakest predictors drop out one by one. The CV-optimal λ sits at the left end of the path — meaning the 5-fold CV found that keeping all (or nearly all) features produces the lowest prediction error on this dataset.

### 3.3 LASSO — Selected Features at Optimal λ

These are the predictors LASSO kept at the CV-optimal penalty. A zero coefficient means the model decided that feature adds no useful information given the others.

```{r fig.height=6}
lasso_coefs_s3 <- lasso_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(term != "(Intercept)")

n_kept_lasso   <- sum(lasso_coefs_s3$estimate != 0)
n_zeroed_lasso <- sum(lasso_coefs_s3$estimate == 0)

cat("=== LASSO — embedded feature selection result ===\n")
cat("Total features available:    ", nrow(lasso_coefs_s3), "\n")
cat("Features selected (non-zero):", n_kept_lasso, "\n")
cat("Features removed (zeroed):   ", n_zeroed_lasso, "\n")

if (n_zeroed_lasso > 0) {
  cat("\nZeroed predictors:\n")
  print(lasso_coefs_s3 %>% filter(estimate == 0) %>% pull(term))
}

lasso_coefs_s3 %>%
  filter(estimate != 0) %>%
  slice_max(abs(estimate), n = 20) %>%
  ggplot(aes(x = reorder(term, abs(estimate)), y = estimate,
             fill = estimate > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
    labels = c("TRUE" = "Positive", "FALSE" = "Negative"),
    name   = "Direction"
  ) +
  labs(
    title    = paste0("LASSO — selected features at λ = ",
                      round(best_lasso$penalty, 5)),
    subtitle = paste0(n_kept_lasso, " features retained, ",
                      n_zeroed_lasso, " removed"),
    x        = NULL,
    y        = "Coefficient estimate"
  ) +
  theme_minimal()
```

### 3.4 Elastic Net — Regularization Path

Elastic Net mixes L1 and L2 penalties. The optimal mixture was found by CV in Scenario 2 (jointly tuning both `penalty` and `mixture`). We show its path using the same approach as LASSO.

```{r}
en_glmnet  <- en_fit %>% extract_fit_engine()
opt_log_en <- log10(best_en$penalty)

path_en <- tibble(
  log_lambda = log10(en_glmnet$lambda),
  n_selected = en_glmnet$df
)

cat("=== Elastic Net — regularization path summary ===\n")
cat("Optimal penalty (λ):", round(best_en$penalty, 6), "\n")
cat("Optimal mixture (α):", round(best_en$mixture, 4),
    " (0 = Ridge, 1 = LASSO)\n")
cat("Features at λ_min (≈ OLS):   ",
    path_en$n_selected[which.min(path_en$log_lambda)], "\n")
cat("Features at λ_opt (CV best): ",
    path_en$n_selected[which.min(abs(path_en$log_lambda - opt_log_en))], "\n")
cat("Features at λ_max (max reg): ",
    path_en$n_selected[which.max(path_en$log_lambda)], "\n")

ggplot(path_en, aes(x = log_lambda, y = n_selected)) +
  geom_line(color = "#27AE60", linewidth = 0.8) +
  geom_vline(xintercept = opt_log_en, color = "tomato", lty = 2) +
  annotate("text",
           x     = opt_log_en + 0.15,
           y     = max(path_en$n_selected) * 0.95,
           label = paste0("λ_opt = ", round(best_en$penalty, 5)),
           color = "tomato", hjust = 0, size = 3.5) +
  labs(
    title    = "Elastic Net — number of selected features along the regularization path",
    subtitle = paste0("Red dashed line = CV-optimal λ  |  α = ",
                      round(best_en$mixture, 3), "  (0 = Ridge, 1 = LASSO)"),
    x        = "log₁₀(λ)",
    y        = "Number of non-zero coefficients"
  ) +
  theme_minimal()
```

### 3.5 Elastic Net — Selected Features at Optimal λ

```{r fig.height=6}
en_coefs_s3 <- en_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(term != "(Intercept)")

n_kept_en   <- sum(en_coefs_s3$estimate != 0)
n_zeroed_en <- sum(en_coefs_s3$estimate == 0)

cat("=== Elastic Net — embedded feature selection result ===\n")
cat("Total features available:    ", nrow(en_coefs_s3), "\n")
cat("Features selected (non-zero):", n_kept_en, "\n")
cat("Features removed (zeroed):   ", n_zeroed_en, "\n")

if (n_zeroed_en > 0) {
  cat("\nZeroed predictors:\n")
  print(en_coefs_s3 %>% filter(estimate == 0) %>% pull(term))
}

en_coefs_s3 %>%
  filter(estimate != 0) %>%
  slice_max(abs(estimate), n = 20) %>%
  ggplot(aes(x = reorder(term, abs(estimate)), y = estimate,
             fill = estimate > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
    labels = c("TRUE" = "Positive", "FALSE" = "Negative"),
    name   = "Direction"
  ) +
  labs(
    title    = paste0("Elastic Net — selected features at λ = ",
                      round(best_en$penalty, 5)),
    subtitle = paste0(n_kept_en, " features retained, ",
                      n_zeroed_en, " removed"),
    x        = NULL,
    y        = "Coefficient estimate"
  ) +
  theme_minimal()
```

### 3.6 Comparing LASSO vs. Elastic Net Feature Selection

We compare which features each model kept and where they agree. Features retained by both are the most robustly selected — they carry enough signal to survive both pure L1 and blended L1+L2 penalization.

```{r}
selection_comparison <- lasso_coefs_s3 %>%
  select(term, lasso = estimate) %>%
  left_join(
    en_coefs_s3 %>% select(term, en = estimate),
    by = "term"
  ) %>%
  mutate(
    lasso_selected = lasso != 0,
    en_selected    = en    != 0,
    agreement      = case_when(
      lasso_selected & en_selected   ~ "Both selected",
      lasso_selected & !en_selected  ~ "LASSO only",
      !lasso_selected & en_selected  ~ "EN only",
      TRUE                           ~ "Both zeroed"
    )
  )

cat("=== LASSO vs. Elastic Net — selection agreement ===\n")
print(table(selection_comparison$agreement))

cat("\n--- Features selected by BOTH models (top 10 by |LASSO coef|) ---\n")
selection_comparison %>%
  filter(agreement == "Both selected") %>%
  arrange(desc(abs(lasso))) %>%
  head(10) %>%
  mutate(lasso = round(lasso, 4), en = round(en, 4)) %>%
  print()

selection_comparison %>%
  filter(lasso_selected | en_selected) %>%
  slice_max(abs(lasso) + abs(en), n = 20) %>%
  pivot_longer(cols = c(lasso, en), names_to = "model", values_to = "coef") %>%
  ggplot(aes(x = reorder(term, abs(coef)), y = coef, fill = model)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_manual(
    values = c("lasso" = "steelblue", "en" = "#27AE60"),
    labels = c("lasso" = "LASSO", "en" = "Elastic Net")
  ) +
  labs(
    title    = "LASSO vs. Elastic Net — coefficient comparison",
    subtitle = "Top 20 features selected by at least one model",
    x        = NULL,
    y        = "Coefficient estimate",
    fill     = "Model"
  ) +
  theme_minimal()
```

### 3.7 Test Set Performance

We evaluate both models on the held-out test set, showing prediction accuracy alongside the number of features used — the embedded selection framing.

```{r}
lasso_test_s3 <- lasso_fit %>%
  predict(new_data = test_data) %>%
  bind_cols(test_data %>% select(comb08))

en_test_s3 <- en_fit %>%
  predict(new_data = test_data) %>%
  bind_cols(test_data %>% select(comb08))

s3_comparison <- tibble(
  model      = c("LASSO", "Elastic Net"),
  n_features = c(n_kept_lasso, n_kept_en),
  RMSE = c(
    rmse_vec(truth = lasso_test_s3$comb08, estimate = lasso_test_s3$.pred),
    rmse_vec(truth = en_test_s3$comb08,   estimate = en_test_s3$.pred)
  ),
  MAE = c(
    mae_vec( truth = lasso_test_s3$comb08, estimate = lasso_test_s3$.pred),
    mae_vec( truth = en_test_s3$comb08,   estimate = en_test_s3$.pred)
  ),
  Rsq = c(
    rsq_vec( truth = lasso_test_s3$comb08, estimate = lasso_test_s3$.pred),
    rsq_vec( truth = en_test_s3$comb08,   estimate = en_test_s3$.pred)
  )
) %>%
  mutate(across(c(RMSE, MAE, Rsq), ~round(., 4)))

cat("=== Scenario 3 — Test set performance ===\n")
print(s3_comparison)

cat("\n=== OLS baseline (all", ncol(train_processed) - 1, "features) ===\n")
test_comparison %>%
  filter(model == "Linear Regression") %>%
  mutate(n_features = ncol(train_processed) - 1) %>%
  select(model, n_features, RMSE, MAE, Rsq) %>%
  print()
```

### 3.8 Extension: extended penalty grid (sparsity vs. accuracy)

The original Scenario 2 grid (`penalty(range = c(-4, 0))`) topped out at λ = 1, and CV picked λ = 1e-4 — the smallest value — which zeroed no features. The regularization path plot in §3.2 hinted at what would happen at higher λ, but we never measured the prediction error along that path. Here we re-tune LASSO over a much wider grid (10⁻⁴ to 10¹) and read off, for each λ, both the number of selected features and the test RMSE — the actual sparsity-vs-accuracy tradeoff.

```{r warning=FALSE}
lasso_grid_wide <- grid_regular(penalty(range = c(-4, 1)), levels = 25)

set.seed(123)
lasso_cv_wide <- lasso_wf %>%
  tune_grid(
    resamples = folds,
    grid      = lasso_grid_wide,
    metrics   = eval_metrics
  )

cv_rmse_wide <- collect_metrics(lasso_cv_wide) %>%
  filter(.metric == "rmse") %>%
  select(penalty, cv_rmse = mean)
```

```{r warning=FALSE}
sparsity_path <- map_dfr(lasso_grid_wide$penalty, function(lam) {
  fit_l <- lasso_wf %>%
    finalize_workflow(tibble(penalty = lam)) %>%
    fit(train_data)

  betas <- fit_l %>%
    extract_fit_parsnip() %>%
    tidy() %>%
    filter(term != "(Intercept)")

  pred_l <- predict(fit_l, new_data = test_data) %>%
    bind_cols(test_data %>% select(comb08))

  tibble(
    penalty    = lam,
    log_lambda = log10(lam),
    n_selected = sum(betas$estimate != 0),
    test_rmse  = rmse_vec(truth = pred_l$comb08, estimate = pred_l$.pred)
  )
}) %>%
  left_join(cv_rmse_wide, by = "penalty")

cat("=== LASSO — sparsity vs. accuracy across λ ===\n")
print(sparsity_path %>%
        mutate(across(c(cv_rmse, test_rmse, log_lambda), ~round(., 4))))
```

```{r}
ggplot(sparsity_path, aes(x = log_lambda)) +
  geom_line(aes(y = test_rmse, color = "Test RMSE"), linewidth = 0.9) +
  geom_line(aes(y = cv_rmse,   color = "CV RMSE"),   linewidth = 0.9, lty = 2) +
  geom_vline(xintercept = log10(best_lasso$penalty),
             color = "tomato", lty = 2) +
  scale_color_manual(values = c("Test RMSE" = "steelblue", "CV RMSE" = "#27AE60")) +
  labs(
    title    = "LASSO — RMSE vs. log₁₀(λ)  (extended grid)",
    subtitle = "Red dashed line = original CV-optimal λ",
    x        = "log₁₀(λ)",
    y        = "RMSE (MPG)",
    color    = NULL
  ) +
  theme_minimal()

ggplot(sparsity_path, aes(x = n_selected, y = test_rmse)) +
  geom_line(color = "steelblue", linewidth = 0.9) +
  geom_point(color = "steelblue", size = 2) +
  labs(
    title    = "LASSO — test RMSE vs. number of selected features",
    subtitle = "Each point is one λ on the extended grid",
    x        = "Number of non-zero coefficients",
    y        = "Test RMSE (MPG)"
  ) +
  theme_minimal()
```

```{r}
n_total <- sparsity_path$n_selected[which.min(sparsity_path$log_lambda)]

representative <- sparsity_path %>%
  arrange(log_lambda) %>%
  filter(
    penalty == best_lasso$penalty |
      n_selected == n_total |
      n_selected <= 1 |
      abs(n_selected - round(n_total / 2)) ==
        min(abs(n_selected - round(n_total / 2))) |
      abs(n_selected - 10) == min(abs(n_selected - 10))
  ) %>%
  distinct(n_selected, .keep_all = TRUE) %>%
  arrange(desc(n_selected)) %>%
  mutate(across(c(cv_rmse, test_rmse, penalty), ~round(., 5)))

cat("=== Representative sparsity levels — what does shrinkage cost? ===\n")
print(representative %>%
        select(penalty, log_lambda, n_selected, cv_rmse, test_rmse))
```

The plot shows where the elbow lies: down to a moderate number of features the test RMSE barely moves, then rises sharply once λ becomes large enough to zero out genuinely informative predictors. This is the answer the original narrow-grid Scenario 3 could not give — embedded selection on this data could discard a meaningful fraction of features at almost no cost in accuracy, until it crosses the threshold where the engine-size signal itself starts being shrunk.

---

### Conclusion

#### Embedded selection vs. Scenario 2 analysis

Scenario 2 used LASSO and Elastic Net as prediction models. Scenario 3 looks at the same models from a different angle: which features did they select, and what does the regularization path reveal about the feature space? The path plot is the embedded selector's "work log" — it shows which predictors are robust enough to survive increasing penalization, even if the CV-optimal λ happens to keep most features.

#### Regularization path insight

The CV-optimal λ sits near the far-left end of both paths — close to zero, meaning almost no features are removed. This is consistent with Scenario 2's finding: with 38 735 training observations and only ~60 dummy-encoded predictors, there is very little variance to reduce. Every predictor carries enough signal that the cross-validation penalty is not large enough to zero it out. The path still demonstrates the embedded selection mechanism: at higher λ values, features do drop out in order of their contribution — the path just happens to stop before that region is reached.

#### Which features were consistently selected

Features retained by both LASSO and Elastic Net at the CV-optimal λ are the most robustly embedded-selected predictors. The strongest predictors — `displ`, `cylinders`, `fuel_group_Electricity`, `drive` variants — appear with large coefficients in both models. Where the models differ slightly in coefficient magnitudes, the L2 component of Elastic Net distributes weight more evenly among correlated features (particularly `displ` and `cylinders` with r ≈ 0.91 in the EDA).

#### Performance vs. number of features

Both embedded models produce identical or near-identical RMSE to OLS with all features. This confirms that even where LASSO zeroes a predictor, the removed feature was genuinely redundant — the prediction does not suffer. Embedded selection here acts as a validator: it endorses the full feature set rather than pruning it, which is itself a meaningful result given the manual feature engineering done in preprocessing.

#### Suitability

LASSO is the simpler embedded selector — clean, interpretable, sparse when the data supports it. Elastic Net is a better fit for this dataset because several predictors are correlated (`displ`/`cylinders`, multiple `drive` dummies) — the L2 component stabilizes their joint selection. The CV-tuned mixture value shows how much L2 the data preferred. The same fundamental limitations apply as in Scenario 2: non-linear engine-size effects, PHEV driving-mode behavior, and missing weight/drag features are not addressable by changing the regularization strategy.

---

## Scenario 5: Partial Least Squares (PLS)

PLS (Partial Least Squares) is a supervised dimensionality reduction method for regression. It finds latent components — like PCA axes, but guided by the target — that maximally **correlate with `comb08`**, not just capture variance in X. We reuse the same preprocessed data as Scenario 2, so the comparison is direct: same features, same split, different method.

```{r}
# install.packages("pls") if not yet installed
library(pls)
```

### 5.1 Fitting PLS with Cross-Validation

We fit PLS with up to 15 components and use built-in 5-fold CV to find the optimal number of components. `scale = FALSE` because `train_processed` is already normalized by `step_normalize()` in the Scenario 2 recipe.

```{r}
set.seed(123)
pls_model <- plsr(
  comb08 ~ .,
  data       = train_processed,
  ncomp      = 15,
  validation = "CV",
  segments   = 5,     # 5-fold CV, matching Scenario 2
  scale      = FALSE  # already step_normalize()'d in recipe
)

# CV RMSEP per ncomp — index 1 = 0-component (intercept-only) model
rmsep_cv <- RMSEP(pls_model)$val["CV", 1, ]

# Elbow selection: fewest components whose RMSEP is within 1% of the global minimum.
# which.min() would pick the global minimum (often an overfit plateau); this
# approach favours parsimony while still being close to optimal.
min_rmsep <- min(rmsep_cv)
opt_ncomp <- which(rmsep_cv <= min_rmsep * 1.01)[1] - 1

cat("=== PLS — CV RMSEP by number of components ===\n")
for (k in 0:min(10, length(rmsep_cv) - 1)) {
  marker <- if (k == opt_ncomp) "  <-- selected" else ""
  cat(sprintf("  ncomp = %2d  |  CV RMSEP = %.4f%s\n", k, rmsep_cv[k + 1], marker))
}
cat("\nOptimal number of components:", opt_ncomp, "\n")
```

```{r}
plot(RMSEP(pls_model),
     main = "PLS — CV RMSEP by number of components",
     xlab = "Number of components",
     ylab = "CV RMSEP (MPG)")
abline(v = opt_ncomp, col = "tomato", lty = 2)
```

The elbow shows where adding more components stops improving CV accuracy. Components past the elbow add complexity without reducing error.

### 5.2 Explained Variance per Component

PLS components are ordered by how much covariance with `comb08` they capture. `explvar()` gives variance explained in X; `R2()` gives R² for y — the more important metric for a regression task.

```{r}
x_var <- explvar(pls_model)
y_r2  <- R2(pls_model, estimate = "CV")$val["CV", 1, -1]  # drop intercept (index 1)

cat("=== PLS — Explained variance per component ===\n")
# x_var is incremental (% of X variance added by each component) — cumulate it.
# y_r2 from R2() is already cumulative (R² when using k components total) — display directly.
cat(sprintf("%-6s  %9s  %10s  %12s\n",
    "Comp", "X var %", "Cum X %", "R²(y) CV"))
cumx <- 0
for (k in seq_len(min(opt_ncomp + 3, length(x_var)))) {
  cumx   <- cumx + x_var[k]
  marker <- if (k == opt_ncomp) "  <--" else ""
  cat(sprintf("%-6d  %9.2f  %10.2f  %12.4f%s\n",
      k, x_var[k], cumx, y_r2[k], marker))
}
```

Comp1 captures the dominant engine-size axis — the same linear combination that drove LD1 in LDA. Subsequent components pick up smaller, residual patterns.

### 5.3 Dimensionality Reduction — Projection onto First Two Components

We project all training vehicles onto Comp1 and Comp2. This is the dimensionality reduction result: ~60 normalized features collapse into 2 coordinates. Color shows actual MPG — a clean gradient confirms Comp1 captures the efficiency axis.

```{r}
scores_df <- tibble(
  Comp1  = scores(pls_model)[, 1],
  Comp2  = scores(pls_model)[, 2],
  comb08 = train_data$comb08
)

cat("=== PLS — Score ranges (training data) ===\n")
cat("Comp1:", round(range(scores_df$Comp1), 3), "\n")
cat("Comp2:", round(range(scores_df$Comp2), 3), "\n")

ggplot(scores_df, aes(x = Comp1, y = Comp2, color = comb08)) +
  geom_point(alpha = 0.3, size = 0.6) +
  scale_color_viridis_c(option = "plasma", name = "MPG") +
  labs(
    title    = "PLS — projection of training data onto first two components",
    subtitle = "Color = actual combined MPG (purple = low, yellow = high)",
    x        = "PLS Component 1",
    y        = "PLS Component 2"
  ) +
  theme_minimal()
```

### 5.4 Component Loadings

Loadings show which predictors drive Comp1. Negative loadings push toward lower MPG; positive loadings push toward higher MPG.

```{r fig.height=6}
loadings_df <- as_tibble(loadings(pls_model)[, 1, drop = FALSE],
                          rownames = "predictor") %>%
  rename(Comp1 = 2)

loadings_df %>%
  slice_max(abs(Comp1), n = 15) %>%
  ggplot(aes(x = reorder(predictor, Comp1), y = Comp1, fill = Comp1 > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#27AE60", "FALSE" = "#E74C3C"), guide = "none") +
  labs(
    title    = "PLS — loadings of Component 1",
    subtitle = "Top 15 predictors — green = higher MPG, red = lower MPG",
    x        = NULL,
    y        = "Comp1 loading"
  ) +
  theme_minimal()

cat("=== PLS — Top 5 predictors for Comp1 (absolute loading) ===\n")
loadings_df %>%
  slice_max(abs(Comp1), n = 5) %>%
  mutate(Comp1 = round(Comp1, 4)) %>%
  print()
```

### 5.5 Test Set Evaluation

We predict on the held-out test set using `opt_ncomp` components and compare directly with the four Scenario 2 models — same test set, same metrics.

```{r}
pls_pred <- predict(pls_model, newdata = test_processed, ncomp = opt_ncomp)[, , 1]

pls_metrics <- tibble(
  truth = test_data$comb08,
  pred  = as.numeric(pls_pred)
)

pls_rmse <- rmse_vec(truth = pls_metrics$truth, estimate = pls_metrics$pred)
pls_mae  <- mae_vec( truth = pls_metrics$truth, estimate = pls_metrics$pred)
pls_rsq  <- rsq_vec( truth = pls_metrics$truth, estimate = pls_metrics$pred)

cat("=== PLS — Test set metrics (ncomp =", opt_ncomp, ") ===\n")
cat("RMSE:", round(pls_rmse, 4), "\n")
cat("MAE: ", round(pls_mae,  4), "\n")
cat("R²:  ", round(pls_rsq,  4), "\n")

cat("\n=== All models: PLS vs. Scenario 2 ===\n")
bind_rows(
  test_comparison,
  tibble(model = paste0("PLS (", opt_ncomp, " comp)"),
         RMSE  = pls_rmse,
         MAE   = pls_mae,
         Rsq   = pls_rsq)
) %>%
  arrange(RMSE) %>%
  print()
```

```{r}
ggplot(pls_metrics, aes(x = truth, y = pred)) +
  geom_point(alpha = 0.15, color = "steelblue", size = 0.8) +
  geom_abline(slope = 1, intercept = 0, color = "tomato", linewidth = 0.8) +
  labs(
    title = "PLS — Actual vs. Predicted (test set)",
    x     = "Actual combined MPG",
    y     = "Predicted combined MPG"
  ) +
  theme_minimal()
```

### 5.6 Extension: PCR (unsupervised components + OLS)

PLS chooses components that maximize covariance with `comb08` — a *supervised* projection. PCA chooses components that maximize variance of the predictors alone, ignoring the target — *unsupervised*. PCR is OLS fitted on the top-k PCA components. Comparing PLS at its optimal k = 3 with PCR at the same k tells us whether the supervision actually helped on this dataset, or whether the data's principal axes already happen to align with MPG.

```{r warning=FALSE}
fit_pcr_k <- function(k) {
  rec_pcr_k <- recipe(comb08 ~ ., data = train_data) %>%
    step_unknown(all_nominal_predictors()) %>%
    step_other(all_nominal_predictors(), threshold = 0.01) %>%
    step_impute_median(displ, cylinders) %>%
    step_dummy(all_nominal_predictors()) %>%
    step_zv(all_predictors()) %>%
    step_nzv(all_predictors()) %>%
    step_normalize(all_numeric_predictors()) %>%
    step_pca(all_numeric_predictors(), num_comp = k)

  pcr_wf_k <- workflow() %>%
    add_recipe(rec_pcr_k) %>%
    add_model(linear_reg() %>% set_engine("lm"))

  fit_k  <- pcr_wf_k %>% fit(train_data)
  pred_k <- predict(fit_k, new_data = test_data) %>%
    bind_cols(test_data %>% select(comb08))

  tibble(
    k    = k,
    RMSE = rmse_vec(truth = pred_k$comb08, estimate = pred_k$.pred),
    MAE  = mae_vec( truth = pred_k$comb08, estimate = pred_k$.pred),
    Rsq  = rsq_vec( truth = pred_k$comb08, estimate = pred_k$.pred)
  )
}

pcr_results <- map_dfr(1:15, fit_pcr_k)

cat("=== PCR — test RMSE per number of components ===\n")
print(pcr_results %>% mutate(across(c(RMSE, MAE, Rsq), ~round(., 4))))
```

```{r}
pls_per_k <- tibble(
  k    = 1:15,
  RMSE = sapply(1:15, function(k) {
    p <- predict(pls_model, newdata = test_processed, ncomp = k)[, , 1]
    rmse_vec(truth = test_processed$comb08, estimate = as.numeric(p))
  })
)

ggplot() +
  geom_line(data = pcr_results, aes(x = k, y = RMSE, color = "PCR (PCA + OLS)"),
            linewidth = 0.9) +
  geom_point(data = pcr_results, aes(x = k, y = RMSE, color = "PCR (PCA + OLS)"),
             size = 2) +
  geom_line(data = pls_per_k, aes(x = k, y = RMSE, color = "PLS"),
            linewidth = 0.9) +
  geom_point(data = pls_per_k, aes(x = k, y = RMSE, color = "PLS"),
             size = 2) +
  geom_vline(xintercept = 3, color = "tomato", lty = 2) +
  scale_color_manual(values = c("PCR (PCA + OLS)" = "#27AE60", "PLS" = "steelblue")) +
  labs(
    title    = "PLS vs. PCR — test RMSE by number of components",
    subtitle = "Red dashed line = k = 3 (PLS elbow)",
    x        = "Number of components",
    y        = "Test RMSE",
    color    = "Method"
  ) +
  theme_minimal()
```

```{r}
pcr_k3 <- pcr_results %>% filter(k == 3)
pcr_k15 <- pcr_results %>% filter(k == 15)

pls_pcr_table <- tibble(
  model        = c("PLS", "PCR", "PCR", "OLS (full)"),
  n_components = c("3", "3", "15", "≈ 60 (all features)"),
  RMSE         = c(round(pls_rmse, 4),
                   round(pcr_k3$RMSE, 4),
                   round(pcr_k15$RMSE, 4),
                   test_comparison %>% filter(model == "Linear Regression") %>% pull(RMSE)),
  Rsq          = c(round(pls_rsq, 4),
                   round(pcr_k3$Rsq, 4),
                   round(pcr_k15$Rsq, 4),
                   test_comparison %>% filter(model == "Linear Regression") %>% pull(Rsq))
)

cat("=== PLS vs. PCR vs. OLS — test set comparison ===\n")
print(pls_pcr_table)
```

If PCR at k = 3 matches PLS at k = 3, the dataset's principal axes of variance already point in the direction of `comb08` — supervision adds nothing because the unsupervised projection happens to be aimed correctly. If PCR is meaningfully worse at k = 3 but catches up by k = 15, supervision genuinely helped: PLS finds the MPG-relevant direction faster than PCA does.

---

### Conclusion

#### Dimensionality reduction

With ~60 normalized features as input, PLS selected just **3 components** (1% elbow threshold). Comp1 alone accounts for **R²(y) = 0.597** — one linear combination captures over half the MPG variance. Adding Comp2 brings it to **0.663**, and Comp3 to **0.670**. Everything beyond that is marginal: 11 more components improve R² by only 0.008.

The top Comp1 predictors are `displ` (−0.517) and `cylinders` (−0.483), both negative — larger engines mean lower MPG — followed by front-wheel drive (0.376, positive) and rear-wheel drive (−0.336, negative). This is the engine-size axis: the same dominant direction that appeared in every Scenario 2 model's coefficient table.

#### Predictive performance

With 3 components: **RMSE ≈ 3.31 MPG, R² ≈ 0.670**. With the global minimum at 14 components: RMSE = 3.281, R² = 0.678 — a difference of 0.03 MPG over 11 additional components. Both are well within the margin of noise.

#### Comparison with Scenario 2

PLS (3 components) matches OLS almost exactly: **RMSE 3.28, R² 0.678** vs. OLS RMSE 3.28, R² 0.678 — the difference is in the fourth decimal place. This makes sense: OLS and PLS are both linear models over the same features; PLS just finds a compressed representation first. The comparison table also confirms that Ridge, LASSO, and Elastic Net all landed at the same result — adding regularization or changing the basis (PLS components vs. original features) doesn't help when the binding constraint is linearity, not dimensionality or variance.

The real message of PLS here is the compression story: **3 numbers are enough to predict MPG as accurately as 60 features**. Everything beyond Comp3 is noise.

#### Suitability

PLS is a good fit: continuous target, correlated predictors, large n. The RMSEP curve has a long flat tail after ncomp=3, which is typical when one dominant factor (engine size) drives most of the signal and the rest is diffuse. The same limitations as Scenario 2 apply — non-linear MPG relationships, PHEVs, missing weight and drag. PLS expresses the same linear model more compactly but cannot fix what linear regression already can't fix.

---

## Overall Assessment: Scenarios 2, 3, and 5

### Was the data suitable?

**Scenario 2 (parametric regression):** Yes. Large dataset (48k rows), only 21 predictors, continuous target with strong linear correlations (up to −0.77 with cylinders). Nearly ideal conditions for linear regression.

**Scenario 3 (embedded feature selection):** Yes — LASSO and Elastic Net are exactly the right embedded methods for a regression problem with a continuous target and correlated predictors. The L1 penalty is a natural embedded selector; the data's large n/p ratio means the selection result (keep all features) is itself informative.

**Scenario 5 (PLS):** Yes — PLS is a regression method with the same continuous target as Scenario 2. It handles correlated predictors well and makes no distributional assumptions on the features. The only assumption is that linear combinations of X are sufficient to predict y, which is the same assumption OLS makes.

---

### What worked against each method?

**Scenario 2:**

1. Non-linear relationships — adding a cylinder affects small engines differently than V8s; linear models can't capture this without interaction terms.
2. PHEVs — their MPG depends on driving mode (electric vs. combustion), which no static feature captures.
3. Missing features — vehicle weight and aerodynamic drag are among the strongest determinants of fuel economy and aren't in the dataset.
4. Regularization was irrelevant — all three regularized models converged to penalty = 1e-4 (minimum of the grid). With 48k rows and 21 features, there's nothing to regularize. Zero gain over OLS.

**Scenario 3:**

1. Regularization barely activates — the CV-optimal λ is effectively zero, so LASSO and EN behave like OLS at the operating point. The embedded selection mechanism is visible on the regularization path, but the CV stops before it reaches the feature-pruning region.
2. Correlated predictors — pure LASSO can arbitrarily choose one of a correlated pair; EN's L2 component mitigates this, but the underlying collinearity (displ/cylinders, r ≈ 0.91) still influences which dummy-encoded features receive weight.
3. Same root limitations as Scenario 2 — non-linearity, PHEVs, missing weight/drag features.

**Scenario 5:**

1. Non-linear relationships — same limitation as Scenario 2. PLS finds the best linear latent directions, but can't model the non-constant marginal effect of adding a cylinder.
2. PHEVs and missing features — PLS Comp1 captures the engine-size axis cleanly, but PHEV driving-mode behavior and absent weight/drag features limit accuracy for the same reasons as OLS.
3. Diminishing returns after Comp1 — most of the predictive signal concentrates in the first component; later components explain little y-variance, so PLS converges quickly to the same ceiling as OLS.

---

### Was the result sufficient?

**Scenario 2:** Good baseline. RMSE = 3.36 MPG (≈11% of the 20 MPG median), R² = 0.661. Fine for analysis and feature interpretation, but not precise enough for regulatory use. One-third of variance is unexplained — missing features (weight, drag) are the likely cause, not model weakness.

**Scenario 3:** RMSE identical to Scenario 2 — both LASSO and EN produce the same test performance as OLS. The embedded selection endorsed the full feature set: every predictor carries real signal at this sample size. The regularization path confirms that embedded selection would prune features at higher λ, but the CV-optimal point is too close to OLS for any pruning to pay off in prediction accuracy.

**Scenario 5:** RMSE = 3.28 MPG with 3 components — identical to OLS with all 60 features. The dimensionality reduction is real (60 → 3), but the prediction improvement over OLS is zero because OLS was already variance-free at n/p > 2000.

---

### How to improve?

Three of the most actionable improvements have already been implemented as supplementary subsections inside their parent scenarios — the items below are what remains.

**Scenario 2 (interaction terms — implemented in §2.5):** Adding the `cylinders × displ` cross-term tests the most obvious non-linearity in the data. Remaining: vehicle weight and aerodynamic drag are not in the dataset; gradient boosting would capture non-linearities and interactions automatically but belongs to a different model family than this report's parametric focus, so it is deferred.

**Scenario 3 (extended penalty grid — implemented in §3.8):** The wider grid (10⁻⁴ to 10¹) produces the actual sparsity-vs-RMSE curve that the original narrow grid could not reveal. Remaining: a two-stage LASSO+OLS refit (run LASSO at high λ, then refit OLS on the surviving subset) would test whether the LASSO bias hurts prediction once the feature set is fixed. Adding weight/drag would also let the embedded selector genuinely prune weaker existing predictors.

**Scenario 5 (PCR comparison — implemented in §5.6):** PCR (OLS on PCA components) is the unsupervised counterpart to PLS and directly answers whether the supervision in PLS actually helped on this target. Remaining: kernel PLS would capture non-linear relationships but loses the interpretable loadings that made Scenario 5 readable; weight and drag features remain unavailable.

For this dataset, **regression is the right problem formulation** — Scenarios 2, 3, and 5 all confirm this. Together they show that the dominant signal lives in a small number of engine-size features (Scenario 5: 3 PLS components capture it; Scenario 3: embedded selection endorses every feature because all carry independent signal at n = 38k).
