---
title: "Vehicles dataset"
author: "Dominika Lukašáková, Lukáš Lovás"
date: "2026-04-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("D:/Škola/Ing/2.semester/OZNAL_projekt")
```

## EDA

Vehicles dataset\
Source: <https://fueleconomy.gov/feg/ws/index.shtml>

Based on the column descriptions, we could potentially predict fuel consumption (MPG).

-   **MPG (Miles per gallon)** – a measure of fuel economy in automobiles, higher number= higher efficiency\
-   **Target variable:** `comb08`

```{r}
library(tidyverse)
library(ggcorrplot)
library(patchwork)

rm(list = ls())
data <- read_csv("data/vehicles.csv")
data
View(data)
```

## Basic Structure

First, we check dataset dimensions and data types.

```{r}
dim(data)
```

The dataset has 49,846 rows and 84 columns.

```{r}
glimpse(data)
```

```{r}
data %>%
  summarise(across(everything(), class))
```

The dataset contains columns of type: character, double, logical. Columns createdOn and modifiedOn are dates stored as strings. The dataset has some missing values.

## Missing Values

```{r}
sum(is.na(data))
```

The dataset contains 538 466 missing values.

# Identify columns with missing values

```{r}
missing_summary <- data %>%
  summarise(across(everything(), ~mean(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_pct") %>%
  arrange(desc(missing_pct))

missing_summary %>%
  print(n = 84)
```

Columns with more than 60% missing values:

-   `c240bDscr`, `c240Dscr`, `sCharger`, `rangeA`, `fuelType2`, `guzzler`, `evMotor`,
-   `atvType`, `tCharger`, `trans_dscr`, `startStop`, `mfrCode`

We will not use these columns and remove them later.

-   Column `eng_dscr` has \~37% missing values
-   Columns `cylinders`, `displ`, `drive`, `trany`, `fuelType1` contain only a small amount of missing data (up to \~3%), so they could potentially be imputed.

## Unique Values

Next we identify columns with unique values.

```{r}
data %>%
  summarise(across(everything(), n_distinct)) %>%
  pivot_longer(everything(), 
               names_to = "variable", 
               values_to = "n_unique") %>%
  arrange(desc(n_unique)) %>%
  print(n = 84)
```

From these we filter out non-informative columns. We remove variables with nearly all unique values, as they act as identifiers and do not provide generalizable patterns for prediction. Also variables with very low variability (binary or near-constant features) were excluded, as they provide minimal information for explaining variation in the target variable.

```{r}
data %>%
  summarise(across(everything(), n_distinct)) %>%
  pivot_longer(everything(), 
               names_to = "variable", 
               values_to = "n_unique") %>%
  arrange(desc(n_unique)) %>%
  filter(!(n_unique >= 0.9 * nrow(data))) %>%
  filter(!(n_unique <= 2)) %>%
  print(n = 84)
```

-   We should also remove `model` column, because its extremely high number of unique values makes it impractical for modeling, and `engId,` as it also represents an identifier.
-   The dataset contains a large number of variables with high cardinality, indicating a high-dimensional feature space that requires careful feature selection.

## Numerical Variables

First, we check the distribution of numerical variables.

```{r}
num_data <- data %>% 
  select(where(is.numeric)) %>% 
  select(-id, -engId)

summary(num_data)
```

-   Many columns have median = 0 because most vehicles do not have that feature, for example `highwayE` applies only to electric vehicles. So, these variables might be informative for specific subsets, but not for the general fuel consumption prediction.

-   Some columns have min and median = -1, which is unusual and likely represents missing values. The columns are: `co2`, `co2A`, `feScore`, `ghgScore`, `ghgScoreA`.

-   Although column `youSaveSpend` does not have issues in terms of distribution, its definition reveals that it is directly derived from fuel consumption and fuel cost variables. As a result, it contains information closely related to the target variable and could introduce data leakage and artificially inflating model performance.

For further analysis we will remove: - columns with median = 0 or -1 - column `youSaveSpend`

```{r}
num_data <- num_data %>%
  select(where(~ {
    med <- median(., na.rm = TRUE)
    !(med == 0 | med == -1)
  })) %>%
  select(-youSaveSpend)

summary(num_data)
```

-   We are left with 11 numerical columns.

-   The similarity between mean and median values suggests that most variables are not heavily skewed, although there might be some outliers.

```{r}
num_data %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~variable, scales = "free") +
  theme_minimal()
# A warning is displayed because of NA in columns cylinders and displ
```

Most variables show right-skewed distributions, with the majority of observations with lower values and with a long tail toward higher values.

A few variables, such as `barrels08` and `co2TailpipeGpm` appear approximately normally distributed, though still with slight positive skew. In contrast, `cylinders` is a discrete variable, with clear peaks at common engine configurations - 4, 6, and 8 cylinders.

The variable `year` shows a relatively even but slightly increasing distribution toward more recent years, indicating a higher representation of newer vehicles in the dataset.

```{r}
num_data %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "value") %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()
# A warning is displayed because of NA in columns cylinders and displ
```

-   The variable `fuelCost08` shows a wide IQR and numerous outliers.
-   Column `co2TailpipeGpm` has a relatively compact central distribution but still includes several higher outliers.
-   The `year`column has a narrow spread from 1984 to 2026, indicating that most observations are concentrated within a similar time range, so there are no outliers.

We will plot another boxplot without columns `year`, `fuelCost08`, and `co2TailpipeGpm` due to differences in scale:

```{r}
num_data %>%
  select(-year, -fuelCost08, -co2TailpipeGpm) %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "value") %>%
  ggplot(aes(x = variable, y = value)) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()
# A warning is displayed because of NA in columns cylinders and displ
```

-   All columns have a significant amount of outsiders on the right side.

The boxplots reveal that most variables are right-skewed, as indicated by longer upper whiskers and a large number of high-value outliers. This pattern is especially clear for fuel economy variables (`city08`, `highway08`, `comb08`, `UCity`, `UHighway`), where a small number of highly efficient vehicles extend the upper range.

Variables such as `barrels08` and `displ` exhibit moderate right skewness with some extreme values. In contrast, `cylinders` is a discrete variable, making its boxplot less informative and leading to apparent outliers due to limited possible values.

Overall, the boxplots highlight skewness and the presence of outliers.

## Correlation Analysis

```{r}
cor_matrix <- num_data %>%
  select(where(is.numeric)) %>%
  cor(use = "pairwise.complete.obs")

cor_matrix
ggcorrplot(cor_matrix)
```

-   Target `comb08` is highly positively correlated with:

    -   `city08`, `highway08`, `UCity`, `UHighway` (up to \~0.996), so the usage of these columns in a model would cause data leakage

-   Target `comb08` is strongly negatively correlated with:

    -   `barrels08`, `co2TailpipeGpm`, `cylinders`, `displ`, `fuelCost08` (up to \~ -0.7)
    -   however we know from the columns description that columns `barrels08`, `co2TailpipeGpm` and `fuelCost08` are derived from fuel consumption, so they would also cause data leakage
    -   `cylinders` and `displ` are suitable predictors but they are highly correlated with each other (\~0.91)

So, for the model, we will keep only column **`displ`**, because it has fewer missing values and a stronger correlation with target (\~ -0.71), and column `year`, because it has a moderate positive correlation (\~0.32) with target and has almost no correlation with `displ`.

## Categorical Variables

```{r}
cat_data <- data %>% 
  select(where(is.character))

ncol(cat_data)

names(cat_data)
```

The dataset has 23 categorical variables.

```{r}
cat_overview <- cat_data %>%
  summarise(across(everything(), ~n_distinct(.))) %>%
  pivot_longer(everything(),
               names_to = "variable",
               values_to = "n_levels")

cat_overview
```

Based on the number of unique values and the description of each column, we will further explore only `fuelType`, `drive`, `VClass` and `trany` columns, because they have rational amount of categories and are related to fuel efficiency.

```{r}
data %>%
  select(fuelType, drive, VClass, trany) %>%
  pivot_longer(everything(),
               names_to = "variable",
               values_to = "category") %>%
  filter(!is.na(category)) %>%
  group_by(variable, category) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(variable) %>%
  slice_max(n, n = 10) %>%
  ggplot(aes(x = reorder(category, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  facet_wrap(~variable, scales = "free") +
  theme_minimal()
```

The `fuelType` variable shows a highly imbalanced distribution, with a few dominant categories (`Regular` and `Premium`) accounting for most observations, while several alternative fuel types appear only rarely. We will group rare categories of the fuelType variable into an "Other" class.

The `drive` variable is moderately imbalanced. `Front-Wheel Drive` and `Rear-Wheel Drive` represent the majority of observations, while several four-wheel-drive variants and two-wheel-drive vehicles occur much less frequently. The presence of multiple related four-wheel-drive labels indicates potential redundancy in category definitions, so merging semantically similar categories could reduce noise.

The `trany` variable contains many transmission categories, with a clear dominance of `Automatic 4-speed` and `Manual 5-speed` types.

The VClass has 3 dominant categories - `Compact Cars`, `Midsize Cars`, and `Subcompact Cars`. The other catogories appear less frequently. Since vehicle class captures structural differences in size, it is expected to be an informative predictor and should be retained in the modeling stage.

```{r}
ggplot(data, aes(x = fuelType, y = comb08)) +
  geom_boxplot() +
  coord_flip()
```

The boxplot shows how different fuel type influences fuel consumption. We can see that the electricity type has higher values then gasoline types.

```{r}
data %>%
  filter(!is.na(drive)) %>%
  ggplot(aes(x = drive, y = comb08)) +
  geom_boxplot() +
  coord_flip()
```

Drive type clearly affects fuel efficiency. `Front-Wheel Drive` vehicles tend to have higher combined MPG, while `4-Wheel or All-Wheel Drive` has smaller values.

```{r}
data %>%
  filter(!is.na(VClass)) %>%
  mutate(VClass = fct_lump(VClass, n = 10)) %>%
  ggplot(aes(x = VClass, y = comb08)) +
  geom_boxplot() +
  coord_flip()
```

Vehicle class has a clear relationship with fuel efficiency. Smaller vehicles (`compact` and `subcompact cars`) tend to achieve higher MPG, while pickups, SUVs, and larger vehicles have smaller MPG values, so they are less fuel efficient.

```{r}
data %>%
  filter(!is.na(trany)) %>%
  mutate(trany = fct_lump(trany, n = 10)) %>%
  ggplot(aes(x = trany, y = comb08)) +
  geom_boxplot() +
  coord_flip()
```

Transmission type strongly differentiates vehicle efficiency. `Automatic (A1)` shows the highest MPG values, while other categories have smaller values.

## Tagret variable

```{r}
summary(data$comb08)
```

The target variable `comb08` (combined fuel efficiency) has strongly right-skewed distribution. Most vehicles have between 15 and 30 MPG, with a median value of 20 MPG. However, a long right tail is observed, extending up to 146 MPG.

```{r}
hist <- ggplot(data, aes(x = comb08)) +
  geom_histogram(bins = 40, fill = "steelblue") +
  theme_minimal()

box <- ggplot(data, aes(y = comb08)) +
  geom_boxplot(fill = "steelblue") +
  coord_flip() +
  theme_minimal()

hist + box + plot_layout(ncol = 1, nrow = 2) 
```

Based on the histogram, we can say that the distribution is not normal. The boxplot confirms the presence of numerous high-value outliers. However, these observations are not errors but represent a distinct technological group of vehicles with significantly higher efficiency.

## Preprocessing and Train/Test Split

```{r message=FALSE, warning=FALSE}
library(rsample)
library(recipes)
library(stringr)
library(forcats)
```

### Column Groups to Remove

#### Columns with Too Many Missing Values (`cols_missing`)

These variables have more than 60% missing values and cannot be reliably used in a general model.

```{r}
cols_missing <- c(
  "c240bDscr", "c240Dscr", "sCharger", "rangeA", "fuelType2",
  "guzzler", "evMotor", "atvType", "tCharger", "trans_dscr",
  "startStop", "mfrCode"
)
```

#### Identifier Columns (`cols_ids`)

These variables are identifiers or overly specific strings. Using them would encourage the model to memorize specific vehicles or models.

```{r}
cols_ids <- c("id", "engId", "model")
```

#### Data Leakage Columns (`cols_leakage`)

These variables are either directly equivalent to `comb08` or derived from the same fuel consumption, emissions, or cost calculations. Including them would incorrectly inflate model performance.

```{r}
cols_leakage <- c(
  "city08", "highway08", "UCity", "UHighway",
  "barrels08", "co2TailpipeGpm", "fuelCost08", "youSaveSpend",
  "comb08U", "city08U", "highway08U",
  "combA08", "cityA08", "highwayA08",
  "combA08U", "cityA08U", "highwayA08U",
  "co2", "co2A", "co2TailpipeAGpm",
  "feScore", "ghgScore", "ghgScoreA",
  "phevCity", "phevHwy", "phevComb",
  "combE", "cityE", "highwayE",
  "combinedCD", "combinedUF",
  "cityCD", "cityUF",
  "highwayCD", "highwayUF"
)
```

#### Sparse or Invalid Columns

As identified in the EDA, many numeric columns have median = 0 or contain encoded missing values (−1). These are typically relevant only for electric or hybrid vehicles. Including them in the general regression dataset would introduce large numbers of zeros, encoded missing values, or derived information.

```{r}
cols_sparse_or_invalid <- c(
  "barrelsA08", "charge120", "charge240", "charge240b",
  "city08U", "cityA08", "cityA08U", "cityCD", "cityE", "cityUF",
  "co2", "co2A", "co2TailpipeAGpm",
  "comb08U", "combA08", "combA08U", "combE",
  "combinedCD", "combinedUF",
  "feScore", "fuelCostA08", "ghgScore", "ghgScoreA",
  "highway08U", "highwayA08", "highwayA08U",
  "highwayCD", "highwayE", "highwayUF",
  "hlv", "hpv", "lv2", "lv4", "pv2", "pv4",
  "range", "rangeCity", "rangeCityA", "rangeHwy", "rangeHwyA",
  "UCityA", "UHighwayA",
  "phevCity", "phevHwy", "phevComb"
)
```

#### Unused Columns

- `eng_dscr` has ~37% missing values and is a free-text engine description.
- `createdOn` and `modifiedOn` are metadata with no predictive value.
- `baseModel` would support memorization of specific model lines rather than generalization.
- `fuelType` and `fuelType1` are replaced by the new `fuel_group` variable.

```{r}
cols_unused <- c("createdOn", "modifiedOn", "eng_dscr", "baseModel")
```

#### Combine All Removal Lists

```{r}
cols_to_remove <- unique(c(cols_missing, cols_ids, cols_leakage,
                            cols_sparse_or_invalid, cols_unused))

cat("Total columns scheduled for removal:", length(cols_to_remove), "\n")
```

### Feature Engineering: Electric Vehicles

Electric vehicles do not have a combustion engine and therefore have no `displ` or `cylinders`. Their missing values are structurally meaningful — they signal the absence of an engine, not random data loss. We capture this before any imputation.

- `is_electric`: TRUE if the vehicle runs on electricity only.
- `has_combustion_engine`: TRUE if the vehicle has a combustion engine.

For EVs, `displ` and `cylinders` are set to 0 — meaning zero displacement, not a statistical fill-in. Remaining missing values in non-EV rows will be handled later by the recipe.

```{r}
data_fe <- data %>%
  mutate(
    is_electric = (fuelType == "Electricity") | (fuelType1 == "Electricity"),
    is_electric = replace_na(is_electric, FALSE),
    has_combustion_engine = !is_electric,
    displ     = if_else(is_electric, 0, displ),
    cylinders = if_else(is_electric, 0, cylinders)
  )

cat("Electric vehicles:", sum(data_fe$is_electric), "\n")
cat("Combustion engine vehicles:", sum(data_fe$has_combustion_engine), "\n")
```

### Categorical Variable Simplification

All transformations below use deterministic rules (not statistics learned from data), so they can safely be applied before the train/test split.

#### `fuel_group` — Structured Fuel Type Grouping

Instead of lumping rare levels of `fuelType`, we create a new interpretable variable `fuel_group` that consolidates all fuel categories into six meaningful groups. The raw `fuelType` and `fuelType1` columns remain unchanged in the original `data` object. For modeling, only `fuel_group` is used as a predictor — `fuelType` and `fuelType1` are removed from `vehicles_model` to avoid duplicating the same information across multiple columns.

```{r}
data_fe <- data_fe %>%
  mutate(
    fuel_group = case_when(
      fuelType %in% c("Regular", "Premium", "Midgrade")       ~ "Gasoline",
      fuelType == "Diesel"                                     ~ "Diesel",
      fuelType == "Electricity"                                ~ "Electricity",
      fuelType %in% c(
        "Premium and Electricity",
        "Regular Gas and Electricity",
        "Premium Gas or Electricity",
        "Regular Gas or Electricity",
        "Electricity and Hydrogen"
      )                                                        ~ "Hybrid/Electric mix",
      fuelType %in% c("Gasoline or E85", "Premium or E85")    ~ "Flex-fuel/E85",
      fuelType %in% c(
        "CNG", "Hydrogen",
        "Gasoline or natural gas",
        "Gasoline or propane"
      )                                                        ~ "Alternative fuel",
      TRUE                                                     ~ "Other"
    )
  )

table(data_fe$fuel_group)
```

#### `drive` — Consolidate 4WD/AWD Variants

The EDA identified multiple overlapping 4-wheel and all-wheel drive labels. We consolidate them into three interpretable groups.

```{r}
data_fe <- data_fe %>%
  mutate(drive = case_when(
    str_detect(drive, "4-Wheel|All-Wheel") ~ "AWD/4WD",
    str_detect(drive, "Front-Wheel")       ~ "Front-Wheel Drive",
    str_detect(drive, "Rear-Wheel")        ~ "Rear-Wheel Drive",
    is.na(drive)                           ~ "Unknown",
    TRUE                                   ~ "Other"
  ))

table(data_fe$drive)
```

#### `VClass` — Keep Top 10 Categories

`VClass` has many levels. We retain the 10 most frequent and collapse the rest into "Other".

```{r}
data_fe <- data_fe %>%
  mutate(VClass = fct_lump(as.factor(VClass), n = 10, other_level = "Other"))

table(data_fe$VClass)
```

#### `trany` — Extract Structured Transmission Features

Instead of keeping the raw high-cardinality `trany` string, we extract four interpretable features:

- `transmission_type`: "Automatic", "Manual", or "Other"
- `is_cvt`: TRUE if the transmission uses continuously variable gear ratios
- `n_gears`: number of speeds extracted from text; 0 means "not applicable" (CVT or non-standard transmission)
- `has_discrete_gears`: TRUE if the number of gear steps was successfully extracted from `trany`

`has_discrete_gears` and `n_gears` are designed to be read together: `has_discrete_gears = FALSE` means a CVT or other transmission without a discrete gear count — this is not randomly missing data, but a structural property of the transmission. Setting `n_gears = 0` via `replace_na` provides a numeric sentinel that is safe to use in models without median imputation.

The raw `trany` column is then removed.

```{r}
data_fe <- data_fe %>%
  mutate(
    transmission_type = case_when(
      str_detect(trany, regex("^Automatic", ignore_case = TRUE)) ~ "Automatic",
      str_detect(trany, regex("^Manual",    ignore_case = TRUE)) ~ "Manual",
      TRUE ~ "Other"
    ),
    is_cvt  = str_detect(trany, regex("variable gear ratios", ignore_case = TRUE)),
    is_cvt  = replace_na(is_cvt, FALSE),
    n_gears = as.integer(str_match(trany, "(\\d+)-spd")[, 2]),
    has_discrete_gears = as.integer(!is.na(n_gears)),
    n_gears = replace_na(n_gears, 0)
  )

cat("Transmission type distribution:\n")
print(table(data_fe$transmission_type))
cat("\nCVT vehicles:", sum(data_fe$is_cvt), "\n")
cat("\nn_gears summary:\n")
print(summary(data_fe$n_gears))
cat("\nhas_discrete_gears distribution:\n")
print(table(data_fe$has_discrete_gears))
```

### Build Final Modeling Dataset

We apply all column removals. The raw `trany` is dropped because it is fully replaced by `transmission_type`, `is_cvt`, and `n_gears`. The raw `fuelType` and `fuelType1` are dropped because their information is captured by `fuel_group`.

Pure battery electric vehicles (BEVs) are excluded from the modeling dataset before the train/test split. The target variable `comb08` represents Miles Per Gallon (MPG) for combustion-engine vehicles, but for BEVs the EPA reports MPGe — an artificial conversion (33.7 kWh = 1 gallon equivalent) that does not reflect actual fuel consumption. Including BEVs would mix two structurally different quantities in a single regression target. Moreover, the features that drive MPGe — battery capacity, motor efficiency, regenerative braking — are not present in the dataset after removing high-missingness columns such as `evMotor`. The `is_electric` and `has_combustion_engine` flags are also dropped after filtering, as they become zero-variance constants.

```{r}
n_ev <- sum(data_fe$is_electric)

vehicles_model <- data_fe %>%
  filter(!is_electric) %>%
  select(-any_of(cols_to_remove)) %>%
  select(-any_of(c("trany", "fuelType", "fuelType1", "is_electric", "has_combustion_engine")))

cat("Electric vehicles excluded:", n_ev, "\n")
cat("Remaining vehicles:        ", nrow(vehicles_model), "\n")
cat("Dimensions of vehicles_model:", nrow(vehicles_model), "×", ncol(vehicles_model), "\n")
cat("\nColumns retained:\n")
print(names(vehicles_model))
```

`cylinders` is kept as a candidate predictor despite its strong correlation with `displ` (~0.91). Multicollinearity between the two was expected, but LASSO (Scenario 2) selected `penalty = 1e-4` — too small to zero any coefficient — and retained both predictors. With n/p > 2 000, OLS variance is low enough that the multicollinearity does not destabilize estimates. Both `cylinders` and `displ` also appear as the two strongest discriminators in LDA (Scenario 5), confirming they each carry independent signal: `displ` reflects total engine displacement while `cylinders` captures engine architecture (e.g. a turbocharged 4-cylinder can have large displacement but few cylinders). Dropping either variable based solely on the correlation would discard real predictive information.

### Sanity Checks

```{r}
cat("Duplicate rows in vehicles_model:", sum(duplicated(vehicles_model)), "\n\n")

cat("summary(comb08):\n")
print(summary(vehicles_model$comb08))

cat("\nYear range:", range(vehicles_model$year, na.rm = TRUE), "\n")

cat("\ncomb08 <= 0:", sum(vehicles_model$comb08 <= 0, na.rm = TRUE), "\n")
cat("displ < 0:",    sum(vehicles_model$displ    < 0, na.rm = TRUE), "\n")
cat("cylinders < 0:", sum(vehicles_model$cylinders < 0, na.rm = TRUE), "\n")

missing_before <- colSums(is.na(vehicles_model))
cat("\nColumns with missing values (before recipe):\n")
print(missing_before[missing_before > 0])
cat("Total missing values:", sum(is.na(vehicles_model)), "\n")
```

After excluding BEVs, the remaining `comb08` values reflect actual MPG from combustion-engine and hybrid vehicles. Extreme values at the high end (diesel hybrids, efficient compact cars) are retained as valid observations.

### Train/Test Split

We split the data into 80% training and 20% test. The test set is only used at the very end to evaluate final model performance — we do not touch it during training or tuning. We stratify on `comb08` so both sets have a similar distribution of MPG values.

```{r}
set.seed(123)

data_split <- initial_split(vehicles_model, prop = 0.8, strata = comb08)

train_data <- training(data_split)
test_data  <- testing(data_split)

cat("Train:", nrow(train_data), "×", ncol(train_data), "\n")
cat("Test: ", nrow(test_data),  "×", ncol(test_data),  "\n")
```

### Check Train/Test Distribution

```{r}
cat("Train comb08:\n"); print(summary(train_data$comb08))
cat("Test  comb08:\n"); print(summary(test_data$comb08))
```

```{r}
bind_rows(
  train_data %>% select(comb08) %>% mutate(split = "train"),
  test_data  %>% select(comb08) %>% mutate(split = "test")
) %>%
  ggplot(aes(x = comb08, fill = split)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("train" = "steelblue", "test" = "tomato")) +
  labs(
    title = "Distribution of comb08: Train vs. Test",
    x     = "Combined MPG (comb08)",
    y     = "Density"
  ) +
  theme_minimal()
```

The density curves confirm that stratification produces very similar `comb08` distributions in both sets.

### Cross-Validation Folds

Ridge, LASSO, and Elastic Net need a `penalty` hyperparameter tuned by comparing performance across different values. We use 5-fold cross-validation for this: the training data is split into 5 folds, the model trains on 4 and is evaluated on the remaining one, rotating through all folds. This gives a reliable performance estimate without using the test set.

```{r}
folds <- vfold_cv(train_data, v = 5, strata = comb08)

cat("Number of CV folds:", nrow(folds), "\n")
print(folds)
```

### Preprocessing Recipe

We define preprocessing as a `recipe` to prevent data leakage — all transformations (normalization, imputation, dummy encoding) are learned only from the training data and then applied to the test set using those saved values. This also works correctly inside cross-validation, where each fold fits its own preprocessing independently.

- `step_unknown()` — handles unseen factor levels that may appear in the test set
- `step_other()` — collapses any residual rare categories below 1%
- `step_impute_median(displ, cylinders)` — fills the small number of numeric NAs in engine parameters for non-BEV rows; `n_gears` is intentionally set to 0 in feature engineering and does not need imputation
- `step_dummy()` — one-hot encodes all nominal predictors; required by linear regression methods
- `step_zv()` / `step_nzv()` — removes zero- and near-zero-variance columns, consistent with EDA findings on low-variability variables
- `step_normalize()` — standardizes all numeric predictors; required for ridge, lasso, elastic net, kNN, and PCA/PCR/PLS

```{r}
model_recipe <- recipe(comb08 ~ ., data = train_data) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%
  step_impute_median(displ, cylinders) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_nzv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

model_recipe
```

### Fit and Apply Recipe

The recipe is fitted exclusively on `train_data`. The same fitted statistics (medians, modes, dummy levels) are then applied to both sets.

```{r}
prep_recipe <- prep(model_recipe, training = train_data)

train_processed <- bake(prep_recipe, new_data = NULL)
test_processed  <- bake(prep_recipe, new_data = test_data)
```

### Verify Preprocessing Output

```{r}
cat("train_processed:", nrow(train_processed), "×", ncol(train_processed), "\n")
cat("test_processed: ", nrow(test_processed),  "×", ncol(test_processed),  "\n")

cat("\nMissing values — train_processed:", sum(is.na(train_processed)), "\n")
cat("Missing values — test_processed: ", sum(is.na(test_processed)),  "\n")

cat("\ncomb08 in train_processed:", "comb08" %in% names(train_processed), "\n")
cat("comb08 in test_processed: ", "comb08" %in% names(test_processed),  "\n")

n_predictors <- ncol(train_processed) - 1
cat("\nNumber of predictors after dummy encoding:", n_predictors, "\n")

cat("\nFirst 10 column names:\n")
print(head(names(train_processed), 10))
```

Both processed datasets contain no missing values in predictors, and `comb08` is retained as the regression target. The objects `vehicles_model`, `train_data`, `test_data`, `folds`, `model_recipe`, `prep_recipe`, `train_processed`, and `test_processed` are ready for regression scenarios 2, 3, and 5.

## Scenario 2: Parametric vs. Non-Parametric Models — Parametric Part

Parametric models assume that the relationship between predictors and the response follows a specific, fixed mathematical form — in regression, a linear combination of weighted features.

We compare four parametric regression methods:

1. **Linear Regression (OLS)** — the classical least-squares baseline, no regularization
2. **Ridge Regression** — OLS plus an L2 penalty that shrinks all coefficients toward zero
3. **LASSO** — OLS plus an L1 penalty that can shrink some coefficients to exactly zero, selecting features
4. **Elastic Net** — a blend of Ridge and LASSO penalties, combining both approaches

All four methods share the same preprocessing recipe defined in the previous section. We use the 5-fold cross-validation folds for hyperparameter tuning and model comparison, and evaluate final performance on the held-out test set.

```{r message=FALSE, warning=FALSE}
library(tidymodels)
library(broom)
library(glmnet)
```

We define the three evaluation metrics once and reuse them across every model.

```{r}
eval_metrics <- metric_set(rmse, rsq, mae)
```

---

### 1. Linear Regression (OLS)

Ordinary least squares minimizes the sum of squared residuals and imposes no constraint on coefficient size. It is our reference point — any benefit from regularization will appear as an improvement over this baseline.

```{r}
lm_spec <- linear_reg() %>%
  set_engine("lm") %>%
  set_mode("regression")

lm_wf <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(lm_spec)
```

We wrap the recipe and model in a `workflow` so preprocessing is applied automatically during cross-validation and when predicting on new data.

```{r warning=FALSE}
set.seed(123)

lm_cv <- lm_wf %>%
  fit_resamples(
    resamples = folds,
    metrics   = eval_metrics,
    control   = control_resamples(save_pred = TRUE)
  )

lm_cv_metrics <- collect_metrics(lm_cv)
cat("=== OLS — 5-fold CV ===\n")
cat("RMSE:", round(lm_cv_metrics %>% filter(.metric == "rmse") %>% pull(mean), 4),
    "  ±", round(lm_cv_metrics %>% filter(.metric == "rmse") %>% pull(std_err), 4), "\n")
cat("MAE: ", round(lm_cv_metrics %>% filter(.metric == "mae")  %>% pull(mean), 4), "\n")
cat("R²:  ", round(lm_cv_metrics %>% filter(.metric == "rsq")  %>% pull(mean), 4), "\n")
```

Now we fit on the full training set to inspect which predictors drive the predictions.

```{r}
lm_fit <- lm_wf %>% fit(data = train_data)

lm_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(term != "(Intercept)") %>%
  slice_max(abs(estimate), n = 20) %>%
  ggplot(aes(x = reorder(term, abs(estimate)), y = estimate)) +
  geom_col(aes(fill = estimate > 0)) +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
    labels = c("TRUE" = "Positive", "FALSE" = "Negative"),
    name   = "Direction"
  ) +
  labs(
    title = "OLS: Top 20 Predictors by Absolute Coefficient",
    x     = NULL,
    y     = "Coefficient estimate"
  ) +
  theme_minimal()
```

Blue bars indicate predictors that increase `comb08` (e.g. electric drivetrain), red bars indicate predictors that decrease it (e.g. larger engine displacement).

---

### 2. Ridge Regression

Ridge regression adds an L2 penalty — λ × Σβ² — to the OLS loss. This forces the optimiser to keep coefficients small, reducing variance at the cost of some bias. Ridge never removes a predictor entirely: all coefficients are shrunk toward zero, but remain non-zero. It is especially effective when predictors are correlated — a situation we identified between `displ` and `cylinders` in the EDA.

The regularization strength λ is controlled by `penalty`. We tune it over a log-spaced grid.

```{r}
ridge_spec <- linear_reg(penalty = tune(), mixture = 0) %>%
  set_engine("glmnet") %>%
  set_mode("regression")

ridge_wf <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(ridge_spec)
```

`glmnet` fits the entire regularization path in a single pass, so evaluating 50 penalty values has almost no extra cost compared to evaluating one.

```{r}
ridge_grid <- grid_regular(penalty(range = c(-4, 2)), levels = 50)
```

```{r warning=FALSE}
set.seed(123)

ridge_cv <- ridge_wf %>%
  tune_grid(
    resamples = folds,
    grid      = ridge_grid,
    metrics   = eval_metrics
  )

autoplot(ridge_cv)
```

At very low penalty (left end), Ridge behaves like OLS. As penalty increases, coefficients shrink and the model becomes more biased but less variable. The optimal point is where RMSE on the validation folds is minimized.

```{r}
best_ridge <- ridge_cv %>% select_best(metric = "rmse")

ridge_best_row <- show_best(ridge_cv, metric = "rmse", n = 1)
cat("=== Ridge — best hyperparameter ===\n")
cat("Best penalty:", round(best_ridge$penalty, 6), "\n")
cat("CV RMSE:     ", round(ridge_best_row$mean, 4),
    "  ±", round(ridge_best_row$std_err, 4), "\n")
```

```{r}
ridge_final_wf <- ridge_wf %>% finalize_workflow(best_ridge)
ridge_fit       <- ridge_final_wf %>% fit(data = train_data)
```

```{r}
ridge_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(term != "(Intercept)") %>%
  slice_max(abs(estimate), n = 20) %>%
  ggplot(aes(x = reorder(term, abs(estimate)), y = estimate)) +
  geom_col(aes(fill = estimate > 0)) +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
    labels = c("TRUE" = "Positive", "FALSE" = "Negative"),
    name   = "Direction"
  ) +
  labs(
    title = paste0("Ridge: Top 20 Predictors (penalty = ",
                   round(best_ridge$penalty, 5), ")"),
    x     = NULL,
    y     = "Coefficient estimate"
  ) +
  theme_minimal()
```

Compared to OLS, Ridge produces smaller coefficient magnitudes across all predictors. The ordering and direction of the most important features generally remains the same, but correlated predictors share their weight more evenly.

---

### 3. LASSO Regression

LASSO (Least Absolute Shrinkage and Selection Operator) uses an L1 penalty — λ × Σ|β|. The key geometric property of L1 is that the regularized solution often lands exactly at zero for many coefficients. LASSO therefore performs simultaneous regression and feature selection: it automatically identifies the most informative predictors and removes the rest.

```{r}
lasso_spec <- linear_reg(penalty = tune(), mixture = 1) %>%
  set_engine("glmnet") %>%
  set_mode("regression")

lasso_wf <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(lasso_spec)
```

LASSO typically needs smaller penalties than Ridge to achieve similar shrinkage, so we search a narrower range.

```{r}
lasso_grid <- grid_regular(penalty(range = c(-4, 0)), levels = 50)
```

```{r warning=FALSE}
set.seed(123)

lasso_cv <- lasso_wf %>%
  tune_grid(
    resamples = folds,
    grid      = lasso_grid,
    metrics   = eval_metrics
  )

autoplot(lasso_cv)
```

```{r}
best_lasso <- lasso_cv %>% select_best(metric = "rmse")

lasso_best_row <- show_best(lasso_cv, metric = "rmse", n = 1)
cat("=== LASSO — best hyperparameter ===\n")
cat("Best penalty:", round(best_lasso$penalty, 6), "\n")
cat("CV RMSE:     ", round(lasso_best_row$mean, 4),
    "  ±", round(lasso_best_row$std_err, 4), "\n")
```

```{r}
lasso_final_wf <- lasso_wf %>% finalize_workflow(best_lasso)
lasso_fit       <- lasso_final_wf %>% fit(data = train_data)
```

One of LASSO's main advantages is sparsity. Let's see how many predictors it retained.

```{r}
lasso_coefs <- lasso_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(term != "(Intercept)")

cat("=== LASSO — sparsity ===\n")
cat("Total predictors available:", nrow(lasso_coefs), "\n")
cat("Kept by LASSO (non-zero):  ", sum(lasso_coefs$estimate != 0), "\n")
cat("Removed by LASSO (zero):   ", sum(lasso_coefs$estimate == 0), "\n")
if (sum(lasso_coefs$estimate == 0) > 0) {
  cat("Zeroed out predictors:\n")
  print(lasso_coefs %>% filter(estimate == 0) %>% pull(term))
}
```

```{r}
lasso_coefs %>%
  filter(estimate != 0) %>%
  slice_max(abs(estimate), n = 20) %>%
  ggplot(aes(x = reorder(term, abs(estimate)), y = estimate)) +
  geom_col(aes(fill = estimate > 0)) +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
    labels = c("TRUE" = "Positive", "FALSE" = "Negative"),
    name   = "Direction"
  ) +
  labs(
    title = paste0("LASSO: Top 20 Non-Zero Predictors (penalty = ",
                   round(best_lasso$penalty, 5), ")"),
    x     = NULL,
    y     = "Coefficient estimate"
  ) +
  theme_minimal()
```

LASSO retains only the most informative predictors, which makes the resulting model more parsimonious and directly interpretable compared to OLS or Ridge. Predictors that survive the L1 penalty are the ones that the algorithm judged worth the cost of adding a parameter.

---

### 4. Elastic Net

Elastic Net combines both L1 and L2 penalties: λ × [α × Σ|β| + (1 − α) × Σβ²]. The `mixture` parameter in tidymodels corresponds to α:  `mixture = 0` gives pure Ridge, `mixture = 1` gives pure LASSO. By tuning both `penalty` and `mixture`, Elastic Net finds the best blend — it can perform feature selection like LASSO while also handling correlated predictors like Ridge.

```{r}
en_spec <- linear_reg(penalty = tune(), mixture = tune()) %>%
  set_engine("glmnet") %>%
  set_mode("regression")

en_wf <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(en_spec)
```

We use a regular 10 × 10 grid — 10 penalty values and 10 mixture values — giving 100 combinations. `glmnet` evaluates all penalty values for each mixture in a single path, so this is still fast.

```{r}
en_grid <- grid_regular(
  penalty(range = c(-4, 0)),
  mixture(range = c(0, 1)),
  levels = 10
)
```

```{r warning=FALSE}
set.seed(123)

en_cv <- en_wf %>%
  tune_grid(
    resamples = folds,
    grid      = en_grid,
    metrics   = eval_metrics
  )

autoplot(en_cv)
```

```{r}
best_en <- en_cv %>% select_best(metric = "rmse")

en_best_row <- show_best(en_cv, metric = "rmse", n = 1)
cat("=== Elastic Net — best hyperparameters ===\n")
cat("Best penalty:", round(best_en$penalty, 6), "\n")
cat("Best mixture:", round(best_en$mixture, 4),
    " (0 = Ridge, 1 = LASSO)\n")
cat("CV RMSE:     ", round(en_best_row$mean, 4),
    "  ±", round(en_best_row$std_err, 4), "\n")
```

The optimal `mixture` value tells us how the data prefers to be regularized. If the best mixture is close to 1, LASSO-like behavior is preferred; if it is close to 0, Ridge-like behavior is preferred.

```{r}
en_final_wf <- en_wf %>% finalize_workflow(best_en)
en_fit       <- en_final_wf %>% fit(data = train_data)
```

---

### Comparing All Parametric Models

We evaluate all four models on the held-out test set. Because each model is a full `workflow()` that includes the preprocessing recipe, we pass the raw `test_data` directly and the recipe transformations are applied automatically.

We write a small helper function so we do not repeat the same prediction + metric code four times.

```{r}
get_test_metrics <- function(fit, model_name) {
  fit %>%
    predict(new_data = test_data) %>%
    bind_cols(test_data %>% select(comb08)) %>%
    summarise(
      model = model_name,
      RMSE  = rmse_vec(truth = comb08, estimate = .pred),
      MAE   = mae_vec( truth = comb08, estimate = .pred),
      Rsq   = rsq_vec( truth = comb08, estimate = .pred)
    )
}
```

```{r}
test_comparison <- bind_rows(
  get_test_metrics(lm_fit,    "Linear Regression"),
  get_test_metrics(ridge_fit, "Ridge"),
  get_test_metrics(lasso_fit, "LASSO"),
  get_test_metrics(en_fit,    "Elastic Net")
) %>%
  arrange(RMSE)

test_comparison
```

```{r}
test_comparison %>%
  pivot_longer(cols = c(RMSE, MAE, Rsq),
               names_to  = "metric",
               values_to = "value") %>%
  ggplot(aes(x = reorder(model, value), y = value, fill = model)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~metric, scales = "free_x") +
  coord_flip() +
  labs(
    title = "Test Set Performance: Parametric Models",
    x     = NULL,
    y     = "Value"
  ) +
  theme_minimal()
```

Lower RMSE and MAE are better; higher R² is better.

#### Cross-Validation vs. Test Set RMSE

We compare CV RMSE with test RMSE. A small gap means no overfitting — the model generalizes to new data roughly as well as it performed during CV.

```{r}
cv_rmse_comparison <- tibble(
  model   = c("Linear Regression", "Ridge", "LASSO", "Elastic Net"),
  cv_rmse = c(
    collect_metrics(lm_cv)    %>% filter(.metric == "rmse") %>% pull(mean),
    show_best(ridge_cv, metric = "rmse", n = 1) %>% pull(mean),
    show_best(lasso_cv, metric = "rmse", n = 1) %>% pull(mean),
    show_best(en_cv,    metric = "rmse", n = 1) %>% pull(mean)
  )
) %>%
  left_join(
    test_comparison %>% select(model, test_rmse = RMSE),
    by = "model"
  ) %>%
  mutate(gap = abs(test_rmse - cv_rmse)) %>%
  arrange(cv_rmse)

cv_rmse_comparison
```

A small `gap` confirms no overfitting.

#### Actual vs. Predicted — All Models

Actual vs. predicted on the test set. Points on the red diagonal = perfect prediction.

```{r}
all_test_preds <- bind_rows(
  lm_fit    %>% predict(new_data = test_data) %>% bind_cols(test_data %>% select(comb08)) %>% mutate(model = "Linear Regression"),
  ridge_fit %>% predict(new_data = test_data) %>% bind_cols(test_data %>% select(comb08)) %>% mutate(model = "Ridge"),
  lasso_fit %>% predict(new_data = test_data) %>% bind_cols(test_data %>% select(comb08)) %>% mutate(model = "LASSO"),
  en_fit    %>% predict(new_data = test_data) %>% bind_cols(test_data %>% select(comb08)) %>% mutate(model = "Elastic Net")
)

all_test_preds %>%
  ggplot(aes(x = comb08, y = .pred)) +
  geom_point(alpha = 0.15, color = "steelblue", size = 0.8) +
  geom_abline(slope = 1, intercept = 0, color = "tomato", linewidth = 0.8) +
  facet_wrap(~model) +
  labs(
    title = "Actual vs. Predicted — Test Set (Parametric Models)",
    x     = "Actual combined MPG",
    y     = "Predicted combined MPG"
  ) +
  theme_minimal()
```

With BEVs excluded, all values are real MPG. The fan-shaped distortion from MPGe outliers is gone. Remaining scatter at high MPG values is mostly PHEVs and efficient compacts — the model slightly underpredicts these.

---

### Summary: Parametric Models

All four models assume a linear relationship. They differ only in regularization:

| Model | Penalty | Sparsity |
|---|---|---|
| OLS | None | No |
| Ridge | L2 | No |
| LASSO | L1 | Yes |
| Elastic Net | L1 + L2 | Yes |

On large datasets the differences are small. Regularization stabilizes coefficients; LASSO additionally zeroes out weak predictors.

---

### Conclusion

#### Effect of removing electric vehicles

Removing 1 425 BEVs (whose `comb08` values are MPGe, not MPG) dropped RMSE from 8.11 to **3.36 MPG** (−59%) and MAE from 5.15 to **2.19 MPG**. R² only moved from 0.642 to 0.661 — this is expected, since R² is scale-independent and both the errors and the total variance dropped together.

#### Model results

All four models gave nearly the same result on the test set: **RMSE ≈ 3.36 MPG, MAE ≈ 2.19 MPG, R² ≈ 0.661**. With a median `comb08` of 20 MPG, an average error of 2.19 MPG is roughly 11% — reasonable for a linear model with 21 features.

#### Why regularization did not help

All three regularized models selected `penalty = 0.0001` — the minimum of the search grid. With 38 735 training observations and only 21 predictors, there is very little variance to reduce, so regularization has almost nothing to do. LASSO also kept all 21 predictors for the same reason — at such a small penalty, no predictor is weak enough to be zeroed out. This is not a bad result; it means the feature set is clean and every variable carries real signal.

#### Where the models fall short

- The relationship between engine size and MPG is not perfectly linear — adding a cylinder to a small 4-cylinder engine is different from adding one to a V8.
- PHEVs and mild hybrids behave differently depending on driving mode, which no single feature captures.
- Vehicle weight and aerodynamic drag are not in the dataset, even though they strongly affect fuel economy.

The CV–test gap is below 0.005 for all models, so there is no overfitting. The models are a solid baseline — interpretable, stable, and easy to explain.

---

## Scenario 5: Linear Discriminant Analysis (LDA)

LDA (Linear Discriminant Analysis) is a classification method that also reduces dimensionality. We convert `comb08` into three efficiency classes and let LDA find **discriminant axes** — linear combinations of predictors that best separate the classes. The projection onto these axes is the dimensionality reduction part: many features compressed into 2 coordinates.

```{r}
library(discrim)
```

### 5.1 Creating the Classification Target

We divide `comb08` into three classes using quantile-based cuts (tertiles). Equal-sized classes help LDA estimate the class means and boundaries reliably.

```{r}
efficiency_cuts <- quantile(vehicles_model$comb08, probs = c(0, 1/3, 2/3, 1))

vehicles_lda <- vehicles_model %>%
  mutate(
    efficiency_class = cut(
      comb08,
      breaks        = efficiency_cuts,
      labels        = c("Low", "Medium", "High"),
      include.lowest = TRUE
    )
  ) %>%
  select(-comb08)

cat("=== Efficiency class distribution ===\n")
print(table(vehicles_lda$efficiency_class))
cat("\nMPG cut points (tertiles):\n")
cat("Low:   ", round(efficiency_cuts[1], 1), "-", round(efficiency_cuts[2], 1), "MPG\n")
cat("Medium:", round(efficiency_cuts[2], 1), "-", round(efficiency_cuts[3], 1), "MPG\n")
cat("High:  ", round(efficiency_cuts[3], 1), "-", round(efficiency_cuts[4], 1), "MPG\n")
cat("\nClass proportions:\n")
print(round(prop.table(table(vehicles_lda$efficiency_class)), 3))
```

### 5.2 Train/Test Split and Cross-Validation Folds

Same split strategy as Scenario 2 — 80/20 stratified split and 5-fold CV to estimate accuracy before final fitting.

```{r}
set.seed(123)
lda_split <- initial_split(vehicles_lda, prop = 0.8, strata = efficiency_class)
lda_train  <- training(lda_split)
lda_test   <- testing(lda_split)
lda_folds  <- vfold_cv(lda_train, v = 5, strata = efficiency_class)

cat("=== LDA — data split ===\n")
cat("Training rows:", nrow(lda_train), "\n")
cat("Test rows:    ", nrow(lda_test),  "\n")
cat("CV:            5-fold, stratified by efficiency_class\n")
```

### 5.3 Recipe and Model Specification

Same preprocessing as Scenario 2, just with `efficiency_class` as the target. LDA needs normalized predictors, so `step_normalize()` is important here.

```{r}
lda_recipe <- recipe(efficiency_class ~ ., data = lda_train) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%
  step_impute_median(displ, cylinders) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_nzv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

lda_spec <- discrim_linear() %>%
  set_engine("MASS") %>%
  set_mode("classification")

lda_wf <- workflow() %>%
  add_recipe(lda_recipe) %>%
  add_model(lda_spec)
```

### 5.4 Cross-Validation

LDA has no hyperparameters, so CV just gives us a reliable accuracy estimate before we fit on the full training set.

```{r warning=FALSE}
lda_cv <- fit_resamples(
  lda_wf,
  resamples = lda_folds,
  metrics   = metric_set(accuracy),
  control   = control_resamples(save_pred = FALSE)
)

lda_cv_metrics <- collect_metrics(lda_cv)

cat("=== LDA — 5-fold CV ===\n")
cat("CV Accuracy:", round(lda_cv_metrics %>% filter(.metric == "accuracy") %>% pull(mean), 4),
    " ±", round(lda_cv_metrics %>% filter(.metric == "accuracy") %>% pull(std_err), 4), "\n")
```

### 5.5 Fitting the Model on the Full Training Set

```{r}
lda_fit <- lda_wf %>% fit(data = lda_train)

mass_lda <- lda_fit %>%
  extract_fit_parsnip() %>%
  pluck("fit")

cat("=== LDA — Prior probabilities ===\n")
print(round(mass_lda$prior, 4))

cat("\n=== LDA — Proportion of variance explained by each discriminant axis ===\n")
prop_trace <- mass_lda$svd^2 / sum(mass_lda$svd^2)
names(prop_trace) <- paste0("LD", seq_along(prop_trace))
print(round(prop_trace, 4))
```

### 5.6 Dimensionality Reduction — Projection onto Discriminant Axes

We project all training vehicles onto LD1 and LD2 — the dimensionality reduction step. Dozens of normalized features collapse into 2 coordinates that maximally separate the classes.

```{r}
prepped     <- prep(lda_recipe, training = lda_train)
train_baked <- bake(prepped, new_data = lda_train)

lda_proj <- predict(mass_lda, newdata = select(train_baked, -efficiency_class))

lda_train_proj <- tibble(
  LD1              = lda_proj$x[, 1],
  LD2              = lda_proj$x[, 2],
  efficiency_class = lda_train$efficiency_class
)

cat("=== LDA — discriminant axis ranges (training data) ===\n")
cat("LD1:", round(range(lda_train_proj$LD1), 3), "\n")
cat("LD2:", round(range(lda_train_proj$LD2), 3), "\n")

ggplot(lda_train_proj, aes(x = LD1, y = LD2, color = efficiency_class)) +
  geom_point(alpha = 0.25, size = 0.6) +
  stat_ellipse(level = 0.95, linewidth = 1.2) +
  scale_color_manual(
    values = c("Low" = "#E74C3C", "Medium" = "#F39C12", "High" = "#27AE60")
  ) +
  labs(
    title    = "LDA — projection of training data onto discriminant axes",
    subtitle = "95% confidence ellipses per efficiency class",
    x        = "LD1 (1st discriminant axis)",
    y        = "LD2 (2nd discriminant axis)",
    color    = "Class"
  ) +
  theme_minimal()
```

### 5.7 Discriminant Axis Loadings

Each loading shows how much a predictor contributes to LD1. Negative = pulls toward Low efficiency; positive = pulls toward High.

```{r fig.height=6}
lda_loadings <- as_tibble(mass_lda$scaling, rownames = "predictor")

lda_loadings %>%
  slice_max(abs(LD1), n = 15) %>%
  ggplot(aes(x = reorder(predictor, LD1), y = LD1, fill = LD1 > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "#27AE60", "FALSE" = "#E74C3C"),
    guide  = "none"
  ) +
  labs(
    title    = "LDA — loadings of the 1st discriminant axis (LD1)",
    subtitle = "Top 15 predictors — green = higher MPG, red = lower MPG",
    x        = NULL,
    y        = "LD1 loading"
  ) +
  theme_minimal()

cat("=== LDA — Top 5 predictors for LD1 (absolute loading) ===\n")
lda_loadings %>%
  slice_max(abs(LD1), n = 5) %>%
  select(predictor, LD1) %>%
  mutate(LD1 = round(LD1, 4)) %>%
  print()
```

### 5.8 Classification Performance on the Test Set

```{r}
lda_test_pred <- predict(lda_fit, new_data = lda_test) %>%
  bind_cols(lda_test %>% select(efficiency_class)) %>%
  rename(pred = .pred_class, truth = efficiency_class)

lda_acc  <- accuracy(lda_test_pred,  truth = truth, estimate = pred)
lda_prec <- precision(lda_test_pred, truth = truth, estimate = pred, estimator = "macro")
lda_rec  <- recall(lda_test_pred,    truth = truth, estimate = pred, estimator = "macro")

cat("=== LDA — Classification metrics (test set) ===\n")
cat("Accuracy  (overall): ", round(lda_acc$.estimate,  4), "\n")
cat("Precision (macro):   ", round(lda_prec$.estimate, 4), "\n")
cat("Recall    (macro):   ", round(lda_rec$.estimate,  4), "\n")

cm <- conf_mat(lda_test_pred, truth = truth, estimate = pred)
cat("\n=== Confusion matrix ===\n")
print(cm)

cat("\n=== Per-class accuracy ===\n")
lda_test_pred %>%
  group_by(truth) %>%
  summarise(
    n        = n(),
    correct  = sum(pred == truth, na.rm = TRUE),
    accuracy = round(correct / n, 4),
    .groups  = "drop"
  ) %>%
  print()

autoplot(cm, type = "heatmap") +
  scale_fill_gradient(low = "#f7fbff", high = "#2171b5") +
  labs(
    title = "LDA — confusion matrix (test set)",
    x     = "Predicted class",
    y     = "True class"
  ) +
  theme_minimal()
```

### Conclusion

#### Dimensionality reduction

LD1 alone explains **98.2% of the between-class variance** — essentially all the class separation fits into one dimension. The top predictors are engine displacement (−0.765), cylinders (−0.618), and discrete gears (−0.615), all negative: larger engines = lower efficiency. FWD (0.504) and number of gears (0.350) push toward High. This matches intuition.

#### Classification performance

Test accuracy: **78.0%**, CV was 77.5% — a 0.5 pp gap, no overfitting.

| Class  | MPG range | Accuracy |
|--------|-----------|----------|
| Low    | 7–18      | 81.4%    |
| Medium | 18–22     | 68.8%    |
| High   | 22–74     | 83.0%    |

Medium is hardest — it spans only 4 MPG. Vehicles near the 18 and 22 MPG boundaries are nearly identical, and any linear classifier will confuse them. Almost all errors are between adjacent classes.

#### Comparison with Scenario 2

Regression RMSE was 3.36 MPG. The Medium class is only 4 MPG wide, so a typical regression error already covers most of that range. The 68.8% Medium accuracy is a direct consequence — not a failure of LDA, but a reflection of how hard the boundary region is.

#### Suitability

LDA works well here. LD1 dominance (98.2%) confirms a linear boundary is enough. The weak point is the tertile-based class cuts — the 18 and 22 MPG boundaries are arbitrary and make Medium artificially hard to classify.

---

## Overall Assessment: Scenarios 2 and 5

### Was the data suitable?

**Scenario 2 (regression):** Yes. Large dataset (48k rows), only 21 predictors, continuous target with strong linear correlations (up to −0.77 with cylinders). Nearly ideal conditions for linear regression.

**Scenario 5 (LDA):** Mostly yes. LDA's assumptions (multivariate normality, equal covariance per class) weren't formally tested, but large samples make LDA robust to violations. The bigger issue is the **target construction** — MPG is continuous and the tertile cuts are arbitrary.

---

### What worked against each method?

**Scenario 2:**

1. Non-linear relationships — adding a cylinder affects small engines differently than V8s; linear models can't capture this without interaction terms.
2. PHEVs — their MPG depends on driving mode (electric vs. combustion), which no static feature captures.
3. Missing features — vehicle weight and aerodynamic drag are among the strongest determinants of fuel economy and aren't in the dataset.
4. Regularization was irrelevant — all three regularized models converged to penalty = 1e-4 (minimum of the grid). With 48k rows and 21 features, there's nothing to regularize. Zero gain over OLS.

**Scenario 5:**

1. Medium class is only 4 MPG wide — regression RMSE alone (3.36 MPG) covers most of this range, so any classifier will struggle near the boundaries.
2. Unequal class widths — High spans 52 MPG, Low spans 11 MPG; the equal-covariance assumption is likely violated.
3. Discretization loses information — 17.9 MPG and 18.1 MPG end up in different classes despite being essentially the same vehicle.

---

### Was the result sufficient?

**Scenario 2:** Good baseline. RMSE = 3.36 MPG (≈11% of the 20 MPG median), R² = 0.661. Fine for analysis and feature interpretation, but not precise enough for regulatory use. One-third of variance is unexplained — missing features (weight, drag) are the likely cause, not model weakness.

**Scenario 5:** Good for a classification demo. 77.9% on 3 balanced classes is reasonable. But classification isn't the natural framing for this problem — users need actual MPG values, not bins.

---

### How to improve?

**Scenario 2:** Add vehicle weight and drag coefficient. Add interaction terms (`cylinders × displ`). Try gradient boosting — it captures non-linear effects and interactions automatically.

**Scenario 5:** Use domain-based class cuts (e.g., EPA categories) instead of tertiles. Try QDA — it relaxes the equal-covariance assumption. Multinomial logistic regression is another natural alternative.

For this dataset, **regression is the right problem formulation**. LDA is useful here as a dimensionality reduction demo and pedagogical contrast, but not as a replacement for Scenario 2.
